### 2023-08-03 README.Rmd
Initial commit

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# Mali.decor.networks

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/AHWA-Lab-Frankfurt/Mali.decor.networks/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2023-08-03 analysis/R/hull geom.Rmd
Initial commit

---
title: "network plotting"
author: "Nikolas Gestrich"
date: '2022-09-09'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
```

# Convex hulls

extending ggplot to make convex hull stat for network communities

```{r}
StatChull <- ggproto("StatChull", Stat,
  compute_group = function(data, scales) {
    data[chull(data$x, data$y), , drop = FALSE]
  },

  required_aes = c("x", "y")
)

geom_chull <- function(mapping = NULL, data = NULL, geom = "polygon",
                       position = "identity", na.rm = FALSE, show.legend = NA, 
                       inherit.aes = TRUE, ...) {
  ggplot2::layer(
    stat = StatChull, data = data, mapping = mapping, geom = geom, 
    position = position, show.legend = show.legend, inherit.aes = inherit.aes,
    params = list(na.rm = na.rm, ...)
  )
}
```

### 2023-08-03 analysis/paper/data analysis.Rmd
Initial commit

---
title: "analysis"
author: "Nikolas Gestrich"
date: '2022-05-02'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(knitr.kable.NA = "")

# libraries
library(dplyr)
library(knitr)
library(patchwork)
library(ggplot2)
library(tidygraph)
library(ggraph)
library(tibble)
library(igraph)
library(tidyr)
library(sf)
```

## Data import
We are importing data on pottery decors from a variety of sources. All of the data represents phases of sites that are dated to have been occupied at some point in the 11th century AD.

XXX: put together the different periods. 
Update: or maybe not, so that we can name the nodes the same

```{r}
library(readr)
# import decor data
decors <- read_csv2(here::here("analysis/data/raw_data/decors_late.csv"))
# import information on sites
sites <- read_csv2(here::here("analysis/data/raw_data/sites_all.csv"))
```

show the data:

```{r fig.cap="decor data used for network construction"}
library(kableExtra)
kbl(decors) |> 
  kable_styling(font_size = 7, html_font = "Cambria") 
```

show the map points:

```{r fig.cap="location of sites in the analysis", warning=FALSE}
source(here::here("analysis/R/map plots.R"))

sitepoints <- st_as_sf(sites, coords = c("lon", "lat"), crs = 4326, agr = "constant")

sitemap <- ggplot() +
  map_bw_grid +
  geom_sf(data = sitepoints)

sitemap + inset_element(inset, 0.72, 0.72, 1, 1)
```

We can see here that sites are geographically unequally distributed. This is due to research history as well as past cultural dynamics. Certainly, the IND and the areas bordering it (notiably the Lakes region) were strongly populated at the time, maybe more so than other areas. The cluster of sites around the Bandiagara escarpment is due to the extraordinarily detailed and long-term research of the University of Geneva's research teams. Blank spaces do not imply a lack of people, in this case people making and using pottery. Global politics and research priorities over the last 50 years have meant that these are areas we do not currently have excavated and dated ceramics for, though we know that sites exist, often in abundance. This means that, when we construct networks, we do so with a geographically unrepresentative sample, which leads to incomplete networks. Sindbaek has pointed out the difficulty in this situation: the actual network is not known, only its effects, from which the network can be partially reconstructed. Accordingly, this study should be seen as an exercise first in the reconstruction of a possible network that led to the data preserved in the archaeological record: a tiny sample of a tiny sample, influenced by an unknown amount of mostly unknowable pre- and post-depositional biases. Nevertheless, we will attempt to analyse the reconstructed network in order to formulate some hypotheses about the nature of innovation and spread in ceramic decors. Future generations of researchers are encouraged to improve upon this initial attempt.

# Construct network

```{r}
# import functions for network construction
source(here::here("analysis/R/edgelist functions.R"))

# import custom functions for network analysis
source(here::here("analysis/R/network functions.R"))

# import custom functions for data visualisation
source(here::here("analysis/R/datavis themes.R"))

# apply function to transform frequency table into edge list
edgelist <- make_edgelist(decors)

# construct network, remove extraneous sites
decor_graph <- tbl_graph(sites, edgelist, directed = FALSE) %>%
  activate(nodes) %>%
  filter(!node_is_isolated())

decor_graph
```

# Plot network

```{r total network, fig.cap="Plot of 11th century AD network of decorative practices. The plot shows the very strongly connected structure of the network."}
ggraph(decor_graph) +
  coloured_edges +
  theme_graph()
```

## Network structure

High number of connections due to the construction of the network. This network is very much unsuitable for further network analysis, but it already allows us to see one major thing about pottery decor interactions: the networks in which individual decor types are propagated are overlapping. There are no two decor networks that cover one another entirely.

```{r total network sample, fig.cap="Partially overlapping networks of four decorative practices"}
decor_graph %>%
  activate(edges) %>%
  filter(name %in% c("braided_cord_roulette", "braided_strip_roulette", "paint", "straight_cord_mat")) %>%
  ggraph() +
  coloured_edges +
  theme_graph()
```

## Calculate difference between network layers

We do this using the `multinet` package.

```{r import to multinet}
library(multinet)
# reduce graph to types with over 5 edges
good_edges <- decor_graph %>%
  activate(edges) %>%
  as_tibble() %>%
  count(name) %>%
  filter(n > 5)

reduced_graph <- decor_graph %>%
  activate(edges) %>%
  filter(name %in% good_edges$name)

# prepare graph and make a vector of different layer names
prep.graph <- reduced_graph %>%
  activate(edges) %>%
  rename(layer = name) %>%
  activate(nodes) %>%
  rename(name = label)

layers <- prep.graph %>%
  activate(edges) %>%
  as_tibble() %>%
  distinct(layer) %>%
  pull(layer)

# get a list of individual networks using the onelayer function (/R/network functions.R)
graphlist <- purrr::map(layers, onelayer)
names(graphlist) <- layers

# add layers into a multinet object
multi <- ml_empty()

for (i in seq_along(layers)) {
  add_igraph_layer_ml(multi, graphlist[[i]], paste(layers[[i]]))
}
```

```{r layer comparison, fig.cap="Jaccard similarity of nodes present in the networks of individual decorative practices"}
# perform layer compariosn using jaccard similarity between nodes
lc <- layer_comparison_ml(multi, method = "jaccard.actors")

# make heatmap
heatmap_matrix(lc)
```

This shows us the jaccard similarity of the actors present on the different levels of the network. We can see that there are more similar and more dissimilar layers in the network.Very high are the similarities between folded strip roulette and braided cord roulette, two common decors. Also high is the similarity in common decorative techniques of channeling and applied decors. The fact that there are similar and dissimilar layers suggests that we might be able to identify groups in the network.

## Clustering

We use the Louvain clustering algorithm (Blondel et al 2008) to find communities in the network.

```{r community detection, hide = 'results', echo=FALSE}
# use louvain clustering method to assign clusters to nodes
decor_graph <- decor_graph %>%
  activate(nodes) %>%
  mutate(louvain = as.character(group_louvain()))

# calculate modularity of louvain clustering
modularity <- decor_graph %>%
  activate(nodes) %>%
  mutate(modularity = graph_modularity(group = as.factor(louvain))) %>%
  pull(modularity) %>%
  head(1)
```

The louvain clustering has a modularity of `r modularity` .

```{r plot-clusters, fig.cap="Louvain communities within the decor network"}
# set plot colours for communities
communitycolor <-
  setNames(
    c("tomato", "cornflowerblue", "darkgreen"),
    unique(V(decor_graph)$louvain)
  )

# disable s2 geometry to plot on map
sf_use_s2(FALSE)

ggraph(decor_graph, x = lon, y = lat) +
  geom_sf(data = mali_adm0, fill = NA) +
  geom_sf(data = rivers, colour = "black") +
  edges_community +
  ggforce::geom_mark_hull(aes(x = lon, y = lat, fill = louvain, colour = louvain),
                          expand = unit(1.5, "mm"),
                          radius = unit(1.5, "mm")) +
  scale_fill_manual(values = communitycolor)+
  scale_colour_manual(values = communitycolor)+
  nodes_community +
  theme_community +
  coord_sf(xlim = c(-7, 2), ylim = c(13, 19), expand = FALSE) +
  theme_graph()
```

```{r setting up colors and multilevel community, hide = 'results', echo=FALSE}
## in the final form this part will be much shorter because the colors are set elsewhere
# we need a vector of colors that is named after the value we want to be plotted (community, name of the decor)
communitycolor <-
  setNames(
    c("tomato", "cornflowerblue", "darkgreen"),
    unique(V(decor_graph)$louvain)
  )

# because decor is longer we need to create a palette with the proper amount of colors (the whole code is still not perfect, but i think it is the proper way that can still be optimized)
col_21 <- colorRampPalette(brewer.pal(8, "Set2"))(21)

# then we create a vector containing all decortypes
decortypes <- decor_graph %>%
  activate(edges) %>%
  as_data_frame(what = "edges") %>%
  select(name) %>%
  unique()

decortypes <- as.vector(decortypes[, 1])

# finally we make a named vector of the colors
decorcolor <- setNames(col_21, decortypes)
```

There are some recent studies in archaeology that use community detection on archaeological networks. Radivojevic and Gruci (2018) have clustered sites on the Balkan based on the source of their copper, Mazucato (2019) has identified communities of building sites in Çatalhöyük to analyse the process of urbanisation, while Vargas and Lanata (2020) used community detection to find distinctive clusters of rock art sites in Patagonia. These studies differ significantly regarding the research object and question, but are similar in data structure and methodological approach. All three used the Louvain method to detect groups of sites that share a lot of edges with each other and have fewer edges with nodes outside of their community. The Louvain method is especially useful for archaeological datasets that typically contain few but heavily connected nodes (Blondel et al. 2008/ Yang et al. 2016).

Here, we employ the same clustering algorithm, which results in three communities with a modularity of 0.3. Community 1 contains 7 Sites in the Dogon Area. Community 2 is in the Niger bend and the IND, it also contains 7 sites. Both only include sites south of the Niger. Community 3 is the biggest and most widely distributed community. Reaching from Gao to Togu Missiri in the Ségu area it seems to mostly contain sites around the Niger river. The nodes of community 3 often connect with sites of community 1 and 2, while those communities are not connected strongly.

There are two major geographic features influencing the clusters. The first one is obviously the Niger river connecting the sites of community 3. The bend of the Niger river and the Inner Niger delta are the location of various sites of community 2 and probably a zone of contact between this community and community 3. The other geographic landmark that is strongly related to the communities is the Bandiagara Plateau. It is the border between community 1 and 2. Both communities overlap in this area. TMD, a site of community 3, lies in the north of the mountains making them the region where all 3 communities meet each other.

```{r cluster multilevel, echo=FALSE}
nodes <- decor_graph %>%
  activate(nodes) %>%
  as_tibble()
hull1 <- nodes %>%
  filter(louvain == 1) %>%
  slice(chull(lon, lat)) %>%
  rename(x = lon, y = lat)

hull2 <- nodes %>%
  filter(louvain == 2) %>%
  slice(chull(lon, lat)) %>%
  rename(x = lon, y = lat)

hull3 <- nodes %>%
  filter(louvain == 3) %>%
  slice(chull(lon, lat)) %>%
  rename(x = lon, y = lat)

# override s2 use for geographic plotting
sf_use_s2(FALSE)

ggraph(decor_graph, x = lon, y = lat) +
  # plot on map background
  bw.dem.map +
  # set edge parameters
  geom_edge_fan(color = "gray", aes(edge_alpha = weight, edge_width = weight)) +
  scale_edge_width(range = c(0.2, 2)) +
  scale_edge_alpha(range = c(0.3, 0.6)) +
  # colour nodes according to cluster and label
  geom_node_point(aes(color = louvain)) +
  scale_color_manual(values = communitycolor) +
  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
  # draw cluster hulls
  geom_polygon(data = hull1, aes(x = x, y = y), alpha = 0.5, fill = "cornflowerblue") +
  geom_polygon(data = hull2, aes(x = x, y = y), alpha = 0.5, fill = "tomato") +
  geom_polygon(data = hull3, aes(x = x, y = y), alpha = 0.5, fill = "darkgreen") +
  # set colour scales
  labs(color = "Community") +
  guides(color = guide_legend(override.aes = list(size = 0.5)), 
         alpha = "none", size = "none", edge_alpha = "none", 
         edge_width = guide_legend(override.aes = list(size = 0.3))) +
  ggtitle("Network with multilevel communities") +
  # set map extent
  coord_sf(xlim = c(-7, 2), ylim = c(13, 19), expand = FALSE) +
  theme_graph()
```

```{r relative edge weights and centralization by cluster}
# Relative edge weights (how important are the connections within the cluster)
rel_edge_weight <- vector("double", length(3))
for (i in 1:3) {
  rel_edge_weight[[i]] <- decor_graph %>%
    # filter community
    activate(nodes) %>%
    filter(louvain == i) %>%
    # relative edge weight (sum of weights / number of edges)
    activate(edges) %>%
    as.data.frame() %>%
    {
      sum(.$weight) / nrow(.)
    }
}

# Centralization of the cluster
centralization <- vector("double", length(3))
for (i in 1:3) {
  centralization[[i]] <- decor_graph %>%
    # filter community
    activate(nodes) %>%
    filter(louvain == i) %>%
    # degree centralization
    centr_degree(mode = "all", normalized = TRUE) %>%
    as_tibble() %>%
    pull(centralization) %>%
    head(1)
}

# put together
tibble(rel_edge_weight, centralization) %>%
  mutate(cluster = 1:3) %>%
  mutate(rel_edge_weight = rel_edge_weight / 100) %>%
  rename("relative edge weight" = rel_edge_weight) %>%
  pivot_longer(!cluster, names_to = "key", values_to = "value") %>%
  # plot
  ggplot(aes(x = cluster, y = value, fill = key)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_grey() +
  theme(axis.text.x = element_text(angle = 90)) +
  theme_publish()
```

```{r edges by cluster, eval = FALSE}

edges.com1 <- decor_graph %>%
  activate(nodes) %>%
  filter(louvain == 1) %>%
  activate(edges) %>%
  as_data_frame(what = "edges") %>%
  group_by(name) %>%
  summarise(n = sum(weight)) %>%
  filter(n > 0) %>%
  mutate(community = 1)

edges.com2 <- decor_graph %>%
  activate(nodes) %>%
  filter(louvain == 2) %>%
  activate(edges) %>%
  as_data_frame(what = "edges") %>%
  group_by(name) %>%
  summarise(n = sum(weight)) %>%
  filter(n > 0) %>%
  mutate(community = 2)

edges.com3 <- decor_graph %>%
  activate(nodes) %>%
  filter(louvain == 3) %>%
  activate(edges) %>%
  as_data_frame(what = "edges") %>%
  group_by(name) %>%
  summarise(n = sum(weight)) %>%
  filter(n > 0) %>%
  mutate(community = 3)

# edges.comcrossing <- decor_graph %>%
# activate(edges) %>%
#  mutate(crossing = igraph::crossing(louvain,.)) %>%
#  filter(crossing == TRUE) %>%
#  as_data_frame(what = "edges") %>%
# group_by(name) %>%
#  summarise(n = sum(weight)) %>%
#  filter(n > 0) %>%
#    mutate(community = "crossing")

plyr::rbind.fill(
  edges.com1, edges.com2, edges.com3
  # , edges.comcrossing
) %>%
  filter(n != 0) %>%
  mutate(community = as.factor(community)) %>%
  ggplot(aes(x = reorder(name, -n, sum), y = n, fill = community)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("tomato", "cornflowerblue", "darkgreen", "yellow")) +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_discrete(name = "decor type") +
  scale_y_continuous(name = "sum of the weight") +
  ggtitle("decors by clusters")

str.1 <- edges.com1 %>%
  select(n) %>%
  sum()

str.2 <- edges.com2 %>%
  select(n) %>%
  sum()

str.3 <- edges.com3 %>%
  select(n) %>%
  sum()

com.edges <- c(str.1 / nrow(edges.com1), str.2 / nrow(edges.com2), str.3 / nrow(edges.com3)) / 10000

library(tibble)
cent.1 <- decor_graph %>%
  filter(louvain == 1) %>%
  activate(nodes) %>%
  centr_degree() %>%
  as.tibble() %>%
  select(centralization) %>%
  transmute(centralization.weight = (centralization / nrow(.))) %>%
  slice(1)

cent.2 <- decor_graph %>%
  activate(nodes) %>%
  filter(louvain == 2) %>%
  centr_degree() %>%
  as.tibble() %>%
  select(centralization) %>%
  transmute(centralization.weight = (centralization / nrow(.))) %>%
  slice(1)

cent.3 <- decor_graph %>%
  activate(nodes) %>%
  filter(louvain == 3) %>%
  centr_degree() %>%
  as.tibble() %>%
  select(centralization) %>%
  transmute(centralization.weight = (centralization / nrow(.))) %>%
  slice(1)

library(tidyr)
community.edges <- bind_rows(cent.1, cent.2, cent.3) %>%
  add_column("sum of com.edges" = com.edges) %>%
  gather(key) %>%
  add_column(community = c(1:3, 1:3))

ggplot(community.edges, aes(x = community, y = value, fill = key)) +
  geom_bar(stat = "identity", position = "dodge") +
  # scale_fill_manual(values = c("tomato", "cornflowerblue", "darkgreen"))+
  theme(axis.text.x = element_text(angle = 90))
```
If we look at the edges (divided by 10000 to make it comparable) and the centrality of sites in regard to the communities, we see that while the sites of community 3 are the most central ones in the network the weight of the edges within the community is relatively small. For community 1 and 2 it is vice versa even though community 2 has weaker edges within and is better connected to the whole network. The more isolated a community is from the network, the stronger is its connection within the community. 
--maybe the bar_plots of the most common decors of all clusters is enough? maybe the first barplot is enough? – maybe the network of every cluster is too much
If we take a closer look at the "internal" connections of community 1, we see that "braided_strip_roulette" is present at all sites. The other main decor "straight_cord_mat" on the other hand is missing on 2 sites: Oudalan and Prom. The two decors are spread across the whole area of the community. The only other decor that is common in the community is the "twisted_cord_roulette" which is also abundant across the whole research area. It makes up a rather large part of the assemblage in Oudalan and Prom but was not found at the sites of Sadia and Ambere Dougon.
The fact that SAO has a high degree of centrality shows that the location is not decisive for the connectedness within this community. This is underlined by the fact that the less central sites are distributed across the whole area. There is an ambivalence between a relative consistency across the whole area and the diversity between sites that are close to each other. One reason for this phenom can be the contact to other communities in the mountain area. Damassogou exemplifies this as the site contains the defining decors of community 1 and 2. But the difference between Oudalan and SAO that are geographically isolated suggest that there might be more reasons for this diversity.
Community 2 is less diverse with "braided_cord_roulette" as the only outstanding decor type. The weaker connections of the sites Kokolo and Damassogou are based on data bias (the proportion is counted in regard to all sherds not only the decorated ones), the decor was probably equally abundant there. Other decors spread across the community are "folded_strip_roulette" and "dragged_comb". The first one is missing in KAN 4 and TOD H while the second one is missing at Kokolo and Damassogou. The other indicator for the homogeneous character of the community is the nearly equal degree of centrality of every site.
As mentioned above community 3 is more diverse but also has weaker connections. Instead of one defining decor there are several ones but none is abundant across the whole community. The spatial distribution differs between the decors. While "paint" and "twisted_cord_roulette" appear from Gao to TOG 2, "folded_strip_roulette", "channel" and "incised" are located in the western part of the community's area. Essouk and TMD have the highest degree of centrality. Even though it is located at the margin of the research area, its character as a "market"-site explains the amount of links it has. The role of TMD for the network is elaborated elsewhere (!).

```{r crossings}
decor_graph %>%
  activate(edges) %>%

  # find the edges that cross between communities
  mutate(crossing = community_crossing(weights = weight)) %>%
  filter(
    crossing == TRUE,
    weight > 200
  ) %>%
  activate(nodes) %>% 
  filter(!node_is_isolated()) %>%

  #plot
  ggraph(layout = "nicely") +

  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
  # scale_edge_color_manual(values = decorcolor) +
  scale_edge_width(range = c(0.2, 2)) +
  geom_node_point(aes(size = centrality_eigen(weights = weight), color = louvain), show.legend = FALSE) +
  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
  scale_color_manual(values = communitycolor)+

  theme_graph() +
  theme_community

```

```{r crossings between communities, eval = FALSE}

## this plot is hard to read but contains a lot of information, it shows only edges that connect the 2 communities. the color of the nodes shows their community while the size shows the betweenness centrality. This way we can show which sites are on the at the "frontier" of the communities that have strong connections to the other community. only edgest that are crossings between communities are shown, the widest ones are those that are those with an higher betweenness. in this way the graph shows the relation between the communities and which decor is shared/crossing community boundaries

nodes.c1 <- decor.graph %>%
  activate(nodes) %>%
  filter(communityl == 1) %>%
  as_data_frame(what = "vertices") %>%
  pull(id)

nodes.c2 <- decor.graph %>%
  activate(nodes) %>%
  filter(communityl == 2) %>%
  as_data_frame(what = "vertices") %>%
  pull(id)

nodes.c3 <- decor.graph %>%
  activate(nodes) %>%
  filter(communityl == 3) %>%
  as_data_frame(what = "vertices") %>%
  pull(id)

c1_3 <- decor.graph %>%
  activate(edges) %>%
  mutate(crossing = crossing(cl, .)) %>%
  filter(crossing == TRUE) %>%
  as_data_frame(what = "edges") %>%
  filter(from %in% c(nodes.c3) | to %in% c(nodes.c3)) %>%
  filter(from %in% c(nodes.c1) | to %in% c(nodes.c1)) %>%
  group_by(name) %>%
  summarise(n = sum(weight)) %>%
  filter(n > 0) %>%
  mutate(crossing = "c1_3")

c2_3 <- decor.graph %>%
  activate(edges) %>%
  mutate(crossing = crossing(cl, .)) %>%
  filter(crossing == TRUE) %>%
  as_data_frame(what = "edges") %>%
  filter(from %in% c(nodes.c3) | to %in% c(nodes.c3)) %>%
  filter(from %in% c(nodes.c2) | to %in% c(nodes.c2)) %>%
  group_by(name) %>%
  summarise(n = sum(weight)) %>%
  filter(n > 0) %>%
  mutate(crossing = "c2_3")

c1_2 <- decor.graph %>%
  activate(edges) %>%
  mutate(crossing = crossing(cl, .)) %>%
  filter(crossing == TRUE) %>%
  as_data_frame(what = "edges") %>%
  filter(from %in% c(nodes.c1) | to %in% c(nodes.c1)) %>%
  filter(from %in% c(nodes.c2) | to %in% c(nodes.c2)) %>%
  group_by(name) %>%
  summarise(n = sum(weight)) %>%
  filter(n > 0) %>%
  mutate(crossing = "c1_2")

plyr::rbind.fill(c1_2, c1_3, c2_3) %>%
  filter(n != 0) %>%
  mutate(crossing = as.factor(crossing)) %>%
  ggplot(aes(x = reorder(name, -n, sum), y = n, fill = crossing)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("tomato", "cornflowerblue", "darkgreen", "yellow")) +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_x_discrete(name = "decor type") +
  scale_y_continuous(name = "sum of the weight") +
  ggtitle("decors by clusters")
```

Not only the connections within but also those between the communities should be examined. Therefore, only the edges that "cross" the two communities will be considered. While there is no obvious "broker"-site, community 3 could be described as a "broker"-community. There are few strong crossings between community 1 and 2. Even though most decors that cross communities connect all three, those are mostly decors that are most abundant at the sites of community 3.

The crossings also show the limits of the community detection at the research area. The clusters that were found are congruent with the results of earlier works and help to understand the network, but they do not capture all the networks and connections of the region. To grasp the dynamics of pottery production we must go beyond those boundaries.

For an archaeological interpretation of the results, we need to take a step back and examine what our edges represent. If we take decoration techniques as the result of cultural practices, the communities we detect may be interpreted as communities of practice. Those are the result of the exchange of knowledge with various facets. Intergenerational teaching is one of them and might link our communities to ethnic groups (as Mayor did). But this is only one aspect and we should also consider communities as trading spheres where the demands influence the production. The exchange of goods is also fuelling the exchange of ideas that might cross ethnic boundaries. Hence our communities are foremost spheres of similar cultural knowledge that was intertemporally and -spatially exchanged.

Community 1 and 2 are spheres were the two decors respectively one decor dominates. The cultural practice was revolving around them. Even though there were other decors present at the sites, they might have played a local role but were not cons. Behind this might be a teaching network and demand, that at least is similar at the sites if not crossing their boundaries. The relatively weak links towards the outside show that these groups of practice were somewhat resilient. The difference regarding the centrality of the sites within community 1 demonstrates that the grade of involvement in this group of practice alter in this sphere.

Community 3 is another type of sphere cultural knowledge. Instead of a distinct decoration technique the variety of those techniques was crucial. The exchange of knowledge on the production side but also the choices of consumers vary strongly. While there are some decors that belong only to this exchange network, especially the sites in the west are part of other spheres of cultural practice.

Outcomes: three communities of different scales are identified. Two of these appear to be geographically bounded, and are maybe representative of intensive information exchange on a regional level, if not even within cultural groups. This ethno-linguistic dimension remains to be discussed re: Swiss research. The last community is not locally bounded and instead represents the exchange of information on ceramic decors through long-distance contacts. We might want to find out how well connected this network is in comparison to the two others. A lower network density or weight of ties would suggest that contacts are not as intensive.
membership in these networks is not mutually exclusive, there are crossing points in all of them. As our initial look at the network structure shows, the networks of each individual decor overlap, and these groups should not be seen as bounded entities, but rather as nodes with a higher propensity for sharing information. 
The next step then examines the pathways through which information is likely to have travelled.

# Sparsification

# Simplified network backbone

we can now try to extract a backbone from the simplified network

```{r flatten}
decor.flat <- flatten(decor_graph, weighted = TRUE)
```

extract the binary backbone of the flattened graph and play around with the alpha to get a good representation

```{r backbone extraction}
# extract binary backbones at decreasing levels of significance
decor.05 <- extract_backbone(decor.flat, 0.05)
decor.15 <- extract_backbone(decor.flat, 0.15)
decor.20 <- extract_backbone(decor.flat, 0.20)

# plot
p1 <- plot.bb(decor.05)
p2 <- plot.bb(decor.15)
p3 <- plot.bb(decor.20) + geom_node_text(aes(label = label), repel = TRUE)

(p1 | p2) / p3 + plot_annotation(tag_levels = "a")
```

The three communities from the previous section are also visible in the extracted backbone, albeit with some major differences in configuration. The site of Kokolo NE Phase 2 is placed in the closely connected cluster of sites in the IND and Lakes region, featuring the sites of the Jenné cluster and Toguére Galia, Toguére Doupwil, and Kawinza. The spread-out cluster of long-distance contact now also includes the sites at Dia, while the "Dogon" cluster remains largely intact. Akumbu is isolated from the network, and remains so until the significance is lowered to 0.2.

On this backbone, whose ties are no longer weighted, we can perform a number of network calculations that were meaningless in the raw network due to its large number of connections. We will begin by asking which nodes are the most influential in the network, i.e. possess the highest potential for spreading information. This, calculated through eigenvalue centrality, gives us a first idea of how pottery decor knowledge might have spread.

```{r backbone eigenvalue}
decor.15 <- decor.15 %>% activate(nodes) %>% filter(!node_is_isolated())

decor.15 %>%
  activate(nodes) %>%
  mutate(eigen = centrality_eigen()) %>%
  select(label, eigen) %>%
  arrange(desc(eigen)) %>%
  kable()

ggraph(decor.15, layout = "stress") +
  geom_node_point(aes(size = centrality_eigen())) +
  geom_edge_fan(color = "red", alpha = 0.8) +
  scale_edge_width(range = c(0.2, 3)) +
  geom_node_text(aes(label = label), repel = TRUE) +
  theme_graph()
```

The result shows that the highest influence on the network is exerted by the sites in the Eastern IND and Lakes cluster. We should not forget that the network is partial, i.e. many connections to many sites are missing, and that the central position of the IND has to do with our sample selection and research history. Nevertheless, it suggests this region specifically as a centre through which information and thus also innovation in pottery decoration was disseminated in this period. The other network clusters are more homogenous, at this time probably less diverse and more stable in their decors.

Betweenness centrality shows nodes in brokerage positions within the network. These are interesting to us in finding out where the different network cliques connect, and which sites are at crossroads positions in the pottery knowledge exchange. These are likely to be places with diverse markets, where producers belonging to different networks meet, and where information thus "jumps" between network cliques.

```{r backbone betweenness}
decor.15 %>%
  activate(nodes) %>%
  mutate(eigen = centrality_betweenness()) %>%
  select(label, eigen) %>%
  arrange(desc(eigen)) %>%
  kable()

ggraph(decor.15, layout = "stress") +
  geom_node_point(aes(size = centrality_betweenness())) +
  geom_edge_fan(color = "red", alpha = 0.8) +
  scale_edge_width(range = c(0.2, 3)) +
  geom_node_text(aes(label = label), repel = TRUE) +
  theme_graph()
```

We can see that Hambarketolo appears as the main gatekeeper to the Jenné site cluster, and to the eastern IND network clique, while Tongo Maaré Diabal, and to a lesser extent also Gao-Saney and Togu Missiri, stand in a brokerage role between the IND and the greater Niger cluster. The Dogon cluster exists at two removes from the IND, passing through Dammassogou and Ambéré-Dougon.At alpha = 0.2, this cluster is also connected to the large niger cluster via the Sangha "tellem" cave sites.

Also interesting is the edge betweenness value. This measures how important the edge between two nodes is for traversing the network. In our case it might indicate areas that were important axes of knowledge exchange.
```{r backbone edge betweenness}
ggraph(decor.15, layout = "stress") +
  geom_node_point() +
  geom_edge_fan(color = "red", alpha = 0.8, aes(width = centrality_edge_betweenness())) +
  scale_edge_width(range = c(0.2, 3)) +
  geom_node_text(aes(label = label), repel = TRUE) +
  theme_graph()
```

The edge betweenness measure shows a central path through the network, connecting the sites of Togu Missiri, Tongo Maaré Diabal, Hambarketolo, Damassogou, Ambéré-Dougon, and Sadia, as well as a prominent side route connecting Gao and Kawinza.These might roughly represent two routes of knowledge exchange: one important one passing through the eastern side of the Inland Niger Delta and below the Niger Bend, in which the Jenné site cluster plays a pivotal role as the meeting place between the three networks of knowledge exchange, and another, though less important to the network we have sampled, along the western side of the IND and north of the Niger Bend to Gao.

## Check for robustness of centrality measures

In order to know whether these centrality measures are valid, rather than random effects of the backbone sparsification, we implement a variation on Peeples' suggested method for estimating the robustness of centrality measures in archaeological networks. In order to do so, the edges of the original graph are sampled at 10 % intervals, from 10 to 90 %. The binary backbone of each of these samples is then extracted and the centrality measures calculated. A Spearman's rank correlation test is run between the samples to see how the deletion of edges influences the  centrality within the backbone. Figure \@ref(fig:MST-betweenness-validity) shows how there is a relatively significant correlation between the betweenness ranks of nodes in the original graph and the sampled graphs, all the way down to the 30% sample. Only on the 20% sample does the correlation statistic drop lower. We can take this to mean that, for our case, the betweenness centrality shown in the MST is dependable.

```{r validity test, warning = FALSE}

centrality.samples <- function(s, graph, centrality) {

  # get size of fraction to sample
  y <- (s / 100) * igraph::gsize(graph)

  # sample graph
  g <- graph %>%
    activate(edges) %>%
    tidygraph::sample_n(size = y) %>%
    flatten(weighted = TRUE) %>%
    as.igraph()

  # extract backbone
  bb <- backbone::disparity(g, alpha = 0.15, narrative = FALSE) %>%
    tidygraph::as_tbl_graph()

  # calculate and extract  centrality
  if (centrality == "eigen") {
    q <- bb %>%
      as_tbl_graph() %>%
      activate(nodes) %>%
      mutate(centr = centrality_eigen(weights = weight)) %>%
      as_tibble() %>%
      dplyr::select(centr) %>%
      dplyr::rename(!!paste0("s", s) := centr)
  }

  if (centrality == "betweenness") {
    q <- bb %>%
      as_tbl_graph() %>%
      activate(nodes) %>%
      mutate(centr = centrality_betweenness(weights = weight)) %>%
      as_tibble() %>%
      dplyr::select(centr) %>%
      dplyr::rename(!!paste0("s", s) := centr)
  }

  return(q)
}

spearman_samples <- function(graph, centrality) {

  # sample graph at 10 percent intervals
  samples <- seq(10, 100, 10)

  # iterate function over intervals
  centralities <-
    samples %>%
    purrr::map(centrality.samples, graph = graph, centrality = centrality) %>%
    dplyr::bind_cols()

  # 2. Spearman's rank correlation

  # get sample names
  cols <- centralities %>%
    select(-s100) %>%
    names()

  # rank correlation function
  rank.correlation <- function(x) {
    cor.test(centralities$s100, centralities[[x]], method = "spearman")
  }

  # iterate for samples
  purrr::map(cols, rank.correlation) %>%
    purrr::map_df(`[`, c("estimate")) %>%
    pull(estimate)
}

test_centralities <- function(graph, centrality, reps) {

  # check for correct arguments
  if (centrality %in% c("eigen", "betweenness") == F) {
    stop("centrality must be 'eigen' or 'betweenness'")
  }

  # get sample names
  s <- seq(10, 90, 10)
  samples <- paste("s", s, sep = "")

  # replicate the test
  replicate(reps, spearman_samples(graph, centrality)) %>%
    as_tibble() %>%
    mutate(sample = samples) %>%
    pivot_longer(!sample, names_to = "iteration", values_to = "value") %>%
    select(-iteration) %>%
    group_by(sample) %>%
    summarise(mean_rho = mean(value)) %>%
    # plot
    ggplot() +
    geom_histogram(aes(x = sample, y = mean_rho), stat = "identity") +
    labs(y = expression(paste("mean Spearman's ", rho))) +
    ylim(0, 1) +
    theme_publish()
}
```

```{r apply validity test, warning = FALSE, message = FALSE}
p1 <- test_centralities(decor_graph, "eigen", 100)
p2 <- test_centralities(decor_graph, "betweenness", 100)

(p1 | p2) + plot_annotation(tag_levels = "a")
```

eigenvector centrality is fine, betweenness less impressive. 
For edge betweenness, the validation is much harder, as the flattening procedure means that edges cannot reliably be numbered, and thus correlation is hard to achieve. We can, however, manually inspect the result.

```{r warning = FALSE, message=FALSE}
bb.edge.betw <- function(x) {

  # sample graph
  g <- decor_graph %>%
    activate(edges) %>%
    tidygraph::sample_n(size = x) %>%
    flatten(weighted = TRUE) %>%
    as.igraph()

  # extract backbone
  bb <- backbone::disparity(g, alpha = 0.15, narrative = FALSE) %>%
    tidygraph::as_tbl_graph()

  # put node attributes back
  decor.nodes <- decor_graph %>%
    activate(nodes) %>%
    as_tibble()
  bb <- bb %>%
    activate(nodes) %>%
    mutate(label = decor.nodes$label, lat = decor.nodes$lat, lon = decor.nodes$lon)

  # plot
  ggraph(bb, layout = "stress") +
    geom_node_point() +
    geom_edge_fan(color = "red", alpha = 0.8, aes(width = centrality_edge_betweenness())) +
    scale_edge_width(range = c(0.2, 3)) +
    geom_node_text(data = . %>% filter(centrality_betweenness(weights = weight) > mean(centrality_betweenness(weights = weight))), aes(label = label), repel = TRUE) +
    theme_graph() +
    theme(legend.position = "none")
}

p1 <- bb.edge.betw(90) + ggtitle("90%")
p2 <- bb.edge.betw(80) + ggtitle("80%")
p3 <- bb.edge.betw(70) + ggtitle("70%")
p4 <- bb.edge.betw(60) + ggtitle("60%")

p1 + p2 + p3 + p4
```
What is clear is that the backbone is, predictably, quite sensitive to random edge deletion, and therefore edge_betweenness is very much dependent on whether an edge exists or not. For the purposes of generating hypotheses for archaeological testing, we have still found this method superior to other methods of finding paths through networks, such as an inverted version of Kruskal's minimum spanning tree (for details and code, see Annex)

## Spanning tree

We can use Kruskal's concept of the minimum spanning tree, simply transformed to a maximum spanning tree (MST) here. This method has its drawbacks, as described by [@Serranoetal]:

>*"One of the big limitations of this method is that spanning trees are by construction acyclic. This means that reduced networks obtained by this algorithm are overly structural simplifications that destroy local cycles, clustering coefficient, and the clustering hierarchies often present in real world networks." *

In our case, however, we can use the maximum spanning tree especially to gain access to a measure that is not accessible or meaningful in hairball networks: betweenness centrality. This is particularly interesting in this case, as high betweenness centrality in a network indicates a brokerage position. If we would like to know which places within the network are particularly well placed to pass on information, we can find out using the MST.

```{r MST, fig.cap="The maximum spanning tree of the Niger pottery decor network, showing edge weights and betweenness centrality"}
# invert edge weights
g <- decor_graph %>%
  activate(edges) %>%
  mutate(weight = 0 - weight) %>%
  as.igraph()

# create MST and reconvert edge weights
mst <- mst(g, weights = E(g)$weight) %>%
  as_tbl_graph() %>%
  activate(edges) %>%
  mutate(weight = 0 - weight)

# plot
mst %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(alpha = weight), colour = "blue") +
  geom_node_point(aes(size = centrality_betweenness(weights = weight))) +
  # geom_node_text(data = . %>% filter(centrality_betweenness(weights = weight) > mean(centrality_betweenness(weights = weight))), aes(label = label), repel = TRUE)+
  geom_node_text(aes(label = label), repel = TRUE) +
  theme_graph()
```

## Check for robustness of betweenness measures

XXX: This needs to be done for the backbone now, instead of the spanning tree.

In order to know whether these centrality measures are valid, rather than random effects of the MST transformation, we implement a variation on Peeples' suggested method for estimating the robustness of centrality measures in archaeological networks. In order to do so, the edges of the original graph are sampled at 10 % intervals, from 10 to 90 %. Each of these samples is then converted to a maximum spanning tree and the betweenness measures are extracted. A Spearman's rank correlation test is run between the samples to see how the deletion of edges influences the betweenness centrality within the MST. Figure \@ref(fig:MST-betweenness-validity) shows how there is a relatively significant correlation between the betweenness ranks of nodes in the original graph and the sampled graphs, all the way down to the 30% sample. Only on the 20% sample does the correlation statistic drop lower. We can take this to mean that, for our case, the betweenness centrality shown in the MST is dependable.

```{r MST-betweenness-validity,fig.cap="Spearman's rank-correlation coefficient between the MST of the original graph and those sampled at 10 to 90 percent."}
# 1. calculate betweenness centrality of sampled graphs
set.seed(6789)
# sample graph at 10 percent intervals
samples <- seq(10, 100, 10)

centrality.samples <- function(s) {

  # get size of fraction to sample
  y <- (s / 100) * igraph::gsize(decor_graph)

  # sample graph and invert weights
  g <- decor_graph %>%
    activate(edges) %>%
    tidygraph::sample_n(size = y) %>%
    mutate(weight = 0 - weight) %>%
    as.igraph()

  # create weighted MST
  mst <- mst(g, weights = E(g)$weight)

  # reconvert weights, calculate and extract betweenness centrality
  q <- mst %>%
    as_tbl_graph() %>%
    activate(edges) %>%
    mutate(weight = 0 - weight) %>%
    activate(nodes) %>%
    mutate(betw = centrality_betweenness(weights = weight)) %>%
    as_tibble() %>%
    dplyr::select(betw) %>%
    dplyr::rename(!!paste0("s", s) := betw)
}

# iterate function over intervals
centralities <- samples %>%
  purrr::map(centrality.samples) %>%
  dplyr::bind_cols()

# 2. Spearman's rank correlation

# get sample names
cols <- centralities %>%
  select(-s100) %>%
  names()

# rank correlation function
rank.correlation <- function(x) {
  cor.test(centralities$s100, centralities[[x]], method = "spearman")
}

# iterate for samples
rho <- purrr::map(cols, rank.correlation)

# 3. plot
rho %>%
  purrr::map_df(`[`, c("estimate", "p.value")) %>%
  mutate(sample = cols) %>%
  ggplot() +
  geom_histogram(aes(x = sample, y = estimate), stat = "identity") +
  labs(y = expression(paste("Spearman's ", rho))) +
  ylim(0, 1) +
  theme_minimal()
```

# Diachronic analysis

## NB clean this part and update the site list.

```{r import early sites}
library(readr)
decorsearly <- read_csv2(here::here("analysis/data/raw_data/decors_early.csv"))

# apply function to transform frequency table into edge list
edgelist <- make_edgelist(decorsearly)

# construct network, remove extraneous sites
decor_graph_early <- tbl_graph(sites, edgelist, directed = FALSE) %>%
  activate(nodes) %>%
  filter(!node_is_isolated())

decor_graph_early
```

## Backbone extraction

```{r extract early backbone}
# flatten with weights
early_flat <- flatten(decor_graph_early, weighted = TRUE)

library(backbone)
# extract backbone
early.flat.bb <- backbone::disparity(as.igraph(early_flat), alpha = 0.31, narrative = TRUE) %>%
  tidygraph::as_tbl_graph()

# igraph transformation loses the node attributes, so we put them back
early.flat.nodes <- early_flat %>%
  activate(nodes) %>%
  as_tibble()
early.flat.bb <- early.flat.bb %>%
  activate(nodes) %>%
  mutate(label = early.flat.nodes$label, lat = early.flat.nodes$lat, lon = early.flat.nodes$lon)

ggraph(early.flat.bb, layout = "stress") +
  geom_node_point() +
  geom_edge_fan(color = "red", alpha = 0.8) +
  scale_edge_width(range = c(0.2, 3)) +
  geom_node_text(aes(label = label), repel = TRUE) +
  theme_graph()
```

This backbone looks significantly different from the later one, though we still distinguish an IND group, a very clear Dogon group, and one that includes "others": Kawinza, TMD, Tiebala, Akumbu. Essouk is isolated. Let's try this with louvain communities.

```{r cluster early}
decor_graph_early <- decor_graph_early %>%
  activate(nodes) %>%
  mutate(louvain = as.character(group_louvain()))

# calculate modularity of louvain clustering
modularity <- decor_graph_early %>%
  activate(nodes) %>%
  mutate(modularity = graph_modularity(group = as.factor(louvain))) %>%
  pull(modularity) %>%
  head(1)

ggraph(decor_graph_early, x = lon, y = lat) +
  geom_sf(data = mali_adm0, fill = NA) +
  geom_sf(data = rivers, colour = "black") +
  edges_community +
  ggforce::geom_mark_hull(aes(x = lon, y = lat, fill = louvain, colour = louvain),
                          expand = unit(1.5, "mm"),
                          radius = unit(1.5, "mm")) +
  scale_fill_manual(values = communitycolor)+
  scale_colour_manual(values = communitycolor)+
  nodes_community +
  theme_community +
  coord_sf(xlim = c(-7, 2), ylim = c(13, 19), expand = FALSE) +
  theme_graph()
```

From these, the hypothesis: The Dogon network in the early period is very densely connected and more isolated from the Niger network. The Niger network on the other hand, is more diffuse, harder to cluster, maybe more generalised.
How to calculate this?

```{r}
# Relative edge weights (how important are the connections within the cluster)
rel_edge_weight_early <- vector("double", length(3))
for (i in 1:2) {
  rel_edge_weight_early[[i]] <- decor_graph_early %>%
    # filter community
    activate(nodes) %>%
    filter(louvain == i) %>%
    # relative edge weight (sum of weights / number of edges)
    activate(edges) %>%
    as.data.frame() %>%
    {
      sum(.$weight) / nrow(.)
    }
}

# Centralization of the cluster
centralization_early <- vector("double", length(3))
for (i in 1:2) {
  centralization_early[[i]] <- decor_graph_early %>%
    # filter community
    activate(nodes) %>%
    filter(louvain == i) %>%
    # degree centralization
    centr_degree(mode = "all", normalized = TRUE) %>%
    as_tibble() %>%
    pull(centralization) %>%
    head(1)
}

# put together
early <- tibble(rel_edge_weight, centralization) %>%
  mutate(cluster = 1:3) %>%
  mutate(rel_edge_weight = rel_edge_weight / 100) %>%
  rename("relative edge weight" = rel_edge_weight) %>%
  pivot_longer(!cluster, names_to = "key", values_to = "value") |> 
  ggplot(aes(x = cluster, y = value, fill = key)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_grey() +
  theme(axis.text.x = element_text(angle = 90)) +
  theme_publish()

late <- tibble(rel_edge_weight_early, centralization_early) %>%
  mutate(cluster = 1:2) %>%
  mutate(rel_edge_weight_early = rel_edge_weight_early / 100) %>%
  rename("relative edge weight early" = rel_edge_weight_early) %>%
  pivot_longer(!cluster, names_to = "key", values_to = "value") |>

  # plot
  ggplot(aes(x = cluster, y = value, fill = key)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_grey() +
  theme(axis.text.x = element_text(angle = 90)) +
  theme_publish()

early + late
```

Clustering Coefficient: "Transitivity measures the probability that the adjacent vertices of a vertex are connected. This is sometimes also called the clustering coefficient."
Diameter
Average Distance

```{r}
# Transitivity(clustering coefficient)

igraph::transitivity(decor_graph_early)
igraph::transitivity(decor_graph)

# density - doesn't work for multiple graphs
igraph::edge_density(decor_graph)
igraph::edge_density(decor_graph_early)

# Diameter
igraph::diameter(decor_graph,
  directed = FALSE,
  # weights = E(decor_graph)$weight
  weights = NA
)
igraph::diameter(early_flat,
  directed = FALSE,
  # weights = E(decor_graph_early)$weight
  weights = NA
)
```
Transitivity and density: The earlier network has a higher transitivity and lower density than the later one (though in both cases, not by much). We can take this to mean that groups in the earlier network are marginally stronger, and that it is less closely connected overall. This reflect, for instance, the strength of the Bandiagara group and the exposed position of Essouk in this dataset.How does this change when the networks are simplified?
There is a big difference in the diameter of the networks, with the earlier one at 237 and the later one at 13. What does this mean? Is it true? The weights make this difficult, don't they?

Following Wasserman and Faust, p.180-183, we report the degree centralization of the graph, along with its average degree.
```{r centralization}

flat <- decor_graph %>% flatten(weighted = TRUE)
igraph::centr_eigen(flat)

flatearly <- decor_graph_early %>% flatten(weighted = TRUE)
igraph::centr_degree(flatearly)

# Transitivity(clustering coefficient)

igraph::transitivity(decor_graph_early)
igraph::transitivity(decor_graph)

# density - doesn't work for multiple graphs
igraph::edge_density(flat)
igraph::edge_density(flatearly)
```

Mean distance

```{r}
# mean distance
igraph::mean_distance(decor_graph, directed = FALSE, weights = NA)
igraph::mean_distance(decor_graph_early, directed = FALSE, weights = NA)

# Dogon cluster only
decor_graph %>%
  activate(nodes) %>%
  filter(louvain == 2) %>%
  igraph::mean_distance(., directed = FALSE, weights = NA)

decor_graph_early %>%
  activate(nodes) %>%
  filter(louvain == 2) %>%
  igraph::mean_distance(., directed = FALSE, weights = NA)
```
If we are to use weighted distances, then we should probably negate the values.https://ona-book.org/paths-distance.html . It might be more appropriate to use unweighted distances.

```{r}
# Modularity:
# how high is the modularity using our louvain clusters?

decor_graph %>% graph_modularity(weights = weight)
```

We can also try global measures for centrality. These should be distributions that are compared.

```{r compare centrality distributions}
pull_centralities <- function(graph) {
  graph %>%
    activate(nodes) %>%
    mutate(deg = centrality_degree(weights = weight), eig = centrality_eigen(weights = weight)) %>%
    select(deg, eig) %>%
    as_tibble()
}

early <- pull_centralities(decor_graph_early) %>% mutate(period = "early")
# %>% mutate(eig = eig/n(), deg = deg/n())
late <- pull_centralities(decor_graph) %>% mutate(period = "late")
# %>% mutate(eig = eig/n(), deg = deg/n())

centralities <- bind_rows(early, late)

a <- ggplot(centralities) +
  geom_boxplot(aes(x = period, y = deg, fill = period)) +
  theme_minimal() +
  ylab("degree centrality")

b <- ggplot(centralities) +
  geom_boxplot(aes(x = period, y = eig, fill = period)) +
  theme_minimal() +
  ylab("eigenvector centrality")

a + b

centralities %>%
  group_by(period) %>%
  summarise(meandeg = mean(deg), meaneig = mean(eig))
```
Degree centrality is clearly lower in the early period, suggesting a less well-connected network. This is an effect of an increasing diversification of decorative techniques. The eigenvector centrality, on the other hand, is relatively equal, but slightly higher and with a lower range in the early period, suggesting that influence on the network was somewhat more evenly distributed. When these values are adjusted for the number of nodes in each network, things change: degree centrality is fairly even, but eigenvector centrality drops drastically in the later period, reflecting a more uneven distribution of influence on the growing network.

Secondly, we can use netdis as a graphlet-based approach.
Work with this: https://github.com/alan-turing-institute/network-comparison/blob/master/vignettes/quickstart_netdis_2graphs.Rmd

```{r}
library(netdist)
# flatten graphs
decor_early_flat <- flatten(decor_graph_early)
decor_flat <- flatten(decor_graph)

# set netdis parameters
# Maximum graphlet size to calculate counts and netdis statistic for.
max_graphlet_size <- 4
# Ego network neighbourhood size
neighbourhood_size <- 2
# Minimum size of ego networks to consider
min_ego_nodes <- 3
min_ego_edges <- 1
# Ego network density binning parameters
min_bin_count <- 5
num_bins <- 100

## Generate ego networks

# Get ego networks for query graphs and reference graph
ego_1 <- make_named_ego_graph(decor_early_flat,
  order = neighbourhood_size,
  min_ego_nodes = min_ego_nodes,
  min_ego_edges = min_ego_edges
)
ego_2 <- make_named_ego_graph(decor_flat,
  order = neighbourhood_size,
  min_ego_nodes = min_ego_nodes,
  min_ego_edges = min_ego_edges
)

# Count graphlets for ego networks in query and reference graphs
graphlet_counts_1 <- ego_to_graphlet_counts(ego_1, max_graphlet_size = max_graphlet_size)
graphlet_counts_2 <- ego_to_graphlet_counts(ego_2, max_graphlet_size = max_graphlet_size)

# Load reference graph
ref_path <- system.file(file.path("extdata", "random", "ER_1250_10_1"),
  package = "netdist"
)
ref_graph <- read_simple_graph(ref_path, format = "ncol")
ego_ref <- make_named_ego_graph(ref_graph,
  order = neighbourhood_size,
  min_ego_nodes = min_ego_nodes,
  min_ego_edges = min_ego_edges
)
graphlet_counts_ref <- ego_to_graphlet_counts(ego_ref, max_graphlet_size = max_graphlet_size)
# Scale ego-network graphlet counts by dividing by total number of k-tuples in
# ego-network (where k is graphlet size)
scaled_graphlet_counts_ref <- scale_graphlet_counts_ego(
  graphlet_counts_ref,
  max_graphlet_size
)
# Get ego-network densities
densities_ref <- ego_network_density(graphlet_counts_ref)
# Adaptively bin ref ego-network densities
binned_densities <- binned_densities_adaptive(densities_ref,
  min_counts_per_interval = min_bin_count,
  num_intervals = num_bins
)
ref_ego_density_bins <- binned_densities$breaks
# Average ref graphlet counts across density bins
ref_binned_graphlet_counts <- mean_density_binned_graphlet_counts(
  scaled_graphlet_counts_ref,
  binned_densities$interval_indexes
)

# Calculate expected graphlet counts (using ref graph ego network density bins)
exp_graphlet_counts_1 <- netdis_expected_counts(graphlet_counts_1,
  ref_ego_density_bins,
  ref_binned_graphlet_counts,
  max_graphlet_size,
  scale_fn = count_graphlet_tuples
)
exp_graphlet_counts_2 <- netdis_expected_counts(graphlet_counts_2,
  ref_ego_density_bins,
  ref_binned_graphlet_counts,
  max_graphlet_size,
  scale_fn = count_graphlet_tuples
)
# Centre graphlet counts by subtracting expected counts
centred_graphlet_counts_1 <- netdis_subtract_exp_counts(
  graphlet_counts_1,
  exp_graphlet_counts_1,
  max_graphlet_size
)
centred_graphlet_counts_2 <- netdis_subtract_exp_counts(
  graphlet_counts_2,
  exp_graphlet_counts_2,
  max_graphlet_size
)

## Sum centred graphlet counts across all ego networks
sum_graphlet_counts_1 <- colSums(centred_graphlet_counts_1)
sum_graphlet_counts_2 <- colSums(centred_graphlet_counts_2)

## Calculate netdis statistics
netdis_result <- netdis_uptok(
  sum_graphlet_counts_1,
  sum_graphlet_counts_2,
  max_graphlet_size
)

print(netdis_result)
```
The result shows us the difference between the two networks using three- and four-node graphlets. The result shows that the three-node graphlet distance is far lower than the four-node one. What does it mean?

We could also see whether the clusters are still there and how the crossings change.

Graph centralization:
Wasserman & Faust p.180-183

### 2023-08-03 analysis/paper/paper.Rmd
Initial commit

--- 2023-08-03 analysis/paper/data analysis.Rmd
@@
 ---
-title: "analysis"
-author: "Nikolas Gestrich"
-date: '2022-05-02'
-output: html_document
+title: "Pottery decor as networks on the Middle Niger"
+author:
+  - Nikolas Gestrich:
+      email: gestrich@uni-frankfurt.de
+      institute: [FI]
+      correspondence: true
+  - Juan-Marco Puerta-Schardt:
+      email: jmpuertaschardt@outlook.de
+      institute: [GU]
+      correspondence: false
+institute:
+  - FI: Frobenius Institute
+  - GU: Goethe University
+date: "`r format(Sys.time(), '%d %B, %Y')`"
+output: 
+    bookdown::word_document2:
+      fig_caption: yes
+      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
+      #pandoc_args:
+      #- --lua-filter=../templates/scholarly-metadata.lua
+      #- --lua-filter=../templates/author-info-blocks.lua
+      #- --lua-filter=../templates/pagebreak.lua
+bibliography: references.bib
+csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+abstract: |
+  Text of abstract
+keywords: |
+  keyword 1; keyword 2; keyword 3
+highlights: |
+  These are the highlights. 
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-
-options(knitr.kable.NA = "")
-
+<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->
+
+<!-- With the following code you can access and display values from the yml header above. -->
+
+Keywords: `r rmarkdown::metadata$keywords`
+
+Highlights: `r rmarkdown::metadata$highlights`
+
+<!-- The following code chunk defines some general settings how code chunks should behave. -->
+
+```{r knitr-setup, echo = FALSE}
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  warning = FALSE,
+  message = FALSE,
+  echo = FALSE,
+  comment = "#>",
+  fig.path = "../figures/",
+  dpi = 300
+)
+```
+
+```{r libraries, echo = FALSE}
 # libraries
@@
 
-## Data import
-We are importing data on pottery decors from a variety of sources. All of the data represents phases of sites that are dated to have been occupied at some point in the 11th century AD.
-
-XXX: put together the different periods. 
-Update: or maybe not, so that we can name the nodes the same
-
-```{r}
+<!-- The actual document text starts here: -->
+
+# Introduction
+
+For much of the history of the archaeological discipline, ceramics have been used as markers of similarity and difference.In many areas and periods of the past, they are the most ubiquitous archaeological find category, and have thus contributed greatly to the recognition of archaeological artefact groups, or archaeological cultures.
+It is not rare to see find complexes and periods named after particular ceramic types.
+Though the current scholarly consensus is that archaeological cultures do not represent social groups as experienced by past humans, and that pots do not equal people, a large amount of work on archaeological ceramics is engaged in grouping ceramics and distinguishing groups from one another.
+There are a number of approaches for doing so and the rationales behind them vary.
+The basis for the creation of archaeological cultures, especially current in works of the first two thirds of the twentieth century, is the 'traditional' typological approach to ceramics and other artefacts.
+This approach is nowadays most used in the best-explored regions with the longest research traditions, and groups ceramics mainly according to outward appearances, such as shape and surface treatment.
+The distribution, both spatially and temporally, of such types then becomes the subject of research, which subsequent studies can augment through the recognition of sub-types and variants.
+A different approach takes as its basis the sequence of manufacture of artefacts, the *chaîne opératoire*.
+This approach proposes that the manufacture of an artefact involves a series of choices constrained by certain outside factors.
+These choices are seen as being cultural, so that a similar set of choices under similar constraints are markers of shared cultural knowledge.
+Here, the archaeological research goal also lies in the identification of spheres of cultural commonality which can be distinguished from one another.
+
+In the archaeology of West Africa, and especially in the wider areas surrounding the upper and middle reaches of the Niger, there has been a strong research focus on ceramics.
+From the 1st millennium BC onwards, they occur in extremely high numbers on archaeological sites, and their appearance, especially their surface treatment, is highly variable.
+In contrast to the Great Lakes region of Eastern Africa, for example, the number of different roulette decorations, painted, incised, and applied decors, but also the variety in form, is very high.
+Consequently, there has been considerable emphasis on ceramics in archaeological and also in ethnoarchaeological work.
+While much of the ethnoarchaeological work has been based on exploring *chaînes opèratoires*, the archaeological work has mainly relied on attribute-based analyses (the system employed by S.K.
+McIntosh at Jenné-jeno has been very influential in subsequent work).
+In a region where large archaeological sites abound and researchers are relatively thin on the ground, the main hope of the analyses usually is to create chrono-stratigraphical sequences in order to be able to give approximate periods to surface assemblages.
+Despite all the variability, it has been found that a great number of ceramic attributes are not particularly time-sensitive.
+On the contrary, most sites appear to have diverse assemblages in which the percentages of individual attributes vary, but persist over several centuries, sometimes millennia.
+
+Some researchers have used the attribute-based approach to define types, statistically derived for the most part, which typify the ceramics of a certain place of time (MacDonald's Facies, McIntosh's timeline).
+The spatiotemporal distribution of individual attributes has also been the subject of some important work, such as individual rouletted decors (Livingstone-Smith) or base-shaping techniques (Huysecom).
+The most comprehensive treatment so far is in the work of Anne Mayor, combining both shaping techniques and decorative practices from an archaeological and ethnoarchaeological point of view.
+Focused on the Bandiagara highlands in central Mali, Mayor uses her data to construct an impressive overview of the development of different techniques and the establishment of ceramic traditions, which she closely relates to ethno-linguistic groups.
+Mayor's work builds on a large body of data and decades of ethnoarchaeological fieldwork, and there is little to add to her conclusions.
+Here, however, we wish to show that an almost identical dataset as available to Mayor can generate quite a different perspective when analysed as a network, and when connectivity is emphasised over boundaries.
+
+# Archaeological ceramics as networks
+
+In recent years, the use of network analysis in archaeology has been growing significantly.
+A large number of these studies focus on ceramics, but they are geographically mainly limited to North America (e.g.
+Hart 2012, Chiu 2019, Barker and Young 2017, Crabtree 2015, Mills et al.
+2013, Peeples and Haas 2013) and the Mediterranean (e.g.
+Sternini 2019, Iacono 2016, Brughmans and Poblome 2016 torrence & Swadling 2008, de Groot 2019).
+The archaeological records of these regions are amongst the best studied in the world, with fine-grained pottery typologies and extensive excavations.
+To our knowledge, this is the first application of a network approach to archaeological material from sub-Saharan Africa, where the conditions are somewhat different.
+
+Ethnographic studies of pottery manufacturing in Africa suggest that the way that pottery is made depends largely on systems of learning and teaching the craft (Gosselain), on local market consumer choices (Dietler & Herbich), or on marriage patterns amongst manufacturing groups (ibid.).
+In short, the processes underlying the propagation of technical practices through time and across geographical space are dependent on communication and, more specifically, knowledge exchange.
+Such exchange occurs between humans and we can therefore understand similarities in the *chaine opératoire* as evidence that knowledge was shared.
+However, different elements of the pottery making process are thought to be reflective of different kinds of social interaction.
+Gosselain (REF), whose research in the years between XX and XX explored this question, concludes that, on the whole, forming techniques are more stable, reflective of durable cultural ties, sometimes even ethnic groups.
+On the other end of the spectrum, decorations are thought to be influenced by far more ephemeral contact, in which seeing someone else's pot might be enough to change a potter's production, as could indirect contact via the wishes of customers.
+
+SNA shifts the focus "from spatial relations to social relations" (cf. Knappett 2011)
+Mills et al.
+(2015) have been able to compare similarity in wares across sites, but this cannot be done here.
+They also highlight Knappet's (2011) idea of different scales in archaeological networks.
+Because of the specific use case scenario here, we are conflating two fo his scales.
+He sees the community of practice as a meso-scale effect, across maybe a few households.
+However, we can see in the archaeological ceramic data from Mali that communities of practice in individual stages of the *chaîne operatoire* is often much larger.
+Thus, certain parts of the pottery manufacturing sequence are distributed across larger or smaller networks, but rarely remain at the household level.
+Also, we do not really have the luxury of being able to pinpoint certain wares, since typology is not as established.
+
+We need to be aware of the fact that we are conflating consumption with production here.
+This is why we will need to threshold the networks, to make sure that a site represents the significant potting communities within its area.
+
+# Methods
+
+```{r import 11c decor data, echo = FALSE}
 library(readr)
@@
 
-show the data:
-
-```{r fig.cap="decor data used for network construction"}
+```{r datatable, fig.cap="decor data used for network construction"}
 library(kableExtra)
-kbl(decors) |> 
-  kable_styling(font_size = 7, html_font = "Cambria") 
-```
-
-show the map points:
-
-```{r fig.cap="location of sites in the analysis", warning=FALSE}
-source(here::here("analysis/R/map plots.R"))
-
-sitepoints <- st_as_sf(sites, coords = c("lon", "lat"), crs = 4326, agr = "constant")
-
-sitemap <- ggplot() +
-  map_bw_grid +
-  geom_sf(data = sitepoints)
-
-sitemap + inset_element(inset, 0.72, 0.72, 1, 1)
-```
-
+kable(decors, digits = 1, "html") %>% 
+  kable_styling("condensed") %>% 
+  save_kable(here::here("analysis/figures/datatable.html"))
+```
+
+In this paper, we use pottery decor data from contexts occupied in the 11th century AD to attempt a reconstruction of networks through which pottery decorative practices were shared.
+The data is in a common format of percentages of prevalence in the assemblage (see \@ref(tab:datatable) This particular time slice, which represents a cumulative view of a great many individual interactions, will then be compared to earlier contexts in the same wider area, in order to demonstrate the diachronic possibilities of the approach.The sites we have been able to include in the network model are shown in \@ref(fig:sitemap).
 We can see here that sites are geographically unequally distributed.
 This is due to research history as well as past cultural dynamics.
-Certainly, the IND and the areas bordering it (notiably the Lakes region) were strongly populated at the time, maybe more so than other areas.
+Certainly, the IND and the areas bordering it (notably the Lakes region) were strongly populated at the time, maybe more so than other areas.
 The cluster of sites around the Bandiagara escarpment is due to the extraordinarily detailed and long-term research of the University of Geneva's research teams.
@@
 
-# Construct network
-
-```{r}
+```{r sitemap, fig.cap="location of sites in the analysis", echo = FALSE}
+source(here::here("analysis/R/map plots.R"))
+
+sitepoints <- st_as_sf(sites, coords = c("lon", "lat"), crs = 4326, agr = "constant")
+
+sitemap <- ggplot()+
+  map_bw_grid +
+  geom_sf(data = sitepoints)
+
+sitemap + inset_element(inset, 0.72, 0.72, 1, 1)
+```
+
+The network is constructed from standard archaeological frequency data, in which the proportions of decorative practices are represented across the selected site phases (REF FIG XXX).
+The decorative practices encode both the tool used and the action performed, for instance a twisted cord (tool) roulette (action).
+This data is transformed into a network by deciding that every decor that is shared between two sites leads to a link between these sites.
+The link is weighted by multiplying the proportion at one site with the proportion at the other.
+This leads to a network in which the nodes (sites) can be connected in more than one manner, known as a multiplex or multilayer network.
+As is shown in \@ref(fig:total-network), this leads to a very tightly connected network, known as a "hairball".
+However, the multiplex nature of the network allows us to make an important initial observations on its structure: the networks in which individual decor types are propagated are overlapping.
+There are no two decor networks that cover one another entirely.
+This emerges more clearly in \@ref(fig:total-network-sample), when the number of decorative practices is reduced.
+This indicates that the adoption of decorative techniques ...
+
+```{r make decor network, echo = FALSE}
 # import functions for network construction
@@
   filter(!node_is_isolated())
-
-decor_graph
-```
-
-# Plot network
-
-```{r total network, fig.cap="Plot of 11th century AD network of decorative practices. The plot shows the very strongly connected structure of the network."}
-ggraph(decor_graph) +
+```
+
+```{r total-network, fig.cap="Plot of 11th century AD network of decorative practices. The plot shows the very strongly connected structure of the network."}
+ggraph(decor_graph, layout = "stress") +
   coloured_edges +
@@
 
-## Network structure
-
-High number of connections due to the construction of the network.
-This network is very much unsuitable for further network analysis, but it already allows us to see one major thing about pottery decor interactions: the networks in which individual decor types are propagated are overlapping.
-There are no two decor networks that cover one another entirely.
-
-```{r total network sample, fig.cap="Partially overlapping networks of four decorative practices"}
+```{r total-network-sample, fig.cap="Partially overlapping networks of four decorative practices", echo = FALSE}
 decor_graph %>%
@@
   filter(name %in% c("braided_cord_roulette", "braided_strip_roulette", "paint", "straight_cord_mat")) %>%
-  ggraph() +
+  ggraph(layout = "stress") +
   coloured_edges +
@@
 
-## Calculate difference between network layers
-
-We do this using the `multinet` package.
-
-```{r import to multinet}
+Using the `multinet` package, the statistical distance (Jaccard similarity) between the different decors' networks were calculated based on the nodes they contained.
+We see in \@ref(fig:layer-comparison) that some networks are more similar to one another than others.
+This leads us to ask whether there might be groups of decors that were shared within similar networks, and thus, whether there might be groups of sites predominantly sharing specific decors.
+
+```{r import to multinet, echo = FALSE}
 library(multinet)
@@
 
-```{r layer comparison, fig.cap="Jaccard similarity of nodes present in the networks of individual decorative practices"}
+```{r layer-comparison, fig.cap="Jaccard similarity of nodes present in the networks of individual decorative practices"}
 # perform layer compariosn using jaccard similarity between nodes
@@
 
-This shows us the jaccard similarity of the actors present on the different levels of the network.
-We can see that there are more similar and more dissimilar layers in the network.Very high are the similarities between folded strip roulette and braided cord roulette, two common decors.
-Also high is the similarity in common decorative techniques of channeling and applied decors.
-The fact that there are similar and dissimilar layers suggests that we might be able to identify groups in the network.
-
-## Clustering
+# Clustering
 
 We use the Louvain clustering algorithm (Blondel et al 2008) to find communities in the network.
-
 ```{r community detection, hide = 'results', echo=FALSE}
@@
 ```
-
 The louvain clustering has a modularity of `r modularity` .
 
-```{r plot-clusters, fig.cap="Louvain communities within the decor network"}
+```{r plot-clusters, fig.cap="Louvain communities within the decor network", echo = FALSE}
 # set plot colours for communities
@@
 
-```{r setting up colors and multilevel community, hide = 'results', echo=FALSE}
-## in the final form this part will be much shorter because the colors are set elsewhere
-# we need a vector of colors that is named after the value we want to be plotted (community, name of the decor)
-communitycolor <-
-  setNames(
-    c("tomato", "cornflowerblue", "darkgreen"),
-    unique(V(decor_graph)$louvain)
-  )
-
-# because decor is longer we need to create a palette with the proper amount of colors (the whole code is still not perfect, but i think it is the proper way that can still be optimized)
-col_21 <- colorRampPalette(brewer.pal(8, "Set2"))(21)
-
-# then we create a vector containing all decortypes
-decortypes <- decor_graph %>%
-  activate(edges) %>%
-  as_data_frame(what = "edges") %>%
-  select(name) %>%
-  unique()
-
-decortypes <- as.vector(decortypes[, 1])
-
-# finally we make a named vector of the colors
-decorcolor <- setNames(col_21, decortypes)
-```
-
 There are some recent studies in archaeology that use community detection on archaeological networks.
@@
 Here, we employ the same clustering algorithm, which results in three communities with a modularity of 0.3.
-Community 1 contains 7 Sites in the Dogon Area.
+See \@ref(fig:plot-clusters) Community 1 contains 7 Sites in the Dogon Area.
 Community 2 is in the Niger bend and the IND, it also contains 7 sites.
@@
 
-```{r cluster multilevel, echo=FALSE}
-nodes <- decor_graph %>%
-  activate(nodes) %>%
-  as_tibble()
-hull1 <- nodes %>%
-  filter(louvain == 1) %>%
-  slice(chull(lon, lat)) %>%
-  rename(x = lon, y = lat)
-
-hull2 <- nodes %>%
-  filter(louvain == 2) %>%
-  slice(chull(lon, lat)) %>%
-  rename(x = lon, y = lat)
-
-hull3 <- nodes %>%
-  filter(louvain == 3) %>%
-  slice(chull(lon, lat)) %>%
-  rename(x = lon, y = lat)
-
-# override s2 use for geographic plotting
-sf_use_s2(FALSE)
-
-ggraph(decor_graph, x = lon, y = lat) +
-  # plot on map background
-  bw.dem.map +
-  # set edge parameters
-  geom_edge_fan(color = "gray", aes(edge_alpha = weight, edge_width = weight)) +
-  scale_edge_width(range = c(0.2, 2)) +
-  scale_edge_alpha(range = c(0.3, 0.6)) +
-  # colour nodes according to cluster and label
-  geom_node_point(aes(color = louvain)) +
-  scale_color_manual(values = communitycolor) +
-  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
-  # draw cluster hulls
-  geom_polygon(data = hull1, aes(x = x, y = y), alpha = 0.5, fill = "cornflowerblue") +
-  geom_polygon(data = hull2, aes(x = x, y = y), alpha = 0.5, fill = "tomato") +
-  geom_polygon(data = hull3, aes(x = x, y = y), alpha = 0.5, fill = "darkgreen") +
-  # set colour scales
-  labs(color = "Community") +
-  guides(color = guide_legend(override.aes = list(size = 0.5)), 
-         alpha = "none", size = "none", edge_alpha = "none", 
-         edge_width = guide_legend(override.aes = list(size = 0.3))) +
-  ggtitle("Network with multilevel communities") +
-  # set map extent
-  coord_sf(xlim = c(-7, 2), ylim = c(13, 19), expand = FALSE) +
-  theme_graph()
-```
-
-```{r relative edge weights and centralization by cluster}
+```{r cluster-comparison, fig.cap="Relative edge weights and centralization by cluster"}
 # Relative edge weights (how important are the connections within the cluster)
@@
   pivot_longer(!cluster, names_to = "key", values_to = "value") %>%
+
   # plot
@@
 
-```{r edges by cluster, eval = FALSE}
-
-edges.com1 <- decor_graph %>%
-  activate(nodes) %>%
-  filter(louvain == 1) %>%
-  activate(edges) %>%
-  as_data_frame(what = "edges") %>%
-  group_by(name) %>%
-  summarise(n = sum(weight)) %>%
-  filter(n > 0) %>%
-  mutate(community = 1)
-
-edges.com2 <- decor_graph %>%
-  activate(nodes) %>%
-  filter(louvain == 2) %>%
-  activate(edges) %>%
-  as_data_frame(what = "edges") %>%
-  group_by(name) %>%
-  summarise(n = sum(weight)) %>%
-  filter(n > 0) %>%
-  mutate(community = 2)
-
-edges.com3 <- decor_graph %>%
-  activate(nodes) %>%
-  filter(louvain == 3) %>%
-  activate(edges) %>%
-  as_data_frame(what = "edges") %>%
-  group_by(name) %>%
-  summarise(n = sum(weight)) %>%
-  filter(n > 0) %>%
-  mutate(community = 3)
-
-# edges.comcrossing <- decor_graph %>%
-# activate(edges) %>%
-#  mutate(crossing = igraph::crossing(louvain,.)) %>%
-#  filter(crossing == TRUE) %>%
-#  as_data_frame(what = "edges") %>%
-# group_by(name) %>%
-#  summarise(n = sum(weight)) %>%
-#  filter(n > 0) %>%
-#    mutate(community = "crossing")
-
-plyr::rbind.fill(
-  edges.com1, edges.com2, edges.com3
-  # , edges.comcrossing
-) %>%
-  filter(n != 0) %>%
-  mutate(community = as.factor(community)) %>%
-  ggplot(aes(x = reorder(name, -n, sum), y = n, fill = community)) +
-  geom_bar(stat = "identity") +
-  scale_fill_manual(values = c("tomato", "cornflowerblue", "darkgreen", "yellow")) +
-  theme(axis.text.x = element_text(angle = 90)) +
-  scale_x_discrete(name = "decor type") +
-  scale_y_continuous(name = "sum of the weight") +
-  ggtitle("decors by clusters")
-
-str.1 <- edges.com1 %>%
-  select(n) %>%
-  sum()
-
-str.2 <- edges.com2 %>%
-  select(n) %>%
-  sum()
-
-str.3 <- edges.com3 %>%
-  select(n) %>%
-  sum()
-
-com.edges <- c(str.1 / nrow(edges.com1), str.2 / nrow(edges.com2), str.3 / nrow(edges.com3)) / 10000
-
-library(tibble)
-cent.1 <- decor_graph %>%
-  filter(louvain == 1) %>%
-  activate(nodes) %>%
-  centr_degree() %>%
-  as.tibble() %>%
-  select(centralization) %>%
-  transmute(centralization.weight = (centralization / nrow(.))) %>%
-  slice(1)
-
-cent.2 <- decor_graph %>%
-  activate(nodes) %>%
-  filter(louvain == 2) %>%
-  centr_degree() %>%
-  as.tibble() %>%
-  select(centralization) %>%
-  transmute(centralization.weight = (centralization / nrow(.))) %>%
-  slice(1)
-
-cent.3 <- decor_graph %>%
-  activate(nodes) %>%
-  filter(louvain == 3) %>%
-  centr_degree() %>%
-  as.tibble() %>%
-  select(centralization) %>%
-  transmute(centralization.weight = (centralization / nrow(.))) %>%
-  slice(1)
-
-library(tidyr)
-community.edges <- bind_rows(cent.1, cent.2, cent.3) %>%
-  add_column("sum of com.edges" = com.edges) %>%
-  gather(key) %>%
-  add_column(community = c(1:3, 1:3))
-
-ggplot(community.edges, aes(x = community, y = value, fill = key)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  # scale_fill_manual(values = c("tomato", "cornflowerblue", "darkgreen"))+
-  theme(axis.text.x = element_text(angle = 90))
-```
 If we look at the edges (divided by 10000 to make it comparable) and the centrality of sites in regard to the communities, we see that while the sites of community 3 are the most central ones in the network the weight of the edges within the community is relatively small.
@@
 --maybe the bar_plots of the most common decors of all clusters is enough? maybe the first barplot is enough? – maybe the network of every cluster is too much
+
+ see \@ref(fig:cluster-comparison)
 If we take a closer look at the "internal" connections of community 1, we see that "braided_strip_roulette" is present at all sites.
@@
 
-```{r crossings}
+```{r crossings, fig.cap="Network of edges which cross between clusters"}
 decor_graph %>%
@@
   geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
-  # scale_edge_color_manual(values = decorcolor) +
   scale_edge_width(range = c(0.2, 2)) +
-  geom_node_point(aes(size = centrality_eigen(weights = weight), color = louvain), show.legend = FALSE) +
+  geom_node_point(aes(size = centrality_degree(weights = weight), color = louvain), show.legend = FALSE) +
   geom_node_text(aes(label = label), col = "black", repel = TRUE) +
   scale_color_manual(values = communitycolor)+
-
   theme_graph() +
   theme_community
-
-```
-
-```{r crossings between communities, eval = FALSE}
-
-## this plot is hard to read but contains a lot of information, it shows only edges that connect the 2 communities.
-the color of the nodes shows their community while the size shows the betweenness centrality.
-This way we can show which sites are on the at the "frontier" of the communities that have strong connections to the other community.
-only edgest that are crossings between communities are shown, the widest ones are those that are those with an higher betweenness.
-in this way the graph shows the relation between the communities and which decor is shared/crossing community boundaries
-
-nodes.c1 <- decor.graph %>%
-  activate(nodes) %>%
-  filter(communityl == 1) %>%
-  as_data_frame(what = "vertices") %>%
-  pull(id)
-
-nodes.c2 <- decor.graph %>%
-  activate(nodes) %>%
-  filter(communityl == 2) %>%
-  as_data_frame(what = "vertices") %>%
-  pull(id)
-
-nodes.c3 <- decor.graph %>%
-  activate(nodes) %>%
-  filter(communityl == 3) %>%
-  as_data_frame(what = "vertices") %>%
-  pull(id)
-
-c1_3 <- decor.graph %>%
-  activate(edges) %>%
-  mutate(crossing = crossing(cl, .)) %>%
-  filter(crossing == TRUE) %>%
-  as_data_frame(what = "edges") %>%
-  filter(from %in% c(nodes.c3) | to %in% c(nodes.c3)) %>%
-  filter(from %in% c(nodes.c1) | to %in% c(nodes.c1)) %>%
-  group_by(name) %>%
-  summarise(n = sum(weight)) %>%
-  filter(n > 0) %>%
-  mutate(crossing = "c1_3")
-
-c2_3 <- decor.graph %>%
-  activate(edges) %>%
-  mutate(crossing = crossing(cl, .)) %>%
-  filter(crossing == TRUE) %>%
-  as_data_frame(what = "edges") %>%
-  filter(from %in% c(nodes.c3) | to %in% c(nodes.c3)) %>%
-  filter(from %in% c(nodes.c2) | to %in% c(nodes.c2)) %>%
-  group_by(name) %>%
-  summarise(n = sum(weight)) %>%
-  filter(n > 0) %>%
-  mutate(crossing = "c2_3")
-
-c1_2 <- decor.graph %>%
-  activate(edges) %>%
-  mutate(crossing = crossing(cl, .)) %>%
-  filter(crossing == TRUE) %>%
-  as_data_frame(what = "edges") %>%
-  filter(from %in% c(nodes.c1) | to %in% c(nodes.c1)) %>%
-  filter(from %in% c(nodes.c2) | to %in% c(nodes.c2)) %>%
-  group_by(name) %>%
-  summarise(n = sum(weight)) %>%
-  filter(n > 0) %>%
-  mutate(crossing = "c1_2")
-
-plyr::rbind.fill(c1_2, c1_3, c2_3) %>%
-  filter(n != 0) %>%
-  mutate(crossing = as.factor(crossing)) %>%
-  ggplot(aes(x = reorder(name, -n, sum), y = n, fill = crossing)) +
-  geom_bar(stat = "identity") +
-  scale_fill_manual(values = c("tomato", "cornflowerblue", "darkgreen", "yellow")) +
-  theme(axis.text.x = element_text(angle = 90)) +
-  scale_x_discrete(name = "decor type") +
-  scale_y_continuous(name = "sum of the weight") +
-  ggtitle("decors by clusters")
 ```
@@
 Even though most decors that cross communities connect all three, those are mostly decors that are most abundant at the sites of community 3.
+See \@ref(fig:crossings)
 
@@
 
-extract the binary backbone of the flattened graph and play around with the alpha to get a good representation
-
-```{r backbone extraction}
+extract the binary backbone of the flattened graph and play around with the alpha to get a good representation. See \@ref(fig:backbone-extraction)
+
+```{r backbone-extraction, fig.cap="Extraction of binary backbones at decreasing significance levels: a) 0.05, b) 0.15, c) 0.20"}
 # extract binary backbones at decreasing levels of significance
@@
 
-```{r backbone eigenvalue}
+```{r backbone-eigenvalue, fig.cap="Binary backbone at alpha = 0.15 showing eigenvector centralities"}
 decor.15 <- decor.15 %>% activate(nodes) %>% filter(!node_is_isolated())
 
-decor.15 %>%
-  activate(nodes) %>%
-  mutate(eigen = centrality_eigen()) %>%
-  select(label, eigen) %>%
-  arrange(desc(eigen)) %>%
-  kable()
-
-ggraph(decor.15, layout = "stress") +
+#decor.15 %>%
+#  activate(nodes) %>%
+#  mutate(eigen = centrality_eigen()) %>%
+#  select(label, eigen) %>%
+#  arrange(desc(eigen)) %>%
+#  kable()
+
+ggraph(decor.15, layout = "nicely") +
   geom_node_point(aes(size = centrality_eigen())) +
@@
 
-The result shows that the highest influence on the network is exerted by the sites in the Eastern IND and Lakes cluster.
+The result (\@ref(fig:backbone-eigenvalue)) shows that the highest influence on the network is exerted by the sites in the Eastern IND and Lakes cluster.
 We should not forget that the network is partial, i.e.
@@
 
-```{r backbone betweenness}
-decor.15 %>%
-  activate(nodes) %>%
-  mutate(eigen = centrality_betweenness()) %>%
-  select(label, eigen) %>%
-  arrange(desc(eigen)) %>%
-  kable()
-
-ggraph(decor.15, layout = "stress") +
+```{r backbone-betweenness, fig.cap="Binary backbone at alpha = 0.15 showing betweenness centralities"}
+#decor.15 %>%
+#  activate(nodes) %>%
+#  mutate(eigen = centrality_betweenness()) %>%
+#  select(label, eigen) %>%
+#  arrange(desc(eigen)) %>%
+#  kable()
+
+ggraph(decor.15, layout = "nicely") +
   geom_node_point(aes(size = centrality_betweenness())) +
@@
 
-We can see that Hambarketolo appears as the main gatekeeper to the Jenné site cluster, and to the eastern IND network clique, while Tongo Maaré Diabal, and to a lesser extent also Gao-Saney and Togu Missiri, stand in a brokerage role between the IND and the greater Niger cluster.
+We can see (\@ref(fig:backbone-betweenness)) that Hambarketolo appears as the main gatekeeper to the Jenné site cluster, and to the eastern IND network clique, while Tongo Maaré Diabal, and to a lesser extent also Gao-Saney and Togu Missiri, stand in a brokerage role between the IND and the greater Niger cluster.
 The Dogon cluster exists at two removes from the IND, passing through Dammassogou and Ambéré-Dougon.At alpha = 0.2, this cluster is also connected to the large niger cluster via the Sangha "tellem" cave sites.
@@
 In our case it might indicate areas that were important axes of knowledge exchange.
-```{r backbone edge betweenness}
+```{r backbone-edge-betweenness, fig.cap="Binary backbone at alpha = 0.15 showing edge betweenness centralities"}
 ggraph(decor.15, layout = "stress") +
@@
   theme_graph()
-```
-
-The edge betweenness measure shows a central path through the network, connecting the sites of Togu Missiri, Tongo Maaré Diabal, Hambarketolo, Damassogou, Ambéré-Dougon, and Sadia, as well as a prominent side route connecting Gao and Kawinza.These might roughly represent two routes of knowledge exchange: one important one passing through the eastern side of the Inland Niger Delta and below the Niger Bend, in which the Jenné site cluster plays a pivotal role as the meeting place between the three networks of knowledge exchange, and another, though less important to the network we have sampled, along the western side of the IND and north of the Niger Bend to Gao.
+
+ggraph(decor.15, x = lon, y= lat) +
+  geom_sf(data = mali_adm0, fill = NA) +
+  geom_sf(data = rivers, colour = "black") +
+  geom_node_point() +
+  geom_edge_fan(color = "red", alpha = 0.8, aes(width = centrality_edge_betweenness())) +
+  scale_edge_width(range = c(0.2, 3)) +
+  geom_node_text(aes(label = label), repel = TRUE) +
+  coord_sf(xlim = c(-7, 2), ylim = c(13, 19), expand = FALSE) +
+  theme_graph()
+```
+
+The edge betweenness measure (\@ref(fig:backbone-edge-betweenness)) shows a central path through the network, connecting the sites of Togu Missiri, Tongo Maaré Diabal, Hambarketolo, Damassogou, Ambéré-Dougon, and Sadia, as well as a prominent side route connecting Gao and Kawinza.These might roughly represent two routes of knowledge exchange: one important one passing through the eastern side of the Inland Niger Delta and below the Niger Bend, in which the Jenné site cluster plays a pivotal role as the meeting place between the three networks of knowledge exchange, and another, though less important to the network we have sampled, along the western side of the IND and north of the Niger Bend to Gao.
 
@@
 A Spearman's rank correlation test is run between the samples to see how the deletion of edges influences the  centrality within the backbone.
-Figure \@ref(fig:MST-betweenness-validity) shows how there is a relatively significant correlation between the betweenness ranks of nodes in the original graph and the sampled graphs, all the way down to the 30% sample.
+Figure \@ref(fig:validity-test) shows how there is a relatively significant correlation between the betweenness ranks of nodes in the original graph and the sampled graphs, all the way down to the 30% sample.
 Only on the 20% sample does the correlation statistic drop lower.
@@
 
-```{r validity test, warning = FALSE}
-
-centrality.samples <- function(s, graph, centrality) {
-
-  # get size of fraction to sample
-  y <- (s / 100) * igraph::gsize(graph)
-
-  # sample graph
-  g <- graph %>%
-    activate(edges) %>%
-    tidygraph::sample_n(size = y) %>%
-    flatten(weighted = TRUE) %>%
-    as.igraph()
-
-  # extract backbone
-  bb <- backbone::disparity(g, alpha = 0.15, narrative = FALSE) %>%
-    tidygraph::as_tbl_graph()
-
-  # calculate and extract  centrality
-  if (centrality == "eigen") {
-    q <- bb %>%
-      as_tbl_graph() %>%
-      activate(nodes) %>%
-      mutate(centr = centrality_eigen(weights = weight)) %>%
-      as_tibble() %>%
-      dplyr::select(centr) %>%
-      dplyr::rename(!!paste0("s", s) := centr)
-  }
-
-  if (centrality == "betweenness") {
-    q <- bb %>%
-      as_tbl_graph() %>%
-      activate(nodes) %>%
-      mutate(centr = centrality_betweenness(weights = weight)) %>%
-      as_tibble() %>%
-      dplyr::select(centr) %>%
-      dplyr::rename(!!paste0("s", s) := centr)
-  }
-
-  return(q)
-}
-
-spearman_samples <- function(graph, centrality) {
-
-  # sample graph at 10 percent intervals
-  samples <- seq(10, 100, 10)
-
-  # iterate function over intervals
-  centralities <-
-    samples %>%
-    purrr::map(centrality.samples, graph = graph, centrality = centrality) %>%
-    dplyr::bind_cols()
-
-  # 2. Spearman's rank correlation
-
-  # get sample names
-  cols <- centralities %>%
-    select(-s100) %>%
-    names()
-
-  # rank correlation function
-  rank.correlation <- function(x) {
-    cor.test(centralities$s100, centralities[[x]], method = "spearman")
-  }
-
-  # iterate for samples
-  purrr::map(cols, rank.correlation) %>%
-    purrr::map_df(`[`, c("estimate")) %>%
-    pull(estimate)
-}
-
-test_centralities <- function(graph, centrality, reps) {
+```{r validity-test, warning = FALSE, message = FALSE}
+p1 <- test_centralities(decor_graph, "eigen", 100)
+p2 <- test_centralities(decor_graph, "betweenness", 100)
+
+(p1 | p2) + plot_annotation(tag_levels = "a")
+```
+
+```{r}
+#make boxplots instead
+test_centralities2 <- function(graph, centrality, reps) {
 
@@
     select(-iteration) %>%
-    group_by(sample) %>%
-    summarise(mean_rho = mean(value)) %>%
+    #group_by(sample) %>%
+    #summarise(mean_rho = mean(value)) %>%
     # plot
     ggplot() +
-    geom_histogram(aes(x = sample, y = mean_rho), stat = "identity") +
+    geom_boxplot(aes(x = sample)) +
     labs(y = expression(paste("mean Spearman's ", rho))) +
@@
 }
-```
-
-```{r apply validity test, warning = FALSE, message = FALSE}
-p1 <- test_centralities(decor_graph, "eigen", 100)
-p2 <- test_centralities(decor_graph, "betweenness", 100)
-
-(p1 | p2) + plot_annotation(tag_levels = "a")
+
+test_centralities2(decor_graph, "eigen", 100)
 ```
@@
 For edge betweenness, the validation is much harder, as the flattening procedure means that edges cannot reliably be numbered, and thus correlation is hard to achieve.
-We can, however, manually inspect the result.
-
-```{r warning = FALSE, message=FALSE}
+We can, however, manually inspect the result (\@ref(fig:edge-betweenness-validity)).
+
+```{r edge-betweenness-validity, fig.cap="edge bewteenness validity",warning = FALSE, message=FALSE}
 bb.edge.betw <- function(x) {
@@
 
-## Spanning tree
-
-We can use Kruskal's concept of the minimum spanning tree, simply transformed to a maximum spanning tree (MST) here. This method has its drawbacks, as described by [@Serranoetal]:
-
->*"One of the big limitations of this method is that spanning trees are by construction acyclic.
-This means that reduced networks obtained by this algorithm are overly structural simplifications that destroy local cycles, clustering coefficient, and the clustering hierarchies often present in real world networks." *
-
-In our case, however, we can use the maximum spanning tree especially to gain access to a measure that is not accessible or meaningful in hairball networks: betweenness centrality.
-This is particularly interesting in this case, as high betweenness centrality in a network indicates a brokerage position.
-If we would like to know which places within the network are particularly well placed to pass on information, we can find out using the MST.
-
-```{r MST, fig.cap="The maximum spanning tree of the Niger pottery decor network, showing edge weights and betweenness centrality"}
-# invert edge weights
-g <- decor_graph %>%
-  activate(edges) %>%
-  mutate(weight = 0 - weight) %>%
-  as.igraph()
-
-# create MST and reconvert edge weights
-mst <- mst(g, weights = E(g)$weight) %>%
-  as_tbl_graph() %>%
-  activate(edges) %>%
-  mutate(weight = 0 - weight)
-
-# plot
-mst %>%
-  ggraph(layout = "fr") +
-  geom_edge_link(aes(alpha = weight), colour = "blue") +
-  geom_node_point(aes(size = centrality_betweenness(weights = weight))) +
-  # geom_node_text(data = . %>% filter(centrality_betweenness(weights = weight) > mean(centrality_betweenness(weights = weight))), aes(label = label), repel = TRUE)+
-  geom_node_text(aes(label = label), repel = TRUE) +
-  theme_graph()
-```
-
-## Check for robustness of betweenness measures
-
-XXX: This needs to be done for the backbone now, instead of the spanning tree.
-
-In order to know whether these centrality measures are valid, rather than random effects of the MST transformation, we implement a variation on Peeples' suggested method for estimating the robustness of centrality measures in archaeological networks.
-In order to do so, the edges of the original graph are sampled at 10 % intervals, from 10 to 90 %.
-Each of these samples is then converted to a maximum spanning tree and the betweenness measures are extracted.
-A Spearman's rank correlation test is run between the samples to see how the deletion of edges influences the betweenness centrality within the MST.
-Figure \@ref(fig:MST-betweenness-validity) shows how there is a relatively significant correlation between the betweenness ranks of nodes in the original graph and the sampled graphs, all the way down to the 30% sample.
-Only on the 20% sample does the correlation statistic drop lower.
-We can take this to mean that, for our case, the betweenness centrality shown in the MST is dependable.
-
-```{r MST-betweenness-validity,fig.cap="Spearman's rank-correlation coefficient between the MST of the original graph and those sampled at 10 to 90 percent."}
-# 1. calculate betweenness centrality of sampled graphs
-set.seed(6789)
-# sample graph at 10 percent intervals
-samples <- seq(10, 100, 10)
-
-centrality.samples <- function(s) {
-
-  # get size of fraction to sample
-  y <- (s / 100) * igraph::gsize(decor_graph)
-
-  # sample graph and invert weights
-  g <- decor_graph %>%
-    activate(edges) %>%
-    tidygraph::sample_n(size = y) %>%
-    mutate(weight = 0 - weight) %>%
-    as.igraph()
-
-  # create weighted MST
-  mst <- mst(g, weights = E(g)$weight)
-
-  # reconvert weights, calculate and extract betweenness centrality
-  q <- mst %>%
-    as_tbl_graph() %>%
-    activate(edges) %>%
-    mutate(weight = 0 - weight) %>%
-    activate(nodes) %>%
-    mutate(betw = centrality_betweenness(weights = weight)) %>%
-    as_tibble() %>%
-    dplyr::select(betw) %>%
-    dplyr::rename(!!paste0("s", s) := betw)
-}
-
-# iterate function over intervals
-centralities <- samples %>%
-  purrr::map(centrality.samples) %>%
-  dplyr::bind_cols()
-
-# 2. Spearman's rank correlation
-
-# get sample names
-cols <- centralities %>%
-  select(-s100) %>%
-  names()
-
-# rank correlation function
-rank.correlation <- function(x) {
-  cor.test(centralities$s100, centralities[[x]], method = "spearman")
-}
-
-# iterate for samples
-rho <- purrr::map(cols, rank.correlation)
-
-# 3. plot
-rho %>%
-  purrr::map_df(`[`, c("estimate", "p.value")) %>%
-  mutate(sample = cols) %>%
-  ggplot() +
-  geom_histogram(aes(x = sample, y = estimate), stat = "identity") +
-  labs(y = expression(paste("Spearman's ", rho))) +
-  ylim(0, 1) +
-  theme_minimal()
-```
+The edge betweenness measure shows a central path through the network, connecting the sites of Togu Missiri, Tongo Maaré Diabal, Hambarketolo, Damassogou, Ambéré-Dougon, and Sadia, as well as a prominent side route connecting Gao and Kawinza.These might roughly represent two routes of knowledge exchange: one important one passing through the eastern side of the Inland Niger Delta and below the Niger Bend, in which the Jenné site cluster plays a pivotal role as the meeting place between the three networks of knowledge exchange, and another, though less important to the network we have sampled, along the western side of the IND and north of the Niger Bend to Gao.
 
@@
 
-```{r import early sites}
+```{r import-new-data}
 library(readr)
@@
 
-```{r extract early backbone}
+```{r early-backbone, fig.cap="early backbone"}
 # flatten with weights
@@
 
-This backbone looks significantly different from the later one, though we still distinguish an IND group, a very clear Dogon group, and one that includes "others": Kawinza, TMD, Tiebala, Akumbu.
+This backbone (\@ref(fig:early-backbone))looks significantly different from the later one, though we still distinguish an IND group, a very clear Dogon group, and one that includes "others": Kawinza, TMD, Tiebala, Akumbu.
 Essouk is isolated.
-Let's try this with louvain communities.
-
-```{r cluster early}
-decor_graph_early <- decor_graph_early %>%
-  activate(nodes) %>%
-  mutate(louvain = as.character(group_louvain()))
-
-# calculate modularity of louvain clustering
-modularity <- decor_graph_early %>%
-  activate(nodes) %>%
-  mutate(modularity = graph_modularity(group = as.factor(louvain))) %>%
-  pull(modularity) %>%
-  head(1)
-
-ggraph(decor_graph_early, x = lon, y = lat) +
-  geom_sf(data = mali_adm0, fill = NA) +
-  geom_sf(data = rivers, colour = "black") +
-  edges_community +
-  ggforce::geom_mark_hull(aes(x = lon, y = lat, fill = louvain, colour = louvain),
-                          expand = unit(1.5, "mm"),
-                          radius = unit(1.5, "mm")) +
-  scale_fill_manual(values = communitycolor)+
-  scale_colour_manual(values = communitycolor)+
-  nodes_community +
-  theme_community +
-  coord_sf(xlim = c(-7, 2), ylim = c(13, 19), expand = FALSE) +
-  theme_graph()
-```
-
-From these, the hypothesis: The Dogon network in the early period is very densely connected and more isolated from the Niger network.
-The Niger network on the other hand, is more diffuse, harder to cluster, maybe more generalised.
-How to calculate this?
+Clustering similarly revelas these two communities.
+
+A more in-depth treatment of the temporal development of the pottery networks in this area will be the subject of a future publication.
+For now, we might draw the following preliminary conclusions: The comparison of the two networks appears to show a process of increasing fragmentation between the two periods reflected by the graphs, but also the growth of netwroks to incorporate more diverse influences.
+# Results
+
+# Discussion
+
+Potter communities and the propagation of knowledge within them, has been the subject of several very detailed ethnoarchaeological and archaeological studies in this region.
+This study has left these out, quite purposefully.
+Their boundaries might be seen in other elements of the pottery making sequence, notably in the forming technique, and they do not necessarily overlap with decors.
+Our network model instead is intended to represent communication of knowledge about pottery that transcends smaller communities of practice.
+As such, it shows a first picture of how decorative techniques might have developed and what mechanisms are at the basis of the patterns in archaeological ceramics that are seen on sites along the Middle Niger.
+
+The results of our analyses suggest that the propagation of pottery decor motifs occurs within networks of differing magnitudes.
+Some of these networks are local, geographically and possibly also culturally bounded.
+This can, for instance, be seen in the group of sites on the Bandiagara massif and in the Seno plain, or in the eastern part of the IND.
+Cross-cutting such local developments, however, are long-distance contacts along the Niger River, which led to similarities between very distant sites.
+This underlines a key aspect of past cultural dynamics in the wider region: there is both a high level of cultural diversity, of human groups and languages, and overarching regional connectivity.
+These two processes, the local and the global, have been in exchange with one another shaping the Middle Niger cultural landscape.
+Every site assemblage can therefore be seen to be reflective of how strongly innovation and tradition influenced local potting communities.
+Changes in the diversity of decors in an assemblage, noted for instance at XXX, might therefore reflect changes to the degree in which the population supplying pottery to a site was exposed and open to new influences.
+
+It is of key importance in this approach to note where this interaction between network groups takes place.
+While the archaeological record is necessarily only a sample and many nodes are missing, we can nevertheless gain a preliminary insight into the channels along which information on pottery decoration was shared, identify those sites within the record which occupy key positions in this exchange, and to suggest key routes.
+The latter, built on edge betweenness, is significant in that it might suggest those areas (for instance between Kawinza and Gao) which are not represented by nodes in the model, but were nevertheless important to the exchange network.
+
+The diachronic comparison of two networks from different periods begins to show us processes of innovation, and how these change networks.
+This must be seen as a work in progress, and a fuller picture will be the subject of future publications.Nevertheless, we can detect how the sites within the IND became a major centre for diversification and diffusion of decorative techniques, while the close-knit and formerly more isolated "Bandiagara cluster" opened up to outside influences.
+Overall, the later period is marked by more ties, an intensifying network.
+This is likely due to an increased frequency of contact between heterogenous potter communities, perhaps as a result of a combination of increases in population size, the establishment of new markets, and an increased flow of goods contained in pottery.
+Mayor (REF) has also suggested political factors behind the spread of some decors, though we are not entirely sure of the causality behind this hypothesis.
+This pattern is less likely to stem from an overall higher mobility of individual potters, since the overlapping structure of the individual technique networks suggests gradual change.
+
+Chronological and spatial approaches to pottery decor in the study region have been largely frustrating, with changes only in percentages of decors which rarely provided the basis for clear phase changes or cultural boundaries.
+This approach emphasises that this picture of gradual differences points to a strongly connected society, in which pottery producers were routinely exposed to their neighbours' productions, and in which the decorative techniques were mostly able to propagate without impediment.
+Nevertheless, within this picture, local traditions can be discerned as cliques within the networks, whose degree of isolation varies, but which are never without outside contact.
+
+# Conclusion
+
+Modelling pottery decorative techniques along the Middle Niger as a knowledge exchange network presents a change of perspective on the subject.
+Previous work has focused on distribution areas and relative prevalence at different sites.
+We have taken the same information to indicate the degree to which pottery producers supplying the various sites were part of contact networks in which knowledge of decors could spread.
+While this approach is certainly as reductive as any other, we draw the novel conclusion that the relatively static record of pottery decor in this area is due to a diverse and very well connected population of pottery producers.
+They exchanged their techniques most frequently on a local level, where geographically bounded closer networks shared intensively.
+Such exchange might have taken many forms, such as marriage-related mobility, kinship-based learning networks, much more ephemeral exchanges in marketplaces, or even indirect contact by seeing another potter's product.
+The latter form is the most likely basis for the way in which these local networks corresponded to a larger one: all along the Middle Niger, from our southernmost sample in the Ségou area to Gao and the Saharan entrepot of Essouk, pottery and the knowledge of certain decorative techniques appears to have travelled.
+Our model suggests that two routes are likely to have been important in this long-distance contact: one along the western IND, the Lakes Region and north of the Niger Bend, and another along the IND's eastern edge and below the Niger Bend.
+
+These results feed into long-standing discussions on the Middle Niger, and West Africa in general, as a highly connected and interdependent cultural space in the late first and early second millennium AD (for a recent discussion, see Gestrich et al.
+2021).
+
+# Acknowledgements
+
+<!-- The following line inserts a page break  -->
+
+\newpage
+
+# References
+
+<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+
+<div id="refs"></div>
+
+\newpage
+
+### Colophon
+
+This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+
+```{r colophon, cache = FALSE}
+# which R packages and versions?
+#if ("devtools" %in% installed.packages()) devtools::session_info()
+```
+
+The current Git commit details are:
 
 ```{r}
-# Relative edge weights (how important are the connections within the cluster)
-rel_edge_weight_early <- vector("double", length(3))
-for (i in 1:2) {
-  rel_edge_weight_early[[i]] <- decor_graph_early %>%
-    # filter community
-    activate(nodes) %>%
-    filter(louvain == i) %>%
-    # relative edge weight (sum of weights / number of edges)
-    activate(edges) %>%
-    as.data.frame() %>%
-    {
-      sum(.$weight) / nrow(.)
-    }
-}
-
-# Centralization of the cluster
-centralization_early <- vector("double", length(3))
-for (i in 1:2) {
-  centralization_early[[i]] <- decor_graph_early %>%
-    # filter community
-    activate(nodes) %>%
-    filter(louvain == i) %>%
-    # degree centralization
-    centr_degree(mode = "all", normalized = TRUE) %>%
-    as_tibble() %>%
-    pull(centralization) %>%
-    head(1)
-}
-
-# put together
-early <- tibble(rel_edge_weight, centralization) %>%
-  mutate(cluster = 1:3) %>%
-  mutate(rel_edge_weight = rel_edge_weight / 100) %>%
-  rename("relative edge weight" = rel_edge_weight) %>%
-  pivot_longer(!cluster, names_to = "key", values_to = "value") |> 
-  ggplot(aes(x = cluster, y = value, fill = key)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  scale_fill_grey() +
-  theme(axis.text.x = element_text(angle = 90)) +
-  theme_publish()
-
-late <- tibble(rel_edge_weight_early, centralization_early) %>%
-  mutate(cluster = 1:2) %>%
-  mutate(rel_edge_weight_early = rel_edge_weight_early / 100) %>%
-  rename("relative edge weight early" = rel_edge_weight_early) %>%
-  pivot_longer(!cluster, names_to = "key", values_to = "value") |>
-
-  # plot
-  ggplot(aes(x = cluster, y = value, fill = key)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  scale_fill_grey() +
-  theme(axis.text.x = element_text(angle = 90)) +
-  theme_publish()
-
-early + late
-```
-
-Clustering Coefficient: "Transitivity measures the probability that the adjacent vertices of a vertex are connected. This is sometimes also called the clustering coefficient."
-Diameter
-Average Distance
-
-```{r}
-# Transitivity(clustering coefficient)
-
-igraph::transitivity(decor_graph_early)
-igraph::transitivity(decor_graph)
-
-# density - doesn't work for multiple graphs
-igraph::edge_density(decor_graph)
-igraph::edge_density(decor_graph_early)
-
-# Diameter
-igraph::diameter(decor_graph,
-  directed = FALSE,
-  # weights = E(decor_graph)$weight
-  weights = NA
-)
-igraph::diameter(early_flat,
-  directed = FALSE,
-  # weights = E(decor_graph_early)$weight
-  weights = NA
-)
-```
-Transitivity and density: The earlier network has a higher transitivity and lower density than the later one (though in both cases, not by much).
-We can take this to mean that groups in the earlier network are marginally stronger, and that it is less closely connected overall.
-This reflect, for instance, the strength of the Bandiagara group and the exposed position of Essouk in this dataset.How does this change when the networks are simplified?
-There is a big difference in the diameter of the networks, with the earlier one at 237 and the later one at 13. What does this mean? Is it true? The weights make this difficult, don't they?
-
-Following Wasserman and Faust, p.180-183, we report the degree centralization of the graph, along with its average degree.
-```{r centralization}
-
-flat <- decor_graph %>% flatten(weighted = TRUE)
-igraph::centr_eigen(flat)
-
-flatearly <- decor_graph_early %>% flatten(weighted = TRUE)
-igraph::centr_degree(flatearly)
-
-# Transitivity(clustering coefficient)
-
-igraph::transitivity(decor_graph_early)
-igraph::transitivity(decor_graph)
-
-# density - doesn't work for multiple graphs
-igraph::edge_density(flat)
-igraph::edge_density(flatearly)
-```
-
-Mean distance
-
-```{r}
-# mean distance
-igraph::mean_distance(decor_graph, directed = FALSE, weights = NA)
-igraph::mean_distance(decor_graph_early, directed = FALSE, weights = NA)
-
-# Dogon cluster only
-decor_graph %>%
-  activate(nodes) %>%
-  filter(louvain == 2) %>%
-  igraph::mean_distance(., directed = FALSE, weights = NA)
-
-decor_graph_early %>%
-  activate(nodes) %>%
-  filter(louvain == 2) %>%
-  igraph::mean_distance(., directed = FALSE, weights = NA)
-```
-If we are to use weighted distances, then we should probably negate the values.https://ona-book.org/paths-distance.html . It might be more appropriate to use unweighted distances.
-
-```{r}
-# Modularity:
-# how high is the modularity using our louvain clusters?
-
-decor_graph %>% graph_modularity(weights = weight)
-```
-
-We can also try global measures for centrality. These should be distributions that are compared.
-
-```{r compare centrality distributions}
-pull_centralities <- function(graph) {
-  graph %>%
-    activate(nodes) %>%
-    mutate(deg = centrality_degree(weights = weight), eig = centrality_eigen(weights = weight)) %>%
-    select(deg, eig) %>%
-    as_tibble()
-}
-
-early <- pull_centralities(decor_graph_early) %>% mutate(period = "early")
-# %>% mutate(eig = eig/n(), deg = deg/n())
-late <- pull_centralities(decor_graph) %>% mutate(period = "late")
-# %>% mutate(eig = eig/n(), deg = deg/n())
-
-centralities <- bind_rows(early, late)
-
-a <- ggplot(centralities) +
-  geom_boxplot(aes(x = period, y = deg, fill = period)) +
-  theme_minimal() +
-  ylab("degree centrality")
-
-b <- ggplot(centralities) +
-  geom_boxplot(aes(x = period, y = eig, fill = period)) +
-  theme_minimal() +
-  ylab("eigenvector centrality")
-
-a + b
-
-centralities %>%
-  group_by(period) %>%
-  summarise(meandeg = mean(deg), meaneig = mean(eig))
-```
-Degree centrality is clearly lower in the early period, suggesting a less well-connected network.
-This is an effect of an increasing diversification of decorative techniques.
-The eigenvector centrality, on the other hand, is relatively equal, but slightly higher and with a lower range in the early period, suggesting that influence on the network was somewhat more evenly distributed.
-When these values are adjusted for the number of nodes in each network, things change: degree centrality is fairly even, but eigenvector centrality drops drastically in the later period, reflecting a more uneven distribution of influence on the growing network.
-
-Secondly, we can use netdis as a graphlet-based approach.
-Work with this: https://github.com/alan-turing-institute/network-comparison/blob/master/vignettes/quickstart_netdis_2graphs.Rmd
-
-```{r}
-library(netdist)
-# flatten graphs
-decor_early_flat <- flatten(decor_graph_early)
-decor_flat <- flatten(decor_graph)
-
-# set netdis parameters
-# Maximum graphlet size to calculate counts and netdis statistic for.
-max_graphlet_size <- 4
-# Ego network neighbourhood size
-neighbourhood_size <- 2
-# Minimum size of ego networks to consider
-min_ego_nodes <- 3
-min_ego_edges <- 1
-# Ego network density binning parameters
-min_bin_count <- 5
-num_bins <- 100
-
-## Generate ego networks
-
-# Get ego networks for query graphs and reference graph
-ego_1 <- make_named_ego_graph(decor_early_flat,
-  order = neighbourhood_size,
-  min_ego_nodes = min_ego_nodes,
-  min_ego_edges = min_ego_edges
-)
-ego_2 <- make_named_ego_graph(decor_flat,
-  order = neighbourhood_size,
-  min_ego_nodes = min_ego_nodes,
-  min_ego_edges = min_ego_edges
-)
-
-# Count graphlets for ego networks in query and reference graphs
-graphlet_counts_1 <- ego_to_graphlet_counts(ego_1, max_graphlet_size = max_graphlet_size)
-graphlet_counts_2 <- ego_to_graphlet_counts(ego_2, max_graphlet_size = max_graphlet_size)
-
-# Load reference graph
-ref_path <- system.file(file.path("extdata", "random", "ER_1250_10_1"),
-  package = "netdist"
-)
-ref_graph <- read_simple_graph(ref_path, format = "ncol")
-ego_ref <- make_named_ego_graph(ref_graph,
-  order = neighbourhood_size,
-  min_ego_nodes = min_ego_nodes,
-  min_ego_edges = min_ego_edges
-)
-graphlet_counts_ref <- ego_to_graphlet_counts(ego_ref, max_graphlet_size = max_graphlet_size)
-# Scale ego-network graphlet counts by dividing by total number of k-tuples in
-# ego-network (where k is graphlet size)
-scaled_graphlet_counts_ref <- scale_graphlet_counts_ego(
-  graphlet_counts_ref,
-  max_graphlet_size
-)
-# Get ego-network densities
-densities_ref <- ego_network_density(graphlet_counts_ref)
-# Adaptively bin ref ego-network densities
-binned_densities <- binned_densities_adaptive(densities_ref,
-  min_counts_per_interval = min_bin_count,
-  num_intervals = num_bins
-)
-ref_ego_density_bins <- binned_densities$breaks
-# Average ref graphlet counts across density bins
-ref_binned_graphlet_counts <- mean_density_binned_graphlet_counts(
-  scaled_graphlet_counts_ref,
-  binned_densities$interval_indexes
-)
-
-# Calculate expected graphlet counts (using ref graph ego network density bins)
-exp_graphlet_counts_1 <- netdis_expected_counts(graphlet_counts_1,
-  ref_ego_density_bins,
-  ref_binned_graphlet_counts,
-  max_graphlet_size,
-  scale_fn = count_graphlet_tuples
-)
-exp_graphlet_counts_2 <- netdis_expected_counts(graphlet_counts_2,
-  ref_ego_density_bins,
-  ref_binned_graphlet_counts,
-  max_graphlet_size,
-  scale_fn = count_graphlet_tuples
-)
-# Centre graphlet counts by subtracting expected counts
-centred_graphlet_counts_1 <- netdis_subtract_exp_counts(
-  graphlet_counts_1,
-  exp_graphlet_counts_1,
-  max_graphlet_size
-)
-centred_graphlet_counts_2 <- netdis_subtract_exp_counts(
-  graphlet_counts_2,
-  exp_graphlet_counts_2,
-  max_graphlet_size
-)
-
-## Sum centred graphlet counts across all ego networks
-sum_graphlet_counts_1 <- colSums(centred_graphlet_counts_1)
-sum_graphlet_counts_2 <- colSums(centred_graphlet_counts_2)
-
-## Calculate netdis statistics
-netdis_result <- netdis_uptok(
-  sum_graphlet_counts_1,
-  sum_graphlet_counts_2,
-  max_graphlet_size
-)
-
-print(netdis_result)
-```
-The result shows us the difference between the two networks using three- and four-node graphlets.
-The result shows that the three-node graphlet distance is far lower than the four-node one.
-What does it mean?
-
-We could also see whether the clusters are still there and how the crossings change.
-
-Graph centralization:
-Wasserman & Faust p.180-183
+# what commit is this file at? 
+if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+```

### 2023-08-03 analysis/supplementary_materials/map backgrounds.Rmd
Initial commit

---
title: "prepare map backgrounds"
author: "Nikolas Gestrich"
date: '2022-05-02'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rnaturalearth)
library(rnaturalearthdata)
library(dplyr)
library(ggplot2)
library(sf)
```

This document is intended to make transparent the processes behind the creation of maps in the article associated with this repository. While this is not strictly necessary for scientific reproducibility, it might nevertheless be interesting to some readers.

# Country and continent inset data

## Continent data

For the outline of Africa, we import a shapefile

```{r}
africa_outline <- sf::st_read(here::here("analysis", "data", "raw_data", "Africa_outline.shp")) %>% sf::st_crop(xmax = 54, xmin = -20, ymax = 40, ymin = -36)
```

## country outline

Here, we subset a free dataset of administrative boundaries downloaded from naturalearthdata.com.

```{r subset adm0}
mali_adm0 <- ne_countries(scale = "medium", returnclass = "sf") %>% filter(admin == "Mali")
```

These can be used for inset maps inset following manner

```{r plot adm0}
ggplot()+
  geom_sf(data = mali_adm0)+
  geom_sf_text(data = mali_adm0, aes(label = admin))

ggplot(africa_outline)+
  geom_sf()
```

# Hydrology

```{r}
rivers <- ne_download(scale = 50, type = "rivers_lake_centerlines", category = "physical", returnclass = "sf") %>% sf::st_intersection(mali_adm0) %>% filter (name == "Niger")

```

```{r plot hydro}
ggplot()+
  geom_sf(data = mali_adm0)+
  geom_sf_text(data = mali_adm0, aes(label = admin))+
  geom_sf(data = rivers, fill = '#9ecae1')+
  geom_sf(data = osm_lakes.sf)
```

# Relief

## DEM

```{r}
library(raster)
#Mali outline
mali_sp <- sf::as_Spatial(mali_adm0)
#get SRTM elevation data
srtm1 <- getData("SRTM", lat = 14, lon = -3, download = TRUE)
srtm2 <- getData("SRTM", lat = 16, lon = -3, download = TRUE)
srtm3 <- getData("SRTM", lat = 14, lon = -6, download = TRUE)
srtm4 <- getData("SRTM", lat = 16, lon = -6, download = TRUE)
srtm5 <- getData("SRTM", lat = 22, lon = -6, download = TRUE)
srtm6 <- getData("SRTM", lat = 22, lon = -3, download = TRUE)
srtm7 <- getData("SRTM", lat = 18, lon = 2, download = TRUE)
srtm8 <- getData("SRTM", lat = 14, lon = -12, download = TRUE)
srtm9 <- getData("SRTM", lat = 16, lon = -12, download = TRUE)
srtm10 <- getData("SRTM", lat = 22, lon = 2, download = TRUE)

#Mosaic/merge srtm tiles
srtmmosaic <- mosaic(srtm1, srtm2, srtm3, srtm4, srtm5, srtm6, srtm7, srtm8, srtm9, srtm10, fun=mean)
#crop to Mali
dem1 <- crop(srtmmosaic, extent(mali_sp))
mali_srtm <- mask(dem1, mali_sp)
#downsample
mali_srtm_a <- terra::aggregate(mali_srtm, fact = 10)
#convert to df
dem.df <- rasterToPoints(mali_srtm_a)
dem.df <- data.frame(dem.df)
colnames(dem.df) = c("lon", "lat", "alt")
```

## Hillshade

```{r}
# extract slope and aspect
slope.raster <- terrain(mali_srtm_a, opt='slope')
aspect.raster <- terrain(mali_srtm_a, opt='aspect')
# calculate hill shade
hill.raster <- hillShade(slope.raster, aspect.raster, 40, 270)
# convert to df
hill.m <- rasterToPoints(hill.raster)
hill.df <-  data.frame(hill.m)
colnames(hill.df) <- c("lon", "lat", "hill")
```

```{r}
ggplot()+
  #geom_sf(data = mali_adm0)+
  geom_raster(data = hill.df, aes(lon, lat, fill = hill)) +
  scale_fill_gradientn(colours = grey.colors(100)) +
  geom_raster(data = dem.df, aes(lon, lat, fill = alt), alpha = 0.8)+
  scale_fill_gradientn(colours = terrain.colors(100))+
  geom_sf_text(data = mali_adm0, aes(label = admin))+
  geom_sf(data = rivers, colour = "blue")+
  theme_minimal()

```

# Save data
```{r save geodata}
save(africa_outline, mali_adm0, rivers, dem.df, hill.df,
     file = here::here("analysis", "data", "derived_data", "geodata.rda"))
```

From this data, we create several thematic map backgrounds for use in the article. These are stored in the `/R` folder, and are called to the paper file as source scripts.

### 2023-08-03 analysis/R/datavis themes.R

@@ -1,0 +1,97 @@
+
+#1. publication theme for ggplot (from the envalysis package, author: Zacharias Steinmetz)
+theme_publish <- function(base_size = 12, base_family = "",
+                          line_size = 0.25, ...) {
+  half_line <- base_size / 2
+  small_rel <- 0.8
+  small_size <- small_rel * base_size
+
+  theme_bw(base_size = base_size, base_family = base_family, ...) %+replace%
+    theme(
+      rect = element_rect(fill = "transparent", colour = NA, color = NA,
+                          size = 0, linetype = 0),
+      text = element_text(family = base_family, face = "plain",
+                          colour = "black", size = base_size, hjust = 0.5,
+                          vjust = 0.5, angle = 0, lineheight = 0.9,
+                          margin = ggplot2::margin(), debug = F),
+
+      axis.text = element_text(size = small_size),
+      axis.text.x = element_text(margin = ggplot2::margin(t = small_size/4),
+                                 vjust = 1),
+      axis.text.y = element_text(margin = ggplot2::margin(r = small_size/4),
+                                 hjust = 1),
+      axis.title.x = element_text(margin = ggplot2::margin(t = small_size,
+                                                           b = small_size)),
+      axis.title.y = element_text(angle = 90,
+                                  margin = ggplot2::margin(r = small_size,
+                                                           l = small_size/4)),
+      axis.ticks = element_line(colour = "black", size = line_size),
+      axis.ticks.length = unit(0.25, 'lines'),
+
+      axis.line = element_line(colour = "black", size = line_size),
+      axis.line.x = element_line(colour = "black", size = line_size),
+      axis.line.y = element_line(colour = "black", size = line_size),
+
+      legend.spacing = unit(base_size/4, "pt"),
+      legend.key = element_blank(),
+      legend.key.size = unit(1 * base_size, "pt"),
+      legend.key.width = unit(1.5 * base_size, 'pt'),
+      legend.text = element_text(size = rel(small_rel)),
+      legend.title = element_text(size = rel(small_rel), face = 'bold'),
+      legend.position = 'bottom',
+      legend.box = 'horizontal',
+
+      panel.spacing = unit(1, "lines"),
+      panel.background = element_blank(),
+      panel.border = element_blank(),
+      panel.grid.major = element_blank(),
+      panel.grid.minor = element_blank(),
+
+      strip.text = element_text(size = base_size),
[…]

### 2023-08-03 analysis/R/edgelist functions.R

@@ -1,0 +1,101 @@
+#####################################################
+# Functions to make edgelists for various data types#
+#####################################################
+library(dplyr)
+
+# 1.function to make a percentage table from presence/absence data
+
+make_presabs_prop <- function(data, col_node){
+
+  col_node <- enquo(col_node)
+
+  data %>%
+    tidyr::pivot_longer(!(!!col_node), names_to = "type", values_to = "count") %>%
+    group_by((!!col_node), type) %>%
+    count(count) %>%
+    mutate(freq = n / sum(n) *100) %>%
+    filter(count == 1) %>%
+    select((!!col_node), type, freq) %>%
+    tidyr::pivot_wider(names_from = type, values_from = freq) %>%
+    tibble::as_tibble()
+}
+
+##############
+
+# 2.Make a proportions table from raw counts, to feed to the edgelist function:
+
+make_count_prop <- function (x){
+
+  x_gather <- x %>%
+    tidyr::gather(key = "type", value = "type_count", -Site)
+
+  x_sum <- x_gather %>%
+   group_by(Site) %>%
+   summarise(Site_sum = sum(type_count))
+
+  x_prop <- x_gather %>%
+    left_join(x_sum, by = "Site") %>%
+    mutate(prop = type_count / Site_sum *100) %>%
+    select(-type_count, -Site_sum) %>%
+    mutate(type = paste0(type)) %>%
+    tidyr::spread(key = type, value = prop)
+
+  return(x_prop)
+}
+
+###############
+# 3. make edgelist from percentage data listed by site and variable
+
+make_edgelist <- function(data){
+  data <- data%>% select_if(function(x){!all(is.na(x))})  #only select those coumns that have values in them
+
+  data_long <- data %>% tidyr::pivot_longer(-1, names_to = "name") %>% dplyr::rename(Var1 = 1)  #make data long
+
+  sites <- data %>%                                                    #get all pairs of sites
+    tidyr::expand(x = .[[1]], y = .[[1]]) %>%
+    filter(x != y)%>%
+    rowwise() %>%
+    mutate(id = paste0(sort(c(x, y)), collapse = " ")) %>%
+    distinct(id, .keep_all = TRUE) %>%
+    dplyr::select(-id)
+
+  Type <- tibble::tibble(name = colnames(data[-1]))                             #get all edge types
+
[…]

### 2023-08-03 analysis/R/map plots.R

@@ -1,0 +1,93 @@
+#########################################################
+# Map background
+#########################################################
+# This produces background maps for plotting sites
+# and networks.
+
+library(ggplot2)
+library(sf)
+library(ggspatial)
+library(terrainr)
+library(terra)
+library(dplyr)
+
+# Load geographic data
+load(here::here("analysis", "data", "derived_data", "geodata.rda"))
+
+# 1. Greyscale map with scale bar
+
+maptheme_bw_scale = theme(panel.grid = element_blank()) +
+  theme(axis.text = element_blank()) +
+  theme(axis.ticks = element_blank()) +
+  theme(axis.title = element_blank()) +
+  theme(legend.position = "bottom") +
+  theme(panel.grid = element_blank()) +
+  theme(panel.background = element_rect(fill = "white")) +
+  theme(plot.margin = unit(c(0, 0, 0.5, 0), 'cm'))+
+  theme(panel.border = element_rect(colour = "black", fill=NA, size=2))
+
+map_bw_scale <- list(geom_sf(data = mali_adm0, fill = NA),
+                     geom_sf(data = rivers, fill = "grey", colour = "grey") ,
+                     ggspatial::annotation_scale(aes(location = "br", style ="ticks")),
+                     maptheme_bw_scale
+                      )
+
+# 2. Greyscale map with degree ticks
+
+maptheme_bw_grid = theme(panel.grid = element_blank()) +
+  theme(axis.text.x=element_text(angle=90, hjust=1)) +
+  theme(axis.title = element_blank()) +
+  theme(legend.position = "bottom") +
+  theme(panel.grid = element_blank()) +
+  theme(panel.background = element_rect(fill = "white")) +
+  theme(plot.margin = unit(c(0, 0, 0.5, 0), 'cm'))+
+  theme(panel.border = element_rect(colour = "black", fill=NA, size=1.5))
+
+map_bw_grid <- list(geom_sf(data = mali_adm0, fill = NA),
+                    geom_sf(data = rivers, fill = "grey", colour = "grey") ,
+                    maptheme_bw_grid
+)
+
+# 3. create an inset map
+
+inset <- ggplot()+
+  geom_sf(data = africa_outline, fill = "white")+
+  geom_sf(data = mali_adm0, fill = NA, size = 0.6)+
+  geom_sf_text(data = mali_adm0, aes(label = admin), position = position_nudge(x = 15, y = 1), size = 3)+
+  theme_void()
+
[…]

### 2023-08-03 analysis/R/network functions.R

@@ -1,0 +1,192 @@
+##################################################
+# Functions for network manipulation and analysis#
+##################################################
+
+library(igraph)
+library(rlang)
+library(dplyr)
+library(tidyr)
+library(tidygraph)
+library(ggplot2)
+
+# 1. function to mark edges crossing between louvain clusters
+
+community_crossing <- function(weights = NULL) {
+  #expect_nodes()
+  weights <- enquo(weights)
+  weights <- eval_tidy(weights, .E())
+  if (is.null(weights)) {
+    weights <- NA
+  }
+  group <- igraph::crossing(cluster_louvain(graph = .G(), weights = weights), .G())
+}
+
+# 2. function to extract individual layer
+
+onelayer <- function(lay){
+  layx <- rlang::sym(lay)
+  prep.graph %>% activate(edges) %>% filter(layer == layx) %>% activate(nodes) %>% filter(!node_is_isolated()) %>% as.igraph
+}
+
+# 3. function to make a heatmap from matrix
+heatmap_matrix <- function(matrix){
+  matrix[lower.tri(matrix, diag = TRUE)] <- NA
+  cor_tri <- as.data.frame(matrix) %>%
+    mutate(Var1 = factor(row.names(.), levels=row.names(.))) %>%
+    tidyr::gather(key = Var2, value = value, -Var1, na.rm = TRUE, factor_key = TRUE)
+
+  ggplot(data = cor_tri, aes(Var2, Var1, fill = value)) +
+    geom_tile()+
+    scale_fill_viridis_c()+
+    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
+          axis.title.x = element_blank(),
+          axis.title.y = element_blank())
+}
+
+# 4. flatten a multilevel graph
+flatten <- function(graph, weighted = FALSE){
+  if(weighted == FALSE) {
+    x <- graph %>%
+      igraph::simplify(edge.attr.comb = "ignore") %>%
+      tidygraph::as_tbl_graph()
+    return(x)
+  } else {
+    graph %>%
+      igraph::simplify(edge.attr.comb = list(weight = "sum", "ignore")) %>%
+      tidygraph::as_tbl_graph()
+  }
+}
+
+
+# 5. Functions to test the validity of eigenvector and betweenness centrality in binary backbones
+centrality.samples <- function(s, graph, centrality) {
+
[…]

### 2023-08-03 analysis/paper/paper.Rmd
initial commit

@@
   - Juan-Marco Puerta-Schardt:
-      email: jmpuertaschardt@outlook.de
+      email: Puerta-Schardt@em.uni-frankfurt.de
       institute: [GU]
@@
       #- --lua-filter=../templates/pagebreak.lua
-bibliography: references.bib
-csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
 abstract: |
-  Text of abstract
+  This paper reports on an attempt to analyse decorative techniques on archaeological pottery from the Middle Niger as knowledge exchange networks.
+The 12th-century CE state of this network is modelled and analysed, and then compared to its 9th-century counterpart.
+The analysis shows how knowledge about pottery decoration was propagated within networks of different sizes and intensities, from local, well-connected networks to more loosely tied long-distance connections.
+The article shows how the perspective of network analysis differs from, and can fruitfully complement, previous treatments of ceramic décor distribution in the region.
+It adds a novel perspective to what the distribution of archaeological ceramics in this region reflects, and contributes to the generation of hypotheses that can be further tested by fieldwork.
 keywords: |
-  keyword 1; keyword 2; keyword 3
-highlights: |
-  These are the highlights. 
+  Mali; network analysis; pottery; archaeological ceramics
 ---
@@
 Keywords: `r rmarkdown::metadata$keywords`
-
-Highlights: `r rmarkdown::metadata$highlights`
 
@@
 # libraries
+
+# Load the dplyr package for data manipulation
 library(dplyr)
+
+# Load the knitr package for generating reports using RMarkdown
 library(knitr)
+
+# Load the patchwork package for combining multiple plots into a single plot
 library(patchwork)
+
+# Load the ggplot2 package for creating graphics and plots
 library(ggplot2)
+
+# Load the tidygraph package for working with tidygraph objects
 library(tidygraph)
+
+# Load the ggraph package for visualizing graph data
 library(ggraph)
+
+# Load the tibble package for creating and manipulating tabular data
 library(tibble)
+
+# Load the igraph package for working with network data
 library(igraph)
+
+# Load the tidyr package for data tidying and reshaping
 library(tidyr)
+
+# Load the sf package for reading, writing, and manipulating spatial data
 library(sf)
+
+# Load the multinet package for analysis of multilayer networks
+library(multinet)
+
+# Load the kableExtra package for nicer table design
+library(kableExtra)
+
+# Load the readr package to read excel-data
+library(readr)
 ```
@@
 
-For much of the history of the archaeological discipline, ceramics have been used as markers of similarity and difference.In many areas and periods of the past, they are the most ubiquitous archaeological find category, and have thus contributed greatly to the recognition of archaeological artefact groups, or archaeological cultures.
-It is not rare to see find complexes and periods named after particular ceramic types.
-Though the current scholarly consensus is that archaeological cultures do not represent social groups as experienced by past humans, and that pots do not equal people, a large amount of work on archaeological ceramics is engaged in grouping ceramics and distinguishing groups from one another.
-There are a number of approaches for doing so and the rationales behind them vary.
-The basis for the creation of archaeological cultures, especially current in works of the first two thirds of the twentieth century, is the 'traditional' typological approach to ceramics and other artefacts.
-This approach is nowadays most used in the best-explored regions with the longest research traditions, and groups ceramics mainly according to outward appearances, such as shape and surface treatment.
-The distribution, both spatially and temporally, of such types then becomes the subject of research, which subsequent studies can augment through the recognition of sub-types and variants.
-A different approach takes as its basis the sequence of manufacture of artefacts, the *chaîne opératoire*.
-This approach proposes that the manufacture of an artefact involves a series of choices constrained by certain outside factors.
-These choices are seen as being cultural, so that a similar set of choices under similar constraints are markers of shared cultural knowledge.
-Here, the archaeological research goal also lies in the identification of spheres of cultural commonality which can be distinguished from one another.
+The 40 years between 1970 and 2010 were a period of increasing archaeological research in Mali, West Africa.
+In the region around the middle reaches of the river Niger, this research documented the remains of a dynamic civilization which, from the last millennium BC onwards, developed urbanism, craft specialization, and long-distance trade (McIntosh 1998), and was highly influential not only in changing narratives about the African past but also in global debates on sociocultural dynamics (McIntosh 1999).
+One of the characteristics of many settlement sites of the first-early second millennium AD in the region is the high volume of ceramic sherds, which sometimes challenged the capacities of research projects (McIntosh 1995, pp.
+132--133).
+Featuring a large variety of decorative motifs, forms, and production techniques, the pottery of the region has been the subject of numerous studies, both from an archaeological and ethnoarchaeological perspective.
+Three key points have emerged from this research history.
+Firstly, establishing chronologically sensitive typologies has been very difficult, with contemporary variability often more significant than diachronic change.
+Secondly, base forming techniques, which vary throughout the region, are likely to correspond to larger cultural and maybe also linguistic spheres.
+Thirdly, pottery vessels appear to have been objects of trade, at times over long distances.
 
-In the archaeology of West Africa, and especially in the wider areas surrounding the upper and middle reaches of the Niger, there has been a strong research focus on ceramics.
-From the 1st millennium BC onwards, they occur in extremely high numbers on archaeological sites, and their appearance, especially their surface treatment, is highly variable.
-In contrast to the Great Lakes region of Eastern Africa, for example, the number of different roulette decorations, painted, incised, and applied decors, but also the variety in form, is very high.
-Consequently, there has been considerable emphasis on ceramics in archaeological and also in ethnoarchaeological work.
-While much of the ethnoarchaeological work has been based on exploring *chaînes opèratoires*, the archaeological work has mainly relied on attribute-based analyses (the system employed by S.K.
-McIntosh at Jenné-jeno has been very influential in subsequent work).
-In a region where large archaeological sites abound and researchers are relatively thin on the ground, the main hope of the analyses usually is to create chrono-stratigraphical sequences in order to be able to give approximate periods to surface assemblages.
-Despite all the variability, it has been found that a great number of ceramic attributes are not particularly time-sensitive.
-On the contrary, most sites appear to have diverse assemblages in which the percentages of individual attributes vary, but persist over several centuries, sometimes millennia.
+In this paper, we consider decorative practices on 12^th^ century AD ceramics of this region.
+We use a network approach to ask how such practices might have spread, and, even when they are not useful as chronological markers, how they might still inform us about sociocultural dynamics of the ancient Middle Niger.
+This research aim reflects a long-standing concern of archaeologists in the area, but the network approach we adopt here shifts the perspective from ceramic production as representing sociocultural entities to the processes by which techniques spread.
+This results in connections, rather than boundaries, as the subject of research.
+Using network models, we ask what ceramics can tell us about the processes that integrated the Middle Niger as a cultural area despite its diversity, and where cultural exchange might have been facilitated or impeded, and why.
 
-Some researchers have used the attribute-based approach to define types, statistically derived for the most part, which typify the ceramics of a certain place of time (MacDonald's Facies, McIntosh's timeline).
-The spatiotemporal distribution of individual attributes has also been the subject of some important work, such as individual rouletted decors (Livingstone-Smith) or base-shaping techniques (Huysecom).
-The most comprehensive treatment so far is in the work of Anne Mayor, combining both shaping techniques and decorative practices from an archaeological and ethnoarchaeological point of view.
-Focused on the Bandiagara highlands in central Mali, Mayor uses her data to construct an impressive overview of the development of different techniques and the establishment of ceramic traditions, which she closely relates to ethno-linguistic groups.
-Mayor's work builds on a large body of data and decades of ethnoarchaeological fieldwork, and there is little to add to her conclusions.
-Here, however, we wish to show that an almost identical dataset as available to Mayor can generate quite a different perspective when analysed as a network, and when connectivity is emphasised over boundaries.
+## Middle Niger Ceramics:
 
-# Archaeological ceramics as networks
+In many contemporary societies in the region, pottery is made by occupational specialist groups (Bambara: *nyamakala*).
+Though there are exceptions, it is the cultural norm in the wider Middle Niger region that pottery is made by women of groups in which the men are specialists in metallurgy.
+There is some debate as to how far in the past this specialization developed, but it currently seems likely that it has its roots in the first millennium AD (McIntosh 1998, pp.
+177--178), and that it had spread throughout the region by the mid second millennium (Tamari 1991, 1995).
+There is no doubt that potting and metallurgical experts were of great significance to the economy of the Middle Niger during this period of urbanism and emerging statehood, since they made basic products used in most parts of everyday life, from food production to the religious sphere.
+The cultural innovations linked to pottery making, both in social terms (endogamous craft specialists) and in technology (pottery and metal products), are a key factor in the development of the Middle Niger civilization (McIntosh 1998).
 
-In recent years, the use of network analysis in archaeology has been growing significantly.
-A large number of these studies focus on ceramics, but they are geographically mainly limited to North America (e.g.
-Hart 2012, Chiu 2019, Barker and Young 2017, Crabtree 2015, Mills et al.
-2013, Peeples and Haas 2013) and the Mediterranean (e.g.
-Sternini 2019, Iacono 2016, Brughmans and Poblome 2016 torrence & Swadling 2008, de Groot 2019).
-The archaeological records of these regions are amongst the best studied in the world, with fine-grained pottery typologies and extensive excavations.
-To our knowledge, this is the first application of a network approach to archaeological material from sub-Saharan Africa, where the conditions are somewhat different.
+From the 1st millennium BC onwards, ceramics occur in high numbers on archaeological sites on the Middle Niger, and their appearance, especially their surface treatment, is highly variable.
+The number of different roulette decorations, painted, incised, and applied decors, but also the variety in form, is very high, especially when attention is paid to the techniques and tools of decoration, rather than exclusively to their effects.
+Consequently, there has been considerable emphasis on ceramics in archaeological and in ethnoarchaeological work.
+Early studies (Bedaux 1980; e.g.
+Curdy 1982; and particularly Bedaux and Lange 1983) noted both regional differences in the appearance of pottery as well as striking widespread similarities in individual forms or attributes.
+Subsequent work has seen both in-depth studies on ceramics of individual sites (e.g.
+McIntosh 1995; Schmidt et al.
+2005; Gallin 2011; van Doosselaere 2014) as well as studies on the spread of particular decorative practices (Livingston Smith 2007) or shaping techniques (e.g.
+Sterner and David 2003; Mayor 2011a).
+Three main features of the wider region's ceramic record are important to this study: 1) the relative lack of clear temporal change, 2) high contemporary variability, and 3) pottery as exchange items.
 
-Ethnographic studies of pottery manufacturing in Africa suggest that the way that pottery is made depends largely on systems of learning and teaching the craft (Gosselain), on local market consumer choices (Dietler & Herbich), or on marriage patterns amongst manufacturing groups (ibid.).
+### Lack of temporal markers:
+
+Susan K.
+McIntosh's analyses of the ceramics of Jenné-jeno and surrounding sites resulted in a first detailed pottery chronology on the Middle Niger.
+In two studies (McIntosh and McIntosh 1980; McIntosh 1995), McIntosh was able to show how some pottery shapes, fabrics and decorations showed clear temporal trends.
+There was, however, also quite strong stability, particularly in many of the decorative techniques (McIntosh and McIntosh 1980, pp.
+233--244).
+Subsequent studies on other sites have also shown change in ceramic production, and particularly in its decorative aspects, to have been very gradual, and stability to have been the norm (e.g.
+Schmidt et al.
+2005).
+This contrasts with the archaeological record in many other regions of the world, where the outward appearance of pottery has been far more changeable over time and can therefore be used as chronological markers.
+This has not been possible for most periods and areas of the Middle Niger region, where decorative aspects tend only to vary in their proportion through time.
+
+### Ceramics and social boundaries:
+
+The more striking differences between archaeological ceramics in the area are not diachronic, but contemporary.
+Pottery from different areas varies in its forming techniques, its shapes (though the most pottery is globular or ovoid in its basic shape), and in decorative practices.
+Occasionally, as at Tongo Maaré Diabal (Gestrich and MacDonald 2018) or Kumbi Saleh (van Doosselaere 2014), different productions can be clearly separated.
+The mechanisms underlying these differences have been the subject of extensive ethnoarchaeological and archaeological research by the University of Geneva.
+Led from the outset by the goal of relating ceramic production to ethnic groups (Gallay et al.
+1998, pp.
+15--22), this extensive research documented how certain parts of the ceramic production -- notably base forming techniques and tools employed in the potting process -- could be employed to create typologies that were patterned by the ethnic identity, economic milieu, and ecological zone of the producers (Gallay et al.
+1998, pp.
+106--107; Mayor 2011b, p.
+14).
+But certain decorative techniques were also found to have a certain correlation to the ethnicity-based typology created by the Geneva team (Mayor 2011b, pp.
+40--78).
+The main limiting factor here, as Mayor stresses, is that "stylistic borrowings are relatively easy and frequent" (Mayor 2011b, p.
+78), but given the fact that some of the decors do correlate with ethnic groups --- or at least with the types assigned to ethnic groups --- Mayor hints at the suggestion that decorative techniques can historically be assigned to one or another group from which they originated (ibid.).
+
+Based on their extensive fieldwork, Mayor's hypothesis is that pottery making is learnt within stable and relatively impermeable family relationships.
+This is in keeping with some of Gosselain's results from Niger, where he has suggested that this initial approach to the shaping of a pot was the most influenced by teaching and learning (Gosselain 2000, p.
+193).
+Since this usually happens from an early age within the family, it makes sense that kinship relationships, and thus often group and linguistic belonging, would influence the propagation of such techniques.
+These processes of knowledge exchange, limited to endogamous groups and ethnic milieus, are what explains the commonalities that underlie the typologies created by the Geneva group.
+Exchange outside of these limits is seen as the exception, rather than the norm.
+Nevertheless, Mayor gives the following scenarios for the diffusion of pottery traits (Mayor 2011b, p.
+83): Firstly, when a technique is simple and expedient, thus easy and profitable to learn.
+Secondly, when an ethnic group is in the minority in a certain area or when communities find themselves at the geographical margins of their ethnic group.
+Thirdly, rare cases in which the social order is transgressed by marriage outside of the ethnic group.
+Mayor thus expects that shaping and decorative techniques would have had a single and ethnically distinct origin and that it is "several centuries of borrowing" (ibid.) that have led to the less demarcated situation that the Geneva team observed in the 1990s and 2000s.
+
+### Ceramics as trade goods:
+
+Even though most ceramic production appears to have been made for local markets, there is also evidence of pottery being used in trade over longer distances.
+Historically, the practice of pottery being moved on the Niger by boat and sold widely throughout the region has been documented in an account from the 18^th^ century (Park 1799, p.
+219).
+From the archaeological evidence, it seems as though this was not confined to later periods, but at least occasionally practiced also in the early second millennium and before.
+Finds of imported ceramics have, for instance, been made at Essouk (Nixon 2017, p.
+143,146-147), and at Kumbi Saleh (van Doosselaere 2014, pp.
+228--233).
+We must therefore always consider the possibility that archaeological assemblages do not only reflect local production, and add the indirect contact from long-distance trade to the possible mechanisms of change in ceramic practices.
+
+If the goal of the work by Gallay and Mayor and their teams was to identify the borders between groups in the past, this study asks a different question: how was pottery knowledge transmitted?
+What were the factors limiting or encouraging the flow of information?
+We thus begin from a null hypothesis of unlimited sharing, rather than of delimited groups, to see how the results might complement those of the Geneva research teams.
+Using pottery décor as a relatively easily borrowed attribute, we aim to generate hypotheses on how pottery making knowledge spread along the ancient Middle Niger.
+Further, we wish to demonstrate the use of network analyses for approaching West African archaeological ceramics and for making such connections visible and explorable.
+This technique presents a change of perspective on the ceramics of this region, allowing questions of connection to take precedence over those of boundaries.
+
+## Ceramic decors as networks
+
+In recent years, the use of network analysis in archaeology has been growing significantly (see e.g.
+Knappett 2021).
+A large number of these studies focus on ceramics, but geographically most important are North America (Hart 2012; Mills et al.
+2013; Crabtree 2015; e.g.
+Barker and Young 2017; Vavrasek 2020) and the Mediterranean (Torrence and Swadling 2008; Brughmans and Poblome 2016; Iacono 2016; de Groot 2019; Sternini 2019).
+The archaeological records of these regions are amongst the best and longest studied in the world, with fine-grained pottery typochronologies and extensive excavations.
+To our knowledge, this is the first application of a network approach to archaeological ceramic material from sub-Saharan Africa, where the conditions are somewhat different (McIntosh 1995, pp.
+130--131).
+That is not to say that archaeological pottery on the Middle Niger is poorly studied.
+Most researchers in the wider area have detailed publications on their ceramic assemblages, efforts have been made at systematizing nomenclature (Haour et al.
+2010) and at synthetic overviews (Mayor 2011b).
+A large body of ethnographic and ethnoarchaeological literature provides the theoretical basis that informs the network model.
+
+Ethnographic studies of pottery manufacturing in Africa suggest that the way that pottery is made depends largely on systems of learning and teaching the craft (Gosselain 1998), on local market consumer choices (Dietler and Herbich 1994), or on marriage patterns among manufacturing groups (ibid.).
 In short, the processes underlying the propagation of technical practices through time and across geographical space are dependent on communication and, more specifically, knowledge exchange.
-Such exchange occurs between humans and we can therefore understand similarities in the *chaine opératoire* as evidence that knowledge was shared.
+Such exchange occurs between humans and we can therefore understand similarities in the *production* as evidence that knowledge was shared.
 However, different elements of the pottery making process are thought to be reflective of different kinds of social interaction.
-Gosselain (REF), whose research in the years between XX and XX explored this question, concludes that, on the whole, forming techniques are more stable, reflective of durable cultural ties, sometimes even ethnic groups.
+Gosselain (2000) concludes that, on the whole, forming techniques are more stable, reflective of durable cultural ties, sometimes even ethnic groups.
 On the other end of the spectrum, decorations are thought to be influenced by far more ephemeral contact, in which seeing someone else's pot might be enough to change a potter's production, as could indirect contact via the wishes of customers.
 
-SNA shifts the focus "from spatial relations to social relations" (cf. Knappett 2011)
-Mills et al.
-(2015) have been able to compare similarity in wares across sites, but this cannot be done here.
-They also highlight Knappet's (2011) idea of different scales in archaeological networks.
-Because of the specific use case scenario here, we are conflating two fo his scales.
-He sees the community of practice as a meso-scale effect, across maybe a few households.
-However, we can see in the archaeological ceramic data from Mali that communities of practice in individual stages of the *chaîne operatoire* is often much larger.
-Thus, certain parts of the pottery manufacturing sequence are distributed across larger or smaller networks, but rarely remain at the household level.
-Also, we do not really have the luxury of being able to pinpoint certain wares, since typology is not as established.
+Decoration on pottery vessels is a topic that has been central to archaeological debates for a very long time.
+To construct an overly simplified history of ceramic decors in African archaeology, we might distinguish an early period analogous to other parts of the world, in which decors (as a main constituent of typologies) served to distinguish human groups.
+This position became more elaborate and changed rationale over time, arguably culminating in Hodder's "Symbols in Action" (1982).
+David et al.'s "Why pots are decorated" (1988) did much to remove confidence in this idea and to situate decorative practices as part of chains of production and expressions of often unconscious propagation of social values.
+For them, "designs on pottery, far from being 'mere decoration,' art for art's sake, or messages consciously emblemic of ethnicity, are low-technology channels through which society implants its values in the individual --- every day at mealtimes" (David et al.
+1988, p.
+379).
+Over the intervening years, the comparatively strong ethnoarchaeological research in Africa has developed an increasingly complex picture of ceramic décor.
+Gosselain's (2011) "pourquoi le decorer", part review, part programmatic article, gives a good overview : Decorations on pottery occur within cultural constraints and logics, are expressions of individuality, are market-oriented, have practical considerations, are made for visual effect, have profound meanings while being thought of as trivial, are subject to fashion and taste, stand for group membership, mean different things to different people.
+In short, pottery decoration is not anchored in any stable way to any cultural phenomenon.
+More importantly for this study, however, Gosselain points out that the propagation, the process of spreading, of decorative practices is a more fruitful topic for study (ibid.: 12).
+Various decorative practices presuppose various kinds of contact.
+Some of these, as in the case of a particularly intricate rouletted tool, require the sharing of cultural knowledge and probably reflect shared practices in other cultural domains as well.
+Others are transmitted more easily, by sight alone.
+Whatever these mechanisms, there are constraints on transmission and innovation.
+As Herbich (1987, p.
+202) notes, potters innovate and express themselves with reference to other potters they interact with.
+Therefore, the innovation and spread of decorative practices does not occur freely, but is helped by geographical proximity, social contact, and cultural factors.
+It is likewise hindered by geographic distance and cultural differences.
 
-We need to be aware of the fact that we are conflating consumption with production here.
-This is why we will need to threshold the networks, to make sure that a site represents the significant potting communities within its area.
+This, then, is the basis on which our social network model is conceived.
+It supposes that co-occurrence of a given practice of decorating pottery is emblematic of a form of sustained social contacts, within the limitations mentioned above.
+These contacts are not as close as we would expect from a teaching and learning environment, which would be better reflected in forming techniques, but are sustained and meaningful enough for there to have been an exchange of cultural aesthetic values and an exchange of knowledge of how to decorate pottery.
+These exchanges are therefore what the ties in our network represent.
+
+In choosing sites as the nodes of the network, the data come mainly from contexts in which pottery was consumed, rather than produced.
+The network agents in the model are therefore communities of consumers.
+Their ties, based on shared decors on their pottery, are indirect: they must be understood to have been connected via producers (individuals and communities) and possibly also merchant middlemen which remain largely unseen in the archaeological record.
+Since we have established at least occasional long-distance trade in pottery in our research area (see above), this is clearly a factor limiting the model's explanatory power.
+Nevertheless, we can use the work carried out by the Geneva MAESAO project to suggest that the majority of pottery found at a site will generally be locally produced (Mayor 2011b, p.
+84), which might reasonably allow some conflation between pottery producers and consumers in the network nodes.
+A low level of imported ceramics can be adjusted for in the network model, but should not be excluded, as these "weak ties" have been shown to be important for introducing innovation in social networks (Granovetter 1973), including technical innovation in ceramic production (Roux et al.
+2018).
+
+The analysis of the network will attempt to generate hypotheses on how such knowledge exchanges were structured, the level of interaction between inferred producers of ceramics, and the overall level and structure of cultural contact in the region.
+This differs somewhat from the rationales for pottery décor networks in studies from other parts of the world.
+Hart and Engelbrecht (2012, pp.
+324--330) justify their building of networks based on pottery décor similarity by arguing for décors as signals in which both group membership and individual skill are expressed.
+Iacono (2016), on the other hand, takes the co-occurrence of pottery decoration as analogous to other forms of social encounters, and thus as representative of them, conveying the agency of individuals "simply through sensory recognition" (Iacono 2016, p.
+126).
+We would like to stress here that these differences are in no small part due to the history of research and the nature of the archaeological record in different parts of the world, and that the building of network models on archaeological data always requires a well-theorised foundation appropriate to the historical context of the studied material.
 
@@
 # import decor data
-decors <- read_csv2(here::here("analysis/data/raw_data/decors_late.csv"))
+
+decors <- read_csv2(here::here("analysis/data/raw_data/decors_late.csv")) |>
+  select(-1)
+
 # import information on sites
@@
 
-```{r datatable, fig.cap="decor data used for network construction"}
-library(kableExtra)
-kable(decors, digits = 1, "html") %>% 
-  kable_styling("condensed") %>% 
-  save_kable(here::here("analysis/figures/datatable.html"))
-```
+## Data
 
-In this paper, we use pottery decor data from contexts occupied in the 11th century AD to attempt a reconstruction of networks through which pottery decorative practices were shared.
-The data is in a common format of percentages of prevalence in the assemblage (see \@ref(tab:datatable) This particular time slice, which represents a cumulative view of a great many individual interactions, will then be compared to earlier contexts in the same wider area, in order to demonstrate the diachronic possibilities of the approach.The sites we have been able to include in the network model are shown in \@ref(fig:sitemap).
+In this paper, we use pottery decor data from contexts occupied in the 12th century AD to attempt a reconstruction of networks through which pottery decorative practices were shared.
+The network is constructed from standard archaeological frequency data, in which the proportions of decorative practices are represented across the selected site phases (the raw data are available as Annex 1).
+The decorative practices encode both the tool used and the action performed, for instance a twisted cord (tool) roulette (action).
+This data is partly taken from existing literature, partly relies on our own fieldwork (TMD and TOG2) or raw data raised by other researchers (TBZ), and was partly documented in collections (Gao-Saney, Tellem).
+This time slice, which represents a cumulative view of a great many individual interactions, will then be compared to earlier contexts in the same wider area, in order to demonstrate the diachronic possibilities of the approach.
+The sites we have been able to include in the network model are shown in \@ref(fig:sitemap).
 We can see here that sites are geographically unequally distributed.
@@
 Global politics and research priorities over the last 50 years have meant that these are areas we do not currently have excavated and dated ceramics for, though we know that sites exist, often in abundance.
-This means that, when we construct networks, we do so with a geographically unrepresentative sample, which leads to incomplete networks.
-Sindbaek has pointed out the difficulty in this situation: the actual network is not known, only its effects, from which the network can be partially reconstructed.
-Accordingly, this study should be seen as an exercise first in the reconstruction of a possible network that led to the data preserved in the archaeological record: a tiny sample of a tiny sample, influenced by an unknown amount of mostly unknowable pre- and post-depositional biases.
-Nevertheless, we will attempt to analyse the reconstructed network in order to formulate some hypotheses about the nature of innovation and spread in ceramic decors.
+This means that in addition to the inherent difficulty of archaeological network models -the actual network is not known, only its effects, from which the network can be partially reconstructed (Sindbaek (2013)-, the effects of missing areas or sites on the network need to be calculated (Peeples et al.
+2016).Accordingly, this study should be understood as an exercise in the reconstruction of a possible network that led to the data preserved in the archaeological record: a tiny, unrepresentative sample of a tiny unrepresentative sample, influenced by an unknown amount of mostly unknowable pre- and post-depositional biases.
+Nevertheless, we will attempt to analyse the reconstructed network to formulate some hypotheses about the nature of innovation and spread in ceramic decors.
 Future generations of researchers are encouraged to improve upon this initial attempt.
@@
 ```{r sitemap, fig.cap="location of sites in the analysis", echo = FALSE}
-source(here::here("analysis/R/map plots.R"))
+source(here::here("analysis/R/map plots.R")) # Load the required R script
 
-sitepoints <- st_as_sf(sites, coords = c("lon", "lat"), crs = 4326, agr = "constant")
+#saving site location as sf
+sitepoints <- st_as_sf(sites, coords = c("lon", "lat"),
+                       crs = 4326, agr = "constant")
 
-sitemap <- ggplot()+
+#creating a map of mali with the sites
+sitemap <- ggplot() +
   map_bw_grid +
@@
 
+#ploting map with inset
 sitemap + inset_element(inset, 0.72, 0.72, 1, 1)
@@
 
-The network is constructed from standard archaeological frequency data, in which the proportions of decorative practices are represented across the selected site phases (REF FIG XXX).
-The decorative practices encode both the tool used and the action performed, for instance a twisted cord (tool) roulette (action).
-This data is transformed into a network by deciding that every decor that is shared between two sites leads to a link between these sites.
-The link is weighted by multiplying the proportion at one site with the proportion at the other.
-This leads to a network in which the nodes (sites) can be connected in more than one manner, known as a multiplex or multilayer network.
-As is shown in \@ref(fig:total-network), this leads to a very tightly connected network, known as a "hairball".
-However, the multiplex nature of the network allows us to make an important initial observations on its structure: the networks in which individual decor types are propagated are overlapping.
+## Network construction
+
+Our theoretical approach to what the network represents informs the way in which the network model is constructed.
+So far, the most frequent form of making ties in archaeological ceramic data has been by calculating the statistical similarity between assemblages.
+(Hart and Engelbrecht 2012) use the Brainerd-Robinson similarity coefficient, as do, for instance, (Mills et al.
+2015), while (de Groot 2019) uses the Jaccard and Kulczyinski-2 similarity measures.
+The similarity value is then used as the basis and weight for forming a tie between the two assemblage nodes.
+Since we did not only want to base our network on overall similarity but wanted to be able to closer analyze the networks of individual decors, we have constructed a multiplex network.
+This is a network in which edges can have different types, and thus the network can be envisaged as having layers.
+A multiplex network is distinct from a multilayer network, in which connections can also exist between layers, and which has recently seen its first application to archaeological ceramics by (Upton 2019).
+In our multiplex network, each decorative practice is represented as a distinct layer.
+The layers share the site assemblages as nodes, though not every node is necessarily represented on every layer.
+The ties are weighted by the product of the percentages of a given décor in two assemblages.
+This network construction is implemented in R (R Core Team 2022) using the tidygraph package (Pedersen 2022).
+
+# Analysis
+
+## Multiplex structure
+
+As is shown in \@ref(fig:total-network), the initial structure of the data leads to a very tightly connected network, known as a "hairball".
+This is normal for archaeological networks based on co-occurrence or similarity, since even rare occurrences lead to a tie.
+However, the multiplex nature of the network allows us to make an important initial observations on its structure: the networks in which individual decor types are propagated are only partly overlapping.
 There are no two decor networks that cover one another entirely.
-This emerges more clearly in \@ref(fig:total-network-sample), when the number of decorative practices is reduced.
-This indicates that the adoption of decorative techniques ...
+This pattern emerges more clearly in \@ref(fig:total-network-sample), when the number of decorative practices represented on the graph is reduced.
+This structure suggests that the networks allowing decorative practices to propagate were at least partially different in every case.
+Remembering the constraints on spread identified by Gosselain (2011, see section 2), we might propose that external factors, such as geography and cultural differences, provided "frontiers" to the spread, albeit different ones in each case.
+Some decors are very widespread, and thus had fewer inhibitions to their spread, while others occur less frequently.
+ We can now proceed to further analyse the network in order to identify where such frontiers and catalysts of spread might have been and how they were structured.
 
@@
 # construct network, remove extraneous sites
-decor_graph <- tbl_graph(sites, edgelist, directed = FALSE) %>%
-  activate(nodes) %>%
+decor_graph <- tbl_graph(sites, edgelist, directed = FALSE) |>
+  activate(nodes) |>
   filter(!node_is_isolated())
@@
 ```{r total-network, fig.cap="Plot of 11th century AD network of decorative practices. The plot shows the very strongly connected structure of the network."}
+# Creating a multi-layer graph
 ggraph(decor_graph, layout = "stress") +
@@
 ```{r total-network-sample, fig.cap="Partially overlapping networks of four decorative practices", echo = FALSE}
-decor_graph %>%
-  activate(edges) %>%
-  filter(name %in% c("braided_cord_roulette", "braided_strip_roulette", "paint", "straight_cord_mat")) %>%
+decor_graph |>
+  activate(edges) |>
+  # Filter "edges" based on 4 specific decorations
+  filter(name %in% c("braided_cord_roulette",
+                     "braided_strip_roulette",
+                     "paint",
+                     "straight_cord_mat")) |>
+
+  #plot the graph
   ggraph(layout = "stress") +
@@
 
-Using the `multinet` package, the statistical distance (Jaccard similarity) between the different decors' networks were calculated based on the nodes they contained.
-We see in \@ref(fig:layer-comparison) that some networks are more similar to one another than others.
-This leads us to ask whether there might be groups of decors that were shared within similar networks, and thus, whether there might be groups of sites predominantly sharing specific decors.
+In the first place, we will determine whether there are decors that share similar knowledge exchange networks and others that are dissimilar.
+Using the multinet package (Magnani et al.
+2021), the statistical distance (Jaccard similarity) between the different decors' networks were calculated based on the nodes they contained.
+
+\@ref(fig:layer-comparison) shows that some networks are more similar to one another than others.
+For instance, the practice of rouletting with a braided cord  (McIntosh and Gueye 2010) appears to have a significant shared network with that of rouletting with a folded strip (Haour and Keita 2010).
+This means that knowledge of the two tools necessary for these roulettes (for details, see Haour et al.
+2010) spread in a similar geographic and cultural sphere, without being impeded.
+On the other hand, there appears to have been quite strong boundaries in place between the networks sharing knowledge on rouletting with a braided strip and dragging a comb along the surface of the pot or using painted decoration.
+There thus appear to have been groups of decors that were shared within similar networks, and we can thus ask whether there might have been communities in the network predominantly sharing specific decors.
 
@@
 # reduce graph to types with over 5 edges
-good_edges <- decor_graph %>%
-  activate(edges) %>%
-  as_tibble() %>%
-  count(name) %>%
+good_edges <- decor_graph |>
+  activate(edges) |>
+  as_tibble() |>
+  count(name) |>
   filter(n > 5)
 
-reduced_graph <- decor_graph %>%
-  activate(edges) %>%
+#creating a reduced graph with only those edges
+reduced_graph <- decor_graph |>
+  activate(edges) |>
   filter(name %in% good_edges$name)
 
-# prepare graph and make a vector of different layer names
-prep.graph <- reduced_graph %>%
-  activate(edges) %>%
-  rename(layer = name) %>%
-  activate(nodes) %>%
+# prepare graph for multinet by renaming
+prep_graph <- reduced_graph |>
+  activate(edges) |>
+  rename(layer = name) |>
+  activate(nodes) |>
   rename(name = label)
 
-layers <- prep.graph %>%
-  activate(edges) %>%
-  as_tibble() %>%
-  distinct(layer) %>%
+#make a vector of different layer names
+layers <- prep_graph |>
+  activate(edges) |>
+  as_tibble() |>
+  distinct(layer) |>
   pull(layer)
 
-# get a list of individual networks using the onelayer function (/R/network functions.R)
+# get a list of individual networks using
+#the onelayer function (/R/network functions.R)
 graphlist <- purrr::map(layers, onelayer)
@@
 
-# Clustering
+## Community detection
 
-We use the Louvain clustering algorithm (Blondel et al 2008) to find communities in the network.
+We approach this question by clustering, or community analysis, which a number of recent studies have successfully used on archaeological netoworks.
+Radivojevic and Grujic (2018) have clustered sites on the Balkan based on the source of their copper, Mazzucato (2019) has identified communities of building sites in Çatalhöyük to analyse the process of urbanisation, while Vargas et al.
+(2020) used community detection to find distinctive clusters of rock art sites in Patagonia.
+These studies differ significantly as regards the object of research and their questions, but are similar in data structure and methodological approach.
+All three used the Louvain clustering algorithm (Blondel et al.
+2008) to detect groups of sites that share a lot of edges with each other and have fewer edges with nodes outside of their community.
+The Louvain method is especially useful for archaeological datasets that typically contain few but heavily connected nodes (Blondel et al.
+2008; Yang et al.
+2016).
+
 ```{r community detection, hide = 'results', echo=FALSE}
 # use louvain clustering method to assign clusters to nodes
-decor_graph <- decor_graph %>%
-  activate(nodes) %>%
+decor_graph <- decor_graph |>
+  activate(nodes) |>
   mutate(louvain = as.character(group_louvain()))
@@
 # calculate modularity of louvain clustering
-modularity <- decor_graph %>%
-  activate(nodes) %>%
-  mutate(modularity = graph_modularity(group = as.factor(louvain))) %>%
-  pull(modularity) %>%
+modularity <- decor_graph |>
+  activate(nodes) |>
+  mutate(modularity = graph_modularity(group = as.factor(louvain))) |>
+  pull(modularity) |>
   head(1)
 ```
-The louvain clustering has a modularity of `r modularity` .
+
+Here, we employ the same clustering algorithm, which results in three communities with a modularity of `r modularity`.
+[Modularity shows how clearly the communities can be defined]{.underline} on a scale [from -1 to 1]{.underline} (1 [being a network with completely separated sub-networks]{.underline})[.]{.underline} As is shown in \@ref(fig:plot-clusters), community 1 contains seven sites in an area centered on the Inland Niger Delta.
+Community 2 also contains seven sites, but located in the so-called "Dogon Country", on the eastern edge of the IND, on the Bandiagara plateau and in the Seno plain.
+Community 3 has the fewest nodes, but geographically the largest extent, containing the southernmost site of the study near Ségou, Togu Missiri, as well as the northernmost at Essouk in the Sahara.
+The nodes of community 3 show frequent connections to sites of communities 1 and 2, while those communities are not as strongly connected between each other.
 
@@
 
+# creating a palette with the proper amount of colors
+col_21 <- colorRampPalette(brewer.pal(8, "Set2"))(21)
+
+# then we create a vector containing all decortypes
+decortypes <- decor_graph %>%
+  activate(edges) %>%
+  as_data_frame(what = "edges") %>%
+  select(name) %>%
+  unique()
+
+decortypes <- as.vector(decortypes[, 1])
+
+# finally we make a named vector of the colors
+decorcolor <- setNames(col_21, decortypes)
+
+#plot:
 # disable s2 geometry to plot on map
@@
   edges_community +
-  ggforce::geom_mark_hull(aes(x = lon, y = lat, fill = louvain, colour = louvain),
+  ggforce::geom_mark_hull(aes(x = lon, y = lat,
+                              fill = louvain, colour = louvain),
                           expand = unit(1.5, "mm"),
                           radius = unit(1.5, "mm")) +
-  scale_fill_manual(values = communitycolor)+
-  scale_colour_manual(values = communitycolor)+
+  scale_fill_manual(values = communitycolor) +
+  scale_colour_manual(values = communitycolor) +
   nodes_community +
@@
 
-There are some recent studies in archaeology that use community detection on archaeological networks.
-Radivojevic and Gruci (2018) have clustered sites on the Balkan based on the source of their copper, Mazucato (2019) has identified communities of building sites in Çatalhöyük to analyse the process of urbanisation, while Vargas and Lanata (2020) used community detection to find distinctive clusters of rock art sites in Patagonia.
-These studies differ significantly regarding the research object and question, but are similar in data structure and methodological approach.
-All three used the Louvain method to detect groups of sites that share a lot of edges with each other and have fewer edges with nodes outside of their community.
-The Louvain method is especially useful for archaeological datasets that typically contain few but heavily connected nodes (Blondel et al.
-2008/ Yang et al.
-2016).
+The clustering of the network appears to be influenced quite strongly by geographic features.
+In the case of community 1, this is the IND and its adjacent areas, which were heavily populated at this time (McIntosh 1995, pp.
+373--377).
+The population density is likely to have been a significant factor in the spread of pottery decors, as Gosselain has suggested.
+\@ref(fig:cluster1) shows that there is a very well connected group of geographically close sites in the eastern IND (Jenné-jeno (J-J), Hambarketolo (Hamb), and Kaniana are part of the same wider site cluster).
+This community mainly shared braided cord roulettes, dragged comb incisions.Interestingly, this also includes the site of Kawinza in the Malian Lakes Region (Raimbault and Sanogo 1991).
+Sites on the western margin of the IND, Dia and Akumbu, are less well connected in this period, conforming to the observations of McIntosh (1995, pp.
+369--370), but these remain connected via folded strip roulettes and painted decors.
 
-Here, we employ the same clustering algorithm, which results in three communities with a modularity of 0.3.
-See \@ref(fig:plot-clusters) Community 1 contains 7 Sites in the Dogon Area.
-Community 2 is in the Niger bend and the IND, it also contains 7 sites.
-Both only include sites south of the Niger.
-Community 3 is the biggest and most widely distributed community.
-Reaching from Gao to Togu Missiri in the Ségu area it seems to mostly contain sites around the Niger river.
-The nodes of community 3 often connect with sites of community 1 and 2, while those communities are not connected strongly.
+```{r cluster1, fig.cap="Internal connections within community 1"}
+decor_graph %>%
+  activate(nodes) %>%
+  mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 1) %>%
+  ggraph(layout = "graphopt") +
+  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
+  scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+  scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "cornflowerblue") +
+  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
+  theme_graph()
+```
 
-There are two major geographic features influencing the clusters.
-The first one is obviously the Niger river connecting the sites of community 3.
-The bend of the Niger river and the Inner Niger delta are the location of various sites of community 2 and probably a zone of contact between this community and community 3.
-The other geographic landmark that is strongly related to the communities is the Bandiagara Plateau.
-It is the border between community 1 and 2.
-Both communities overlap in this area.
-TMD, a site of community 3, lies in the north of the mountains making them the region where all 3 communities meet each other.
+The main geographic factor affecting community 2, and isolating it from community 1, is the mountainous region of the Bandiagara plateau.
+The cluster has strong links, mostly because of the common use of braided strip roulettes and impressions of a straight-woven cord mat (\@ref(fig:cluster2).
+The networks exchanging these decors, which are in part linked to base forming techniques, appear to have had a spatial component, with mat impressions linking sites in the Seno plain while braided strip roulettes are more strongly shared on the Bandiagara plateau.
+These spheres, however, had considerable overlap, most notably at the Sangha cave sites (Tellem) and at Sadia.
+
+```{r cluster2, fig.cap="Internal connections within community 2"}
+decor_graph %>%
+  activate(nodes) %>%
+  mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 2) %>%
+  ggraph(layout = "graphopt") +
+  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
+  scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+  scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "tomato") +
+  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
+  theme_graph()
+```
+
+Community 3 appears to be linked through the wider Niger river system and the long-distance exchange routes that linked into it.
+This cluster is the most diverse one regarding the types of decoration techniques shared (\@ref(fig:cluster3)), yet the weak nature of the resulting ties shows that this might be due to more exchange of artifacts and less in common shared knowledge across artisans.
+This is particularly visible at Essouk, which has the weakest connections that, due to its position and cultural history (Nixon 2017), are probably due to imports more than a shared cultural sphere.
+All of the sites in the community have been linked to long-distance trade (Insoll 1996; Nixon 2017; Gestrich and MacDonald 2018; Gestrich et al.
+2021), though this applies just as well to some of the sites in the other communities.
+The fact that these cluster together is likely due to the fact that their closer local communities are not part of the network model.
+This, however, allows us to see such ephemeral, long-distance connections based on weaker ties that might be obscured by more intense local exchange.
+
+```{r cluster3, fig.cap="Internal connections within community 3"}
+decor_graph %>%
+  activate(nodes) %>%
+  mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 3) %>%
+  ggraph(layout = "graphopt") +
+  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
+  scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+  scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "darkgreen") +
+  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
+  theme_graph()
+```
+
+The measures of network centralization and the mean weight of the edges relative to their number show us that communities 1 and 2 are more equally connected with higher weights, whereas community 3's network is more centralized, thus more uneven, but also much less intense (@ref(fig:cluster-comparison)).
+This suggests to us a fundamental difference in the nature of the networks.
+Communities 1 and 2 may have been founded on close cultural ties, frequent exchanges, and maybe even kinship bonds within specialized potter communities.
+Cultural frontiers, accentuated by geographical boundaries, hindered the spread of decorative techniques between them.
+Community 3 seems to have been a network based on far less intense contacts, such as sporadic movement of individuals in the context of trade or migration or indirect, down-the-line, exchange of knowledge.
+\@ref(fig:cluster_centralities) shows the distribution of the weighted degree and eigenvector centralities for each community within the overall network.
+Clearly, community 1 consists of nodes that both have more and stronger connections and consequently a higher potential for influence on the network.
+This reflects a number of points in the archaeological record.
+Firstly, the ceramics in the IND are highly decorated with diverse decors during this period.
+The clustered urban sites in this area are likely to have been culturally diverse, both drawing in populations from surrounding areas and exerting cultural influence across the wider region.
+McIntosh sees the Eastern IND as becoming increasingly culturally distinct from the Western and Northern IND during this period (McIntosh 1995, p.
+372), and the emergence of new material cultural practices may well be reflected in these measures.
 
 ```{r cluster-comparison, fig.cap="Relative edge weights and centralization by cluster"}
-# Relative edge weights (how important are the connections within the cluster)
-rel_edge_weight <- vector("double", length(3))
-for (i in 1:3) {
-  rel_edge_weight[[i]] <- decor_graph %>%
-    # filter community
-    activate(nodes) %>%
-    filter(louvain == i) %>%
-    # relative edge weight (sum of weights / number of edges)
-    activate(edges) %>%
-    as.data.frame() %>%
-    {
-      sum(.$weight) / nrow(.)
-    }
-}
+# adding community/crossing to the edges
+decor_graph <- decor_graph |>
+  add_community_edges()
+
+# mean edge weights
+mean_edge_weight <- decor_graph |>
+  get.data.frame("edges") |>
+  filter(community != "crossing") |>
+  group_by(community) |>
+  summarise(mean_weight = mean(weight))
 
@@
 for (i in 1:3) {
-  centralization[[i]] <- decor_graph %>%
+  centralization[[i]] <- decor_graph |>
     # filter community
-    activate(nodes) %>%
-    filter(louvain == i) %>%
+    activate(nodes) |>
+    filter(louvain == i) |>
     # degree centralization
-    centr_degree(mode = "all", normalized = TRUE) %>%
-    as_tibble() %>%
-    pull(centralization) %>%
+    centr_degree(mode = "all", normalized = TRUE) |>
+    as_tibble() |>
+    pull(centralization) |>
     head(1)
@@
 
-# put together
-tibble(rel_edge_weight, centralization) %>%
-  mutate(cluster = 1:3) %>%
-  mutate(rel_edge_weight = rel_edge_weight / 100) %>%
-  rename("relative edge weight" = rel_edge_weight) %>%
-  pivot_longer(!cluster, names_to = "key", values_to = "value") %>%
-
-  # plot
-  ggplot(aes(x = cluster, y = value, fill = key)) +
+# Graph of mean edge weight of the different communities
+g1 <- mean_edge_weight |>
+  pivot_longer(!community, names_to = "key",
+               values_to = "mean_edge_weight") |>
+  ggplot(aes(x = community, y = mean_edge_weight)) +
   geom_bar(stat = "identity", position = "dodge") +
   scale_fill_grey() +
-  theme(axis.text.x = element_text(angle = 90)) +
+  theme_publish() +
+  ylab("mean edge weight")
+
+# Graph of centralization of the different communities
+g2 <- tibble(centralization) |>
+  mutate(community = 1:3) |>
+  pivot_longer(!community, names_to = "key", values_to = "centralisation") |>
+  ggplot(aes(x = community, y = centralisation)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  scale_fill_grey() +
   theme_publish()
+
+# plot:
+g1 + g2
 ```
 
-If we look at the edges (divided by 10000 to make it comparable) and the centrality of sites in regard to the communities, we see that while the sites of community 3 are the most central ones in the network the weight of the edges within the community is relatively small.
-For community 1 and 2 it is vice versa even though community 2 has weaker edges within and is better connected to the whole network.
-The more isolated a community is from the network, the stronger is its connection within the community.
---maybe the bar_plots of the most common decors of all clusters is enough? maybe the first barplot is enough? – maybe the network of every cluster is too much
+```{r cluster_centralities, fig.cap="Distribution of weighted degree and eigenvector centralities by community."}
 
- see \@ref(fig:cluster-comparison)
-If we take a closer look at the "internal" connections of community 1, we see that "braided_strip_roulette" is present at all sites.
-The other main decor "straight_cord_mat" on the other hand is missing on 2 sites: Oudalan and Prom.
-The two decors are spread across the whole area of the community.
-The only other decor that is common in the community is the "twisted_cord_roulette" which is also abundant across the whole research area.
-It makes up a rather large part of the assemblage in Oudalan and Prom but was not found at the sites of Sadia and Ambere Dougon.
-The fact that SAO has a high degree of centrality shows that the location is not decisive for the connectedness within this community.
-This is underlined by the fact that the less central sites are distributed across the whole area.
-There is an ambivalence between a relative consistency across the whole area and the diversity between sites that are close to each other.
-One reason for this phenom can be the contact to other communities in the mountain area.
-Damassogou exemplifies this as the site contains the defining decors of community 1 and 2.
-But the difference between Oudalan and SAO that are geographically isolated suggest that there might be more reasons for this diversity.
-Community 2 is less diverse with "braided_cord_roulette" as the only outstanding decor type.
-The weaker connections of the sites Kokolo and Damassogou are based on data bias (the proportion is counted in regard to all sherds not only the decorated ones), the decor was probably equally abundant there.
-Other decors spread across the community are "folded_strip_roulette" and "dragged_comb".
-The first one is missing in KAN 4 and TOD H while the second one is missing at Kokolo and Damassogou.
-The other indicator for the homogeneous character of the community is the nearly equal degree of centrality of every site.
-As mentioned above community 3 is more diverse but also has weaker connections.
-Instead of one defining decor there are several ones but none is abundant across the whole community.
-The spatial distribution differs between the decors.
-While "paint" and "twisted_cord_roulette" appear from Gao to TOG 2, "folded_strip_roulette", "channel" and "incised" are located in the western part of the community's area.
-Essouk and TMD have the highest degree of centrality.
-Even though it is located at the margin of the research area, its character as a "market"-site explains the amount of links it has.
-The role of TMD for the network is elaborated elsewhere (!).
+# Calculate centralities for each cluster
+centr <- pull_centralities_clust(decor_graph)
+
+# Display boxplots of centralities for degree centrality
+a <- ggplot(centr) +
+  geom_boxplot(aes(x = louvain, y = deg, fill = louvain)) +
+  scale_fill_grey(start = 0.5, end  = 1) +
+  theme_bw() +
+  theme(legend.position = "none") +
+  labs(y = "degree centrality",
+       x = "community")
+
+# Display boxplots of centralities for eigenvector centrality
+b <- ggplot(centr) +
+  geom_boxplot(aes(x = louvain, y = eig, fill = louvain)) +
+  scale_fill_grey(start = 0.5, end  = 1) +
+  theme_bw() +
+  theme(legend.position = "none") +
+  labs(y = "eigenvector centrality",
+       x = "community")
+
+# Combine the boxplots
+a + b
+```
+
+These three communities, however, were by no means separate from one another.
+If we consider only those edges that "cross" between communities, we can identify the decors that passed between communities, but also the sites that were important crossing points (\@ref(fig:crossings).
+These nodes are indicated by their weighted degree centrality, that is, the sum of the weights of all edges pertaining to the node.
+The sites of cluster 1 are generally better connected, both within the cluster and to sites outside their cluster.
+This might be a measure of the cultural influence and diversity of the IND's urbanized society in this period.
+The strongest node in this measure represents the assemblage from TMD (Tongo Maaré Diabal, Horizon 5 (Gestrich and MacDonald 2018)), which provides a crossing between all three communities.
+These results complement Gestrich's (2013, MacDonald & Gestrich 2018) result from correspondence analysis from which he claims that Tongo Maaré Diabal's ceramic assemblage reflects varying geographic influences.
+Even without the edges that tie the individual communities together, the network of crossing edges remains densely connected, demonstrating just how well connected the populations of the Middle Niger region must have been in this period, and that the limits of the communities we have detected in the model are far from being clear-cut cultural boundaries.
 
 ```{r crossings, fig.cap="Network of edges which cross between clusters"}
-decor_graph %>%
-  activate(edges) %>%
 
-  # find the edges that cross between communities
-  mutate(crossing = community_crossing(weights = weight)) %>%
+#show crossings with a weight over 3
+decor_graph |>
   filter(
-    crossing == TRUE,
-    weight > 200
-  ) %>%
-  activate(nodes) %>% 
-  filter(!node_is_isolated()) %>%
-
+    community == "crossing",
+    weight > 3
+  ) |>
+  activate(nodes) |>
+  filter(!node_is_isolated()) |>
   #plot
-  ggraph(layout = "nicely") +
-
+  ggraph(layout = "stress") +
   geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
   scale_edge_width(range = c(0.2, 2)) +
-  geom_node_point(aes(size = centrality_degree(weights = weight), color = louvain), show.legend = FALSE) +
+  geom_node_point(aes(size = centrality_degree(weights = weight),
+                      color = louvain), show.legend = FALSE) +
   geom_node_text(aes(label = label), col = "black", repel = TRUE) +
-  scale_color_manual(values = communitycolor)+
+  scale_color_manual(values = communitycolor) +
   theme_graph() +
@@
 
-Not only the connections within but also those between the communities should be examined.
-Therefore, only the edges that "cross" the two communities will be considered.
-While there is no obvious "broker"-site, community 3 could be described as a "broker"-community.
-There are few strong crossings between community 1 and 2.
-Even though most decors that cross communities connect all three, those are mostly decors that are most abundant at the sites of community 3.
-See \@ref(fig:crossings)
+## Sparsification
 
-The crossings also show the limits of the community detection at the research area.
-The clusters that were found are congruent with the results of earlier works and help to understand the network, but they do not capture all the networks and connections of the region.
-To grasp the dynamics of pottery production we must go beyond those boundaries.
+### Simplified network backbone
 
-For an archaeological interpretation of the results, we need to take a step back and examine what our edges represent.
-If we take decoration techniques as the result of cultural practices, the communities we detect may be interpreted as communities of practice.
-Those are the result of the exchange of knowledge with various facets.
-Intergenerational teaching is one of them and might link our communities to ethnic groups (as Mayor did).
-But this is only one aspect and we should also consider communities as trading spheres where the demands influence the production.
-The exchange of goods is also fuelling the exchange of ideas that might cross ethnic boundaries.
-Hence our communities are foremost spheres of similar cultural knowledge that was intertemporally and -spatially exchanged.
-
-Community 1 and 2 are spheres were the two decors respectively one decor dominates.
-The cultural practice was revolving around them.
-Even though there were other decors present at the sites, they might have played a local role but were not cons.
-Behind this might be a teaching network and demand, that at least is similar at the sites if not crossing their boundaries.
-The relatively weak links towards the outside show that these groups of practice were somewhat resilient.
-The difference regarding the centrality of the sites within community 1 demonstrates that the grade of involvement in this group of practice alter in this sphere.
-
-Community 3 is another type of sphere cultural knowledge.
-Instead of a distinct decoration technique the variety of those techniques was crucial.
-The exchange of knowledge on the production side but also the choices of consumers vary strongly.
-While there are some decors that belong only to this exchange network, especially the sites in the west are part of other spheres of cultural practice.
-
-Outcomes: three communities of different scales are identified.
-Two of these appear to be geographically bounded, and are maybe representative of intensive information exchange on a regional level, if not even within cultural groups.
-This ethno-linguistic dimension remains to be discussed re: Swiss research.
-The last community is not locally bounded and instead represents the exchange of information on ceramic decors through long-distance contacts.
-We might want to find out how well connected this network is in comparison to the two others.
-A lower network density or weight of ties would suggest that contacts are not as intensive.
-membership in these networks is not mutually exclusive, there are crossing points in all of them.
-As our initial look at the network structure shows, the networks of each individual decor overlap, and these groups should not be seen as bounded entities, but rather as nodes with a higher propensity for sharing information.
-The next step then examines the pathways through which information is likely to have travelled.
-
-# Sparsification
-
-# Simplified network backbone
-
-we can now try to extract a backbone from the simplified network
+While the hyperconnected hairball network has shown us key elements of the structure of pottery décor sharing, it is of less use for more traditional network analyses.
+These include finding key nodes in the network and key routes through it.
+In our case, such network statistics might tell us which of the sites, on the basis of their assemblages, were influential in the propagation of pottery decorations, and through which channels the knowledge of different décors might have preferentially passed.
+In order to extract such measures, the number of connections needs to be reduced.
+Network sparsification or 'backbone extraction', as this process is also called, is a common challenge in network analysis across its different applications (Serrano et al.
+2009; Saha et al.
+2013).
+In archaeological networks, this has frequently been achieved through thresholding, i.e.
+retaining only the strongest connections and discarding the others (as implemented for example in Hart & Engelbrecht 2012).
+There are, however, a number of other recent approaches to network sparsification.
+For our purposes, we have found the best results to come from a process called the "disparity filter" (Serrano et al.
+2009).
+This evaluates the weight of each edge against a null model in which the overall edge weights of a node are evenly distributed among its edges.
+If an edge is found to diverge from the null model at a set probability, it is retained.
+The backbone package (Domagalski et al.
+2021) implements this in R.
+In order to use it, we must first 'flatten' the multiplex graph, by adding the weights of the edges between each pair of nodes on all layers.
+We then extract the backbone of the graph using the disparity filter at different probability thresholds (\@ref(fig:backbone-extraction)).
 
 ```{r flatten}
-decor.flat <- flatten(decor_graph, weighted = TRUE)
+decor_flat <- flatten(decor_graph, weighted = TRUE)
 ```
 
-extract the binary backbone of the flattened graph and play around with the alpha to get a good representation. See \@ref(fig:backbone-extraction)
-
-```{r backbone-extraction, fig.cap="Extraction of binary backbones at decreasing significance levels: a) 0.05, b) 0.15, c) 0.20"}
+```{r backbone-extraction, fig.cap="Extraction of binary backbones at decreasing significance levels: a) 0.05, b) 0.15, c) 0.18"}
 # extract binary backbones at decreasing levels of significance
-decor.05 <- extract_backbone(decor.flat, 0.05)
-decor.15 <- extract_backbone(decor.flat, 0.15)
-decor.20 <- extract_backbone(decor.flat, 0.20)
+decor_05 <- extract_backbone(decor_flat, 0.05)
+decor_15 <- extract_backbone(decor_flat, 0.15)
+decor_18 <- extract_backbone(decor_flat, 0.18)
 
 # plot
-p1 <- plot.bb(decor.05)
-p2 <- plot.bb(decor.15)
-p3 <- plot.bb(decor.20) + geom_node_text(aes(label = label), repel = TRUE)
+p1 <- plot.bb(decor_05)
+p2 <- plot.bb(decor_15)
+p3 <- plot.bb(decor_18) + geom_node_text(aes(label = label), repel = TRUE)
 
-(p1 | p2) / p3 + plot_annotation(tag_levels = "a")
+(p1 | p2) / p3 +
+  plot_annotation(tag_levels = "a") +
+  plot_layout(heights = unit(c(1, 8), c("cm", "null")))
+
+decor_18 <- decor_18 |>
+  #isolated nodes are removed
+  activate(nodes) |>
+  filter(!node_is_isolated())
 ```
@@
 The three communities from the previous section are also visible in the extracted backbone, albeit with some major differences in configuration.
-The site of Kokolo NE Phase 2 is placed in the closely connected cluster of sites in the IND and Lakes region, featuring the sites of the Jenné cluster and Toguére Galia, Toguére Doupwil, and Kawinza.
-The spread-out cluster of long-distance contact now also includes the sites at Dia, while the "Dogon" cluster remains largely intact.
-Akumbu is isolated from the network, and remains so until the significance is lowered to 0.2.
+The spread-out cluster of long-distance contact now also includes the sites at Akumbu and Kawinza which were placed in the IND cluster.
+The latter is now reduced to the sites of the Eastern IND, while the Dia sites take a central position within the network, and the "Dogon" cluster remains largely intact.
+Dammassogou and Kokolo NE 2 are isolated from the network, and remain so until the significance is lowered to 0.2.
+We will use the backbone at alpha = 0.18 to proceed with our analysis.
+If this selection of significance values seems rather arbitrary, we point out that analogous situations exist in other attempts at archaeological network simplifications.
+Hart & Engelbrecht (2012, p.
+339), for instance, threshold their similarity network to the point where all nodes form a network but for a single outlier.
+Higher significance values also do not necessarily mean we get a more reliable network model, especially in a situation where sampling and recovery bias is as extreme as here.
 
-On this backbone, whose ties are no longer weighted, we can perform a number of network calculations that were meaningless in the raw network due to its large number of connections.
-We will begin by asking which nodes are the most influential in the network, i.e.
-possess the highest potential for spreading information.
-This, calculated through eigenvalue centrality, gives us a first idea of how pottery decor knowledge might have spread.
+### Influential nodes
+
+On this backbone, whose ties are no longer weighted, we can perform network calculations.
+We will begin by asking which nodes are the most influential in the network, i.e. possess the highest potential for spreading information.
+This is calculated through eigenvalue centrality, which is based on the number of nodes a node is linked to and how well those nodes are linked themselves.
+Eigenvalue centrality gives us a first idea of key sites in the spread of pottery decor knowledge.
 
 ```{r backbone-eigenvalue, fig.cap="Binary backbone at alpha = 0.15 showing eigenvector centralities"}
-decor.15 <- decor.15 %>% activate(nodes) %>% filter(!node_is_isolated())
-
-#decor.15 %>%
-#  activate(nodes) %>%
-#  mutate(eigen = centrality_eigen()) %>%
-#  select(label, eigen) %>%
-#  arrange(desc(eigen)) %>%
-#  kable()
-
-ggraph(decor.15, layout = "nicely") +
+ggraph(decor_18, layout = "nicely") +
   geom_node_point(aes(size = centrality_eigen())) +
-  geom_edge_fan(color = "red", alpha = 0.8) +
+  geom_edge_fan(color = "grey", alpha = 0.8) +
   scale_edge_width(range = c(0.2, 3)) +
   geom_node_text(aes(label = label), repel = TRUE) +
-  theme_graph()
+  theme_graph() +
+  labs(size = "eigenvector centrality")
 ```
@@
 The result (\@ref(fig:backbone-eigenvalue)) shows that the highest influence on the network is exerted by the sites in the Eastern IND and Lakes cluster.
-We should not forget that the network is partial, i.e.
-many connections to many sites are missing, and that the central position of the IND has to do with our sample selection and research history.
-Nevertheless, it suggests this region specifically as a centre through which information and thus also innovation in pottery decoration was disseminated in this period.
-The other network clusters are more homogenous, at this time probably less diverse and more stable in their decors.
+This reinforces the pattern already seen in clustering and shows the importance of the urban sites of the IND to Middle Niger culture in this period.
+We should not forget that the network is partial, i.e. many connections to many sites are missing.
+While the IND and the Bandiagara plateau and the Seno plain are both well researched and the networks therefore comparable, sites like Gao-Saney might have been influential in an undocumented local network.
+ Nevertheless, it suggests this region specifically as a center through which information and innovation in pottery decoration were disseminated in this period.
+
+### Brokers
 
@@
 ```{r backbone-betweenness, fig.cap="Binary backbone at alpha = 0.15 showing betweenness centralities"}
-#decor.15 %>%
-#  activate(nodes) %>%
-#  mutate(eigen = centrality_betweenness()) %>%
-#  select(label, eigen) %>%
-#  arrange(desc(eigen)) %>%
-#  kable()
-
-ggraph(decor.15, layout = "nicely") +
+ggraph(decor_18, layout = "nicely") +
   geom_node_point(aes(size = centrality_betweenness())) +
-  geom_edge_fan(color = "red", alpha = 0.8) +
+  geom_edge_fan(color = "grey", alpha = 0.8) +
   scale_edge_width(range = c(0.2, 3)) +
   geom_node_text(aes(label = label), repel = TRUE) +
-  theme_graph()
+  theme_graph() +
+  labs(size = "betweenness centrality")
 ```
 
-We can see (\@ref(fig:backbone-betweenness)) that Hambarketolo appears as the main gatekeeper to the Jenné site cluster, and to the eastern IND network clique, while Tongo Maaré Diabal, and to a lesser extent also Gao-Saney and Togu Missiri, stand in a brokerage role between the IND and the greater Niger cluster.
-The Dogon cluster exists at two removes from the IND, passing through Dammassogou and Ambéré-Dougon.At alpha = 0.2, this cluster is also connected to the large niger cluster via the Sangha "tellem" cave sites.
+Particularly interesting among the results (\@ref(fig:backbone-betweenness) is the position of the two sites at Dia, Shoma and Mara (Bedaux et al.
+2005), as providing a connection between the wider Niger cluster and the Eastern IND.
+It is notable that Shoma clusters more with the former, whereas Mara is more closely linked to the latter, via the site of Toguéré Galia (Bedaux et al.
+1978).
+The Eastern IND sites that had such high eigenvector centrality have particularly low betweenness centrality.
+Also of note is the brokerage position of the Sanga cave sites (Tellem) in relation to the Bandiagara and Seno cluster.
+As the assemblage here notionally stems from a funerary context, their high betweenness centrality might indicate that ceramics using unusual or imported decorations were preferentially deposited here, while the ceramics from contemporary habitation sites tended to have a more restricted repertoire.
 
-Also interesting is the edge betweenness value.
-This measures how important the edge between two nodes is for traversing the network.
-In our case it might indicate areas that were important axes of knowledge exchange.
-```{r backbone-edge-betweenness, fig.cap="Binary backbone at alpha = 0.15 showing edge betweenness centralities"}
-ggraph(decor.15, layout = "stress") +
-  geom_node_point() +
-  geom_edge_fan(color = "red", alpha = 0.8, aes(width = centrality_edge_betweenness())) +
-  scale_edge_width(range = c(0.2, 3)) +
-  geom_node_text(aes(label = label), repel = TRUE) +
-  theme_graph()
+### Evaluating measurements
 
-ggraph(decor.15, x = lon, y= lat) +
-  geom_sf(data = mali_adm0, fill = NA) +
-  geom_sf(data = rivers, colour = "black") +
-  geom_node_point() +
-  geom_edge_fan(color = "red", alpha = 0.8, aes(width = centrality_edge_betweenness())) +
-  scale_edge_width(range = c(0.2, 3)) +
-  geom_node_text(aes(label = label), repel = TRUE) +
-  coord_sf(xlim = c(-7, 2), ylim = c(13, 19), expand = FALSE) +
-  theme_graph()
-```
-
-The edge betweenness measure (\@ref(fig:backbone-edge-betweenness)) shows a central path through the network, connecting the sites of Togu Missiri, Tongo Maaré Diabal, Hambarketolo, Damassogou, Ambéré-Dougon, and Sadia, as well as a prominent side route connecting Gao and Kawinza.These might roughly represent two routes of knowledge exchange: one important one passing through the eastern side of the Inland Niger Delta and below the Niger Bend, in which the Jenné site cluster plays a pivotal role as the meeting place between the three networks of knowledge exchange, and another, though less important to the network we have sampled, along the western side of the IND and north of the Niger Bend to Gao.
-
-## Check for robustness of centrality measures
-
-In order to know whether these centrality measures are valid, rather than random effects of the backbone sparsification, we implement a variation on Peeples' suggested method for estimating the robustness of centrality measures in archaeological networks.
+In order to know whether these centrality measures are valid, rather than random effects of the backbone sparsification, we implement a variation on Peeples' (2017) suggested method for estimating the robustness of centrality measures in archaeological networks.
 In order to do so, the edges of the original graph are sampled at 10 % intervals, from 10 to 90 %.
@@
 A Spearman's rank correlation test is run between the samples to see how the deletion of edges influences the centrality within the backbone.
-Figure \@ref(fig:validity-test) shows how there is a relatively significant correlation between the betweenness ranks of nodes in the original graph and the sampled graphs, all the way down to the 30% sample.
-Only on the 20% sample does the correlation statistic drop lower.
-We can take this to mean that, for our case, the betweenness centrality shown in the MST is dependable.
+This is repeated 100 times.
+\@ref(fig:validity-test) shows that the correlation between the eigenvector and betweenness ranks of nodes in the original graph and the mean of the sampled graphs decreases quite rapidly.
+This shows us that our network is rather fragile, and the inferences we have made on the basis of the extracted backbone should be treated with the appropriate caution.
 
-```{r validity-test, warning = FALSE, message = FALSE}
+```{r validity-test, warning = FALSE, message = FALSE, fig.cap= "Testing the robustness of the bb-network by comparing its centralities to the original graph"}
+#comparing
 p1 <- test_centralities(decor_graph, "eigen", 100)
@@
 
-(p1 | p2) + plot_annotation(tag_levels = "a")
+(p1 | p2) + plot_annotation(tag_levels = "a") +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1))
 ```
 
-```{r}
-#make boxplots instead
-test_centralities2 <- function(graph, centrality, reps) {
+## Diachronic analysis
 
-  # check for correct arguments
-  if (centrality %in% c("eigen", "betweenness") == F) {
-    stop("centrality must be 'eigen' or 'betweenness'")
-  }
-
-  # get sample names
-  s <- seq(10, 90, 10)
-  samples <- paste("s", s, sep = "")
-
-  # replicate the test
-  replicate(reps, spearman_samples(graph, centrality)) %>%
-    as_tibble() %>%
-    mutate(sample = samples) %>%
-    pivot_longer(!sample, names_to = "iteration", values_to = "value") %>%
-    select(-iteration) %>%
-    #group_by(sample) %>%
-    #summarise(mean_rho = mean(value)) %>%
-    # plot
-    ggplot() +
-    geom_boxplot(aes(x = sample)) +
-    labs(y = expression(paste("mean Spearman's ", rho))) +
-    ylim(0, 1) +
-    theme_publish()
-}
-
-test_centralities2(decor_graph, "eigen", 100)
-```
-
-eigenvector centrality is fine, betweenness less impressive. 
-For edge betweenness, the validation is much harder, as the flattening procedure means that edges cannot reliably be numbered, and thus correlation is hard to achieve.
-We can, however, manually inspect the result (\@ref(fig:edge-betweenness-validity)).
-
-```{r edge-betweenness-validity, fig.cap="edge bewteenness validity",warning = FALSE, message=FALSE}
-bb.edge.betw <- function(x) {
-
-  # sample graph
-  g <- decor_graph %>%
-    activate(edges) %>%
-    tidygraph::sample_n(size = x) %>%
-    flatten(weighted = TRUE) %>%
-    as.igraph()
-
-  # extract backbone
-  bb <- backbone::disparity(g, alpha = 0.15, narrative = FALSE) %>%
-    tidygraph::as_tbl_graph()
-
-  # put node attributes back
-  decor.nodes <- decor_graph %>%
-    activate(nodes) %>%
-    as_tibble()
-  bb <- bb %>%
-    activate(nodes) %>%
-    mutate(label = decor.nodes$label, lat = decor.nodes$lat, lon = decor.nodes$lon)
-
-  # plot
-  ggraph(bb, layout = "stress") +
-    geom_node_point() +
-    geom_edge_fan(color = "red", alpha = 0.8, aes(width = centrality_edge_betweenness())) +
-    scale_edge_width(range = c(0.2, 3)) +
-    geom_node_text(data = . %>% filter(centrality_betweenness(weights = weight) > mean(centrality_betweenness(weights = weight))), aes(label = label), repel = TRUE) +
-    theme_graph() +
-    theme(legend.position = "none")
-}
-
-p1 <- bb.edge.betw(90) + ggtitle("90%")
-p2 <- bb.edge.betw(80) + ggtitle("80%")
-p3 <- bb.edge.betw(70) + ggtitle("70%")
-p4 <- bb.edge.betw(60) + ggtitle("60%")
-
-p1 + p2 + p3 + p4
-```
-What is clear is that the backbone is, predictably, quite sensitive to random edge deletion, and therefore edge_betweenness is very much dependent on whether an edge exists or not.
-For the purposes of generating hypotheses for archaeological testing, we have still found this method superior to other methods of finding paths through networks, such as an inverted version of Kruskal's minimum spanning tree (for details and code, see Annex)
-
-The edge betweenness measure shows a central path through the network, connecting the sites of Togu Missiri, Tongo Maaré Diabal, Hambarketolo, Damassogou, Ambéré-Dougon, and Sadia, as well as a prominent side route connecting Gao and Kawinza.These might roughly represent two routes of knowledge exchange: one important one passing through the eastern side of the Inland Niger Delta and below the Niger Bend, in which the Jenné site cluster plays a pivotal role as the meeting place between the three networks of knowledge exchange, and another, though less important to the network we have sampled, along the western side of the IND and north of the Niger Bend to Gao.
-
-# Diachronic analysis
-
-## NB clean this part and update the site list.
+The network of interactions that we have been working with up to this point is built on data taken from archaeological contexts whose date range includes the 12^th^ century AD.
+This means that, in some cases, the period under consideration is actually much longer.
+Change over time within this period is to be expected, but is hard to distinguish given the nature of the data.
+Within wider brackets, the issue of change can be approached from a network perspective.
+While a more systematic diachronic view will be the subject of future publications, we will here consider an early 9^th^ century AD sample in order to show how the temporal development of the network can be approached.
 
@@
 library(readr)
-decorsearly <- read_csv2(here::here("analysis/data/raw_data/decors_early.csv"))
+decorsearly <- 
+  read_csv2(here::here("analysis/data/raw_data/decors_early.csv")) |>
+  select(-1)
 
@@
 # construct network, remove extraneous sites
-decor_graph_early <- tbl_graph(sites, edgelist, directed = FALSE) %>%
-  activate(nodes) %>%
+decor_graph_early <- tbl_graph(sites, edgelist, directed = FALSE) |>
+  activate(nodes) |>
   filter(!node_is_isolated())
-
-decor_graph_early
 ```
-
-## Backbone extraction
 
@@
 # extract backbone
-early.flat.bb <- backbone::disparity(as.igraph(early_flat), alpha = 0.31, narrative = TRUE) %>%
+early_flat_bb <- 
+  backbone::disparity(as.igraph(early_flat), alpha = 0.25, narrative = TRUE) |>
   tidygraph::as_tbl_graph()
@@
 # igraph transformation loses the node attributes, so we put them back
-early.flat.nodes <- early_flat %>%
-  activate(nodes) %>%
+early_flat_nodes <- early_flat |>
+  activate(nodes) |>
   as_tibble()
-early.flat.bb <- early.flat.bb %>%
-  activate(nodes) %>%
-  mutate(label = early.flat.nodes$label, lat = early.flat.nodes$lat, lon = early.flat.nodes$lon)
+early_flat_bb <- early_flat_bb |>
+  activate(nodes) |>
+  mutate(label = early_flat_nodes$label,
+         lat = early_flat_nodes$lat, lon = early_flat_nodes$lon)
 
-ggraph(early.flat.bb, layout = "stress") +
+#plot
+ggraph(early_flat_bb, layout = "stress") +
   geom_node_point() +
-  geom_edge_fan(color = "red", alpha = 0.8) +
+  geom_edge_fan(color = "grey", alpha = 0.8) +
   scale_edge_width(range = c(0.2, 3)) +
@@
 
-This backbone (\@ref(fig:early-backbone))looks significantly different from the later one, though we still distinguish an IND group, a very clear Dogon group, and one that includes "others": Kawinza, TMD, Tiebala, Akumbu.
-Essouk is isolated.
-Clustering similarly revelas these two communities.
+The analytical steps run on the 12^th^ century network were carried out identically on the earlier assemblages.
+The main differences are best visualized on the extracted backbone (\@ref(fig:early-backbone).
+Here, only two network cliques remain.
+The first takes sites within the IND and on its margins, including TMD and Akumbu.
+The other group consists of sites on the Bandiagara plateau and in the Seno plain.
+The latter group is more strongly internally connected than it is in the 12^th^ century network, while the IND group is less closely connected.
+Essouk is isolated in the backbone until alpha is increased to 0.37, indicating that connections to Essouk are not very reliable.
 
-A more in-depth treatment of the temporal development of the pottery networks in this area will be the subject of a future publication.
-For now, we might draw the following preliminary conclusions: The comparison of the two networks appears to show a process of increasing fragmentation between the two periods reflected by the graphs, but also the growth of netwroks to incorporate more diverse influences.
-# Results
+A more in-depth treatment of the temporal development of the pottery networks in this area will be the subject of future publications.
+For now, we might draw the following preliminary conclusions: The comparison of the two networks appears to show several processes.
+One, observable on the Bandiagara-Seno cluster, is a network that is becoming less tightly connected, in other words it opens up to more outside influences between the two periods.
+This, in a sense, is reflective of Gallay et al.'s vision of increasingly mixed traditions, discussed above.
+The second process is observable in the IND cluster, which is less strongly connected in the 9^th^ century network and becomes more coherent and also internally diversified by the 12^th^ century.
+This is more reflective of a dynamic of an area in the process of cultural genesis, rather than fragmentation.
+Thirdly, the long-distance connections emerging from the 12^th^ century networks are very weak, so as not to be statistically detectable, in the 9^th^ century.
+While there are issues of unequal samples between the two periods, we can take this to suggest that the impact of long-distance exchange along the Middle Niger on pottery decoration was not yet very strong in the earlier period.
 
@@
 
-Potter communities and the propagation of knowledge within them, has been the subject of several very detailed ethnoarchaeological and archaeological studies in this region.
-This study has left these out, quite purposefully.
-Their boundaries might be seen in other elements of the pottery making sequence, notably in the forming technique, and they do not necessarily overlap with decors.
-Our network model instead is intended to represent communication of knowledge about pottery that transcends smaller communities of practice.
-As such, it shows a first picture of how decorative techniques might have developed and what mechanisms are at the basis of the patterns in archaeological ceramics that are seen on sites along the Middle Niger.
+Potter communities and the propagation of knowledge within them, have been the subject of several very detailed ethnoarchaeological and archaeological studies in this region and within Africa more widely.
+Their boundaries might be seen in other elements of the pottery making sequence, notably in the forming technique, but decorative practices are recognized not to be straightforwardly related to social identities (Gosselain 2011, p.
+11).
+Our network model is intended to represent communication of knowledge about pottery that transcends smaller communities of practice.
+As such, it permits us to explore what mechanisms of exchange may have been at the basis of the patterns in archaeological ceramic decors on sites along the Middle Niger.
 
-The results of our analyses suggest that the propagation of pottery decor motifs occurs within networks of differing magnitudes.
-Some of these networks are local, geographically and possibly also culturally bounded.
+The results of our analyses suggest that the propagation of pottery decor motifs occurred within networks of differing geographic extent.
+Some of these networks were local, geographically, and possibly also culturally, bounded.
 This can, for instance, be seen in the group of sites on the Bandiagara massif and in the Seno plain, or in the eastern part of the IND.
-Cross-cutting such local developments, however, are long-distance contacts along the Niger River, which led to similarities between very distant sites.
-This underlines a key aspect of past cultural dynamics in the wider region: there is both a high level of cultural diversity, of human groups and languages, and overarching regional connectivity.
-These two processes, the local and the global, have been in exchange with one another shaping the Middle Niger cultural landscape.
+Cross-cutting such local exchange, however, were long-distance contacts along the Niger River, which led to similarities between very distant sites.
+This underlines a key aspect of past cultural dynamics in the wider region: there was, and continues to be, both a high level of cultural diversity, of human groups and languages, and overarching regional connectivity.
+These two processes, the local and the global, interacted to shape the Middle Niger cultural landscape.
 Every site assemblage can therefore be seen to be reflective of how strongly innovation and tradition influenced local potting communities.
-Changes in the diversity of decors in an assemblage, noted for instance at XXX, might therefore reflect changes to the degree in which the population supplying pottery to a site was exposed and open to new influences.
 
-It is of key importance in this approach to note where this interaction between network groups takes place.
-While the archaeological record is necessarily only a sample and many nodes are missing, we can nevertheless gain a preliminary insight into the channels along which information on pottery decoration was shared, identify those sites within the record which occupy key positions in this exchange, and to suggest key routes.
-The latter, built on edge betweenness, is significant in that it might suggest those areas (for instance between Kawinza and Gao) which are not represented by nodes in the model, but were nevertheless important to the exchange network.
+Network analysis allows us to explore the roles of nodes within the network, and thus to show note where this interaction between network groups takes place.
+While the archaeological record is necessarily only a sample and many nodes are missing, we can nevertheless gain a preliminary insight into the channels along which information on pottery decoration was shared and identify those sites within the record which occupy key positions in this exchange.
+In the case of the 12^th^ century Middle Niger, our network model suggests that the clustered urban sites of the eastern IND exerted a large influence on the sharing of pottery decoration techniques, even though other sites, such as Dia, were more important to their spread.
 
 The diachronic comparison of two networks from different periods begins to show us processes of innovation, and how these change networks.
-This must be seen as a work in progress, and a fuller picture will be the subject of future publications.Nevertheless, we can detect how the sites within the IND became a major centre for diversification and diffusion of decorative techniques, while the close-knit and formerly more isolated "Bandiagara cluster" opened up to outside influences.
-Overall, the later period is marked by more ties, an intensifying network.
-This is likely due to an increased frequency of contact between heterogenous potter communities, perhaps as a result of a combination of increases in population size, the establishment of new markets, and an increased flow of goods contained in pottery.
-Mayor (REF) has also suggested political factors behind the spread of some decors, though we are not entirely sure of the causality behind this hypothesis.
-This pattern is less likely to stem from an overall higher mobility of individual potters, since the overlapping structure of the individual technique networks suggests gradual change.
+This must be seen as a work in progress, and a fuller picture will be the subject of future publications.
+Nevertheless, we can detect how the sites within the IND became a major centre for diversification and diffusion of decorative techniques, while the close-knit and formerly more isolated "Bandiagara cluster" opened up to outside influences.
+Overall, the later period is marked by more ties, and therefore an intensifying network.
+This is likely due to an increased frequency of contact between heterogenous potter communities, perhaps as a result of a combination of increases in population size, the establishment of new markets, and an increased flow of goods, some of which will have been contained in pottery.
+Mayor (2011b, p.
+208) has also suggested political factors behind the spread of some decors.
+The difference between the networks is less likely to stem from an overall higher [long-range]{.underline} mobility of individual potters, since the overlapping structure of the individual technique networks suggests gradual change.
 
-Chronological and spatial approaches to pottery decor in the study region have been largely frustrating, with changes only in percentages of decors which rarely provided the basis for clear phase changes or cultural boundaries.
-This approach emphasises that this picture of gradual differences points to a strongly connected society, in which pottery producers were routinely exposed to their neighbours' productions, and in which the decorative techniques were mostly able to propagate without impediment.
+Chronological and spatial approaches to pottery decor in the study region have not often yielded satisfying results, with changes only in percentages of decors which rarely provided the basis for clear phase changes or cultural boundaries.
+Our approach here emphasises that this picture of gradual differences points to a strongly connected society, in which pottery producers were routinely exposed to their neighbours' productions, and in which decorative techniques were mostly able to propagate without impediment.
 Nevertheless, within this picture, local traditions can be discerned as cliques within the networks, whose degree of isolation varies, but which are never without outside contact.
+
+We should note here that the network has artificial boundaries which are likely to impact on how well the model reflects actual social contacts in the period in question.
+Our sample is geographically centered on the IND, where we have one of the network cliques and the most abundant data.
+Especially towards the geographical edges of the sample, the model is probably much less reliable.
+Thus, we know that the Bandiagara -- Seno cluster is only a small part of what would likely be a much larger and internally differentiated cluster below the Niger Bend.
+If this were fully included in the model, it is certain that it would provide new and different linkages to the northeastern sites of our model, Gao and Essouk.
+The latter two sites also have extensive cultural links further north and east, which cannot be captured in this model.
+The same is true of sites in the south and west (Togu Missiri, Akumbu, Kawinza), whose main local networks are probably simply not reflected in this sample.
+All of this has led to network clusters of different kinds, as discussed above: the geographically large Niger river cluster is very likely reflective of very different forms of contact than the smaller, more tightly knit IND cluster is.
+We hypothesise that such smaller clusters could probably be found anywhere in the area if there were enough suitable data, and that the historical spread of pottery decors would have been influenced by both closer local and looser regional networks.
+In Gosselain's (2011) sense, the flow of information in the larger regional network would most likely have been impeded by geographic distance, but also by the force of the closer local networks.
 
@@
 Modelling pottery decorative techniques along the Middle Niger as a knowledge exchange network presents a change of perspective on the subject.
-Previous work has focused on distribution areas and relative prevalence at different sites.
+Previous work, including the author's own, has focused on distribution areas and relative prevalence at different sites.
 We have taken the same information to indicate the degree to which pottery producers supplying the various sites were part of contact networks in which knowledge of decors could spread.
@@
 The latter form is the most likely basis for the way in which these local networks corresponded to a larger one: all along the Middle Niger, from our southernmost sample in the Ségou area to Gao and the Saharan entrepot of Essouk, pottery and the knowledge of certain decorative techniques appears to have travelled.
-Our model suggests that two routes are likely to have been important in this long-distance contact: one along the western IND, the Lakes Region and north of the Niger Bend, and another along the IND's eastern edge and below the Niger Bend.
 
-These results feed into long-standing discussions on the Middle Niger, and West Africa in general, as a highly connected and interdependent cultural space in the late first and early second millennium AD (for a recent discussion, see Gestrich et al.
-2021).
+The study of archaeological material as networks shifts the focus "from spatial relations to social relations"(Knappett 2011).
+Herein lies the main difference of the approach taken in this paper to previous synthetic studies of archaeological ceramics on the Middle Niger.
+Mayor's (2011b) study, in particular, has admirably shown how the ceramics of the wider region can reflect a number of cultural groups and historical influences.
+These preliminary results of our network study, on the other hand, begin to show how the area's pottery reflects webs of exchange, both at smaller local levels and at a larger regional level.
+It makes visible and understandable the way in which the Middle Niger was both an integrated cultural sphere, albeit one consisting of significant differences.
+These two approaches do not invalidate each other but should be seen as complementary.
+The results feed into long-standing discussions on the Middle Niger, and West Africa in general, as a highly connected and interdependent cultural space in the first and early second millennium AD (for recent discussions on this topic, see Gestrich et al.
+2021; Dueppen and Gallagher 2023).
 
 # Acknowledgements
+
+The work presented here was undertaken as part of the project "Borrowed words and shared objects" within the German Research Foundation's (DFG) Priority Programme "Entangled Africa".
+The authors wish to thank the Institut des Sciences Humaines du Mali, the Musée National du Mali, and the Rijksmuseum voor Volkenkunde in Leiden for access to their pottery collections, and Sam Nixon and Douglas Post Park for providing primary data.
+The R compendium for this article was created using the rrtools package.
 
@@
 
-<div id="refs"></div>
+Barker CS, Young LC (2017) Networks of Ceramic Exchange: Comparing Homol’ovi Pueblo III Pithouse and Pueblo Communities. KIVA
+
+Bedaux R (1980) The geographic distribution of footed bowls in the upper and middle Niger region.
+In: Swartz, B., Dumett, R.
+(eds), West African Culture Dynamics: Archaeological and Historical Perspectives.
+pp 247–258
+
+Bedaux R, Lange A (1983) Tellem, reconnaissance archéologique d’une culture de l’Ouest africain au Moyen-Age: la poterie. Journal des Africanistes 53:5–59
+
+Bedaux R, Polet J, Sanogo K (2005) Recherches archéologiques à Dia dans le delta intérieur du Niger (Mali): Bilan des saisons de fouilles 1998-2003. CNWS Publications, Leiden
+
+Bedaux RMA, Constandse-Westermann TS, Haquebord L, et al (1978) Recherches archeologiques dans le Delta Interieur du Niger (Mali). Palaeohistoria Bussum 20:91–220
+
+Blondel VD, Guillaume J-L, Lambiotte R, Lefebvre E (2008) Fast unfolding of communities in large networks. J Stat Mech 2008:P10008. https://doi.org/10.1088/1742-5468/2008/10/P10008
+
+Brughmans T, Poblome J (2016) Pots in space: understanding Roman pottery distribution from confronting exploratory and geographical network analyses.
+In: Barker E, Bouzarovski S, Pelling C, Isaksen L (eds) New worlds from old texts: revisiting ancient space and place.
+Oxford University Press, Oxford, pp 255–280
+
+Crabtree SA (2015) Inferring Ancestral Pueblo Social Networks from Simulation in the Central Mesa Verde. J Archaeol Method Theory 22:144–181. https://doi.org/10.1007/s10816-014-9233-8
+
+Curdy P (1982) Tiébala (Mali): un complexe céramique du 6e siècle après J.-C. Archives suisses d’anthropologie générale 46:183–198
+
+David N, Sterner J, Gavua K (1988) Why pots are decorated. Current Anthropology 29:365–389
+
+de Groot BG (2019) A Diachronic Study of Networks of Ceramic Assemblage Similarity in Neolithic Western Anatolia, the Aegean and the Balkans (c.6600–5500 bc).
+Archaeometry 61:600–613.
+https://doi.org/10.1111/arcm.12450
+
+Dietler M, Herbich I (1994) Ceramics and thnic identity : ethnoarchaeological observations on the distibution of pottery styles and the relationship between the social contexts of production and consumption.
+In: Terre cuite et société : céramique, document technique, économique, culturel.
+Rencontres internationales d’archéologie et d’histoire (14, 21- 23 oct.
+1993, Antibes).
+Eds APDCA (Assoc.
+pour la promotion et la diffusion des connaissances archaéologiques).
+Juan-les-Pins, pp 459–472
+
+Domagalski R, Neal ZP, Sagan B (2021) Backbone: An R package for extracting the backbone of bipartite projections. PLoS ONE 16:e0244363. https://doi.org/10.1371/journal.pone.0244363
+
+Dueppen SA, Gallagher D (2023) Networked Farmers, Ancestral Rituals, Regional Marketplaces, and Salt: New Insights into the Complexity of First Millennium BC/AD Farming Societies in West Africa.
+Afr Archaeol Rev.
+https://doi.org/10.1007/s10437-022-09509-2
+
+Gallay A, Huysecom E, Mayor A (1998) Peuples et céramiques du Delta intérieur du Niger (Mali): un bilan de cinq années de missions (1988-1993). Mainz
+
+Gallin A (2011) Les styles céramiques de Kobadi. Analyse comparative et implications chronoculturelles au Néolithique du Sahel malien. Frankfurt
+
+Gestrich N, Champion L, Keita D, Coulibaly N (2021) Evidence of an 11th century CE cola nitida trade into the Middle Niger region. African Archaeological Review 38:403–418
+
+Gestrich N, MacDonald KC (2018) On the Margins of Ghana and Kawkaw: Four Seasons of Excavation at Tongo Maaré Diabal (AD 500-1150), Mali. Journal of African Archaeology 16:1–30
+
+Gosselain O (2000) Materializing identities : an African perspective. Journal of Archaeological Method and Theory 7:187–217
+
+Gosselain OP (1998) Social and Technical Identity in a Clay Crystal Ball. In: Stark MT (ed) The Archaeology of Social Boundaries
+
+Gosselain OP (2011) Pourquoi le décorer? Quelques observations sur le décor céramique  en Afrique. Azania: Archaeological Research in Africa 46:3–19
+
+Granovetter MS (1973) The Strength of Weak Ties. American Journal of Sociology 78:1360–80
+
+Haour A, Keita D (2010) Folded Strip Roulette/ Roulette de fibre plate pliée.
+In: Haour A, Manning K, Arazi N, et al.
+(eds) African Pottery Roulettes Past and Present.
+Techniques, Identification  and Distribution.
+Oxbow, Oxford, pp 169–176
+
+Haour A, Manning K, Arazi N, et al (eds) (2010) African pottery roulettes past and present: techniques, identification and distribution. Oxbow Books, Oxford
+
+Hart JP (2012) The effects of geographical distances on pottery assemblage similarities: a case study from Northern Iroquoia.
+Journal of Archaeological Science 39:128–134.
+https://doi.org/10.1016/j.jas.2011.09.010
+
+Hart JP, Engelbrecht W (2012) Northern Iroquoian Ethnic Evolution: A Social Network Analysis. J Archaeol Method Theory 19:322–349. https://doi.org/10.1007/s10816-011-9116-1
+
+Herbich I (1987) Learning patterns, potter interaction and ceramic style among the  Luo of Kenya. African Archaeological Review 5:193
+
+Hodder I (1982) Symbols in Action: Ethnoarchaeological Studies of Material Culture. Cambridge University Press, Cambridge
+
+Iacono F (2016) From Networks to Society: Pottery Style and Hegemony in Bronze Age Southern Italy. Cambridge Archaeological Journal 26:121–140. https://doi.org/10.1017/S0959774315000190
+
+Insoll T (1996) Islam, Archaeology and History. Gao Region (Mali) ca. AD 900-1250. Tempus Reparatum, Cambridge Monographs in African Archaeology 39. BAR S647. Oxford
+
+Knappett C (2011) An archaeology of interaction: network perspectives on material culture and society. Oxford University Press, Oxford
+
+Knappett C (2021) Networks in Archaeology. The Oxford Handbook of Social Networks 443–466. https://doi.org/10.1093/oxfordhb/9780190251765.013.26
+
+Livingston Smith A (2007) Histoire du décor à la roulette en Afrique subsaharienne. Journal of African Archaeology 5:189–216
+
+Magnani M, Rossi L, Vega D (2021) Analysis of Multiplex Social Networks with R. Journal of Statistical Software 98:1–30
+
+Mayor A (2011a) Impressions de vanneries et technique du martelage sur forme concave:  Anthropologie et histoire d’une technique dans la boucle du Niger. Azania 46:1:88–109
+
+Mayor A (2011b) Traditions céramiques dans la boucle du Niger. Africa Magna, Cologne
+
+Mazzucato C (2019) Socio-Material Archaeological Networks at Çatalhöyük a Community Detection Approach. Frontiers in Digital Humanities 6:. https://doi.org/10.3389/fdigh.2019.00008
+
+McIntosh RJ (1998) The peoples of the Middle Niger. The island of gold. Blackwell Publishers, Malden, Mass.; London
+
+McIntosh SK (1995) Excavations at Jenné-Jeno, Hambarketolo, and Kaniana, Inland Niger delta, Mali, the 1981 season. University of California press, Berkeley (Calif.) Los Angeles (Calif.) London
+
+McIntosh SK (ed) (1999) Beyond Chiefdoms. Pathways to Complexity in Africa. Cambridge
+
+McIntosh SK, Gueye NS (2010) Braided cord roulette (simple and composite).
+In: Haour A, Manning K, Arazi N, et al.
+(eds) African Pottery Roulettes Past and Present.
+Techniques, Identification  and Distribution.
+Oxbow, Oxford
+
+McIntosh SK, McIntosh RJ (1980) Prehistoric Investigations in the Region of Jenne, Mali. BAR, Oxford
+
+Mills BJ, Peeples MA, Haas, Jr. WR, et al (2015) Multiscalar Perspectives on Social Networks in the Late Prehispanic Southwest. American Antiquity 80:3–24. https://doi.org/10.7183/0002-7316.79.4.3
+
+Mills BJ, Roberts JM, Clark JJ, et al (2013) The dynamics of social networks in the Late Prehispanic U.S.
+Southwest.
+In: Knappett C (ed) Network analysis in archaeology.
+New approaches to regional interaction.
+Oxford University Press, Oxford, pp 181–202
+
+Nixon S (2017) Essouk-Tadmekka: An Early Islamic Trans-Saharan Market Town. Brill
+
+Park M (1799) Travels in the Interior of Africa. W. Bulmer & Co., London
+
+Pedersen T (2022) tidygraph: A Tidy API for Graph Manipulation. https://tidygraph.data-imaginist.com, https://github.com/thomasp85/tidygraph
+
+Peeples MA (2017) Network Science and Statistical Techniques for Dealing with Uncertainties in Archaeological Datasets. http://www.mattpeeples.net/netstats.html. Accessed 18 Oct 2022
+
+Peeples MA, Mills BJ, Haas WRJr, et al (2016) Analytical Challenges for the Application of Social Network Analysis in Archaeology.
+In: Brughmans T, Collar A, Coward F (eds) The Connected Past: challenges to network studies in archaeology and history.
+Oxford University Press, Oxford, pp 59–84
+
+R Core Team (2022) R: A Language and Environment for Statistical Computing. R Foundation for Statistical Computing, Vienna, Austria
+
+Radivojević M, Grujić J (2018) Community structure of copper supply networks in the prehistoric Balkans: An independent evaluation of the archaeological record from the 7th to the 4th millennium BC.
+Journal of Complex Networks 6:106–124.
+https://doi.org/10.1093/comnet/cnx013
+
+Raimbault M, Sanogo K (eds) (1991) Recherches Archéologiques au Mali: Prospections et inventaire,  fouilles et études analytiques en Zone lacustre. Karthala, Paris
+
+Roux V, Bril B, Karasik A (2018) Weak Ties and Expertise: Crossing Technological Boundaries. J Archaeol Method Theory 25:1024–1050. https://doi.org/10.1007/s10816-018-9397-8
+
+Saha T, Rangwala H, Domeniconi C (2013) Sparsification and Sampling of Networks for Collective Classification.
+In: Greenberg AM, Kennedy WG, Bos ND (eds) Social Computing, Behavioral-Cultural Modeling and Prediction.
+Springer, Berlin, Heidelberg, pp 293–302
+
+Schmidt A, Arazi N, MacDonald KC, et al (2005) La poterie.
+In: Bedaux RMA, Polet J, Sanogo K, Schmidt A (eds) Recherches archéologiques a Dia dans le Delta intérieur du Niger  (Mali): bilan des saisons de fouilles 1998-2003.
+CNWS, Leiden, pp 191–256
+
+Serrano MÁ, Boguñá M, Vespignani A (2009) Extracting the multiscale backbone of complex weighted networks.
+Proceedings of the National Academy of Sciences 106:6483–6488.
+https://doi.org/10.1073/pnas.0808904106
+
+Sindbæk SM (2013) Broken links and black boxes: material affiliations and contextual network synthesis in the Viking world.
+Network Analysis in Archaeology: New Approaches to Regional Interaction 71–94
+
+Sterner J, David N (2003) Action on matter: the unique African tamper and concave anvil pot-forming  technique. Journal of African Archaeology 1:3–38
+
+Sternini M (2019) The production centres and river network of Italian terra sigillata between the Arno and Tiber valleys: a geographical point of view.
+Journal of Roman Archaeology 32:485–494.
+https://doi.org/10.1017/S1047759419000254
+
+Tamari T (1991) The development of case systems in West Africa. Journal of African History 32:221–250
+
+Tamari T (1995) Linguistic Evidence for the History of West African "Castes. In: Conrad DC, Frank BE (eds) Status and Identity in West Africa. Indiana University Press, Bloomington, pp 61–85
+
+Upton AJ (2019) Multilayer Network Relationships and Culture Contact in Mississippian West-Central Illinois, A.D. 1200 - 1450. Ph.D., Michigan State University
+
+van Doosselaere B (2014) Le Roi et le Potier: Étude technologique de l’assemblage céramique de Koumbi Saleh, Mauritanie (5e/6e - 17e siècles AD). Africa Magna, Frankfurt am Main
+
+Vargas FE, Lanata JL, Abramson G, et al (2020) Digging the topology of rock art in northwestern Patagonia. Journal of Complex Networks 8:cnz033. https://doi.org/10.1093/comnet/cnz033
+
+Vavrasek JL (2020) Interactions Between Jefferson County Iroquoian Women and Men as Explored Through Social Network Analysis of Pottery and Pipe Decorative Motif Traits.
+Ph.D., State University of New York at Albany
+
+Yang Z, Algesheimer R, Tessone CJ (2016) A Comparative Analysis of Community Detection Algorithms on Artificial Networks. Sci Rep 6:30750. https://doi.org/10.1038/srep30750
 
@@
 # which R packages and versions?
-#if ("devtools" %in% installed.packages()) devtools::session_info()
+# if ("devtools" %in% installed.packages()) devtools::session_info()
 ```
@@
 # what commit is this file at?
-if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+if ("git2r" %in% installed.packages() &
+    git2r::in_repository(path = ".")) 
+  git2r::repository(here::here()) 
 ```

### 2023-08-03 analysis/R/edgelist functions.R

@@ -64,8 +64,9 @@ make_edgelist <- function(data){
   merge(sites, Type) %>%
     left_join(data_long, by = c("x" = "Var1", "name" = "name"))%>%
     left_join(data_long, by = c("y" = "Var1", "name" = "name")) %>%    #merge the data
-    mutate(weight = value.x * value.y) %>%                            #multiply the pairs
-    dplyr::select(-c(value.x, value.y)) %>%
+    rowwise() %>%
+    mutate(weight = min(value.x, value.y)) %>%                            #because of rowwise we now get the minimum perc as weight
+    select(-c(value.x, value.y)) %>%
     tidyr::drop_na() %>%
     filter(weight > 0)
 }

### 2023-08-03 analysis/paper/data analysis.Rmd
code style

@@
 # import decor data
-decors <- read_csv2(here::here("analysis/data/raw_data/decors_late.csv"))
+
+#decors late is the "cleaned" version without Oudalan and SAO. they are still present in the earlier version decors.csv
+decors <- read_csv(here::here("analysis/data/raw_data/decors.csv"))
+
+decors <- decors %>%
+  filter(site != "SAO 3" & site != "Oudalan fer recent")  %>%
+  mutate_at(vars(-site), as.numeric) %>% 
+  mutate(site = case_when(site == "HAMB 4" ~ "Hamb 4",
+         site == "KAN 4" ~ "Kaniana 4",
+         site == "Akumbu 1" ~ "Akumbu late",
+         site == "TMD" ~ "TMD 5",
+         TRUE ~ site)) %>% 
+  select(-slip)
+
+ decors[is.na(decors)] <- 0
+
+ #write.csv2(decors,here::here("analysis/data/raw_data/decors_late.csv"))
+
 # import information on sites
@@
   activate(edges) %>%
-  filter(name %in% c("braided_cord_roulette", "braided_strip_roulette", "paint", "straight_cord_mat")) %>%
+  filter(name %in% c("braided_cord_roulette", "braided_strip_roulette", "paint", "straight_cord_mat"))%>%
   ggraph() +
+  coloured_edges +
+  theme_graph()
+
+decor_graph %>%
+  activate(edges) %>%
+  filter(name %in% c("folded_strip_roulette", "braided_strip_roulette", "paint", "straight_cord_mat", "twisted_cord_roulette", "braided_cord_roulette"))%>%
+  filter(weight >= 3) |> 
+  activate(nodes) |> 
+  filter(!node_is_isolated()) |> 
+  ggraph() +
+  coloured_edges +
+  theme_graph()
+
+decor_graph %>%
+  activate(edges) %>%
+  filter(name %in% c("folded_strip_roulette","braided_strip_roulette", "paint", "straight_cord_mat", "braided_cord_roulette"))%>%
+  filter(weight >= 3) |> 
+  activate(nodes) |> 
+  filter(!node_is_isolated()) |> 
+  ggraph() +
+  coloured_edges +
+  theme_graph()
+
+decor_graph |> 
+    activate(edges) |> 
+  # Filters 'edges' based on 4 specific decorations
+  filter(name %in% c('folded_strip_roulette', 'braided_strip_roulette', 'paint', 'straight_cord_mat')) |> 
+#  filter(weight > 1) |> 
+  #plot the graph
+    ggraph(layout = 'stress') + 
     coloured_edges + 
@@
   head(1)
+
+decor_graph %>%  
+  get.data.frame(what = "vertices") %>% 
+  count(louvain)
+
+nodes <- decor_graph %>% 
+  get.data.frame("vertices")
 ```
@@
 
+The three communities can also be analysed as separate networks. The sites around the Niger Delta are mostly
+
+```{r network 1 centrality in community vs in gen network}
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_degree(weights = weight)) %>%
+  filter(louvain == 1) %>%
+  mutate(deg_c = centrality_degree(weights = weight)) %>%
+ggraph(layout = "graphopt") + 
+  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(aes(size = deg, alpha = deg_c), col = "cornflowerblue", show.legend = FALSE) +
+  geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+      scale_color_manual(values = communitycolor)+
+  theme_graph()+
+  guides(color = guide_legend(override.aes = list(size = 0.5)), alpha = FALSE, edge_alpha = FALSE, edge_width = guide_legend(override.aes = list(size = 0.3)))+
+  ggtitle("network cluster 1 (red)")+
+  theme(legend.title = element_text(size = 8, face = "bold"),
+              legend.text = element_text(size = 7),
+              legend.key.size = unit(0.8, "lines"),
+        plot.title = element_text(size = 10, face= "bold"))
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 1) %>%
+  mutate(deg_c = centrality_eigen(weights = weight)) %>%
+flatten(weighted = TRUE) %>% 
+ggraph(layout = "graphopt") + 
+  geom_edge_fan(aes(edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(aes(size = deg, alpha = deg_c), col = "cornflowerblue", show.legend = FALSE) +
+  geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+      scale_color_manual(values = communitycolor)+
+  theme_graph()+
+  guides(color = guide_legend(override.aes = list(size = 0.5)), alpha = FALSE, edge_alpha = FALSE, edge_width = guide_legend(override.aes = list(size = 0.3)))+
+  ggtitle("network cluster 1 (red)")+
+  theme(legend.title = element_text(size = 8, face = "bold"),
+              legend.text = element_text(size = 7),
+              legend.key.size = unit(0.8, "lines"),
+        plot.title = element_text(size = 10, face= "bold"))
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 1) %>%
+ggraph(layout = "graphopt") + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "cornflowerblue") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 1) %>%
+ggraph(x = lon, y = lat) + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "cornflowerblue") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+```
+
+```{r network 2 centrality in community vs in gen network}
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_degree(weights = weight)) %>%
+  filter(louvain == 2) %>%
+  mutate(deg_c = centrality_degree(weights = weight)) %>%
+ activate(edges) %>%
+ggraph(layout = "graphopt") + 
+  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(aes(size = deg, alpha = deg_c), col = "tomato", show.legend = FALSE) +
+  geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+      scale_color_manual(values = communitycolor)+
+  theme_graph()+
+  guides(color = guide_legend(override.aes = list(size = 0.5)), alpha = FALSE, edge_alpha = FALSE, edge_width = guide_legend(override.aes = list(size = 0.3)))+
+  ggtitle("network cluster 1 (red)")+
+  theme(legend.title = element_text(size = 8, face = "bold"),
+              legend.text = element_text(size = 7),
+              legend.key.size = unit(0.8, "lines"),
+        plot.title = element_text(size = 10, face= "bold"))
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 2) %>%
+  mutate(deg_c = centrality_eigen(weights = weight)) %>%
+flatten(weighted = TRUE) %>% 
+ggraph(layout = "graphopt") + 
+  geom_edge_fan(aes(edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(aes(size = deg, alpha = deg_c), col = "tomato", show.legend = FALSE) +
+  geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+      scale_color_manual(values = communitycolor)+
+  theme_graph()+
+  guides(color = guide_legend(override.aes = list(size = 0.5)), alpha = FALSE, edge_alpha = FALSE, edge_width = guide_legend(override.aes = list(size = 0.3)))+
+  ggtitle("network cluster 1 (red)")+
+  theme(legend.title = element_text(size = 8, face = "bold"),
+              legend.text = element_text(size = 7),
+              legend.key.size = unit(0.8, "lines"),
+        plot.title = element_text(size = 10, face= "bold"))
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 2) %>%
+ggraph(layout = "graphopt") + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "tomato") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 2) %>%
+ggraph(x = lon, y = lat) + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "tomato") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+```
+
+```{r network 3 centrality in community vs in gen network}
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_degree(weights = weight)) %>%
+  filter(louvain == 3) %>%
+  mutate(deg_c = centrality_degree(weights = weight)) %>%
+ activate(edges) %>%
+ggraph(layout = "graphopt") + 
+  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(aes(size = deg, alpha = deg_c), col = "darkgreen", show.legend = FALSE) +
+  geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+      scale_color_manual(values = communitycolor)+
+  theme_graph()+
+  guides(color = guide_legend(override.aes = list(size = 0.5)), alpha = FALSE, edge_alpha = FALSE, edge_width = guide_legend(override.aes = list(size = 0.3)))+
+  ggtitle("network cluster 1 (red)")+
+  theme(legend.title = element_text(size = 8, face = "bold"),
+              legend.text = element_text(size = 7),
+              legend.key.size = unit(0.8, "lines"),
+        plot.title = element_text(size = 10, face= "bold"))
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 3) %>%
+  mutate(deg_c = centrality_eigen(weights = weight)) %>%
+flatten(weighted = TRUE) %>% 
+ggraph(layout = "graphopt") + 
+  geom_edge_fan(aes(edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(aes(size = deg, alpha = deg_c), col = "darkgreen", show.legend = FALSE) +
+  geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+      scale_color_manual(values = communitycolor)+
+  theme_graph()+
+  guides(color = guide_legend(override.aes = list(size = 0.5)), alpha = FALSE, edge_alpha = FALSE, edge_width = guide_legend(override.aes = list(size = 0.3)))+
+  ggtitle("network cluster 1 (red)")+
+  theme(legend.title = element_text(size = 8, face = "bold"),
+              legend.text = element_text(size = 7),
+              legend.key.size = unit(0.8, "lines"),
+        plot.title = element_text(size = 10, face= "bold"))
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 3) %>%
+ggraph(layout = "graphopt") + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "darkgreen") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 3) %>%
+ggraph(x = lon, y = lat) + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "darkgreen") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+```
+
+## dont use this
 ```{r relative edge weights and centralization by cluster}
@@
 
+## checking the centrality of the nodes of a cluster
+
+centrality_deg.cluster <- decor_graph %>%
+    # filter community
+    activate(nodes) %>%
+    #filter(louvain == i) %>%
+    # degree centralization
+  #  mutate(cent = centrality_degree("all", normalized = TRUE)) %>%
+  mutate(centr = centrality_degree(weights = weight, mode = "all", normalized = TRUE)) %>% 
+#centr_degree(mode = "all", normalized = TRUE, weights = weight) %>%
+    as_tibble()
+
+centrality_eigen.cluster <- decor_graph %>%
+    # filter community
+    activate(nodes) %>%
+    #filter(louvain == i) %>%
+    # degree centralization
+  #  mutate(cent = centrality_degree("all", normalized = TRUE)) %>%
+  mutate(centr = centrality_eigen(weights = weight)) %>% 
+#centr_degree(mode = "all", normalized = TRUE, weights = weight) %>%
+    as_tibble()
+
+centr_deg.comb.cluster <- c()
+for (i in 1:3) {
+
+  centr_deg.comb.cluster[i] <- centrality_deg.cluster %>%
+    # filter community
+    filter(louvain == i) %>%
+    # degree centralization
+    summarise(centr_comb = mean(centr)) %>%
+    as_tibble() %>% 
+    pull(centr_comb)
+
+}
+
+centr_eig.comb.cluster <- c()
+for (i in 1:3) {
+
+  centr_eig.comb.cluster[i] <- centrality_eigen.cluster %>%
+    # filter community
+    filter(louvain == i) %>%
+    # degree centralization
+    summarise(centr_comb = mean(centr)) %>%
+    as_tibble() %>% 
+    pull(centr_comb)
+
+}
+
 # put together
-tibble(rel_edge_weight, centralization) %>%
+tibble(rel_edge_weight, 
+       centralization,
+    centr_deg.comb.cluster, 
+     centr_eig.comb.cluster
+     ) %>%
   mutate(cluster = 1:3) %>%
-  mutate(rel_edge_weight = rel_edge_weight / 100) %>%
+  mutate(rel_edge_weight = rel_edge_weight / 100,
+        centr_deg.comb.cluster = centr_deg.comb.cluster/1000,
+         centr_eig.comb.cluster = centr_eig.comb.cluster/0.1
+        ) %>%
   rename("relative edge weight" = rel_edge_weight) %>%
@@
   theme_publish()
+```
+
+## dont use this
+
+```{r relative edge weights, roles and centralization by cluster as boxplot}
+# Relative edge weights (how important are the connections within the cluster)
+rel_edge_weight <- data.frame()
+#for (i in 1:3) {
+  rel_edge_weight <- decor_graph %>%
+    # filter community
+    activate(nodes) %>%
+  #  filter(louvain == i) %>%
+    # relative edge weight (sum of weights / number of edges)
+    activate(edges) %>%
+    as.data.frame() %>%
+   mutate(louvain = i,
+          parameter = "edge_weight") %>% 
+    rename(value = weight) %>% 
+  select(value, louvain, parameter)
+#}
+
+# Centralization of the cluster
+centralization <- vector("double", length(3))
+for (i in 1:3) {
+  centralization[[i]] <- decor_graph %>%
+    # filter community
+    activate(nodes) %>%
+    filter(louvain == i) %>%
+    # degree centralization
+    centr_degree(mode = "all", normalized = TRUE) %>%
+    as_tibble() %>%
+    pull(centralization) %>%
+    head(1)
+}
+
+centralization <- data.frame(value = centralization, louvain = c(1:3), parameter = "centralization")
+## checking the centrality of the nodes of a cluster
+
+centrality_deg.cluster <- decor_graph %>%
+    # filter community
+    activate(nodes) %>%
+    #filter(louvain == i) %>%
+    # degree centralization
+  #  mutate(cent = centrality_degree("all", normalized = TRUE)) %>%
+  mutate(centr = centrality_degree(weights = weight, mode = "all", normalized = TRUE)) %>% 
+#centr_degree(mode = "all", normalized = TRUE, weights = weight) %>%
+    as_tibble()
+
+centrality_eigen.cluster <- decor_graph %>%
+    # filter community
+    activate(nodes) %>%
+    #filter(louvain == i) %>%
+    # degree centralization
+  #  mutate(cent = centrality_degree("all", normalized = TRUE)) %>%
+  mutate(centr = centrality_eigen(weights = weight)) %>% 
+#centr_degree(mode = "all", normalized = TRUE, weights = weight) %>%
+    as_tibble()
+
+centr_deg.comb.cluster <- c()
+for (i in 1:3) {
+
+  centr_deg.comb.cluster[i] <- centrality_deg.cluster %>%
+    # filter community
+    filter(louvain == i) %>%
+    # degree centralization
+    summarise(centr_comb = mean(centr)) %>%
+    as_tibble() %>% 
+    pull(centr_comb)
+
+}
+
+centr_eig.comb.cluster <- c()
+for (i in 1:3) {
+
+  centr_eig.comb.cluster[i] <- centrality_eigen.cluster %>%
+    # filter community
+    filter(louvain == i) %>%
+    # degree centralization
+    summarise(centr_comb = mean(centr)) %>%
+    as_tibble() %>% 
+    pull(centr_comb)
+
+}
+
+# put together
+tibble(rel_edge_weight, 
+     #  centralization,
+      #centr_deg.comb.cluster, 
+     #  centr_eig.comb.cluster
+     ) %>%
+  mutate(cluster = 1:3) %>%
+  mutate(rel_edge_weight = rel_edge_weight / 100,
+        centr_deg.comb.cluster = centr_deg.comb.cluster/1000#,
+        #  centr_eig.comb.cluster = centr_eig.comb.cluster/0.1
+        ) %>%
+  rename("relative edge weight" = rel_edge_weight) %>%
+  pivot_longer(!cluster, names_to = "key", values_to = "value") %>%
+  # plot
+  ggplot(aes(x = cluster, y = value, fill = key)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  scale_fill_grey() +
+  theme(axis.text.x = element_text(angle = 90)) +
+  theme_publish()
+```
+
+```{r add community to edges}
+
+#mark which edges are in which community or without
+decor_graph <- decor_graph %>% 
+  activate(edges) %>% 
+  mutate(community = case_when(from %in% which(nodes$louvain == 1) & to %in% which(nodes$louvain == 1) ~ "1",
+                               from %in% which(nodes$louvain == 2) & to %in% which(nodes$louvain == 2) ~ "2",
+                               from %in% which(nodes$louvain == 3) & to %in% which(nodes$louvain == 3) ~ "3",
+         TRUE ~ "crossing"))
+
+```
+
+```{r}
+decor_graph %>% 
+  get.data.frame(what = "edges") %>% 
+  count(community)
+
+edge_inside <- decor_graph %>% 
+  activate(edges) %>% 
+  filter(community != "crossing") %>% 
+  activate(nodes) %>% 
+  mutate(centr_comm = centrality_degree(weights = weight, mode = "all")) %>% 
+  get.data.frame("vertices") %>% 
+  mutate(centr_comm_weigthed = case_when(louvain == 1 ~ centr_comm/8,
+                                         louvain == 2 ~ centr_comm/6,
+                                         TRUE ~ centr_comm/3))%>% 
+  mutate(type = "inside")
+
+edge_outside <- decor_graph %>% 
+  activate(edges) %>% 
+  filter(community == "crossing") %>% 
+  activate(nodes) %>% 
+  mutate(centr_comm = centrality_degree(weights = weight, mode = "all")) %>% 
+  get.data.frame("vertices") %>% 
+  mutate(centr_comm_weigthed = case_when(louvain == 1 ~ centr_comm/(19-9),
+                                         louvain == 2 ~ centr_comm/(19-7),
+                                         TRUE ~ centr_comm/(19-4))) %>% 
+  mutate(type = "outside")
+
+data <- rbind(edge_inside, edge_outside) %>% 
+  mutate(louvain = as.character(louvain))
+
+# plot as a boxplot
+ggplot(data, aes(x = louvain, y = centr_comm_weigthed, fill = type)) +
+  geom_boxplot() +
+  scale_fill_manual(values = c("grey80", "black")) +
+  xlab("Community") +
+  ylab("Total Edge Weight") +
+  ggtitle("Weights of Edges Within and Between Communities")
+
+ggplot(data, aes(x = louvain, y = centr_comm, fill = type)) +
+  geom_boxplot() +
+  scale_fill_manual(values = c("grey80", "black")) +
+  xlab("Community") +
+  ylab("Total Edge Weight") +
+  ggtitle("Weights of Edges Within and Between Communities")
+```
+
+```{r}
+data <- rbind(edge_inside, edge_outside) %>% 
+  mutate(louvain_bb = as.character(louvain))
+
+data %>% 
+  group_by(louvain, type) %>% 
+  summarise(centr_comm = mean(centr_comm),
+            centr_comm_weigthed = mean(centr_comm_weigthed)) %>%
+  ggplot(aes(x = louvain, y = centr_comm_weigthed, fill = type)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(x = "Community", y = "Mean centrality") +
+  scale_fill_manual(values = c("red", "green", "blue")) +
+  theme_classic()
+
+data %>% 
+  group_by(louvain, type) %>% 
+  summarise(centr_comm_weigthed = mean(centr_comm_weigthed)) %>%
+  pivot_wider(names_from = type, values_from = centr_comm_weigthed) %>%
+  mutate(ratio_in_out = inside / outside)
+
+```
+
+```{r edge doenst work}
+library(dplyr)
+library(ggplot2)
+
+# group edges by community and whether they are within or outside the community
+edge_groups <- decor_graph %>% 
+  activate(edges) %>%
+  mutate(
+    community = ifelse(
+      from %in% nodes$louvain & to %in% nodes$louvain & nodes$louvain[from] == nodes$louvain[to],
+      as.character(nodes$louvain[from]),
+      "Crossing"
+    ),
+    is_within = ifelse(community == "Crossing", "Crossing", "Within")
+  ) %>%
+  group_by(community, is_within) %>%
+  summarize(weight = sum(weight))
+
+# create boxplot
+ggplot(edge_groups, aes(x = community, y = weight, fill = is_within)) +
+  geom_boxplot() +
+  labs(title = "Weight of Edges Within and Outside Communities",
+       x = "Community",
+       y = "Edge Weight")
+
+```
+
+```{r trash }
+# plot the weight of edges within the community
+ggplot(edges_within_community, aes(x = in_community, y = weight)) +
+  geom_boxplot() +
+  xlab("Community") +
+  ylab("Edge weight")
+
+# calculate edge weights
+g <- g %>% activate(edges) %>% mutate(weight = sample(1:10, n(), replace = TRUE))
+
+g <- decor_graph
+
+# calculate total weights within each community
+in_community <- g %>% 
+  activate(edges) %>%
+  group_by(from, to, louvain) %>%
+  summarise(total_weight = sum(weight)) %>%
+  ungroup()
+
+# calculate total weights between communities
+between_communities <- g %>% 
+  activate(edges) %>%
+  filter(!from %--% to) %>%
+  group_by(from, to) %>%
+  summarise(total_weight = sum(weight)) %>%
+  ungroup()
+
+# combine the data frames for plotting
+data <- rbind(in_community, between_communities)
+
+# plot as a boxplot
+ggplot(data, aes(x = community, y = total_weight, fill = from %--% to)) +
+  geom_boxplot() +
+  scale_fill_manual(values = c("grey80", "black")) +
+  xlab("Community") +
+  ylab("Total Edge Weight") +
+  ggtitle("Weights of Edges Within and Between Communities")
+
+library(tidygraph)
+library(igraph)
+
+# Create a tidygraph object from your data
+
+# Calculate the communities using the Louvain algorithm
+decor_graph <- decor_graph %>% activate(nodes) %>% mutate(louvain = group_louvain())
+
+# Mark edges that are within a community
+#decor_graph <- decor_graph %>% 
+#  activate(edges) %>% 
+  #mutate(in_community = group_edges(decor_graph, "louvain", .f = function(x) all(x == x[1])))
+
+# `in_community` column will be `TRUE` for edges within a community and `FALSE` for edges crossing communities
+decor_graph <- decor_graph %>%
+  activate(edges) %>%
+  mutate(
+    community = ifelse(
+      from %in% nodes$louvain & to %in% nodes$louvain & nodes$louvain[from] == nodes$louvain[to],
+      as.character(nodes$louvain[from]),
+      "crossing"
+    )
+  )
+
 ```
@@
     crossing == TRUE,
-    weight > 200
+    weight >= 10
   ) %>%
@@
 
+new cluster comparison by centralisation and centrality
+
+```{r cluster centralisation, fig.cap="Degree and eigen centralization by cluster"}
+### Centralization
+
+# Function to calculate centralization for each cluster
+calculate_centralization <- function(graph, louvain_column) {
+  # Create an empty tibble to store the centralization results
+  centralization <- tibble(cluster = character(), cent_deg = double(), cent_eigen = double())
+
+  # Get the unique clusters from the specified louvain column
+  unique_clusters <- graph |>
+    pull({{louvain_column}}) |>
+    unique()
+
+  # Loop through each cluster
+  for (cluster in unique_clusters) {
+    # Filter the graph based on the current cluster
+    subgraph <- graph |>
+      filter({{louvain_column}} == cluster)
+
+    # Calculate the centralization measures for the subgraph
+    cent_deg <- centr_degree(subgraph, normalized = TRUE)$centralization
+    cent_eigen <- centr_eigen(subgraph, normalized = TRUE)$centralization
+
+    # Add the results to the centralization tibble
+    centralization <- centralization |>
+      add_row(cluster = as.character(cluster), cent_deg = cent_deg, cent_eigen = cent_eigen)
+  }
+
+  # Arrange the centralization results by cluster and return the tibble
+  centralization <- centralization |>
+    arrange(cluster)
+  return(centralization)
+}
+
+# Calculate centralization for the given graph and louvain column
+calculate_centralization(decor_graph, louvain)
+
+calculate_centralization(decor_graph, louvain) |> 
+  pivot_longer(-cluster, names_to = "cent_type") |> 
+  ggplot() +
+  geom_bar(aes(x = cent_type, y = value, fill = cluster), stat = "identity", position = "dodge") +
+  theme_minimal() +
+  labs(x = "Centrality Type", y = "Centrality Value", fill = "Cluster")
+```
+
+```{r cluster centralities, fig.cap="Mean centralities (Degree & Eigen) by cluster"}
+### Centralities
+
+# Function to calculate centralities for each cluster group
+pull_centralities_clust <- function(graph) {
+  # Activate the nodes in the graph
+  graph |>
+    activate(nodes) |>
+    # Calculate degree centrality and eigenvector centrality
+    mutate(deg = centrality_degree(weights = weight, normalized = TRUE), eig = centrality_eigen(weights = weight)) |>
+    # Select the louvain cluster, degree centrality, and eigenvector centrality columns
+    select(label,louvain, deg, eig) |>
+    # Convert the result to a tibble
+    as_tibble()
+}
+
+# Calculate centralities for each cluster
+centr <- pull_centralities_clust(decor_graph)
+
+# Display boxplots of centralities for degree centrality
+a <- ggplot(centr) +
+  geom_boxplot(aes(x = louvain, y = deg, fill = louvain)) +
+  scale_fill_grey(start = 0.5, end  = 1) +
+  theme_bw() +
+  theme(legend.position = "none")+
+  labs(y = "degree centrality",
+       x = "community")
+
+# Display boxplots of centralities for eigenvector centrality
+b <- ggplot(centr) +
+  geom_boxplot(aes(x = louvain, y = eig, fill = louvain)) +
+  scale_fill_grey(start = 0.5, end  = 1) +
+  theme_bw() +
+  theme(legend.position = "none")+
+  labs(y = "eigenvector centrality",
+       x = "community")
+
+# Combine the boxplots
+a + b
+
+# Summary of centralities per cluster
+centr |>
+  group_by(louvain) |>
+  summarise(meandeg = mean(deg), meaneig = mean(eig))
+```
+
+If we take a closer look at the "internal" connections of community 1, we see that "braided_strip_roulette" is present at all sites.
+The other main decor "straight_cord_mat" on the other hand is missing on 2 sites: Oudalan and Prom.
+The two decors are spread across the whole area of the community.
+The only other decor that is common in the community is the "twisted_cord_roulette" which is also abundant across the whole research area.
+It makes up a rather large part of the assemblage in Oudalan and Prom but was not found at the sites of Sadia and Ambere Dougon.
+The fact that SAO has a high degree of centrality shows that the location is not decisive for the connectedness within this community.
+This is underlined by the fact that the less central sites are distributed across the whole area.
+There is an ambivalence between a relative consistency across the whole area and the diversity between sites that are close to each other.
+One reason for this phenom can be the contact to other communities in the mountain area.
+Damassogou exemplifies this as the site contains the defining decors of community 1 and 2.
+But the difference between Oudalan and SAO that are geographically isolated suggest that there might be more reasons for this diversity.
+Community 2 is less diverse with "braided_cord_roulette" as the only outstanding decor type.
+The weaker connections of the sites Kokolo and Damassogou are based on data bias (the proportion is counted in regard to all sherds not only the decorated ones), the decor was probably equally abundant there.
+Other decors spread across the community are "folded_strip_roulette" and "dragged_comb".
+The first one is missing in KAN 4 and TOD H while the second one is missing at Kokolo and Damassogou.
+The other indicator for the homogeneous character of the community is the nearly equal degree of centrality of every site.
+As mentioned above community 3 is more diverse but also has weaker connections.
+Instead of one defining decor there are several ones but none is abundant across the whole community.
+The spatial distribution differs between the decors.
+While "paint" and "twisted_cord_roulette" appear from Gao to TOG 2, "folded_strip_roulette", "channel" and "incised" are located in the western part of the community's area.
+Essouk and TMD have the highest degree of centrality.
+Even though it is located at the margin of the research area, its character as a "market"-site explains the amount of links it has.
+The role of TMD for the network is elaborated elsewhere (!).
+
 # Sparsification
 
-# Simplified network backbone
+## Simplified network backbone
 
@@
 decor.15 <- extract_backbone(decor.flat, 0.15)
-decor.20 <- extract_backbone(decor.flat, 0.20)
+decor.20 <- extract_backbone(decor.flat, 0.18)
 
@@
 library(readr)
-decorsearly <- read_csv2(here::here("analysis/data/raw_data/decors_early.csv"))
-
+decorsearly <- read_csv2(here::here("analysis/data/raw_data/decors_early.csv")) %>% 
+  select(-1)
+
+#decorsearly <- decorsearly%>%
+#  filter(site != "SAO 3" & site != "Oudalan Fer moyen")  %>%
+ ## mutate_at(vars(-site), as.numeric) %>% 
+ # mutate(site = case_when(site == "Akumbu 1" ~ "Akumbu early",
+     #    site == "TMD" ~ "TMD 3",
+   #      TRUE ~ site))
+
+#write.csv2(decorsearly, here::here("analysis/data/raw_data/decors_early.csv"))
 # apply function to transform frequency table into edge list
@@
 Wasserman & Faust p.180-183
+
+# Sparsified clustering
+
+## late cluster
+### calculating cluster with sparsification
+
+here we test clustering a sparsified network in contrast to the "normal" network. first we have to get the new clusters and then add them as louvain_bb to our decor_graph
+```{r}
+decor.20 <- decor.20 %>%
+  activate(nodes) %>%
+  mutate(louvain = as.character(group_louvain()))
+
+# calculate modularity of louvain clustering
+modularity.20 <- decor.20 %>%
+  activate(nodes) %>%
+  mutate(modularity = graph_modularity(group = as.factor(louvain))) %>%
+  pull(modularity) %>%
+  head(1)
+
+decor.20 %>%  
+  get.data.frame(what = "vertices") %>% 
+  count(louvain)
+
+decor_graph <- decor_graph %>% 
+    activate(nodes) %>%
+  mutate(louvain_bb = as.character(decor.20 %>% 
+  get.data.frame("vertices") %>% 
+  pull(louvain)))
+
+nodes <- decor_graph %>% 
+  get.data.frame("vertices")
+
+```
+
+### mapping the new clusters
+
+```{r plot-clusters, fig.cap="Louvain communities within the decor network"}
+# set plot colours for communities
+communitycolor <-
+  setNames(
+    c("tomato", "cornflowerblue", "darkgreen"),
+    unique(V(decor_graph)$louvain)
+  )
+
+# disable s2 geometry to plot on map
+sf_use_s2(FALSE)
+
+ggraph(decor_graph, x = lon, y = lat) +
+  geom_sf(data = mali_adm0, fill = NA) +
+  geom_sf(data = rivers, colour = "black") +
+  edges_community +
+  ggforce::geom_mark_hull(aes(x = lon, y = lat, fill = louvain_bb, colour = louvain_bb),
+                          expand = unit(1.5, "mm"),
+                          radius = unit(1.5, "mm")) +
+  scale_fill_manual(values = communitycolor)+
+  scale_colour_manual(values = communitycolor)+
+  nodes_community +
+  theme_community +
+  coord_sf(xlim = c(-7, 2), ylim = c(13, 19), expand = FALSE) +
+  theme_graph()
+```
+
+### taking a look at all 3 clusters
+
+he three communities can also be analysed as separate networks. The sites around the Niger Delta are mostly
+
+```{r network 1 centrality in community vs in gen network}
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain_bb == 1) %>%
+ggraph(layout = "graphopt") + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "cornflowerblue") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain_bb == 1) %>%
+ggraph(x = lon, y = lat) + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "cornflowerblue") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+```
+
+```{r network 2 centrality in community vs in gen network}
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain_bb == 2) %>%
+ggraph(layout = "graphopt") + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "tomato") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain_bb == 2) %>%
+ggraph(x = lon, y = lat) + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "tomato") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+```
+
+```{r network 3 centrality in community vs in gen network}
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain_bb == 3) %>%
+ggraph(layout = "graphopt") + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "darkgreen") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+
+decor_graph %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain_bb == 3) %>%
+ggraph(x = lon, y = lat) + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "darkgreen") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+```
+
+### comparing connections within and without
+
+```{r add community to edges}
+
+#mark which edges are in which community or without
+decor_graph <- decor_graph %>% 
+  activate(edges) %>% 
+  mutate(community_bb = case_when(from %in% which(nodes$louvain_bb == 1) & to %in% which(nodes$louvain_bb == 1) ~ "1",
+                               from %in% which(nodes$louvain_bb == 2) & to %in% which(nodes$louvain_bb == 2) ~ "2",
+                               from %in% which(nodes$louvain_bb == 3) & to %in% which(nodes$louvain_bb == 3) ~ "3",
+                               from %in% which(nodes$louvain_bb == 4) & to %in% which(nodes$louvain_bb == 4) ~ "4",
+                               from %in% which(nodes$louvain_bb == 5) & to %in% which(nodes$louvain_bb == 5) ~ "5",
+         TRUE ~ "crossing"))
+
+decor_graph 
+```
+
+```{r crossings}
+decor_graph %>%
+  activate(edges) %>%
+
+  # find the edges that cross between communities
+#  mutate(crossing = community_crossing(weights = weight)) %>%
+  filter(
+    community_bb == "crossing",
+    weight >= 3
+  ) %>%
+  activate(nodes) %>% 
+  filter(!node_is_isolated()) %>%
+
+  #plot
+  ggraph(layout = "nicely") +
+
+  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
+  # scale_edge_color_manual(values = decorcolor) +
+  scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(aes(size = centrality_degree(weights = weight), color = louvain_bb), show.legend = FALSE) +
+  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
+  scale_color_manual(values = communitycolor)+
+
+  theme_graph() +
+  theme_community
+
+```
+
+```{r}
+decor_graph %>% 
+  get.data.frame(what = "edges") %>% 
+  count(community_bb)
+
+edge_inside <- decor_graph %>% 
+  activate(edges) %>% 
+  filter(community_bb != "crossing") %>% 
+  activate(nodes) %>% 
+  mutate(centr_comm = centrality_degree(weights = weight, mode = "all")) %>% 
+  get.data.frame("vertices") %>% 
+  mutate(centr_comm_weigthed = case_when(louvain_bb == 1 ~ centr_comm/7,
+                                         louvain_bb == 2 ~ centr_comm/4,
+                                         TRUE ~ centr_comm/4))%>% 
+  mutate(type = "inside")
+
+edge_outside <- decor_graph %>% 
+  activate(edges) %>% 
+  filter(community_bb == "crossing") %>% 
+  activate(nodes) %>% 
+  mutate(centr_comm = centrality_degree(weights = weight, mode = "all")) %>% 
+  get.data.frame("vertices") %>% 
+  mutate(centr_comm_weigthed = case_when(louvain_bb == 1 ~ centr_comm/(19-8),
+                                         louvain_bb == 2 ~ centr_comm/(19-5),
+                                         louvain_bb == 3 ~ centr_comm/(19-5),
+                                         TRUE ~ centr_comm/(19-1))) %>% 
+  mutate(type = "outside")
+
+data <- rbind(edge_inside, edge_outside) %>% 
+  mutate(louvain_bb = as.character(louvain_bb))
+
+# plot as a boxplot
+ggplot(data, aes(x = louvain_bb, y = centr_comm_weigthed, fill = type)) +
+  geom_boxplot() +
+  scale_fill_manual(values = c("grey80", "black")) +
+  xlab("Community") +
+  ylab("Total Edge Weight") +
+  ggtitle("Weights of Edges Within and Between Communities")
+
+ggplot(data, aes(x = louvain_bb, y = centr_comm, fill = type)) +
+  geom_boxplot() +
+  scale_fill_manual(values = c("grey80", "black")) +
+  xlab("Community") +
+  ylab("Total Edge Weight") +
+  ggtitle("Weights of Edges Within and Between Communities")
+
+```
+the inside vs outside ratio  is 
+```{r}
+data %>% 
+  group_by(louvain_bb, type) %>% 
+  summarise(centr_comm = mean(centr_comm),
+            centr_comm_weigthed = mean(centr_comm_weigthed)) %>%
+  ggplot(aes(x = louvain_bb, y = centr_comm_weigthed, fill = type)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(x = "Community", y = "Mean centrality") +
+  scale_fill_manual(values = c("red", "green", "blue")) +
+  theme_classic()
+
+data %>% 
+  group_by(louvain_bb, type) %>% 
+  summarise(centr_comm_weigthed = mean(centr_comm_weigthed)) %>%
+  pivot_wider(names_from = type, values_from = centr_comm_weigthed) %>%
+  mutate(ratio_in_out = inside / outside)
+```
+
+# sparsification cluster early
+
+# Sparsified clustering
+
+here we test clustering a sparsified network in contrast to the "normal" network. first we have to get the new clusters and then add them as louvain_bb to our decor_graph
+```{r}
+early.flat.bb <- early.flat.bb %>%
+  activate(nodes) %>%
+  mutate(louvain = as.character(group_louvain()))
+
+# calculate modularity of louvain clustering
+modularity.early.bb <- early.flat.bb %>%
+  activate(nodes) %>%
+  mutate(modularity = graph_modularity(group = as.factor(louvain))) %>%
+  pull(modularity) %>%
+  head(1)
+
+early.flat.bb %>%  
+  get.data.frame(what = "vertices") %>% 
+  count(louvain)
+
+decor_graph_early <- decor_graph_early %>% 
+  mutate(louvain_bb = as.character(early.flat.bb %>% 
+  get.data.frame("vertices") %>% 
+  pull(louvain)))
+
+nodes_early <- decor_graph_early %>% 
+  get.data.frame("vertices")
+
+```
+
+the new clusters
+
+```{r plot-clusters, fig.cap="Louvain communities within the decor network"}
+# set plot colours for communities
+communitycolor <-
+  setNames(
+    c("tomato", "cornflowerblue", "darkgreen"),
+    unique(V(decor_graph)$louvain)
+  )
+
+# disable s2 geometry to plot on map
+sf_use_s2(FALSE)
+
+ggraph(decor_graph_early, x = lon, y = lat) +
+  geom_sf(data = mali_adm0, fill = NA) +
+  geom_sf(data = rivers, colour = "black") +
+  edges_community +
+  ggforce::geom_mark_hull(aes(x = lon, y = lat, fill = louvain_bb, colour = louvain_bb),
+                          expand = unit(1.5, "mm"),
+                          radius = unit(1.5, "mm")) +
+  scale_fill_manual(values = communitycolor)+
+  scale_colour_manual(values = communitycolor)+
+  nodes_community +
+  theme_community +
+  coord_sf(xlim = c(-7, 2), ylim = c(13, 19), expand = FALSE) +
+  theme_graph()
+```
+
+he three communities can also be analysed as separate networks. The sites around the Niger Delta are mostly
+
+```{r network 1 centrality in community vs in gen network}
+
+decor_graph_early %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain_bb == 1) %>%
+ggraph(layout = "graphopt") + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph_early)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "cornflowerblue") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+
+decor_graph_early %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain_bb == 1) %>%
+ggraph(x = lon, y = lat) + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph_early)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "cornflowerblue") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+```
+
+```{r network 2 centrality in community vs in gen network}
+
+decor_graph_early %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain_bb == 2) %>%
+ggraph(layout = "graphopt") + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph_early)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "tomato") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+
+decor_graph_early %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain_bb == 2) %>%
+ggraph(x = lon, y = lat) + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph_early)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "tomato") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+```
+
+```{r network 3 centrality in community vs in gen network}
+
+decor_graph_early %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain_bb == 3) %>%
+ggraph(layout = "graphopt") + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph_early)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "darkgreen") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+
+decor_graph_early %>%
+  activate(nodes) %>%
+    mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain_bb == 3) %>%
+ggraph(x = lon, y = lat) + 
+geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) + 
+     scale_edge_color_manual(values = decorcolor[E(decor_graph_early)$name]) +
+ scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(size = 3, colour = "darkgreen") +
+    geom_node_text(aes(label = label), col= "black", repel = TRUE) +
+  theme_graph()
+```
+
+```{r add community to edges}
+
+#mark which edges are in which community or without
+decor_graph_early <- decor_graph_early %>% 
+  activate(edges) %>% 
+  mutate(community_bb = case_when(from %in% which(nodes_early$louvain_bb == 1) & to %in% which(nodes_early$louvain_bb == 1) ~ "1",
+                               from %in% which(nodes_early$louvain_bb == 2) & to %in% which(nodes_early$louvain_bb == 2) ~ "2",
+                               from %in% which(nodes_early$louvain_bb == 3) & to %in% which(nodes_early$louvain_bb == 3) ~ "3",
+                               from %in% which(nodes_early$louvain_bb == 4) & to %in% which(nodes_early$louvain_bb == 4) ~ "4",
+                               from %in% which(nodes_early$louvain_bb == 5) & to %in% which(nodes_early$louvain_bb == 5) ~ "5",
+         TRUE ~ "crossing"))
+
+decor_graph_early 
+```
+
+```{r crossings}
+decor_graph_early %>%
+  activate(edges) %>%
+
+  # find the edges that cross between communities
+#  mutate(crossing = community_crossing(weights = weight)) %>%
+  filter(
+    community_bb == "crossing",
+    weight >= 3
+  ) %>%
+  activate(nodes) %>% 
+  filter(!node_is_isolated()) %>%
+
+  #plot
+  ggraph(layout = "nicely") +
+
+  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
+  # scale_edge_color_manual(values = decorcolor) +
+  scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(aes(size = centrality_degree(weights = weight), color = louvain_bb), show.legend = FALSE) +
+  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
+  scale_color_manual(values = communitycolor)+
+
+  theme_graph() +
+  theme_community
+
+decor_graph_early %>%
+  activate(edges) %>%
+
+  # find the edges that cross between communities
+#  mutate(crossing = community_crossing(weights = weight)) %>%
+  filter(
+   # community_bb == "crossing",
+    weight >= 3
+  ) %>%
+  activate(nodes) %>% 
+  filter(!node_is_isolated()) %>%
+
+  #plot
+  ggraph(layout = "nicely") +
+
+  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
+  # scale_edge_color_manual(values = decorcolor) +
+  scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(aes(size = centrality_degree(weights = weight), color = louvain_bb), show.legend = FALSE) +
+  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
+  scale_color_manual(values = communitycolor)+
+
+  theme_graph() +
+  theme_community
+
+```
+
+```{r}
+decor_graph_early %>% 
+  get.data.frame(what = "edges") %>% 
+  count(community_bb)
+
+edge_inside <- decor_graph_early %>% 
+  activate(edges) %>% 
+  filter(community_bb != "crossing") %>% 
+  activate(nodes) %>% 
+  mutate(centr_comm = centrality_degree(weights = weight, mode = "all")) %>% 
+  get.data.frame("vertices") %>% 
+  mutate(centr_comm_weigthed = case_when(louvain_bb == 1 ~ centr_comm/3,
+                                         louvain_bb == 2 ~ centr_comm/3,
+                                         louvain_bb == 3 ~ centr_comm/3,
+TRUE ~ centr_comm/1))%>% 
+  mutate(type = "inside")
+
+edge_outside <- decor_graph_early %>% 
+  activate(edges) %>% 
+  filter(community_bb == "crossing") %>% 
+  activate(nodes) %>% 
+  mutate(centr_comm = centrality_degree(weights = weight, mode = "all")) %>% 
+  get.data.frame("vertices") %>% 
+  mutate(centr_comm_weigthed = case_when(louvain_bb == 1 ~ centr_comm/(12-4),
+                                         louvain_bb == 2 ~ centr_comm/(12-4),
+                                        louvain_bb == 3 ~ centr_comm/(12-4),
+                                         TRUE ~ centr_comm/(12-1))) %>% 
+  mutate(type = "outside")
+
+data <- rbind(edge_inside, edge_outside) %>% 
+  mutate(louvain_bb = as.character(louvain_bb))
+
+# plot as a boxplot
+ggplot(data, aes(x = louvain_bb, y = centr_comm_weigthed, fill = type)) +
+  geom_boxplot() +
+  scale_fill_manual(values = c("grey80", "black")) +
+  xlab("Community") +
+  ylab("Total Edge Weight") +
+  ggtitle("Weights of Edges Within and Between Communities")
+
+ggplot(data, aes(x = louvain_bb, y = centr_comm, fill = type)) +
+  geom_boxplot() +
+  scale_fill_manual(values = c("grey80", "black")) +
+  xlab("Community") +
+  ylab("Total Edge Weight") +
+  ggtitle("Weights of Edges Within and Between Communities")
+
+```
+
+the inside vs outside ratio  is 
+```{r}
+data %>% 
+  group_by(louvain_bb, type) %>% 
+  summarise(centr_comm = mean(centr_comm),
+            centr_comm_weigthed = mean(centr_comm_weigthed)) %>%
+  ggplot(aes(x = louvain_bb, y = centr_comm_weigthed, fill = type)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(x = "Community", y = "Mean centrality") +
+  scale_fill_manual(values = c("red", "green", "blue")) +
+  theme_classic()
+
+data %>% 
+  group_by(louvain_bb, type) %>% 
+  summarise(centr_comm_weigthed = mean(centr_comm_weigthed)) %>%
+  pivot_wider(names_from = type, values_from = centr_comm_weigthed) %>%
+  mutate(ratio_in_out = inside / outside)
+```
+
+# cluster measures
+## inside outside normal cluster
+### comparing connections within and without
+
+```{r add community to edges}
+
+#mark which edges are in which community or without
+decor_graph <- decor_graph %>% 
+  activate(edges) %>% 
+  mutate(community = case_when(from %in% which(nodes$louvain == 1) & to %in% which(nodes$louvain == 1) ~ "1",
+                               from %in% which(nodes$louvain == 2) & to %in% which(nodes$louvain == 2) ~ "2",
+                               from %in% which(nodes$louvain == 3) & to %in% which(nodes$louvain == 3) ~ "3",
+                               from %in% which(nodes$louvain == 4) & to %in% which(nodes$louvain == 4) ~ "4",
+                               from %in% which(nodes$louvain == 5) & to %in% which(nodes$louvain == 5) ~ "5",
+         TRUE ~ "crossing"))
+
+decor_graph 
+```
+
+```{r crossings}
+decor_graph %>%
+  activate(edges) %>%
+
+  # find the edges that cross between communities
+#  mutate(crossing = community_crossing(weights = weight)) %>%
+  filter(
+    community == "crossing",
+    weight >= 3
+  ) %>%
+  activate(nodes) %>% 
+  filter(!node_is_isolated()) %>%
+
+  #plot
+  ggraph(layout = "nicely") +
+  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
+  # scale_edge_color_manual(values = decorcolor) +
+  scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(aes(size = centrality_degree(weights = weight), color = louvain), show.legend = FALSE) +
+  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
+  scale_color_manual(values = communitycolor)+
+
+  theme_graph() +
+  theme_community
+
+```
+
+```{r}
+decor_graph %>% 
+  get.data.frame(what = "edges") %>% 
+  count(community)
+
+edge_inside <- decor_graph %>% 
+  activate(edges) %>% 
+  filter(community != "crossing") %>% 
+  activate(nodes) %>% 
+  mutate(centr_comm = centrality_degree(weights = weight, mode = "all")) %>% 
+  get.data.frame("vertices") %>% 
+  mutate(centr_comm_weigthed = case_when(louvain == 1 ~ centr_comm/8,
+                                         louvain == 2 ~ centr_comm/6,
+                                         TRUE ~ centr_comm/3))%>% 
+  mutate(type = "inside")
+
+edge_outside <- decor_graph %>% 
+  activate(edges) %>% 
+  filter(community == "crossing") %>% 
+  activate(nodes) %>% 
+  mutate(centr_comm = centrality_degree(weights = weight, mode = "all")) %>% 
+  get.data.frame("vertices") %>% 
+  mutate(centr_comm_weigthed = case_when(louvain == 1 ~ centr_comm/(19-9),
+                                         louvain == 2 ~ centr_comm/(19-7),
+                                         louvain == 3 ~ centr_comm/(19-4),
+                                         TRUE ~ centr_comm/(19-1))) %>% 
+  mutate(type = "outside")
+
+data <- rbind(edge_inside, edge_outside) %>% 
+  mutate(louvain = as.character(louvain))
+
+# plot as a boxplot
+ggplot(data, aes(x = louvain, y = centr_comm_weigthed, fill = type)) +
+  geom_boxplot() +
+  scale_fill_manual(values = c("grey80", "black")) +
+  xlab("Community") +
+  ylab("Total Edge Weight") +
+  ggtitle("Weights of Edges Within and Between Communities")
+
+ggplot(data, aes(x = louvain, y = centr_comm, fill = type)) +
+  geom_boxplot() +
+  scale_fill_manual(values = c("grey80", "black")) +
+  xlab("Community") +
+  ylab("Total Edge Weight") +
+  ggtitle("Weights of Edges Within and Between Communities")
+
+```
+the inside vs outside ratio  is 
+```{r}
+data %>% 
+  group_by(louvain, type) %>% 
+  summarise(centr_comm = mean(centr_comm),
+            centr_comm_weigthed = mean(centr_comm_weigthed)) %>%
+  ggplot(aes(x = louvain, y = centr_comm_weigthed, fill = type)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(x = "Community", y = "Mean centrality") +
+  scale_fill_manual(values = c("red", "green", "blue")) +
+  theme_classic()
+
+data %>% 
+  group_by(louvain, type) %>% 
+  summarise(centr_comm_weigthed = mean(centr_comm_weigthed)) %>%
+  pivot_wider(names_from = type, values_from = centr_comm_weigthed) %>%
+  mutate(ratio_in_out = inside / outside)
+```
+## centralities and centralisation
+
+### centralities by cluster
+
+```{r}
+pull_centralities_clust <- function(graph) {
+  graph %>%
+    activate(nodes) %>%
+    mutate(deg = centrality_degree(weights = weight, normalized = TRUE), eig = centrality_eigen(weights = weight)) %>%
+    select(louvain, deg, eig) %>%
+    as_tibble()
+}
+
+centr_late <- pull_centralities_clust(decor_graph) %>% 
+  mutate(period = "late")
+
+centr_early <- pull_centralities_clust(decor_graph_early) %>%
+  mutate(period = "early")
+
+centr <- bind_rows(centr_early, centr_late)
+
+a <- ggplot(centr) +
+  geom_boxplot(aes(x = period, y = deg, fill = louvain)) +
+  theme_minimal() +
+  ylab("degree centrality")
+
+b <- ggplot(centr) +
+  geom_boxplot(aes(x = period, y = eig, fill = louvain)) +
+  theme_minimal() +
+  ylab("eigenvector centrality")
+
+a + b
+
+centralities %>%
+  group_by(period) %>%
+  summarise(meandeg = mean(deg), meaneig = mean(eig))
+```
+### centralities
+
+```{r}
+pull_centralities_clust <- function(graph) {
+  graph %>%
+    activate(nodes) %>%
+    mutate(deg = centrality_degree(weights = weight, normalized = TRUE), eig = centrality_eigen(weights = weight)) %>%
+    select(louvain, deg, eig) %>%
+    as_tibble()
+}
+
+centr <- pull_centralities_clust(decor_graph)
+
+a <- ggplot(centr) +
+  geom_boxplot(aes(x = louvain, y = deg, fill = louvain)) +
+  theme_minimal() +
+  ylab("degree centrality")
+
+b <- ggplot(centr) +
+  geom_boxplot(aes(x = louvain, y = eig, fill = louvain)) +
+  theme_minimal() +
+  ylab("eigenvector centrality")
+
+a + b
+
+centralities %>%
+  group_by(louvain) %>%
+  summarise(meandeg = mean(deg), meaneig = mean(eig))
+```
+
+### centralisation
+
+```{r}
+library(tidyverse)
+library(igraph)
+
+calculate_centralization <- function(graph, louvain_column) {
+  centralization <- tibble(cluster = character(), cent_deg = double(), cent_eigen = double())
+
+  unique_clusters <- graph %>%
+    pull({{louvain_column}}) %>%
+    unique()
+
+  for (cluster in unique_clusters) {
+    subgraph <- graph %>%
+      filter({{louvain_column}} == cluster)
+
+    cent_deg <- centr_degree(subgraph, normalized = TRUE)$centralization
+    cent_eigen <- centr_eigen(subgraph, normalized = TRUE)$centralization
+
+    centralization <- centralization %>%
+      add_row(cluster = as.character(cluster), cent_deg = cent_deg, cent_eigen = cent_eigen)
+  }
+
+  centralization <- centralization |> 
+    arrange(cluster)
+  return(centralization)
+}
+
+calculate_centralization(decor_graph, louvain)
+```
+
+```{r}
+netmeasurements <- function(graph, type, dist){
+
+  trans <- graph %>%
+    igraph::transitivity()  # calculate transitivity
+
+  dens <- graph %>%
+    igraph::graph.density()  # calculate density
+
+  diam <- graph %>%
+    activate(edges) %>%
+    mutate(weight = (dist-weight)) %>%
+    diameter()
+
+  rad <- graph %>%
+    radius()
+
+  cent_deg <- centr_degree(graph, normalized = TRUE)$centralization
+  cent_eigen <- centr_eigen(graph, normalized = TRUE)$centralization
+
+  graph %>%
+    mutate(meandist = graph_mean_dist(),  # calculate mean distance
+           size = graph_size(),
+           order= graph_order()) %>%  # number of edges
+    select(meandist, size, order) %>%  # select measures
+    as_tibble() %>%  # turn into dataframe
+    head(1) %>%   # select only first row (as data is repetitive)
+    mutate(type = type,
+           transitivity = trans,
+           density = dens,
+           diameter = diam,
+           # radius = rad,
+           centralization.deg = cent_deg,
+           centralization.eigen = cent_eigen) %>%   # add all measures together
+    select(type, everything()) # order phase to the front
+}
+```
+
+### centralisation diachronic comparison
+```{r}
+
+```

### 2023-08-03 analysis/paper/paper.Rmd
code style

@@
   scale_fill_grey() +
-  theme_publish() +
+  theme_bw() +
   ylab("mean edge weight")
@@
   scale_fill_grey() +
-  theme_publish()
+  theme_bw()
 

### 2023-08-03 analysis/R/datavis themes.R

@@ -1,70 +1,6 @@
 
-#1. publication theme for ggplot (from the envalysis package, author: Zacharias Steinmetz)
-theme_publish <- function(base_size = 12, base_family = "",
-                          line_size = 0.25, ...) {
-  half_line <- base_size / 2
-  small_rel <- 0.8
-  small_size <- small_rel * base_size
 
-  theme_bw(base_size = base_size, base_family = base_family, ...) %+replace%
-    theme(
-      rect = element_rect(fill = "transparent", colour = NA, color = NA,
-                          size = 0, linetype = 0),
-      text = element_text(family = base_family, face = "plain",
-                          colour = "black", size = base_size, hjust = 0.5,
-                          vjust = 0.5, angle = 0, lineheight = 0.9,
-                          margin = ggplot2::margin(), debug = F),
-
-      axis.text = element_text(size = small_size),
-      axis.text.x = element_text(margin = ggplot2::margin(t = small_size/4),
-                                 vjust = 1),
-      axis.text.y = element_text(margin = ggplot2::margin(r = small_size/4),
-                                 hjust = 1),
-      axis.title.x = element_text(margin = ggplot2::margin(t = small_size,
-                                                           b = small_size)),
-      axis.title.y = element_text(angle = 90,
-                                  margin = ggplot2::margin(r = small_size,
-                                                           l = small_size/4)),
-      axis.ticks = element_line(colour = "black", size = line_size),
-      axis.ticks.length = unit(0.25, 'lines'),
-
-      axis.line = element_line(colour = "black", size = line_size),
-      axis.line.x = element_line(colour = "black", size = line_size),
-      axis.line.y = element_line(colour = "black", size = line_size),
-
-      legend.spacing = unit(base_size/4, "pt"),
-      legend.key = element_blank(),
-      legend.key.size = unit(1 * base_size, "pt"),
-      legend.key.width = unit(1.5 * base_size, 'pt'),
-      legend.text = element_text(size = rel(small_rel)),
-      legend.title = element_text(size = rel(small_rel), face = 'bold'),
-      legend.position = 'bottom',
-      legend.box = 'horizontal',
-
-      panel.spacing = unit(1, "lines"),
-      panel.background = element_blank(),
-      panel.border = element_blank(),
-      panel.grid.major = element_blank(),
-      panel.grid.minor = element_blank(),
-
-      strip.text = element_text(size = base_size),
[…]

### 2023-08-03 analysis/R/edgelist functions.R

@@ -5,15 +5,14 @@ library(dplyr)
 
 # 1.function to make a percentage table from presence/absence data
 
-make_presabs_prop <- function(data, col_node){
-
+make_presabs_prop <- function(data, col_node) {
   col_node <- enquo(col_node)
 
   data %>%
     tidyr::pivot_longer(!(!!col_node), names_to = "type", values_to = "count") %>%
     group_by((!!col_node), type) %>%
     count(count) %>%
-    mutate(freq = n / sum(n) *100) %>%
+    mutate(freq = n / sum(n) * 100) %>%
     filter(count == 1) %>%
     select((!!col_node), type, freq) %>%
     tidyr::pivot_wider(names_from = type, values_from = freq) %>%
@@ -24,18 +23,17 @@ make_presabs_prop <- function(data, col_node){
 
 # 2.Make a proportions table from raw counts, to feed to the edgelist function:
 
-make_count_prop <- function (x){
-
+make_count_prop <- function(x) {
   x_gather <- x %>%
     tidyr::gather(key = "type", value = "type_count", -Site)
 
   x_sum <- x_gather %>%
-   group_by(Site) %>%
-   summarise(Site_sum = sum(type_count))
+    group_by(Site) %>%
+    summarise(Site_sum = sum(type_count))
 
   x_prop <- x_gather %>%
     left_join(x_sum, by = "Site") %>%
-    mutate(prop = type_count / Site_sum *100) %>%
+    mutate(prop = type_count / Site_sum * 100) %>%
     select(-type_count, -Site_sum) %>%
     mutate(type = paste0(type)) %>%
     tidyr::spread(key = type, value = prop)
@@ -46,33 +44,37 @@ make_count_prop <- function (x){
 ###############
 # 3. make edgelist from percentage data listed by site and variable
 
-make_edgelist <- function(data){
-  data <- data%>% select_if(function(x){!all(is.na(x))})  #only select those coumns that have values in them
+make_edgelist <- function(data) {
+  data <- data %>% select_if(function(x) {
+    !all(is.na(x))
+  }) # only select those coumns that have values in them
 
-  data_long <- data %>% tidyr::pivot_longer(-1, names_to = "name") %>% dplyr::rename(Var1 = 1)  #make data long
+  data_long <- data %>%
+    tidyr::pivot_longer(-1, names_to = "name") %>%
+    dplyr::rename(Var1 = 1) # make data long
 
[…]

### 2023-08-03 analysis/R/map plots.R

@@ -16,76 +16,92 @@ load(here::here("analysis", "data", "derived_data", "geodata.rda"))
 
 # 1. Greyscale map with scale bar
 
-maptheme_bw_scale = theme(panel.grid = element_blank()) +
+maptheme_bw_scale <- theme(panel.grid = element_blank()) +
   theme(axis.text = element_blank()) +
   theme(axis.ticks = element_blank()) +
   theme(axis.title = element_blank()) +
   theme(legend.position = "bottom") +
   theme(panel.grid = element_blank()) +
   theme(panel.background = element_rect(fill = "white")) +
-  theme(plot.margin = unit(c(0, 0, 0.5, 0), 'cm'))+
-  theme(panel.border = element_rect(colour = "black", fill=NA, size=2))
+  theme(plot.margin = unit(c(0, 0, 0.5, 0), "cm")) +
+  theme(panel.border = element_rect(colour = "black", fill = NA, size = 2))
 
-map_bw_scale <- list(geom_sf(data = mali_adm0, fill = NA),
-                     geom_sf(data = rivers, fill = "grey", colour = "grey") ,
-                     ggspatial::annotation_scale(aes(location = "br", style ="ticks")),
-                     maptheme_bw_scale
-                      )
+map_bw_scale <- list(
+  geom_sf(data = mali_adm0, fill = NA),
+  geom_sf(data = rivers, fill = "grey", colour = "grey"),
+  ggspatial::annotation_scale(aes(location = "br", style = "ticks")),
+  maptheme_bw_scale
+)
 
 # 2. Greyscale map with degree ticks
 
-maptheme_bw_grid = theme(panel.grid = element_blank()) +
-  theme(axis.text.x=element_text(angle=90, hjust=1)) +
+maptheme_bw_grid <- theme(panel.grid = element_blank()) +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
   theme(axis.title = element_blank()) +
   theme(legend.position = "bottom") +
   theme(panel.grid = element_blank()) +
   theme(panel.background = element_rect(fill = "white")) +
-  theme(plot.margin = unit(c(0, 0, 0.5, 0), 'cm'))+
-  theme(panel.border = element_rect(colour = "black", fill=NA, size=1.5))
+  theme(plot.margin = unit(c(0, 0, 0.5, 0), "cm")) +
+  theme(panel.border = element_rect(colour = "black", fill = NA, size = 1.5))
 
[…]

### 2023-08-03 analysis/R/network functions.R

@@ -9,118 +9,137 @@ library(tidyr)
 library(tidygraph)
 library(ggplot2)
 
+
 # 1. function to mark edges crossing between louvain clusters
 
-community_crossing <- function(weights = NULL) {
-  #expect_nodes()
-  weights <- enquo(weights)
-  weights <- eval_tidy(weights, .E())
-  if (is.null(weights)) {
-    weights <- NA
-  }
-  group <- igraph::crossing(cluster_louvain(graph = .G(), weights = weights), .G())
+add_community_edges <- function(x) {
+  # extract nodes
+  nodes <- x |>
+    get.data.frame("vertices")
+
+  x <- x |>
+    activate(edges) |>
+    # adding community/ crossing to edges, based on their nodes
+    mutate(community = case_when(
+      from %in% which(nodes$louvain == 1) &
+        to %in% which(nodes$louvain == 1) ~ "1",
+      from %in% which(nodes$louvain == 2) &
+        to %in% which(nodes$louvain == 2) ~ "2",
+      from %in% which(nodes$louvain == 3) &
+        to %in% which(nodes$louvain == 3) ~ "3",
+      TRUE ~ "crossing"
+    ))
+
+  return(x)
 }
+
 
 # 2. function to extract individual layer
 
-onelayer <- function(lay){
+onelayer <- function(lay) {
   layx <- rlang::sym(lay)
-  prep.graph %>% activate(edges) %>% filter(layer == layx) %>% activate(nodes) %>% filter(!node_is_isolated()) %>% as.igraph
+  prep_graph |>
+    activate(edges) |>
+    filter(layer == layx) |>
+    activate(nodes) |>
+    filter(!node_is_isolated()) |>
+    as.igraph()
 }
+
 
 # 3. function to make a heatmap from matrix
-heatmap_matrix <- function(matrix){
+
+heatmap_matrix <- function(matrix) {
   matrix[lower.tri(matrix, diag = TRUE)] <- NA
-  cor_tri <- as.data.frame(matrix) %>%
-    mutate(Var1 = factor(row.names(.), levels=row.names(.))) %>%
-    tidyr::gather(key = Var2, value = value, -Var1, na.rm = TRUE, factor_key = TRUE)
+  cor_tri <- as.data.frame(matrix) |>
+    mutate(Var1 = factor(row.names(matrix), levels = row.names(matrix))) |>
[…]

### 2023-08-03 analysis/supplementary_materials/map backgrounds.Rmd

@@
 ```{r}
-africa_outline <- sf::st_read(here::here("analysis", "data", "raw_data", "Africa_outline.shp")) %>% sf::st_crop(xmax = 54, xmin = -20, ymax = 40, ymin = -36)
+africa_outline <- sf::st_read(here::here("analysis", "data",
+                                         "raw_data", "Africa_outline.shp")) |>
+  sf::st_crop(xmax = 54, xmin = -20, ymax = 40, ymin = -36)
 ```
@@
 ```{r subset adm0}
-mali_adm0 <- ne_countries(scale = "medium", returnclass = "sf") %>% filter(admin == "Mali")
+mali_adm0 <- ne_countries(scale = "medium", returnclass = "sf") |>
+  filter(admin == "Mali")
 ```
@@
 ```{r plot adm0}
-ggplot()+
-  geom_sf(data = mali_adm0)+
+ggplot() +
+  geom_sf(data = mali_adm0) +
   geom_sf_text(data = mali_adm0, aes(label = admin))
 
-ggplot(africa_outline)+
+ggplot(africa_outline) +
   geom_sf()
@@
 ```{r}
-rivers <- ne_download(scale = 50, type = "rivers_lake_centerlines", category = "physical", returnclass = "sf") %>% sf::st_intersection(mali_adm0) %>% filter (name == "Niger")
+rivers <- ne_download(scale = 50, type = "rivers_lake_centerlines",
+                      category = "physical", returnclass = "sf") |>
+  sf::st_intersection(mali_adm0) |> filter(name == "Niger")
 
@@
 ```{r plot hydro}
-ggplot()+
-  geom_sf(data = mali_adm0)+
-  geom_sf_text(data = mali_adm0, aes(label = admin))+
-  geom_sf(data = rivers, fill = '#9ecae1')+
+ggplot() +
+  geom_sf(data = mali_adm0) +
+  geom_sf_text(data = mali_adm0, aes(label = admin)) +
+  geom_sf(data = rivers, fill = '#9ecae1') +
   geom_sf(data = osm_lakes.sf)
@@
 #Mosaic/merge srtm tiles
-srtmmosaic <- mosaic(srtm1, srtm2, srtm3, srtm4, srtm5, srtm6, srtm7, srtm8, srtm9, srtm10, fun=mean)
+srtmmosaic <- mosaic(srtm1, srtm2, srtm3, srtm4, srtm5, srtm6, srtm7, srtm8, srtm9, srtm10, fun = mean)
 #crop to Mali
@@
 dem.df <- data.frame(dem.df)
-colnames(dem.df) = c("lon", "lat", "alt")
+colnames(dem.df) <- c("lon", "lat", "alt")
 ```
@@
 # extract slope and aspect
-slope.raster <- terrain(mali_srtm_a, opt='slope')
-aspect.raster <- terrain(mali_srtm_a, opt='aspect')
+slope.raster <- terrain(mali_srtm_a, opt = 'slope')
+aspect.raster <- terrain(mali_srtm_a, opt = 'aspect')
 # calculate hill shade
@@
 ```{r}
-ggplot()+
-  #geom_sf(data = mali_adm0)+
+ggplot() +
   geom_raster(data = hill.df, aes(lon, lat, fill = hill)) +
   scale_fill_gradientn(colours = grey.colors(100)) +
-  geom_raster(data = dem.df, aes(lon, lat, fill = alt), alpha = 0.8)+
-  scale_fill_gradientn(colours = terrain.colors(100))+
-  geom_sf_text(data = mali_adm0, aes(label = admin))+
-  geom_sf(data = rivers, colour = "blue")+
+  geom_raster(data = dem.df, aes(lon, lat, fill = alt), alpha = 0.8) +
+  scale_fill_gradientn(colours = terrain.colors(100)) +
+  geom_sf_text(data = mali_adm0, aes(label = admin)) +
+  geom_sf(data = rivers, colour = "blue") +
   theme_minimal()

### 2023-08-03 analysis/R/edgelist functions.R
knit

@@ -5,14 +5,15 @@ library(dplyr)
 
 # 1.function to make a percentage table from presence/absence data
 
-make_presabs_prop <- function(data, col_node) {
+make_presabs_prop <- function(data, col_node){
+
   col_node <- enquo(col_node)
 
   data %>%
     tidyr::pivot_longer(!(!!col_node), names_to = "type", values_to = "count") %>%
     group_by((!!col_node), type) %>%
     count(count) %>%
-    mutate(freq = n / sum(n) * 100) %>%
+    mutate(freq = n / sum(n) *100) %>%
     filter(count == 1) %>%
     select((!!col_node), type, freq) %>%
     tidyr::pivot_wider(names_from = type, values_from = freq) %>%
@@ -23,17 +24,18 @@ make_presabs_prop <- function(data, col_node) {
 
 # 2.Make a proportions table from raw counts, to feed to the edgelist function:
 
-make_count_prop <- function(x) {
+make_count_prop <- function (x){
+
   x_gather <- x %>%
     tidyr::gather(key = "type", value = "type_count", -Site)
 
   x_sum <- x_gather %>%
-    group_by(Site) %>%
-    summarise(Site_sum = sum(type_count))
+   group_by(Site) %>%
+   summarise(Site_sum = sum(type_count))
 
   x_prop <- x_gather %>%
     left_join(x_sum, by = "Site") %>%
-    mutate(prop = type_count / Site_sum * 100) %>%
+    mutate(prop = type_count / Site_sum *100) %>%
     select(-type_count, -Site_sum) %>%
     mutate(type = paste0(type)) %>%
     tidyr::spread(key = type, value = prop)
@@ -44,44 +46,34 @@ make_count_prop <- function(x) {
 ###############
 # 3. make edgelist from percentage data listed by site and variable
 
-make_edgelist <- function(data) {
-  data <- data %>% select_if(function(x) {
-    !all(is.na(x))
-  }) # only select those coumns that have values in them
+make_edgelist <- function(data){
+  data <- data%>% select_if(function(x){!all(is.na(x))})  #only select those coumns that have values in them
 
-  data_long <- data %>%
-    tidyr::pivot_longer(-1, names_to = "name") %>%
-    dplyr::rename(Var1 = 1) # make data long
+  data_long <- data %>% tidyr::pivot_longer(-1, names_to = "name") %>% dplyr::rename(Var1 = 1)  #make data long
 
[…]

### 2023-08-03 analysis/R/network functions.R

@@ -177,7 +177,7 @@ test_centralities <- function(graph, centrality, reps) {
     geom_boxplot(aes(sample, value)) +
     labs(y = expression(paste("Spearman's ", rho))) +
     ylim(0, 1) +
-    theme_publish()
+    theme_bw()
 }
 
 
@@ -202,7 +202,7 @@ test_centralities2 <- function(graph, centrality, reps) {
     geom_boxplot(aes(sample, value)) +
     labs(y = expression(paste("mean Spearman's ", rho))) +
     ylim(0, 1) +
-    theme_publish()
+    theme_bw()
 }
 
 

### 2024-03-04 analysis/paper/paper.Rmd
paper merges
Adding the start of diversity analysis

@@
 sites <- read_csv2(here::here("analysis/data/raw_data/sites_all.csv"))
+
 ```
@@
 This network construction is implemented in R (R Core Team 2022) using the tidygraph package (Pedersen 2022).
+## quality of data
+```{r}
+library(iNEXT)
+data("ciliates")
+as.data.frame(decors) |> 
+  column_to_rownames("site") |> 
+  t() |> 
+iNEXT(q=0, datatype="incidence_freq", size=NULL, endpoint=NULL, knots=40, se=TRUE, conf=0.95,
+nboot=50)
+
+```
+```{r}
+ #install_github('AnneChao/iNEXT.4steps') #not currently on CRAN
+ library(iNEXT)
+ library(iNEXT)
+# library(iNEXT.4steps)
+ library(ggplot2)
+ data("ciliates")   #included in package iNEXT
+
+ i <- as.data.frame(decors) |> 
+  column_to_rownames("site") |> 
+   head(3) |> 
+  t() |> 
+  iNEXT(q=0, datatype="abundance", size=NULL, endpoint=NULL, knots=40, se=TRUE, conf=0.95,
+nboot=50)
+
+ ggiNEXT(i, type=1, facet.var="Assemblage")
+
+ i$figure   #automatically generate plots shown below
+ i
+i$summary
+
+i$summary$`STEP2. Asymptotic analysis`
+
+i$summary$`STEP3. Non-asymptotic coverage-based rarefaction and extrapolation analysis`
+
+i$summary$`STEP4. Evenness among species abundances`
+
+i$figure
+i$figure[[1]]
+
+i$figure[[2]]
+
+i$figure[[3]]
+
+i$figure[[4]]
+
+i$figure[[5]]
+
+i$figure[[6]]
+
+# Example data
+data <- read.table(text = "A B C
+                     X 1 0 1
+                     Y 0 1 1
+                     Z 1 1 0", header = TRUE)
+
+# Correspondence Analysis
+library(FactoMineR)
+data <- read.table(text = "A B C
+                     X 1 0 1
+                     Y 0 1 1
+                     Z 1 1 0", header = TRUE)
+ca_result <- CA(data, graph = FALSE)
+
+# Custom plot combining rows and columns
+plot(ca_result$ind$coord, type = "n", xlab = "Dimension 1", ylab = "Dimension 2", main = "CA Rows and Columns")
+
+# Plot rows
+points(ca_result$row$coord, col = "blue", pch = 16)
+
+# Plot columns
+points(ca_result$col$coord, col = "red", pch = 17)
+
+# Add legend
+legend("topright", legend = c("Rows", "Columns"), col = c("blue", "red"), pch = c(16, 17))
+
+```
 

### 2024-03-04 analysis/paper/paper.Rmd
updated data

@@
 
-decors <- read_csv2(here::here("analysis/data/raw_data/decors_late.csv")) |>
-  select(-1)
+decors <- read_csv2(here::here("analysis/data/raw_data/decors12.csv")) |>
+  select(-1) |> 
+  rename(sites = Site) |> 
+  select(-slip, -unidentified_roulette, -"twisted_cord_roulette?")
 
 # import information on sites
-sites <- read_csv2(here::here("analysis/data/raw_data/sites_all.csv"))
+sites <- read_csv2(here::here("analysis/data/raw_data/sites_all3.csv"))
 
@@
   activate(nodes) |>
-  mutate(louvain = as.character(group_louvain()))
+  mutate(louvain = as.character(group_optimal()))
 
@@
 decorsearly <- 
-  read_csv2(here::here("analysis/data/raw_data/decors_early.csv")) |>
+  read_csv2(here::here("analysis/data/raw_data/decors9.csv")) |>
   select(-1)

### 2024-03-04 analysis/paper/paper.Rmd
without twisted

@@
 decor_graph <- decor_graph |>
+  activate(edges) |> 
+  filter(!name == "twisted_cord_roulette") |> 
   activate(nodes) |>
@@
   head(1)
+
+decor_graph
+
 ```
@@
   mutate(label = early_flat_nodes$label,
-         lat = early_flat_nodes$lat, lon = early_flat_nodes$lon)
+         lat = early_flat_nodes$lat, 
+         lon = early_flat_nodes$lon
+         )
 

### 2024-03-04 analysis/paper/paper.Rmd
experimenting with clustering sparsified networks

@@
 library(tidyr)
+library(tidyverse)
 
@@
 
+decor_graph1 <- decor_graph |>
+    activate(edges) |> 
+   filter(str_detect(name, "roulette") | name %in% c("paint", "straight_mat_impression"))  |> 
+  activate(nodes) |>
+  mutate(louvain = as.character(group_louvain()))
+
 # calculate modularity of louvain clustering
@@
 
-decor_graph
-
+decor_graph1 <- decor_18 |>
+  activate(nodes) |>
+  mutate(louvain = as.character(group_louvain()))
+
+decor_graph1 <- decor_graph |>
+  activate(nodes) |>
+  mutate(louvain = as.data.frame(decor_graph1, "vertices")$louvain)
+
+decor_graph1
+
+decor_graph |>
+  activate(edges) |> 
+  filter(str_detect(name, "roulette") | name %in% c("paint", "straight_mat_impression"))  |> 
+  as.data.frame("edges") |> 
+  group_by(name) |> 
+  summarise()
+
+decor_graph |>
+  activate(edges) |> 
+  filter(str_detect(name, "strip_roulette")| str_detect(name, "cord_roulette"))  |> 
+  as.data.frame("edges") |> 
+  group_by(name) |> 
+  summarise()
 ```
@@
   setNames(
-    c("tomato", "cornflowerblue", "darkgreen"),
+    c("tomato", "cornflowerblue", "darkgreen", "yellow"),
     unique(V(decor_graph)$louvain)
@@
 
-ggraph(decor_graph, x = lon, y = lat) +
+ggraph(decor_graph1, x = lon, y = lat) +
   geom_sf(data = mali_adm0, fill = NA) +
@@
 ```{r cluster1, fig.cap="Internal connections within community 1"}
-decor_graph %>%
+decor_graph1 %>%
   activate(nodes) %>%
@@
   scale_edge_width(range = c(0.2, 2)) +
-  geom_node_point(size = 3, colour = "cornflowerblue") +
+  geom_node_point(aes(size = centrality_degree(weights = weight)), colour = "cornflowerblue") +
   geom_node_text(aes(label = label), col = "black", repel = TRUE) +
@@
 ```{r cluster2, fig.cap="Internal connections within community 2"}
-decor_graph %>%
+decor_graph1 %>%
   activate(nodes) %>%
@@
   scale_edge_width(range = c(0.2, 2)) +
-  geom_node_point(size = 3, colour = "tomato") +
+  geom_node_point(aes(size = centrality_degree(weights = weight)), colour = "tomato") +
   geom_node_text(aes(label = label), col = "black", repel = TRUE) +
@@
 ```{r cluster3, fig.cap="Internal connections within community 3"}
-decor_graph %>%
+decor_graph1 %>%
   activate(nodes) %>%
@@
   scale_edge_width(range = c(0.2, 2)) +
-  geom_node_point(size = 3, colour = "darkgreen") +
+  geom_node_point(aes(size = centrality_degree(weights = weight)), colour = "darkgreen") +
+  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
+  theme_graph()
+```
+```{r cluster3, fig.cap="Internal connections within community 3"}
+decor_graph1 %>%
+  activate(nodes) %>%
+  mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 4) %>%
+  ggraph(layout = "graphopt") +
+  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
+  scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+  scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(aes(size = centrality_degree(weights = weight)), colour = "yellow") +
   geom_node_text(aes(label = label), col = "black", repel = TRUE) +
@@
 # adding community/crossing to the edges
-decor_graph <- decor_graph |>
+decor_graph1 <- decor_graph1 |>
   add_community_edges()
@@
 # mean edge weights
-mean_edge_weight <- decor_graph |>
+mean_edge_weight <- decor_graph1 |>
   get.data.frame("edges") |>
@@
 # Centralization of the cluster
-centralization <- vector("double", length(3))
-for (i in 1:3) {
-  centralization[[i]] <- decor_graph |>
+centralization <- vector("double", length(4))
+for (i in 1:4) {
+  centralization[[i]] <- decor_graph1 |>
     # filter community
@@
 g2 <- tibble(centralization) |>
-  mutate(community = 1:3) |>
+  mutate(community = 1:4) |>
   pivot_longer(!community, names_to = "key", values_to = "centralisation") |>
@@
 # Calculate centralities for each cluster
-centr <- pull_centralities_clust(decor_graph)
+centr <- pull_centralities_clust(decor_graph1)
 
@@
 #show crossings with a weight over 3
-decor_graph |>
+decor_graph1 |>
   filter(
@@
 decor_15 <- extract_backbone(decor_flat, 0.15)
-decor_18 <- extract_backbone(decor_flat, 0.18)
+decor_18 <- extract_backbone(decor_flat, 0.20) #20 ideale anzahl?
 

### 2024-03-07 analysis/paper/paper.Rmd
data and data prep + working inext

@@
 
+decorsearly <-  read_csv2(here::here("analysis/data/raw_data/decors9.csv")) |>
+  select(-1) |> 
+  rename(sites = Site) |> 
+  select(-slip, -unidentified_roulette)
+
 # import information on sites
@@
 
+# import decor count per site
+n_decors <- read_csv2(here::here("analysis/data/raw_data/n_decor.csv")) |> 
+  rename(sites = site) |> 
+  select(-1)
 ```
@@
 data("ciliates")
-as.data.frame(decors) |> 
-  column_to_rownames("site") |> 
-  t() |> 
-iNEXT(q=0, datatype="incidence_freq", size=NULL, endpoint=NULL, knots=40, se=TRUE, conf=0.95,
-nboot=50)
-
-```
+data("ant")
+data("bird")
+
+decors_abs <- decors |> 
+  left_join(n_decors) |> 
+    mutate(across(where(is.numeric), ~ .*n/100)) |> 
+  mutate(across(where(is.numeric), ~ round(.))) |> 
+    column_to_rownames("sites") |> 
+  select(-n) |> 
+t()
+
+iNEXT(decors_abs, q=0, datatype="abundance")
+
+```
+
 ```{r}
@@
 
- i <- as.data.frame(decors) |> 
-  column_to_rownames("site") |> 
-   head(3) |> 
-  t() |> 
-  iNEXT(q=0, datatype="abundance", size=NULL, endpoint=NULL, knots=40, se=TRUE, conf=0.95,
-nboot=50)
+ i <- decors_abs[,c(2,5,8)]|> 
+ iNEXT(q=c(0,1,2), datatype="abundance")
 
  ggiNEXT(i, type=1, facet.var="Assemblage")
-
- i$figure   #automatically generate plots shown below
- i
-i$summary
-
-i$summary$`STEP2. Asymptotic analysis`
-
-i$summary$`STEP3. Non-asymptotic coverage-based rarefaction and extrapolation analysis`
-
-i$summary$`STEP4. Evenness among species abundances`
-
-i$figure
-i$figure[[1]]
-
-i$figure[[2]]
-
-i$figure[[3]]
-
-i$figure[[4]]
-
-i$figure[[5]]
-
-i$figure[[6]]
-
-# Example data
-data <- read.table(text = "A B C
-                     X 1 0 1
-                     Y 0 1 1
-                     Z 1 1 0", header = TRUE)
-
-# Correspondence Analysis
-library(FactoMineR)
-data <- read.table(text = "A B C
-                     X 1 0 1
-                     Y 0 1 1
-                     Z 1 1 0", header = TRUE)
-ca_result <- CA(data, graph = FALSE)
-
-# Custom plot combining rows and columns
-plot(ca_result$ind$coord, type = "n", xlab = "Dimension 1", ylab = "Dimension 2", main = "CA Rows and Columns")
-
-# Plot rows
-points(ca_result$row$coord, col = "blue", pch = 16)
-
-# Plot columns
-points(ca_result$col$coord, col = "red", pch = 17)
-
-# Add legend
-legend("topright", legend = c("Rows", "Columns"), col = c("blue", "red"), pch = c(16, 17))
+ggiNEXT(i, type=3, facet.var="Order.q", color.var="Assemblage")
+
+ i <- decors_abs[,c(2)]|> 
+ iNEXT(q=c(0,1,2), datatype="abundance")
+
+ ggiNEXT(i, type=1, facet.var="Assemblage")
+ggiNEXT(i, type=3, facet.var="Order.q", color.var="Assemblage")
 
@@
 ```{r community detection, hide = 'results', echo=FALSE}
-# use louvain clustering method to assign clusters to nodes
-decor_graph <- decor_graph |>
-  activate(edges) |> 
-  filter(!name == "twisted_cord_roulette") |> 
+decor_graph1 <- decor_flat |>
   activate(nodes) |>
@@
 
-decor_graph1 <- decor_graph |>
-    activate(edges) |> 
-   filter(str_detect(name, "roulette") | name %in% c("paint", "straight_mat_impression"))  |> 
+decor_graph1 <- decor_18 |>
   activate(nodes) |>
@@
 
+decor_graph1 <- decor_graph |>
+  activate(nodes) |>
+  mutate(louvain = as.data.frame(decor_graph1, "vertices")$louvain)
+
 # calculate modularity of louvain clustering
-modularity <- decor_graph |>
+modularity <- decor_graph1 |>
   activate(nodes) |>
@@
   head(1)
-
-decor_graph1 <- decor_18 |>
-  activate(nodes) |>
-  mutate(louvain = as.character(group_louvain()))
-
-decor_graph1 <- decor_graph |>
-  activate(nodes) |>
-  mutate(louvain = as.data.frame(decor_graph1, "vertices")$louvain)
 
@@
   summarise()
+
+decor_flat
 ```
@@
     c("tomato", "cornflowerblue", "darkgreen", "yellow"),
-    unique(V(decor_graph)$louvain)
+    unique(V(decor_graph1)$louvain)
   )
@@
   theme_graph()
+
 ```
@@
   theme_graph()
+
+decor_18 %>%
+  ggraph(layout = "graphopt") +
+  geom_edge_fan(aes(edge_width = weight, edge_alpha = weight)) +
+  scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+  scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(aes(size = centrality_degree(weights = weight)), colour = "cornflowerblue") +
+  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
+  theme_graph()
+
+decor_flat %>%
+  ggraph(layout = "graphopt") +
+  geom_edge_fan(aes(edge_width = weight, edge_alpha = weight)) +
+  scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+  scale_edge_width(range = c(0.2, 2)) +
+  geom_node_point(aes(size = centrality_degree(weights = weight)), colour = "cornflowerblue") +
+  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
+  theme_graph()
 ```
@@
 decor_graph1 <- decor_graph1 |>
+ # flatten(weighted = TRUE) |> 
   add_community_edges()
+
+decor_graph1
 
@@
   get.data.frame("edges") |>
-  filter(community != "crossing") |>
-  group_by(community) |>
-  summarise(mean_weight = mean(weight))
+#  filter(community != "crossing") |>
+  group_by(from,to, community) |> 
+  summarise(weight = sum(weight),
+  div = n()) #|> 
+ # group_by(community) |>
+ # summarise(median_weight = median(weight),
+  #          mean_weight = mean(weight))
 
@@
 g1 + g2
+
+ggplot(mean_edge_weight) +
+  geom_boxplot(aes(x = community, y = weight, fill = community)) +
+  scale_fill_grey(start = 0.5, end  = 1) +
+  theme_bw() +
+  theme(legend.position = "none") +
+  labs(y = "degree centrality",
+       x = "community")
+
+ggplot(mean_edge_weight) +
+  geom_boxplot(aes(x = community, y = div, fill = community)) +
+  scale_fill_grey(start = 0.5, end  = 1) +
+  theme_bw() +
+  theme(legend.position = "none") +
+  labs(y = "degree centrality",
+       x = "community")
+
+ggplot(centrali) +
+  boxplot(aes(x = community, y = div, fill = community)) +
+  scale_fill_grey(start = 0.5, end  = 1) +
+  theme_bw() +
+  theme(legend.position = "none") +
+  labs(y = "degree centrality",
+       x = "community")
 ```
@@
   filter(
-    community == "crossing",
+    community == "crossing" &
     weight > 3 
+  # & !name %in% c("channel", "twisted_cord_roulette")
   ) |>
@@
 ```{r flatten}
-decor_flat <- flatten(decor_graph, weighted = TRUE)
+decor_flat <- flatten(decor_graph, weighted = TRUE) |> 
+activate(nodes)
+
 ```
@@
   filter(!node_is_isolated())
+
 ```
@@
 early_flat_bb <- 
-  backbone::disparity(as.igraph(early_flat), alpha = 0.25, narrative = TRUE) |>
+  backbone::disparity(as.igraph(early_flat), alpha = 0.22, narrative = TRUE) |>
   tidygraph::as_tbl_graph()

### 2024-03-07 analysis/supplementary_materials/data preparation.qmd
csv and dataprep

---
title: "Data preparation"
author: "Nikolas Gestrich and Juan Marco Puerta Schardt"
format: html
editor: visual
---

```{r libraries}
library(dplyr)
library(readxl)
library(readr)
library(tibble)
library(tidyr)
library(stringr)
```

## Jenné-jeno

For the 12^th^ century network, Jenne -jeno LXN levels 15-24 were used, which date to the period AD 1030-1200. For the 9^th^ century network, LXN levels 29-40 were used which date to the late 9^th^ early 10^th^ centuries (S. K. McIntosh pers.comm.) The raw sherd data were kindly given to us by Susan Keech McIntosh.

```{r jenne-jeno data prep 1}
# read data
Jj <- read_xlsx(here::here("analysis/data/raw_data/Jenne-jeno 1981 rim sherd data.xlsx"), sheet = 2)

# subset 12th century contexts
Jj12 <- Jj |> 
  filter(PIT == 1,
         Level > 14,
         Level < 25)

# subset 9th century contexts
Jj9 <- Jj |> 
  filter(PIT == 1,
         Level > 28,
         Level < 41)
```

Decors are coded separately in three columns, Twine, Paint, and Plastic. The slip coded in the surface columns is left our due to uneven reporting at other sites. We now prepare percentage tables for each of these, standardise their naming, and then merge them together.

```{r jenne-jeno data prep 2}
# Twine decors
Jj12twine <- Jj12 |> 
  count(Twine) |> 
  mutate(percent = n/143 * 100) |> 
  select(!n) |> 
  pivot_wider(names_from = Twine, values_from = percent) |> 
  rename_with( ~ paste0("tw", .x))

# standardise nomenclature
Jj12twine <- Jj12twine |> 
  mutate(braided_cord_roulette = tw1 + tw2 + tw3, # Twine patterns 1, 2, and 3 are all made by rouletting a braided cord
         folded_strip_roulette = tw4 + tw5, # Twines 4 and 5 are folded strip roulette patterns
         twisted_cord_roulette = tw6 + tw7) |>  # twines 6 and 7 are made using a twisted cord roulette
  select(ends_with("roulette"))

# the code 40 in twine is unclear, and thus left out

# Paint decors
Jj12paint <- Jj12 |> 
  count(Paint) |> 
  mutate(percent = n/143 * 100) |> 
  select(!n) |> 
  pivot_wider(names_from = Paint, values_from = percent) |> 
  rename_with( ~ paste0("x", .x))

Jj12paint <- Jj12paint |>  
  select(-x0)

Jj12paint <- Jj12paint |>    
  mutate(paint = rowSums(Jj12paint)) |> # summing all paint decors regardless of colour
  select(-starts_with("x"))

# Plastic decors
Jj12plastic <- Jj12 |> 
  count(Plastic) |> 
  mutate(percent = n/143 * 100) |> 
  select(!n) |> 
  pivot_wider(names_from = Plastic, values_from = percent) |> 
  rename_with( ~ paste0("x", .x))

Jj12plastic <- Jj12plastic |> 
  mutate(channel = x1+x2, #single and multiple grooves are treated as one category
         nonparallel_grooves = x3,
         impressed_comb = x4+x46+x47,
         finger_arcs = x13,
         dragged_comb = x24+x29+x93,
         punctate = x27, 
         applied_plastic = x91) |> 
  select(-starts_with("x"))

 Jj12decor <- bind_cols(Jj12paint, Jj12plastic, Jj12twine)  
 rownames(Jj12decor) <- "Jenne-jeno12"
```

Same procedure for the 9th century contexts

```{r jenne-jeno data prep 3}
# Twine decors
Jj9twine <- Jj9 |> 
  count(Twine) |> 
  mutate(percent = n/360 * 100) |> 
  select(!n) |> 
  pivot_wider(names_from = Twine, values_from = percent) |> 
  rename_with( ~ paste0("tw", .x))

# standardise nomenclature
Jj9twine <- Jj9twine |> 
  mutate(braided_cord_roulette = tw1 + tw2 + tw3 + tw10, # Twine patterns 1, 2, 3, and 10 are all made by rouletting a braided cord
         folded_strip_roulette = tw4 + tw5, # Twines 4 and 5 are folded strip roulette patterns
         twisted_cord_roulette = tw6 + tw7 + tw12, # twines 6 and 7 are made using a twisted cord roulette
         cord_wrapped_stick_roulette = tw14) |> 
  select(ends_with("roulette"))

# the code 40 in twine is unclear, and thus left out

# Paint decors
Jj9paint <- Jj9 |> 
  count(Paint) |> 
  mutate(percent = n/360 * 100) |> 
  select(!n) |> 
  pivot_wider(names_from = Paint, values_from = percent) |> 
  rename_with( ~ paste0("x", .x))

Jj9paint <- Jj9paint |> 
  select(-x0)

Jj9paint <- Jj9paint |> 
  mutate(paint = rowSums(Jj9paint)) |> # summing all paint decors regardless of colour
  select(-starts_with("x"))

# Plastic decors
Jj9plastic <- Jj9 |> 
  count(Plastic) |> 
  mutate(percent = n/360 * 100) |> 
  select(!n) |> 
  pivot_wider(names_from = Plastic, values_from = percent) |> 
  rename_with( ~ paste0("x", .x))

Jj9plastic <- Jj9plastic |> 
  mutate(channel = x1+x2+x18, #single and multiple grooves are treated as one category
         nonparallel_grooves = x3,
         finger_arcs = x13, 
         applied_plastic = x91,
         applied_ridges = x11,
         incised = x9+x12+x39,
         fingernail_impression = x18) |> 
  select(-starts_with("x"))

 Jj9decor <- bind_cols(Jj9paint, Jj9plastic, Jj9twine)  
 rownames(Jj9decor) <- "Jenne-jeno9"

```

Creating a dataframe with the number of decorations in Jenne-Jeno

```{r}
n_jj <- data.frame(
  site = c("Jenne-jeno9","Jenne-jeno12"),
  n = c(360,143)
)
```

## Togu Missiri

The data for the 12th century network includes rim and body sherds from two site phases, C and D (contexts 12 - 23), which comprised of several midden and two floor level deposits. They are dated with two radiocarbon dates to span a period from 1039 to 1222 cal AD.

```{r Togu data prep 1}
togu <- read_csv2(here::here("analysis/data/raw_data/TOG2 Rim sherds.csv"))

 # subset 12th century contexts
togu12 <- togu |>
  mutate(Context = as.numeric(Context)) |> 
  filter(Context > 11,
         Context < 24)
```

Decors are coded in four columns, beginning with "Motif". These are counted and their percentages of the assemblage calculated. Their naming is then standardised.

```{r Togu data prep 2}
# calculate percentages
togu_decor <- c(togu12$Motif1, togu12$Motif2, togu12$Motif3, togu12$Motif4) |> 
  as_tibble() |> 
  count(value) |> 
  mutate(percent = (n/148)*100) |> 
  select(-n) |> 
  pivot_wider(names_from = value, values_from = percent) |> 
  rename_with( ~ paste0("x", .x))

# standardise naming
togu_decor <- togu_decor |> 
  mutate(
    channel = xCH+xMCH,
    braided_cord_roulette = xCR2,
    folded_strip_roulette = xCR4+xCR4G,
    twisted_cord_roulette = xCR6,
    paint = `xP-R` + `xP-R-L` + `xP-R-L-N` + `xP-R-W` + `xP-W-L`,
    applied_plastic = xPA,
    applied_ridges = xPA3,
    fish_spine_roulette = xVPR
  ) |> 
  select(-starts_with("x"))

rownames(togu_decor) <- "Togu"

#number of decors in togu missiri
n_togu <- data.frame(
  site = c("Togu"),
  n = c(148)
)
```

## **Marakaduguba**

Finds used here come from an unpublished sequence in Unit C, excavated by Mamadou Cissé, Bourahima Ouedraogo and Nafogo Coulibaly. Levels 38, 39, and 40 are used, which consist of construction materials as well as a midden deposit. The levels are dated with two radiocarbon dates (from 40 and 38) to between 770 and 900 cal CE.

```{r MDB data prep 1}
mdb <- read_csv(here::here("analysis/data/raw_data/MDB_rims.csv"))

 # subset 9th century contexts
mdb9 <- mdb |>
  mutate(Context = as.numeric(Context)) |> 
  filter(Context %in% c(38,39,40))
```

Decors are coded in five columns, beginning with "Motif". These are counted and their percentages of the assemblage calculated. Their naming is then standardised.

```{r MDB data prep 2}
# calculate percentages 
mdb_decor <- c(mdb9$Motif1, mdb9$Motif2, mdb9$Motif3, mdb9$Motif4, mdb9$Motif5) |>    
  as_tibble() |>    
  count(value) |>    
  mutate(percent = (n/59)*100) |>    
  select(-n) |>    
  pivot_wider(names_from = value, values_from = percent) |>
  rename_with( ~ paste0("x", .x))

# standardise naming 
mdb_decor <- mdb_decor |>    
  mutate(     
    channel = xCH+xMCH,     
    folded_strip_roulette = xCR4,     
    twisted_cord_roulette = xCR6,     
    paint = `xPR`,     
    incised= xSI2+xSI4,     
    fish_spine_roulette = xVPR   
    ) |>   
  select(-starts_with("x"))

rownames(mdb_decor) <- "Marakaduguba"

#number of decorations in Marakaduguba
n_mdb <- data.frame(
  site = c("Marakaduguba"),
  n = c(59)
)
```

## Tongo Maaré Diabal

Sherds used in the 12^th^ century network stem from the upper levels of Horizon 5, which is dated to 1000-1150. By using only the uppermost layers of this Phase, we attempt to eliminate most of the 11^th^ century material from our sample. For the 9^th^ century network, sherds from Horizon 3 upper (750-900) are used.

```{r TMD data prep 1}
tmd <- read.csv2(here::here("analysis/data/raw_data/TMD93-96 decor.csv"))

# subset 9th and 12th century contexts
tmd12 <- tmd |> 
  filter(Phase == "5 up")

tmd9 <- tmd |> 
  filter(Phase == "3 up")
```

As at Togu Missiri and Duguba, the decorative elements are coded in columns prefixed with "Motif". These are extracted, percentages calculated, and names standardised.

```{r TMD data prep 2}
# 12th c.
# calculate percentages 
tmd12_decor <- c(tmd12$Motif1, tmd12$Motif2, tmd12$Motif3, tmd12$Motif4)|>    
  as_tibble() |> 
  mutate(value = if_else(value=="", NA_character_ ,value))|>    
  count(value) |>    
  mutate(percent = (n/143)*100) |>    
  select(-n) |>    
  pivot_wider(names_from = value, values_from = percent) |>
  rename_with( ~ paste0("x", .x))

# standardise naming 
tmd12_decor <- tmd12_decor |>    
  mutate(  
    applied_ridges = xAP3,
    braided_cord_roulette = xCCRY+xTW1,
    channel = xCH+xMCH, 
    cord_wrapped_stick_roulette = xCSR1+xCSR7,
    dragged_comb = xDC1+xDC2,
    dragged_organics = xGRASS,
    straight_mat_impression = xICS1+xICS4+xICS7,
    folded_strip_roulette = xTW4,     
    twisted_cord_roulette = xTW6+xTW12,     
    paint = `xP`+ `xP B`+`xP B-G`+`xP B-W`+`xP R`+`xP R-W`+`xP W`+   `xP W+B`+ `xP W-G` + `xP W-W`,     
    incised= xSI1+xSI3   
    ) |>   
  select(-starts_with("x"))

rownames(tmd12_decor) <- "TMD12"

# 9th century

# calculate percentages 
tmd9_decor <- c(tmd9$Motif1, tmd9$Motif2, tmd9$Motif3, tmd9$Motif4)|>    
  as_tibble() |> 
  mutate(value = if_else(value=="", NA_character_ ,value))|>    
  count(value) |>    
  mutate(percent = (n/73)*100) |>    
  select(-n) |>    
  pivot_wider(names_from = value, values_from = percent) |>
  rename_with( ~ paste0("x", .x))

# standardise naming 
tmd9_decor <- tmd9_decor |>    
  mutate(  
    braided_cord_roulette = xCCRX+xCCRY,
    channel = xCH+xMCH, 
    cord_wrapped_stick_roulette = xCSR1+xCSR7+xCSR8+xTW14,
    dragged_comb = xDC2,
    dragged_organics = xGRASS,
    straight_mat_impression = xICS1+`xPD/O`,
    twisted_cord_impression = xTI1,
    folded_strip_roulette = xTW4,     
    twisted_cord_roulette = xTW6+xTW12,     
    paint = `xP B`+`xP B-G`+`xP B-W`+`xP R`+`xP R-W`+`xP W`+`xP W-G` + `xP W-W`  
    ) |>   
  select(-starts_with("x"))

rownames(tmd9_decor) <- "TMD9"
```

```{r}
n_tmd <- data.frame(
  site = c("TMD9","TMD12"),
  n = c(73,143)
)
```

## Dia Mara

For the 12th century network, we have, by necessity, used body sherd data for Unit D6 Levels 6-8 (Haskell et al. 1988 Table 15 and approximated data from Figure 36), which McIntosh and McIntosh have recently reinterpreted as mostly falling into the 11th century (McIntosh et al. 2023, 21) with new AMS dates from levels 5 and 8 giving a combined span of 1026-1149 cal CE (McIntosh et al. 2023, 19, Table 7)

For the 9th century network, we have used the Horizon III data reported in Arazi (2005) Table 6.34 and 6.35. This is an agglomerate of rim and body sherd decors from Units M, S and Q. This is somewhat problematic as the Horizon III deposits in Unit M are undated, whereas those in Units Q and S are dated by a single date each (Bedaux et al. 2005; Arazi 2005). The date range for both is from the mid-7th to the mid-10th centuries (Q: GX-28750, 1250 +- 60; S: GX-30007-AMS, 1210+-40). Despite the fact that the main probablity ranges for both dates are in the 9th century, considerable insecurity over the stratigraphic and chronological integrity of this sample must remain.

```{r Mara data prep}
# 12th century
mara_12 <- read_csv2(here::here("analysis/data/raw_data/Mara 12.csv"))

n_mara12 <- mara_12 |> 
  filter(level == "total") |> 
  pull(n_total)

mara_12 <- mara_12|> 
  filter(level == "percent") |> 
  select(-c(n_total, level))

rownames(mara_12) <- "Mara12"

#renaming comb_impression
mara_12 <- mara_12 |> 
  rename(impressed_comb = comb_impression)
# 9th century
mara_9 <- read_csv2(here::here("analysis/data/raw_data/Mara 9.csv"))

#renaming comb_impression
mara_9 <- mara_9 |> 
  rename(impressed_comb = comb_impression,
         incised = incision)

rownames(mara_9) <- "Mara9"

#n of decorations at mara (number of Mara taken from Bedaux et al. 2005, 231)
n_mara <- data.frame(
  site = c("Mara9","Mara12"),
  n = c(281,512)
)
```

## ***Dia Shoma***

For the 12th century network, we use body sherd data from Unit Sh7 levels 6-9 (Haskell et al. 1988, Table 11 and Figure 34) which are dated by an AMS date from level 7 spanning from 1040-1165 cal CE (McIntosh et al. 2023, 19, Table 7). The only other date from occupation layers to fall into the 12th century range is from Unit C (1025-1159 cal CE.), but the corresponding ceramic data could not be isolated from the data published by Bedaux et al. or Arazi. The ceramics from these particular levels in Sh7 is discussed by McIntosh and McIntosh (2023) as being unusual and unknown in the remaining sequence at Dia, and as having clear affinities with the eastern IND. In including this material in the network, we should be aware of the fact that it might not well represent "typical" western IND pottery in this period.

For the 9th century, we did not feel comfortable using Horizon III ceramics, as reported in Bedaux et al. (2005). While most Horizon III dates from Shoma stem from burials, a further indication of the problematic nature of deposits from this period, Units F and B have dated Horizon III occupation layers whose dates span the entire second half of the first millennium CE, making the bracket too wide for meaningful comparison.

```{r Shoma data prep}
# 12th century
shoma_12 <- read_csv2(here::here("analysis/data/raw_data/Shoma 12.csv"))

n_Shoma12 <- shoma_12 |> 
    filter(level == "total") |> 
  rename(n = n_total) |> 
mutate(site = "Shoma12")|> 
select(site, n)

shoma_12 <- shoma_12|> 
  filter(level == "percent") |> 
  select(-c(n_total, level))

rownames(shoma_12) <- "Shoma12"
```

## **Essouk-Tadmekka**

The handmade and unglazed pottery data from Unit Ek-A Periods 3 a and 3b were used in the 12th century network \[\@nixonEssoukTadmekkaEarlyIslamic2017 Appendix E:21\]. It is not entirely certain whether this data refers to rims sherds only, or to rim and body sherds. The overall numbers of studied sherds in table E.21 does not correspond to any of the counts given in Table 12.2, and is somewhat higher than the numbers given for the total number of decorated sherds indicated there. It is therefore unclear what the sample represents. We assume that this an effect of multiple decors on some sherds and have taken the total number of decorated rim and body sherds as a basis for percentage calculation.

Despite the fact that Nixon (2017, 97) includes the 12th and 13th century in the dating for these phases, the radiocarbon dates for the deposits would rather indicate the 11th and 12th centuries as the date range. While only the basal layers of the preceding Period 2 are radiocarbon dated, the beginning of Period 3a has a *terminus post quem* of the 11th century from a sherd of Chinese Qingbai pottery \[\@nixonEssoukTadmekkaEarlyIslamic2017, p. 99-100\]. It therefore seems likely that the bulk of the deposits in these phases stem from the 12th century AD, with some possible admixture from the late 11th and early 13th centuries.

While there are 9th century deposits at Essouk-Tadmekka, they are either too insecurely dated (as in Unit Ek-B), or as in Unit Ek-A Periods 1 and 2, the pottery data presentation does not allow sufficient differentiation of the 9th century deposits from earlier and later ones.

```{r Essouk data prep}
essouk <- read_csv2(here::here("analysis/data/raw_data/Essouk Phase 3 pottery.csv")) |> 
  select(n,`decor description`) |> 
  mutate(percent = (n/450)*100) |> 
  select(-n) |> 
  pivot_wider(names_from = `decor description`, 
              values_from = percent, 
              values_fn = sum)|> 
  rename_with( ~ paste0("x", .x)) |> 
  mutate(across(everything(), as.numeric))

# standardise names
essouk <- essouk |>
  mutate(
    channel = `xsingle channel`+`xmultiple channels`,
    paint = `xpaint white cross-hatch`+
      `xpaint white line`+
      `xpaint whitetriangle`+
      `xpaint white wash`+
      `xpaint black line`+
      `xpaint black wash`+
      `xpaint red line`+
      `xpaint-red wash`,
    finger_impression = `xfinger impressions`+
      `xfingernail impression`, 
    dragged_comb = `xdragged comb linear` + 
      `xdragged comb geometric`,
    cord_wrapped_impression = `ximpressed cord-wrapped` +
      `ximpressed cord-wrapped stick`,
    incised_patterns = `xincised dashed lines`+
      `xincised geometric`+
      `xpunctate circular` +
      `xsideways stylus impression`
  ) |> 
  rename(
    twisted_cord_roulette = `xtwisted cord roulette`, 
    cord_wrapped_roulette = `xcord-wrapped roulette`
  ) |> 
  select(-starts_with("x"))

rownames(essouk) <- "Essouk"

#n of decorations in Essouk
n_ess <- data.frame(
  site = c("Essouk"),
  n = c(450)
)
```

## Sanga

While the cave sites studied by Bedaux were only sporadically dated, some contexts are secure enough to be useful for our analysis. This is in particular Grotte C, where very little disturbance was noted by the excavators, and where three dates on human bones in a collective burial of 170 individuals in the excavated 4,5 m2 have their main probability peaks in the late 11^th^ to late 13^th^ centuries. The two sigma range, however, extends from 979 to 1394. Given the nature of the site and the estimated total number of up to 3500 individual skeletons, a long use must be assumed. Grotte C is associated by the excavators with Grotte D below it, which are assumed to have been used at the same time.

For the 12^th^ century network, we have used ceramics from Grotte D recorded at the Rijksmuseum voor Volkenkunde in Leiden in 2012. We include this data despite its time range and problematic dating because the non-settlement nature of the site provides a counterpoint to the other assemblages. Due caution in the interpretation is necessary.

```{r tellem data prep}
tellem <- read_csv2(here::here("analysis/data/raw_data/LEIDEN_TELLEM.csv")) |> 
  filter(str_detect(Site, "^D"))

# calculate percentages
tellem_decors <- tellem |> 
  select(starts_with("Decor")) |> 
  pivot_longer(cols = everything()) |> 
  count(value) |> 
  mutate(percent = (n/57)*100) |> 
  select(-n) |> 
  pivot_wider(names_from = value, values_from = percent) |> 
  rename_with( ~ paste0("x", .x))

# standardise decors
tellem_decors <- tellem_decors |> 
  mutate(
    channel = xCH + xMCH,
    knotted_cord_roulette = xCR1,
    braided_cord_roulette = xCR2,
    twisted_cord_roulette = xCR6 + `xB4,CR6`,
    dragged_organics = xHERB,
    applied_ridges = xPA4,
    dragged_comb = xPE1,
    cord_wrapped_roulette = xPFR1 + xPFR2,
    incised_patterns = xSI4,
    braided_strip_roulette = xBSR
  ) |> 
  select(-starts_with("x"))

rownames(tellem_decors) <- "Sanga D"

#n of decorations in Essouk
n_sanga <- data.frame(
  site = c("Sanga"),
  n = c(57)
)
```

## Toguéré Doupwil

Sherds from section C, period k are used in the 12^th^ century network. The date GrN-7944 (1046-1275 cal CE) dates the lower layer 3 in section C. This layer consists of laminated sands, charcoal and pottery directly overlying the first evidence of building at the site in section C layer 2. This is summarised as period k in section C, which, also contains material from the underlying level 1, a thin occupational layer directly on floodplain material. While it is unlikely that level 2 is much older than the C14 date indicates, the relationship to level 1 is unclear. It is therefore possible that this sample contains material that significantly predates the 12th century.

In section B, Period e also includes the second half of the 12th century, as dated by GrN-7943 (1165-1283 cal CE), but the main peaks of this date fall in the 13th century and the sherds have therefore been excluded from the sample.

```{r doupwil data prep}
doupwil <- read_csv(here::here("analysis/data/raw_data/Toguere_Doupwil.csv"))

rownames(doupwil) <- "Doupwil"

n_doupwil <- data.frame(
  site = c("Doupwil"),
  n = c(33)
)
```

## Toguéré Galia

The12^th^ century network includes pottery from Period II. This period, comprising of deposit groups C and D, is dated with a single C14 date to 1026--1217 cal CE (GrN-8555). This date was originally rejected by the excavators as being somewhat too late, since the uncalibrated BP range did not overlap with the underlying date for period I (GrN-7945). The latter date's calibrated range of 1038--1383 cal CE, however, removes this objection. Nevertheless, it must be cautioned that the sequence at this site is not securely dated.

```{r galia data prep}
#changing names to fit dataset
galia <- read_csv(here::here("analysis/data/raw_data/Toguere Galia.csv")) |> 
  rename(dragged_comb = drageed_comb,
         incised_patterns = incised_pattern)

rownames(galia) <- "Galia"

n_galia <- data.frame(
  site = c("Galia"),
  n = c(99)
)
```

## Akumbu

Stratigraphically detailed pottery décor data is available for the body sherds from units AK1 (Togola 2008, pp. 55–57) and AK5 (McIntosh 2017). The body sherd decor from AKB4 is reported in Horizons (Cissé et al. 2017, pp. 41–43). Décor data for rim sherds are not published for any of the excavations.

For the 12th century network, we have used selected data from Akumbu Mound A, Unit AK1’s Middle Assemblage layers. In this unit, a living surface dated to 1266-1425 cal. CE (Beta-40441, 640 +- 70) was constructed above a refuse pit cut into mixed deposits, which the excavator believes to have accumulated slowly. The upper portion of the pit was dated to 1175-1329 cal. CE at 82.6% (Beta-40442, 750+-70). A further, undated, construction episode lay between these deposits and a curvilinear structure with a date (Beta-40443, 980+-70) of 954-1219 cal. CE at 92.7%. While the calibrated date range is rather wide, it was decided to use the pit levels (8,10,13,15) in our analysis. Unfortunately, the pottery data for these layers also includes layer 12, a wall collapse layer with uncertain stratigraphic relationship to the pit, rendering the dating of the sample more insecure. A sample more securely dated to the 12th century might have come from the upper levels of AK4, but the pottery was not recorded.

On Akumbu Mound B, the upper levels in the adjacent Units AK3 and AKB4 date to a range including the 9th century. However, the nearby AK5 has a thin 13th midden deposit overlying 3rd-4th century CE deposits. While the AK3 pottery is not described in detail, AKB4 Horizon 1 has two AMS dates, both from Level 2, which date to 771-976 cal. CE at 98.2% and 771-894 cal. CE at 88.3%. However, the nature of the levels is not made clear in the publication, most of the levels of Horizon 1 lie below this dated deposit, and it is unclear whether Horizon 1 includes the 5^th^-7^th^ century Level 8 or not (cf. Cissé et al. 2017, pp. 36 & 41). Trusting that this is not the case, as is also McIntosh’s (McIntosh 2017, p. 197) interpretation, we have used the Horizon 1 body sherd data for the 9th century network.

There are problems with the pottery data, because paint and incisions are not listed separately even though they are mentioned at being present at the site of AKB4.

```{r}
#downloading the data from Cisse& Takezawa for the 9th century network
Akumbu9_decor <- read_csv2(here::here("analysis/data/raw_data/Akumbu early.csv")) |> rename("braided_cord_roulette" = "Cordelette roulee tressee en natte TW1",
          "folded_strip_roulette" = "Fibre plate pliee TW4",
"twisted_cord_roulette" = "Cordelette Torsadee TW6",
"dragged_comb" = "Peigne filete roule", #translation unsure
"slip" = "Engobe",
"other" = "Autre")

#getting the amount of sherds in Akumbu
n_Aku9 <- Akumbu9_decor |> 
  pull(n)

#removing undefinded sherds and n from table 
Aku9_decor <- Akumbu9_decor |> 
  select(-n, -other)|> 
  mutate(Site = "Akumbu9") |> 
  column_to_rownames("Site")

names(Aku9_decor)

```

```{r}
Akumbu12 <- read_csv2(here::here("analysis/data/raw_data/Akumbu_12.csv"))

#absolute numbers into proportion
Akumbu12_decor <- Akumbu12 |> 
  rename(Site = 1) |> 
column_to_rownames("Site")

  n_Aku12 <- Akumbu12_decor |> 
  select_if(is.numeric) |> 
  rowSums()

Aku12_decor <- Akumbu12_decor |> 
  add_column(n_Aku12) |> 
  mutate_if(is.numeric, function(x) x/n_Aku12*100) |> 
  select(-n_Aku12)

```

```{r}
n_akumbu <- data.frame(
  site = c("Akumbu9", "Akumbu12"),
  n = c(n_Aku9, n_Aku12)
)
```

## Ambere-Dougon

```{r}
AD <- read_csv2(here::here("analysis/data/raw_data/Ambere-Dougon.csv")) |> 
  rename(straight_mat_impression = "Impression de Vannerie",
         simple_braided_strip_roulette = "Impression de fibres plates tresses simple",
         alternate_braided_strip_roulette = "Impressions de Fibre plates tresses alterne",
        carved_wooden_roulette = "Cylindre grave",
        slip = "Engobe",
        twisted_cord_roulette = "Cordelette torsadee") |> 
  select(-Context) |> 
  mutate(Site = "Ambere-Dougon")  |> 
    column_to_rownames("Site") |> 
  mutate_all(as.numeric) |> 
  mutate(braided_strip_roulette = simple_braided_strip_roulette + alternate_braided_strip_roulette) |> 
  select(-slip, -simple_braided_strip_roulette, -alternate_braided_strip_roulette)

n_AD <- AD  |> 
  rowSums()

AD12_decor <- AD |> 
 mutate_if(is.numeric, function(x) x/33*100)

n_AD <- data.frame(
  site = c("Ambere-Dougon"),
  n = c(n_AD)
)

names(AD)
```

## Damassogou

```{r}
Dmssg <- read_csv2(here::here("analysis/data/raw_data/Damassogou.csv")) |> 
  rename(straight_mat_impression = "Impression de Vannerie",
         simple_braided_strip_roulette = "Impression de fibres plates tresses simple",
         alternate_braided_strip_roulette = "Impressions de Fibre plates tresses alterne",
        carved_wooden_roulette = "Cylindre grave",
        undecorated = "non decores",
        braided_cord_roulette = "Cordolette tressee simple",
        finger_impression = "Empreinte digitale",
        incised = "Incision")|> 
  mutate(Site = "Damassogou") |> 
    column_to_rownames("Site") |> 
  mutate_all(as.numeric) |> 
  mutate(braided_strip_roulette = simple_braided_strip_roulette + alternate_braided_strip_roulette) |> 
  select(-undecorated, -simple_braided_strip_roulette, -alternate_braided_strip_roulette)

n_Dmssg <- Dmssg  |> 
  rowSums()

Dmssg12_decor <- Dmssg |> 
 mutate_all(function(x) x/179*100)

names(Dmssg)

#n of pottery decors in Damassagou
n_Dmssg <- data.frame(
  site = c("Damassagou"),
  n = c(n_Dmssg)
)
```

## Gao

```{r}
Gao <- read_csv2(here::here("analysis/data/raw_data/Gao.csv")) |> rename("dragged_comb" = "Comb_tot",
    "channel" = "Channel_comb",
          "folded_strip_roulette" = "Folded strip roulette",
"twisted_cord_roulette" = "Twisted twine",
"paint" = "Paint_total", #translation unsure
"unidentified_roulette" = "Unident. Twine",
"cord_wrapped_roulette" = "Sisal", # Cisse (2011, 161) identifies them as cord_wrapped_roulette after Nixon (2008, 183)
"straight_mat_impression" = "Birdfood" # Cisse (2011, 161) identified it as mat, but not completly sure, maybe other mat than straight
) |> 
mutate_if(is.numeric, round, digits=0) #rounding the numbers (because they were extracted from graphics there were minor deviations)

n_gao <- Gao |> 
  select_if(is.numeric) |> 
  rowSums()

Gao <- Gao |> 
  add_column(n_gao) |> 
  mutate_if(is.numeric, function(x) x/n_gao*100)

Gao9_decor <- Gao |> 
  filter(Phase == "lower") |> 
  select(-"Site-Code", -"Phase", -"n_gao") |> 
  mutate(Site = "Gao9") |> 
  column_to_rownames("Site")

Gao12_decor <- Gao |> 
  filter(Phase == "final") |> 
  select(-"Site-Code", -"Phase", -"n_gao") |> 
  mutate(Site = "Gao12") |> 
  column_to_rownames("Site")

names(Gao)

n_gao <- data.frame(
  site = c("Gao12", "Gao9"),
  n = c(n_gao)
)
```

## Kokolo NE

```{r}
Kokolo <- read_csv2(here::here("analysis/data/raw_data/Kokolo NE.csv")) |> 
  rename(folded_strip_roulette = "fibres plates et pliees",
         twisted_cord_roulette = "cordelette torsadee",
         braided_cord_roulette = "tresse simple", #only as "l’impression à la tresse simple" in Keita 2011, pp. 33
        cord_wrapped_roulette = "roulette de cordelette enroulee")

n_Kokolo <- Kokolo |> 
  pull(n)

Kkl12 <- Kokolo |> 
  mutate(Site = "Kokolo")  |> 
  column_to_rownames("Site") |> 
  select(-n, -Context)

names(Kokolo)

n_kkl <- data.frame(
  site = c("Kokolo"),
  n = c(n_Kokolo)
)
```

## Nin-Bére

```{r}
Nin_Bere <- read_csv2(here::here("analysis/data/raw_data/Nin-Bére.csv")) |> 
  rename(folded_strip_roulette = "Impression de fibres plates plies",
         braided_cord_roulette = "Cordolette tressee simple",
         undefined = "Non determines") |> 
  mutate(Site = "Nin_Bere") |> 
  select(-Context) |> 
  column_to_rownames("Site")

n_NB <- Nin_Bere |> 
  rowSums()

Nin_Bere9 <- Nin_Bere |> 
    mutate_if(is.numeric, function(x) x/n_NB*100) |> 
  select(-undefined)

names(Nin_Bere)

n_NB <- data.frame(
  site = c("Nin-Bere"),
  n = c(n_NB)
)
```

## Promontoire site of Ounjougou

```{r}
Ounjg <- read_csv2(here::here("analysis/data/raw_data/Promontoire Ounjougou.csv"))

#use T3 for the 12th century network and removing columns without values for this level
Ounjg12 <- Ounjg |> 
  filter(Level == "T3") |> 
  mutate(Site = "Ounjougou12") |> 
  column_to_rownames("Site") |> 
  select_if(is.numeric) |> 
    select_if(~all(. > 0)) |> 
  select(-"Non decore")

#renaming decorations
Ounjg12 <- Ounjg12 |> 
  rename(braided_strip_roulette = "Fibre plate tressee simple ou alterne",
         twisted_cord_roulette = "Cordelette torsadee",
         incised = "Incicions au couteau seules",
         cord = "Cordons" #this is described as "presénce d'un petit cordon sous un bord 
         )

#amount of sherds
n_ounjg <- Ounjg12 |> 
  rowSums()

# number of decorations as percentage
Ounjg12 <- Ounjg12 |> 
    mutate_if(is.numeric, function(x) x/n_ounjg*100)

names(Ounjg12)

n_ounjg <- data.frame(
  site = c("Ounjougou12"),
  n = c(n_ounjg)
)
```

## Sadia

```{r}
Sadia <- read_csv2(here::here("analysis/data/raw_data/Sadia.csv")) |> 
  rename(straight_mat_impression = "mat impression",
         folded_strip_roulette = "cr4",
         braided_strip_roulette = "braided.strip",
         braided_cord_roulette = "braided.cord",
        unid_c_r_a = "unidentified composite roulette - A",
         unid_c_r_b = "unidentified composite roulette - B") |> 
  mutate(cord_wrapped_stick_roulette = unid_c_r_a + unid_c_r_b) |> 
  select(-plain, -unid_c_r_a, -unid_c_r_b)

#get total sherd number and percentage
n_sadia <- Sadia |> 
  select_if(is.numeric) |> 
  rowSums()

Sadia <- Sadia |> 
  add_column(n_sadia) |> 
  mutate_if(is.numeric, function(x) x/n_sadia*100)

#Selecting SE1 and SE2 for the 12th century network
Sadia12 <- Sadia |> 
  filter(Phase == "SE1" | Phase == "SE2") |> 
  summarise_if(is.numeric, mean) |> 
  select(-others,-n_sadia) |> 
  mutate(Site = "Sadia12") |> 
  column_to_rownames("Site")

#Selecting SE5 for the 9th century network
Sadia9 <- Sadia |> 
  filter(Phase == "SE5")|> 
  mutate(Site = "Sadia9") |> 
  column_to_rownames("Site")|> 
  select(-others, -Phase,-n_sadia)

#

names(Sadia)

n_sadia <- data.frame(
  site = c("Sadia12", "Sadia12", "Sadia9"),
  n = c(n_sadia[c(1, 2, 5)])
) |> 
  group_by(site) |> 
  summarise(n = sum(n))
```

# Collate datasets

Here we create two separate collated dataset for the 12th and 9th century networks by binding together the previously created data for each site and then further cleaning and standardising.

```{r 12c bind}
decors_12 <- bind_rows(Aku12_decor, AD12_decor, Dmssg12_decor, doupwil, essouk, galia, Gao12_decor, Jj12decor, Kkl12, mara_12, Ounjg12, Sadia12, shoma_12, tmd12_decor, togu_decor, tellem_decors) |> 
 mutate_all(~ifelse(is.na(.), 0, .)) |> 
  mutate(incised = incised + incised_patterns) |> 
  select(-incised_patterns) |> 
  rownames_to_column("Site")

```

```{r 9c bind}
decors_9 <- bind_rows(Aku9_decor, Gao9_decor, Jj9decor, mara_9, Nin_Bere9, Sadia9, tmd9_decor, mdb_decor)|> 
 mutate_all(~ifelse(is.na(.), 0, .))|> 
  rownames_to_column("Site")

```

```{r}
n_decor <- rbind(n_akumbu, n_AD, n_Dmssg, n_doupwil, n_ess, n_galia, n_gao, n_jj, n_kkl, n_mara, n_mdb, n_NB, n_ounjg, n_sadia, n_sanga, n_Shoma12, n_tmd, n_togu)
```

# Saving data

We create two csv datasets to be used in further analyses.

```{r}
#12th century decor data
decors_12 |> 
write.csv2(here::here("analysis/data/derived_data/decors12.csv"))

#9th century decor data
decors_9 |> 
write.csv2(here::here("analysis/data/derived_data/decors9.csv"))

n_decor |> 
  write.csv2(here::here("analysis/data/derived_data/n_decor.csv"))

```

### 2024-03-08 analysis/paper/paper.Rmd
script in proper order, workin iNext, communities based on sparsified networks and better community comparison

@@
 This network construction is implemented in R (R Core Team 2022) using the tidygraph package (Pedersen 2022).
+
 ## quality of data
-```{r}
+```{r iNext overview, echo = FALSE}
 library(iNEXT)
-data("ciliates")
-data("ant")
-data("bird")
-
+
+#getting absolute numbers of decorations
 decors_abs <- decors |> 
@@
  library(ggplot2)
- data("ciliates")   #included in package iNEXT
 
@@
 
+## Sparsification
+
+### Simplified network backbone
+
+While the hyperconnected hairball network has shown us key elements of the structure of pottery décor sharing, it is of less use for more traditional network analyses.
+These include finding key nodes in the network and key routes through it.
+In our case, such network statistics might tell us which of the sites, on the basis of their assemblages, were influential in the propagation of pottery decorations, and through which channels the knowledge of different décors might have preferentially passed.
+In order to extract such measures, the number of connections needs to be reduced.
+Network sparsification or 'backbone extraction', as this process is also called, is a common challenge in network analysis across its different applications (Serrano et al.
+2009; Saha et al.
+2013).
+In archaeological networks, this has frequently been achieved through thresholding, i.e.
+retaining only the strongest connections and discarding the others (as implemented for example in Hart & Engelbrecht 2012).
+There are, however, a number of other recent approaches to network sparsification.
+For our purposes, we have found the best results to come from a process called the "disparity filter" (Serrano et al.
+2009).
+This evaluates the weight of each edge against a null model in which the overall edge weights of a node are evenly distributed among its edges.
+If an edge is found to diverge from the null model at a set probability, it is retained.
+The backbone package (Domagalski et al.
+2021) implements this in R.
+In order to use it, we must first 'flatten' the multiplex graph, by adding the weights of the edges between each pair of nodes on all layers.
+We then extract the backbone of the graph using the disparity filter at different probability thresholds (\@ref(fig:backbone-extraction)).
+
+```{r flatten}
+decor_flat <- flatten(decor_graph, weighted = TRUE) |> 
+activate(nodes)
+
+```
+
+```{r backbone-extraction, fig.cap="Extraction of binary backbones at decreasing significance levels: a) 0.05, b) 0.15, c) 0.18"}
+# extract binary backbones at decreasing levels of significance
+decor_05 <- extract_backbone(decor_flat, 0.05)
+decor_15 <- extract_backbone(decor_flat, 0.15)
+decor_18 <- extract_backbone(decor_flat, 0.20) #20 ideale anzahl?
+
+# plot
+p1 <- plot.bb(decor_05)
+p2 <- plot.bb(decor_15)
+p3 <- plot.bb(decor_18) + geom_node_text(aes(label = label), repel = TRUE)
+
+(p1 | p2) / p3 +
+  plot_annotation(tag_levels = "a") +
+  plot_layout(heights = unit(c(1, 8), c("cm", "null")))
+
+decor_18 <- decor_18 |>
+  #isolated nodes are removed
+  activate(nodes) |>
+  filter(!node_is_isolated())
+
+```
+
+The three communities from the previous section are also visible in the extracted backbone, albeit with some major differences in configuration.
+The spread-out cluster of long-distance contact now also includes the sites at Akumbu and Kawinza which were placed in the IND cluster.
+The latter is now reduced to the sites of the Eastern IND, while the Dia sites take a central position within the network, and the "Dogon" cluster remains largely intact.
+Dammassogou and Kokolo NE 2 are isolated from the network, and remain so until the significance is lowered to 0.2.
+We will use the backbone at alpha = 0.18 to proceed with our analysis.
+If this selection of significance values seems rather arbitrary, we point out that analogous situations exist in other attempts at archaeological network simplifications.
+Hart & Engelbrecht (2012, p.
+339), for instance, threshold their similarity network to the point where all nodes form a network but for a single outlier.
+Higher significance values also do not necessarily mean we get a more reliable network model, especially in a situation where sampling and recovery bias is as extreme as here.
+
+### Influential nodes
+
+On this backbone, whose ties are no longer weighted, we can perform network calculations.
+We will begin by asking which nodes are the most influential in the network, i.e. possess the highest potential for spreading information.
+This is calculated through eigenvalue centrality, which is based on the number of nodes a node is linked to and how well those nodes are linked themselves.
+Eigenvalue centrality gives us a first idea of key sites in the spread of pottery decor knowledge.
+
+```{r backbone-eigenvalue, fig.cap="Binary backbone at alpha = 0.15 showing eigenvector centralities"}
+ggraph(decor_18, layout = "nicely") +
+  geom_node_point(aes(size = centrality_eigen())) +
+  geom_edge_fan(color = "grey", alpha = 0.8) +
+  scale_edge_width(range = c(0.2, 3)) +
+  geom_node_text(aes(label = label), repel = TRUE) +
+  theme_graph() +
+  labs(size = "eigenvector centrality")
+```
+
+The result (\@ref(fig:backbone-eigenvalue)) shows that the highest influence on the network is exerted by the sites in the Eastern IND and Lakes cluster.
+This reinforces the pattern already seen in clustering and shows the importance of the urban sites of the IND to Middle Niger culture in this period.
+We should not forget that the network is partial, i.e. many connections to many sites are missing.
+While the IND and the Bandiagara plateau and the Seno plain are both well researched and the networks therefore comparable, sites like Gao-Saney might have been influential in an undocumented local network.
+ Nevertheless, it suggests this region specifically as a center through which information and innovation in pottery decoration were disseminated in this period.
+
+### Brokers
+
+Betweenness centrality shows nodes in brokerage positions within the network.
+These are interesting to us in finding out where the different network cliques connect, and which sites are at crossroads positions in the pottery knowledge exchange.
+These are likely to be places with diverse markets, where producers belonging to different networks meet, and where information thus "jumps" between network cliques.
+
+```{r backbone-betweenness, fig.cap="Binary backbone at alpha = 0.15 showing betweenness centralities"}
+ggraph(decor_18, layout = "nicely") +
+  geom_node_point(aes(size = centrality_betweenness())) +
+  geom_edge_fan(color = "grey", alpha = 0.8) +
+  scale_edge_width(range = c(0.2, 3)) +
+  geom_node_text(aes(label = label), repel = TRUE) +
+  theme_graph() +
+  labs(size = "betweenness centrality")
+```
+
+Particularly interesting among the results (\@ref(fig:backbone-betweenness) is the position of the two sites at Dia, Shoma and Mara (Bedaux et al.
+2005), as providing a connection between the wider Niger cluster and the Eastern IND.
+It is notable that Shoma clusters more with the former, whereas Mara is more closely linked to the latter, via the site of Toguéré Galia (Bedaux et al.
+1978).
+The Eastern IND sites that had such high eigenvector centrality have particularly low betweenness centrality.
+Also of note is the brokerage position of the Sanga cave sites (Tellem) in relation to the Bandiagara and Seno cluster.
+As the assemblage here notionally stems from a funerary context, their high betweenness centrality might indicate that ceramics using unusual or imported decorations were preferentially deposited here, while the ceramics from contemporary habitation sites tended to have a more restricted repertoire.
+
+### Evaluating measurements
+
+In order to know whether these centrality measures are valid, rather than random effects of the backbone sparsification, we implement a variation on Peeples' (2017) suggested method for estimating the robustness of centrality measures in archaeological networks.
+In order to do so, the edges of the original graph are sampled at 10 % intervals, from 10 to 90 %.
+The binary backbone of each of these samples is then extracted and the centrality measures calculated.
+A Spearman's rank correlation test is run between the samples to see how the deletion of edges influences the centrality within the backbone.
+This is repeated 100 times.
+\@ref(fig:validity-test) shows that the correlation between the eigenvector and betweenness ranks of nodes in the original graph and the mean of the sampled graphs decreases quite rapidly.
+This shows us that our network is rather fragile, and the inferences we have made on the basis of the extracted backbone should be treated with the appropriate caution.
+
+```{r validity-test, warning = FALSE, message = FALSE, fig.cap= "Testing the robustness of the bb-network by comparing its centralities to the original graph"}
+#comparing
+p1 <- test_centralities(decor_graph, "eigen", 100)
+p2 <- test_centralities(decor_graph, "betweenness", 100)
+
+(p1 | p2) + plot_annotation(tag_levels = "a") +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1))
+```
+
 ## Community detection
@@
 ```{r community detection, hide = 'results', echo=FALSE}
-decor_graph1 <- decor_flat |>
-  activate(nodes) |>
-  mutate(louvain = as.character(group_optimal()))
-
-decor_graph1 <- decor_18 |>
+decor_graph1 <- decor_20 |>
   activate(nodes) |>
@@
 
-## Sparsification
-
-### Simplified network backbone
-
-While the hyperconnected hairball network has shown us key elements of the structure of pottery décor sharing, it is of less use for more traditional network analyses.
-These include finding key nodes in the network and key routes through it.
-In our case, such network statistics might tell us which of the sites, on the basis of their assemblages, were influential in the propagation of pottery decorations, and through which channels the knowledge of different décors might have preferentially passed.
-In order to extract such measures, the number of connections needs to be reduced.
-Network sparsification or 'backbone extraction', as this process is also called, is a common challenge in network analysis across its different applications (Serrano et al.
-2009; Saha et al.
-2013).
-In archaeological networks, this has frequently been achieved through thresholding, i.e.
-retaining only the strongest connections and discarding the others (as implemented for example in Hart & Engelbrecht 2012).
-There are, however, a number of other recent approaches to network sparsification.
-For our purposes, we have found the best results to come from a process called the "disparity filter" (Serrano et al.
-2009).
-This evaluates the weight of each edge against a null model in which the overall edge weights of a node are evenly distributed among its edges.
-If an edge is found to diverge from the null model at a set probability, it is retained.
-The backbone package (Domagalski et al.
-2021) implements this in R.
-In order to use it, we must first 'flatten' the multiplex graph, by adding the weights of the edges between each pair of nodes on all layers.
-We then extract the backbone of the graph using the disparity filter at different probability thresholds (\@ref(fig:backbone-extraction)).
-
-```{r flatten}
-decor_flat <- flatten(decor_graph, weighted = TRUE) |> 
-activate(nodes)
-
-```
-
-```{r backbone-extraction, fig.cap="Extraction of binary backbones at decreasing significance levels: a) 0.05, b) 0.15, c) 0.18"}
-# extract binary backbones at decreasing levels of significance
-decor_05 <- extract_backbone(decor_flat, 0.05)
-decor_15 <- extract_backbone(decor_flat, 0.15)
-decor_18 <- extract_backbone(decor_flat, 0.20) #20 ideale anzahl?
-
-# plot
-p1 <- plot.bb(decor_05)
-p2 <- plot.bb(decor_15)
-p3 <- plot.bb(decor_18) + geom_node_text(aes(label = label), repel = TRUE)
-
-(p1 | p2) / p3 +
-  plot_annotation(tag_levels = "a") +
-  plot_layout(heights = unit(c(1, 8), c("cm", "null")))
-
-decor_18 <- decor_18 |>
-  #isolated nodes are removed
-  activate(nodes) |>
-  filter(!node_is_isolated())
-
-```
-
-The three communities from the previous section are also visible in the extracted backbone, albeit with some major differences in configuration.
-The spread-out cluster of long-distance contact now also includes the sites at Akumbu and Kawinza which were placed in the IND cluster.
-The latter is now reduced to the sites of the Eastern IND, while the Dia sites take a central position within the network, and the "Dogon" cluster remains largely intact.
-Dammassogou and Kokolo NE 2 are isolated from the network, and remain so until the significance is lowered to 0.2.
-We will use the backbone at alpha = 0.18 to proceed with our analysis.
-If this selection of significance values seems rather arbitrary, we point out that analogous situations exist in other attempts at archaeological network simplifications.
-Hart & Engelbrecht (2012, p.
-339), for instance, threshold their similarity network to the point where all nodes form a network but for a single outlier.
-Higher significance values also do not necessarily mean we get a more reliable network model, especially in a situation where sampling and recovery bias is as extreme as here.
-
-### Influential nodes
-
-On this backbone, whose ties are no longer weighted, we can perform network calculations.
-We will begin by asking which nodes are the most influential in the network, i.e. possess the highest potential for spreading information.
-This is calculated through eigenvalue centrality, which is based on the number of nodes a node is linked to and how well those nodes are linked themselves.
-Eigenvalue centrality gives us a first idea of key sites in the spread of pottery decor knowledge.
-
-```{r backbone-eigenvalue, fig.cap="Binary backbone at alpha = 0.15 showing eigenvector centralities"}
-ggraph(decor_18, layout = "nicely") +
-  geom_node_point(aes(size = centrality_eigen())) +
-  geom_edge_fan(color = "grey", alpha = 0.8) +
-  scale_edge_width(range = c(0.2, 3)) +
-  geom_node_text(aes(label = label), repel = TRUE) +
-  theme_graph() +
-  labs(size = "eigenvector centrality")
-```
-
-The result (\@ref(fig:backbone-eigenvalue)) shows that the highest influence on the network is exerted by the sites in the Eastern IND and Lakes cluster.
-This reinforces the pattern already seen in clustering and shows the importance of the urban sites of the IND to Middle Niger culture in this period.
-We should not forget that the network is partial, i.e. many connections to many sites are missing.
-While the IND and the Bandiagara plateau and the Seno plain are both well researched and the networks therefore comparable, sites like Gao-Saney might have been influential in an undocumented local network.
- Nevertheless, it suggests this region specifically as a center through which information and innovation in pottery decoration were disseminated in this period.
-
-### Brokers
-
-Betweenness centrality shows nodes in brokerage positions within the network.
-These are interesting to us in finding out where the different network cliques connect, and which sites are at crossroads positions in the pottery knowledge exchange.
-These are likely to be places with diverse markets, where producers belonging to different networks meet, and where information thus "jumps" between network cliques.
-
-```{r backbone-betweenness, fig.cap="Binary backbone at alpha = 0.15 showing betweenness centralities"}
-ggraph(decor_18, layout = "nicely") +
-  geom_node_point(aes(size = centrality_betweenness())) +
-  geom_edge_fan(color = "grey", alpha = 0.8) +
-  scale_edge_width(range = c(0.2, 3)) +
-  geom_node_text(aes(label = label), repel = TRUE) +
-  theme_graph() +
-  labs(size = "betweenness centrality")
-```
-
-Particularly interesting among the results (\@ref(fig:backbone-betweenness) is the position of the two sites at Dia, Shoma and Mara (Bedaux et al.
-2005), as providing a connection between the wider Niger cluster and the Eastern IND.
-It is notable that Shoma clusters more with the former, whereas Mara is more closely linked to the latter, via the site of Toguéré Galia (Bedaux et al.
-1978).
-The Eastern IND sites that had such high eigenvector centrality have particularly low betweenness centrality.
-Also of note is the brokerage position of the Sanga cave sites (Tellem) in relation to the Bandiagara and Seno cluster.
-As the assemblage here notionally stems from a funerary context, their high betweenness centrality might indicate that ceramics using unusual or imported decorations were preferentially deposited here, while the ceramics from contemporary habitation sites tended to have a more restricted repertoire.
-
-### Evaluating measurements
-
-In order to know whether these centrality measures are valid, rather than random effects of the backbone sparsification, we implement a variation on Peeples' (2017) suggested method for estimating the robustness of centrality measures in archaeological networks.
-In order to do so, the edges of the original graph are sampled at 10 % intervals, from 10 to 90 %.
-The binary backbone of each of these samples is then extracted and the centrality measures calculated.
-A Spearman's rank correlation test is run between the samples to see how the deletion of edges influences the centrality within the backbone.
-This is repeated 100 times.
-\@ref(fig:validity-test) shows that the correlation between the eigenvector and betweenness ranks of nodes in the original graph and the mean of the sampled graphs decreases quite rapidly.
-This shows us that our network is rather fragile, and the inferences we have made on the basis of the extracted backbone should be treated with the appropriate caution.
-
-```{r validity-test, warning = FALSE, message = FALSE, fig.cap= "Testing the robustness of the bb-network by comparing its centralities to the original graph"}
-#comparing
-p1 <- test_centralities(decor_graph, "eigen", 100)
-p2 <- test_centralities(decor_graph, "betweenness", 100)
-
-(p1 | p2) + plot_annotation(tag_levels = "a") +
-  theme(axis.text.x = element_text(angle = 90, hjust = 1))
-```
-
 ## Diachronic analysis

### 2024-03-08 analysis/paper/paper.Rmd
script in proper order, workin iNext, communities based on sparsified networks and better community comparison

@@
 decor_15 <- extract_backbone(decor_flat, 0.15)
-decor_18 <- extract_backbone(decor_flat, 0.20) #20 ideale anzahl?
+decor_20 <- extract_backbone(decor_flat, 0.20) #20 ideale anzahl?
 
@@
 p2 <- plot.bb(decor_15)
-p3 <- plot.bb(decor_18) + geom_node_text(aes(label = label), repel = TRUE)
+p3 <- plot.bb(decor_20) + geom_node_text(aes(label = label), repel = TRUE)
 
@@
 
-decor_18 <- decor_18 |>
+decor_20 <- decor_20 |>
   #isolated nodes are removed
@@
 ```{r backbone-eigenvalue, fig.cap="Binary backbone at alpha = 0.15 showing eigenvector centralities"}
-ggraph(decor_18, layout = "nicely") +
+ggraph(decor_20, layout = "nicely") +
   geom_node_point(aes(size = centrality_eigen())) +
@@
 ```{r backbone-betweenness, fig.cap="Binary backbone at alpha = 0.15 showing betweenness centralities"}
-ggraph(decor_18, layout = "nicely") +
+ggraph(decor_20, layout = "nicely") +
   geom_node_point(aes(size = centrality_betweenness())) +
@@
 ```{r community detection, hide = 'results', echo=FALSE}
-decor_graph1 <- decor_20 |>
+#sparsifying 
+decor_graph_c <- decor_20 |>
   activate(nodes) |>
@@
 
-decor_graph1 <- decor_graph |>
+decor_graph <- decor_graph |>
   activate(nodes) |>
-  mutate(louvain = as.data.frame(decor_graph1, "vertices")$louvain)
+  mutate(louvain = as.data.frame(decor_graph_c, "vertices")$louvain)
 
 # calculate modularity of louvain clustering
-modularity <- decor_graph1 |>
+modularity <- decor_graph |>
   activate(nodes) |>
@@
   head(1)
-
-decor_graph1
-
-decor_graph |>
-  activate(edges) |> 
-  filter(str_detect(name, "roulette") | name %in% c("paint", "straight_mat_impression"))  |> 
-  as.data.frame("edges") |> 
-  group_by(name) |> 
-  summarise()
-
-decor_graph |>
-  activate(edges) |> 
-  filter(str_detect(name, "strip_roulette")| str_detect(name, "cord_roulette"))  |> 
-  as.data.frame("edges") |> 
-  group_by(name) |> 
-  summarise()
-
-decor_flat
 ```
@@
   setNames(
-    c("tomato", "cornflowerblue", "darkgreen", "yellow"),
-    unique(V(decor_graph1)$louvain)
+    c("cornflowerblue","tomato", "darkgreen", "yellow"),
+    unique(V(decor_graph)$louvain)
   )
@@
 
-ggraph(decor_graph1, x = lon, y = lat) +
+ggraph(decor_graph, x = lon, y = lat) +
   geom_sf(data = mali_adm0, fill = NA) +
@@
 ```{r cluster1, fig.cap="Internal connections within community 1"}
-decor_graph1 %>%
+decor_graph %>%
   activate(nodes) %>%
@@
   theme_graph()
-
-decor_18 %>%
-  ggraph(layout = "graphopt") +
-  geom_edge_fan(aes(edge_width = weight, edge_alpha = weight)) +
-  scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
-  scale_edge_width(range = c(0.2, 2)) +
-  geom_node_point(aes(size = centrality_degree(weights = weight)), colour = "cornflowerblue") +
-  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
-  theme_graph()
-
-decor_flat %>%
-  ggraph(layout = "graphopt") +
-  geom_edge_fan(aes(edge_width = weight, edge_alpha = weight)) +
-  scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
-  scale_edge_width(range = c(0.2, 2)) +
-  geom_node_point(aes(size = centrality_degree(weights = weight)), colour = "cornflowerblue") +
-  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
-  theme_graph()
 ```
@@
 ```{r cluster2, fig.cap="Internal connections within community 2"}
-decor_graph1 %>%
+decor_graph %>%
   activate(nodes) %>%
@@
 ```{r cluster3, fig.cap="Internal connections within community 3"}
-decor_graph1 %>%
+decor_graph %>%
   activate(nodes) %>%
@@
 
-```{r cluster-comparison, fig.cap="Relative edge weights and centralization by cluster"}
+```{r cluster-comparison, fig.cap="Relative edge weights, edge diversity and centralization by cluster"}
 # adding community/crossing to the edges
-decor_graph1 <- decor_graph1 |>
+decor_graph <- decor_graph |>
  # flatten(weighted = TRUE) |> 
@@
 
-decor_graph1
-
 # mean edge weights
-mean_edge_weight <- decor_graph1 |>
+mean_edge_weight <- decor_graph |>
   get.data.frame("edges") |>
@@
 for (i in 1:4) {
-  centralization[[i]] <- decor_graph1 |>
+  centralization[[i]] <- decor_graph |>
     # filter community
@@
 # Graph of mean edge weight of the different communities
-g1 <- mean_edge_weight |>
-  pivot_longer(!community, names_to = "key",
-               values_to = "mean_edge_weight") |>
-  ggplot(aes(x = community, y = mean_edge_weight)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  scale_fill_grey() +
+g1 <- ggplot(mean_edge_weight) +
+  geom_boxplot(aes(x = community, y = weight, fill = community)) +
+  scale_fill_grey(start = 0.5, end  = 1) +
   theme_bw() +
-  ylab("mean edge weight")
+  theme(legend.position = "none") +
+  labs(y = "degree centrality",
+       x = "community")
 
 # Graph of centralization of the different communities
-g2 <- tibble(centralization) |>
+g2 <- ggplot(mean_edge_weight) +
+  geom_boxplot(aes(x = community, y = div, fill = community)) +
+  scale_fill_grey(start = 0.5, end  = 1) +
+  theme_bw() +
+  theme(legend.position = "none") +
+  labs(y = "degree centrality",
+       x = "community")
+
+#Graph of centralization of the communities
+g3 <- tibble(centralization) |>
   mutate(community = 1:4) |>
@@
 # plot:
-g1 + g2
-
-ggplot(mean_edge_weight) +
-  geom_boxplot(aes(x = community, y = weight, fill = community)) +
-  scale_fill_grey(start = 0.5, end  = 1) +
-  theme_bw() +
-  theme(legend.position = "none") +
-  labs(y = "degree centrality",
-       x = "community")
-
-ggplot(mean_edge_weight) +
-  geom_boxplot(aes(x = community, y = div, fill = community)) +
-  scale_fill_grey(start = 0.5, end  = 1) +
-  theme_bw() +
-  theme(legend.position = "none") +
-  labs(y = "degree centrality",
-       x = "community")
-
-ggplot(centrali) +
-  boxplot(aes(x = community, y = div, fill = community)) +
-  scale_fill_grey(start = 0.5, end  = 1) +
-  theme_bw() +
-  theme(legend.position = "none") +
-  labs(y = "degree centrality",
-       x = "community")
+g1 + g2 +g3
 ```
@@
 # Calculate centralities for each cluster
-centr <- pull_centralities_clust(decor_graph1)
+centr <- pull_centralities_clust(decor_graph)
 
@@
 #show crossings with a weight over 3
-decor_graph1 |>
+decor_graph |>
   filter(
@@
 early_flat_bb <- 
-  backbone::disparity(as.igraph(early_flat), alpha = 0.22, narrative = TRUE) |>
+  backbone::disparity(as.igraph(early_flat), alpha = 0.21, narrative = TRUE) |>
   tidygraph::as_tbl_graph()

### 2024-03-08 analysis/paper/paper.Rmd

@@
 
-decor_20 <- decor_20 |>
+decor_20 <- decor_20 
   #isolated nodes are removed
@@
 ```{r community detection, hide = 'results', echo=FALSE}
+
 #sparsifying 

### 2024-03-11 analysis/supplementary_materials/data preparation.qmd
data preparation now complete

@@
 
-# the code 40 in twine is unclear, and thus left out
+# the code 40 in twine is unclear from the literature, and thus left out
 
@@
 Jj12plastic <- Jj12plastic |> 
-  mutate(channel = x1+x2, #single and multiple grooves are treated as one category
+  mutate(channel = x1+x2, #single and multiple channels are treated as one category
          nonparallel_grooves = x3,
@@
 ```{r doupwil data prep}
-doupwil <- read_csv(here::here("analysis/data/raw_data/Toguere_Doupwil.csv"))
+doupwil <- read_csv(here::here("analysis/data/raw_data/Toguere_Doupwil.csv")) |> 
+  rename(twisted_cord_roulette = 'twisted_cord_roulette?')
 
@@
 Stratigraphically detailed pottery décor data is available for the body sherds from units AK1 (Togola 2008, pp.
-55–57) and AK5 (McIntosh 2017).
+55--57) and AK5 (McIntosh 2017).
 The body sherd decor from AKB4 is reported in Horizons (Cissé et al.
 2017, pp.
-41–43).
+41--43).
 Décor data for rim sherds are not published for any of the excavations.
 
-For the 12th century network, we have used selected data from Akumbu Mound A, Unit AK1’s Middle Assemblage layers.
+For the 12th century network, we have used selected data from Akumbu Mound A, Unit AK1's Middle Assemblage layers.
 In this unit, a living surface dated to 1266-1425 cal.
@@
 36 & 41).
-Trusting that this is not the case, as is also McIntosh’s (McIntosh 2017, p.
+Trusting that this is not the case, as is also McIntosh's (McIntosh 2017, p.
 197) interpretation, we have used the Horizon 1 body sherd data for the 9th century network.
@@
          incised = "Incicions au couteau seules",
-         cord = "Cordons" #this is described as "presénce d'un petit cordon sous un bord 
+         channel = "Cordons" #this is described as "presénce d'un petit cordon sous un bord 
          )
@@
 
-Here we create two separate collated dataset for the 12th and 9th century networks by binding together the previously created data for each site and then further cleaning and standardising.
+Here we create two separate collated datasets for the 12th and 9th century networks by binding together the previously created data for each site and then further cleaning and standardising.
 
@@
  mutate_all(~ifelse(is.na(.), 0, .)) |> 
-  mutate(incised = incised + incised_patterns) |> 
-  select(-incised_patterns) |> 
+  mutate(incised = incised + incised_patterns,
+         cord_wrapped_roulette = cord_wrapped_roulette + cord_wrapped_stick_roulette,
+         applied_plastic = applied_plastic + applied_ridges,
+         punctate = punctate + impressed_stick,
+         comb_impression = impressed_comb) |> 
+  select(-c(incised_patterns,
+            cord_wrapped_stick_roulette,
+            slip, # suffers from uneven reporting
+            unidentified_roulette,
+            nonparallel_grooves,
+            finger_arcs,
+            applied_ridges,
+            impressed_stick,
+            impressed_comb)) |> 
   rownames_to_column("Site")
@@
   mutate_all(~ifelse(is.na(.), 0, .))|> 
+  mutate(applied_plastic = applied_plastic + applied_ridges,
+         cord_wrapped_roulette = cord_wrapped_roulette + cord_wrapped_stick_roulette) |> 
+  select(-c(slip,
+            unidentified_roulette,
+            nonparallel_grooves,
+            applied_ridges,
+            cord_wrapped_stick_roulette)) |> 
   rownames_to_column("Site")

### 2024-03-11 analysis/paper/paper.Rmd
small typo changes

@@
 
-decor_20 <- decor_20 
+decor_20 <- decor_20 |> 
   #isolated nodes are removed
@@
   theme(legend.position = "none") +
-  labs(y = "degree centrality",
+  labs(y = "mean edge weight",
        x = "community")

### 2024-04-07 analysis/paper/paper.Rmd
inext update

--- 2024-03-08 analysis/paper/paper.Rmd
@@
   select(-1)
+
 ```
@@
 
+i <- iNEXT(decors_abs, q=0, datatype="abundance")
+
+i
+
+ i <- decors_abs[,c(2,4,11)]|> 
+ iNEXT(q=c(0,1,2), datatype="abundance")
+ ggiNEXT(i, type=1, facet.var="Assemblage")
+
+  i <- decors_abs[,c(4)]|> 
+ iNEXT(q=c(0), datatype="abundance")
+ ggiNEXT(i, type=1, facet.var="Assemblage")
+
+ i <- decors_abs[,c(8,15,16)]|> 
+ iNEXT(q=c(0,1,2), datatype="abundance")
+  ggiNEXT(i, type=1, facet.var="Assemblage")
+
+  i <- decors_abs[,c(8)]|> 
+ iNEXT(q=c(0), datatype="abundance")
+  ggiNEXT(i, type=1, facet.var="Assemblage")
+
+    i <- decors_abs[,c(13)]|> 
+ iNEXT(q=c(0), datatype="abundance")
+  ggiNEXT(i, type=1, facet.var="Assemblage")
+
+decors_abs <- decorsearly |> 
+  left_join(n_decors) |> 
+    mutate(across(where(is.numeric), ~ .*n/100)) |> 
+  mutate(across(where(is.numeric), ~ round(.))) |> 
+    column_to_rownames("sites") |> 
+  select(-n) |> 
+t()
+
 iNEXT(decors_abs, q=0, datatype="abundance")
-
 ```
@@
 
- i <- decors_abs[,c(2,5,8)]|> 
+ i <- decors_abs[,c(2,4,11)]|> 
  iNEXT(q=c(0,1,2), datatype="abundance")
+ ggiNEXT(i, type=1, facet.var="Assemblage")
+
+ i <- decors_abs[,c(8,15,16)]|> 
+ iNEXT(q=c(0,1,2), datatype="abundance")
+  ggiNEXT(i, type=1, facet.var="Assemblage")
 
@@
 
+```
+```{r}
+i$DataInfo |> 
+  arrange(n) |> 
+  mutate_at(vars(2:14), as.numeric) |> 
+  ggplot(aes(x = Assemblage))+
+  geom_point(aes(y = n)) +
+  geom_point(aes(x = Assemblage, y = SC))
 ```

### 2024-04-08 analysis/supplementary_materials/data preparation.qmd
gao new data

@@
 "cord_wrapped_roulette" = "Sisal", # Cisse (2011, 161) identifies them as cord_wrapped_roulette after Nixon (2008, 183)
-"straight_mat_impression" = "Birdfood" # Cisse (2011, 161) identified it as mat, but not completly sure, maybe other mat than straight
-) |> 
+"straight_mat_impression" = "Birdfood", # Cisse (2011, 161) identified it as mat, but not completly sure, maybe other mat than straight
+"n_gao" = "n") |> 
+  filter(Phase != "lower") |> 
+  mutate(Site = "Gao") |> 
 mutate_if(is.numeric, round, digits=0) #rounding the numbers (because they were extracted from graphics there were minor deviations)
@@
 n_gao <- Gao |> 
-  select_if(is.numeric) |> 
-  rowSums()
+  select(n_gao)
 
 Gao <- Gao |>
-  add_column(n_gao) |> 
-  mutate_if(is.numeric, function(x) x/n_gao*100)
+  mutate_if(is.numeric, function(x) (x/n_gao)*100)
 
 Gao9_decor <- Gao |> 
-  filter(Phase == "lower") |> 
+  filter(Phase == "upper") |> 
   select(-"Site-Code", -"Phase", -"n_gao") |> 
@@
 
-names(Gao)
+names(Gao12_decor)
 
@@
 )
+
+Gao
 ```

### 2024-04-08 analysis/supplementary_materials/data preparation.qmd
implementation of changed gao data

--- 2024-03-11 analysis/supplementary_materials/data preparation.qmd
@@
 ) |> 
+  filter(Phase != "lower") |> 
+  mutate(Site = "Gao") |> 
 mutate_if(is.numeric, round, digits=0) #rounding the numbers (because they were extracted from graphics there were minor deviations)
+
+Gao <- as.data.frame(Gao)
 
@@
 Gao9_decor <- Gao |> 
-  filter(Phase == "lower") |> 
-  select(-"Site-Code", -"Phase", -"n_gao") |> 
+  filter(Phase == "upper") |> 
+  select(-"Site-Code", -"Phase") |> 
   mutate(Site = "Gao9") |> 
-  column_to_rownames("Site")
+  column_to_rownames("Site") |> 
+  select(-"n_gao")
 
@@
 
-names(Gao)
-
 n_gao <- data.frame(

### 2024-04-08 analysis/supplementary_materials/data preparation.qmd
added commentary

@@
 ) |> 
-  filter(Phase != "lower") |> 
+  filter(Phase != "lower") |> #removing the lowest phase as it is probably to old
   mutate(Site = "Gao") |> 

### 2024-04-30 analysis/paper/paper.Rmd
cleaned repo, knit works

--- 2024-03-11 analysis/paper/paper.Rmd
@@
 
-#getting absolute numbers of decorations
+#getting absolute numbers of decorations for the 12th century sites
 decors_abs <- decors |> 
@@
 
+# getting a table of the results of the analysis
+i <- iNEXT(decors_abs, q=0, datatype="abundance")
+
+i
+
+# Fractured and extrapolated diversity of Doupwil
+  i <- decors_abs[,c(4)]|> 
+ iNEXT(q=c(0), datatype="abundance")
+ ggiNEXT(i, type=1, facet.var="Assemblage")+
+  labs(title = "Doupwil")
+
+# Fractured and extrapolated diversity of Jenne-jeno
+ i <- decors_abs[,c(8)]|> 
+ iNEXT(q=c(0), datatype="abundance")
+  ggiNEXT(i, type=1, facet.var="Assemblage")+
+  labs(title = "Jenne-jeno")
+
+#getting absolute numbers of decorations for the 9th century sites
+decors_abs <- decorsearly |> 
+  left_join(n_decors) |> 
+    mutate(across(where(is.numeric), ~ .*n/100)) |> 
+  mutate(across(where(is.numeric), ~ round(.))) |> 
+    column_to_rownames("sites") |> 
+  select(-n) |> 
+t()
+
+# getting a table of the results of the analysis
 iNEXT(decors_abs, q=0, datatype="abundance")
-
-```
-
-```{r}
- #install_github('AnneChao/iNEXT.4steps') #not currently on CRAN
- library(iNEXT)
- library(iNEXT)
-# library(iNEXT.4steps)
- library(ggplot2)
-
- i <- decors_abs[,c(2,5,8)]|> 
- iNEXT(q=c(0,1,2), datatype="abundance")
-
- ggiNEXT(i, type=1, facet.var="Assemblage")
-ggiNEXT(i, type=3, facet.var="Order.q", color.var="Assemblage")
-
- i <- decors_abs[,c(2)]|> 
- iNEXT(q=c(0,1,2), datatype="abundance")
-
- ggiNEXT(i, type=1, facet.var="Assemblage")
-ggiNEXT(i, type=3, facet.var="Order.q", color.var="Assemblage")
-
 ```
@@
 ```
+```{r weighted flat decor}
+flat_edges <- decor_flat |> 
+  get.data.frame("edges")
+
+decor_20 <- decor_20 |> 
+  activate(edges) |> 
+  select(-weight) |> 
+  left_join(flat_edges)
+
+ggraph(decor_20, layout = "nicely") +
+  geom_node_point(aes(size = centrality_eigen())) +
+  geom_edge_fan(color = "grey", alpha = 0.8, aes(edge_width = weight)) +
+  scale_edge_width(range = c(0.2, 3)) +
+  geom_node_text(aes(label = label), repel = TRUE) +
+  theme_graph() +
+  labs(size = "eigenvector centrality")
+```
 
@@
   geom_node_point(aes(size = centrality_eigen())) +
-  geom_edge_fan(color = "grey", alpha = 0.8) +
+   geom_edge_fan(color = "grey", alpha = 0.8, aes(edge_width = weight)) +
   scale_edge_width(range = c(0.2, 3)) +
@@
   geom_node_point(aes(size = centrality_betweenness())) +
-  geom_edge_fan(color = "grey", alpha = 0.8) +
+   geom_edge_fan(color = "grey", alpha = 0.8, aes(edge_width = weight)) +
   scale_edge_width(range = c(0.2, 3)) +
@@
 # calculate modularity of louvain clustering
-modularity <- decor_graph |>
+modularity <- decor_20 |>
   activate(nodes) |>
-  mutate(modularity = graph_modularity(group = as.factor(louvain))) |>
+  mutate(modularity = graph_modularity(group = group_louvain())) |>
   pull(modularity) |>
@@
   filter(louvain == 1) %>%
+  filter(label != "Sanga D") |> 
+  activate(edges) |> 
+  #filter(weight > 3) |> 
   ggraph(layout = "graphopt") +
+  geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
+  scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+  scale_edge_width(range = c(0.5, 2)) +
+  geom_node_point(aes(size = centrality_degree(weights = weight)), colour = "cornflowerblue") +
+  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
+  theme_graph()
+
+decor_graph %>%
+ flatten(weighted = TRUE) |> 
+ activate(nodes) %>%
+  mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 1) %>%
+flatten(weighted = TRUE) |> 
+  ggraph(layout = "graphopt") +
+  geom_edge_fan(aes(edge_width = weight, edge_alpha = weight)) +
+  scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+  scale_edge_width(range = c(0.5, 2)) +
+  geom_node_point(aes(size = centrality_degree(weights = weight)), colour = "cornflowerblue") +
+  geom_node_text(aes(label = label), col = "black", repel = TRUE) +
+  theme_graph()
+
+decor_graph %>%
+  activate(nodes) %>%
+  mutate(deg = centrality_eigen(weights = weight)) %>%
+  filter(louvain == 1) %>%
+  ggraph(x = lon, y = lat) +
   geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
@@
 ```
-```{r cluster3, fig.cap="Internal connections within community 3"}
+```{r cluster4, fig.cap="Internal connections within community 3"}
 decor_graph %>%
@@
   theme(legend.position = "none") +
-  labs(y = "mean edge weight",
+  labs(y = "edge weight",
        x = "community")
@@
   theme(legend.position = "none") +
-  labs(y = "degree centrality",
+  labs(y = "edge div",
        x = "community")

### 2024-04-30 analysis/paper/paper.Rmd

@@
   theme(legend.position = "none") +
-  labs(y = "edge weight",
+  labs(y = "mean edge weight",
        x = "community")

### 2024-04-30 analysis/supplementary_materials/data preparation.qmd

@@
 
+The site was excavated by Guindo (Guindo 2011, 243f).
+At the base of the uppermost layer, a piece of charcoal was found, dating to 1029-1209 cal.
+CE (Ly-3712, 915 ± 30 BP). We used the pottery from this Layer for our 12^th^-century network.
+The other date from the excavation comes from layer 5, the oldest of the excavation, dated to 430-599 cal.
+CE (Ly-3968, 1545 ± 30 BP).
+Unfortunately, the other layers cannot be precisely chronologically assigned, which makes it possible that one or more of these layers could correspond to our 9^th^-century network.
+However, for a reliable assignment, the corresponding radiocarbon d
+
+During the recording of ceramic decorations, no distinction was made between red painting and slip. However, the author's descriptions lean more towards slip.
+
 ```{r}
@@
 ## Damassogou
+
+The site was excavated by Néma Guindo (2011, 171).
+For the 12^th^-century network, pottery from Layer 2 was utilized.
+At the bottom of Layer 2 there is a structure that is only described as friable.
+The layer is dated to 925 ± 30 BP, corresponding to 1024-1206 AD44 (Ly 3189).
+
+During the recording of ceramic decorations, no distinction was made between red painting and slip. However, the author's descriptions lean more towards slip.
 
@@
 
+The Site was excavated by Keita (2011, 74).
+At the structure in Horizon 2 of the trench section A of Kokolo a radiocarbon date was taken (790 ± 35 BP 1190-1278 cal.
+CE).
+The author uses this date to contextualize horizons in other trenches to the same settlement Phase in the 12^th^ and 13^th^ century.
+Nevertheless, we will only use pottery from the dated horizon itself for our 12^th^ century network model.
+
 ```{r}
@@
 ## Nin-Bére
+
+Nin-Bére was excavated by Guindo (2011, 227).From the Nin-Bèrè archaeological.
+site, there are two radiocarbon datings: one from Layer 2 dated to 730 ± 40 BP 1225-1376 cal.
+CE (Ly-13211) and another from Layer 5, which represents the latest excavation, dated to 1095 ± 35 BP 889-1016 cal.
+CE (Ly-13212).
+Since there are no structures providing a more precise chronological.
+placement for the other layers, only pottery from Layer 5 can be considered for inclusion in the 9^th^ century network.
+
+During the recording of ceramic decorations, no distinction was made between red painting and slip. However, the author's descriptions lean more towards slip.
 
@@
 ## Promontoire site of Ounjougou
+
+The excavation at the Promontoire site of the Ounjougou site complex was conducted by Downing as part of a diploma thesis.
+Ounjougou comprises seven levels, sometimes divisible into different horizons, with their summits appearing to correspond to occupation levels (Mayor et al., 2005).
+Level 4 is composed of two horizons.
+T4.2 was dated to 1315 ± 45 BP 660--720 cal.
+CE (Ly-1284 OXA), making it unsuitable for our 9^th^ century network.
+The other horizon of this level, T4.1, dates to 1005 ± 30 BP 990--1040 cal.
+CE (Ly-9340), thus being not beeintpart of our 12^th^ century network.
+We only used level T3, dating to 885 ± 50 BP 1040--1220 cal.
+CE (Ly-10150), for this network.
+
+For our network models, we utilize pottery decoration data published by Mayor (2010).
+Her reassessment of the sherds aids in making them comparable to those from other sites within our model.
+But some decorations that are mentioned in other publications, like incisions will therefore not be taken into account.
 
@@
 ## Sadia
+
+The site was excavated as part of the \"Human Population and Paleoenvironment in Africa\" by Huysecom et al..
+The dates of the different stratigraphic ensembles of the site are the results of a bayesian model based on 27 radiocarbon dates.
+For our 12^th^ century network we used pottery from the stratigraphic ensembles SE1 (1068-1209) and SE2 (1030-1133).
+In our 7^th^-9^th^ century network pottery from the earliest stratigraphic ensemble, SE5 (792-887), was included.
+
+The ceramic data was taken from figure 9 of Huysecom et al. (2017).
+
+It is problematic that for some SE the proportion of not specifally named decorations is high (especially for SE5 where this is the case for 25% of the decorations).
 

### 2024-05-15 analysis/paper/paper.Rmd
old changes

--- 2024-04-07 analysis/paper/paper.Rmd
@@
   select(-1)
-
 ```
@@
 
-i <- iNEXT(decors_abs, q=0, datatype="abundance")
-
-i
-
- i <- decors_abs[,c(2,4,11)]|> 
- iNEXT(q=c(0,1,2), datatype="abundance")
- ggiNEXT(i, type=1, facet.var="Assemblage")
-
-  i <- decors_abs[,c(4)]|> 
- iNEXT(q=c(0), datatype="abundance")
- ggiNEXT(i, type=1, facet.var="Assemblage")
-
- i <- decors_abs[,c(8,15,16)]|> 
- iNEXT(q=c(0,1,2), datatype="abundance")
-  ggiNEXT(i, type=1, facet.var="Assemblage")
-
-  i <- decors_abs[,c(8)]|> 
- iNEXT(q=c(0), datatype="abundance")
-  ggiNEXT(i, type=1, facet.var="Assemblage")
-
-    i <- decors_abs[,c(13)]|> 
- iNEXT(q=c(0), datatype="abundance")
-  ggiNEXT(i, type=1, facet.var="Assemblage")
-
-decors_abs <- decorsearly |> 
-  left_join(n_decors) |> 
-    mutate(across(where(is.numeric), ~ .*n/100)) |> 
-  mutate(across(where(is.numeric), ~ round(.))) |> 
-    column_to_rownames("sites") |> 
-  select(-n) |> 
-t()
-
 iNEXT(decors_abs, q=0, datatype="abundance")
+
 ```
@@
 
- i <- decors_abs[,c(2,4,11)]|> 
+ i <- decors_abs[,c(2,5,8)]|> 
  iNEXT(q=c(0,1,2), datatype="abundance")
- ggiNEXT(i, type=1, facet.var="Assemblage")
-
- i <- decors_abs[,c(8,15,16)]|> 
- iNEXT(q=c(0,1,2), datatype="abundance")
-  ggiNEXT(i, type=1, facet.var="Assemblage")
 
@@
 
-```
-```{r}
-i$DataInfo |> 
-  arrange(n) |> 
-  mutate_at(vars(2:14), as.numeric) |> 
-  ggplot(aes(x = Assemblage))+
-  geom_point(aes(y = n)) +
-  geom_point(aes(x = Assemblage, y = SC))
 ```
@@
                      "paint",
-                     "straight_cord_mat")) |>
+                     "straight_mat_impression")) |>
 
@@
 decor_05 <- extract_backbone(decor_flat, 0.05)
-decor_15 <- extract_backbone(decor_flat, 0.15)
+decor_15 <- extract_backbone(decor_flat, 0.14)
 decor_20 <- extract_backbone(decor_flat, 0.20) #20 ideale anzahl?

### 2024-05-29 analysis/paper/paper.Rmd
changes accroding to paper revisions

@@
 
+The full text of the article is available in the published version in "Azania: archaeological research in Africa".
+This document contains the code for all analyses and figures in the text,as well as some supplementary analzses and figures.
+It is structured by section to ease overview.
+
+For details on the data and its preparation, see the document in the folder "analysis/supplementary_materials".
+
 # Introduction
-
-The 40 years between 1970 and 2010 were a period of increasing archaeological research in Mali, West Africa.
-In the region around the middle reaches of the river Niger, this research documented the remains of a dynamic civilization which, from the last millennium BC onwards, developed urbanism, craft specialization, and long-distance trade (McIntosh 1998), and was highly influential not only in changing narratives about the African past but also in global debates on sociocultural dynamics (McIntosh 1999).
-One of the characteristics of many settlement sites of the first-early second millennium AD in the region is the high volume of ceramic sherds, which sometimes challenged the capacities of research projects (McIntosh 1995, pp.
-132--133).
-Featuring a large variety of decorative motifs, forms, and production techniques, the pottery of the region has been the subject of numerous studies, both from an archaeological and ethnoarchaeological perspective.
-Three key points have emerged from this research history.
-Firstly, establishing chronologically sensitive typologies has been very difficult, with contemporary variability often more significant than diachronic change.
-Secondly, base forming techniques, which vary throughout the region, are likely to correspond to larger cultural and maybe also linguistic spheres.
-Thirdly, pottery vessels appear to have been objects of trade, at times over long distances.
-
-In this paper, we consider decorative practices on 12^th^ century AD ceramics of this region.
-We use a network approach to ask how such practices might have spread, and, even when they are not useful as chronological markers, how they might still inform us about sociocultural dynamics of the ancient Middle Niger.
-This research aim reflects a long-standing concern of archaeologists in the area, but the network approach we adopt here shifts the perspective from ceramic production as representing sociocultural entities to the processes by which techniques spread.
-This results in connections, rather than boundaries, as the subject of research.
-Using network models, we ask what ceramics can tell us about the processes that integrated the Middle Niger as a cultural area despite its diversity, and where cultural exchange might have been facilitated or impeded, and why.
 
@@
 
-In many contemporary societies in the region, pottery is made by occupational specialist groups (Bambara: *nyamakala*).
-Though there are exceptions, it is the cultural norm in the wider Middle Niger region that pottery is made by women of groups in which the men are specialists in metallurgy.
-There is some debate as to how far in the past this specialization developed, but it currently seems likely that it has its roots in the first millennium AD (McIntosh 1998, pp.
-177--178), and that it had spread throughout the region by the mid second millennium (Tamari 1991, 1995).
-There is no doubt that potting and metallurgical experts were of great significance to the economy of the Middle Niger during this period of urbanism and emerging statehood, since they made basic products used in most parts of everyday life, from food production to the religious sphere.
-The cultural innovations linked to pottery making, both in social terms (endogamous craft specialists) and in technology (pottery and metal products), are a key factor in the development of the Middle Niger civilization (McIntosh 1998).
-
-From the 1st millennium BC onwards, ceramics occur in high numbers on archaeological sites on the Middle Niger, and their appearance, especially their surface treatment, is highly variable.
-The number of different roulette decorations, painted, incised, and applied decors, but also the variety in form, is very high, especially when attention is paid to the techniques and tools of decoration, rather than exclusively to their effects.
-Consequently, there has been considerable emphasis on ceramics in archaeological and in ethnoarchaeological work.
-Early studies (Bedaux 1980; e.g.
-Curdy 1982; and particularly Bedaux and Lange 1983) noted both regional differences in the appearance of pottery as well as striking widespread similarities in individual forms or attributes.
-Subsequent work has seen both in-depth studies on ceramics of individual sites (e.g.
-McIntosh 1995; Schmidt et al.
-2005; Gallin 2011; van Doosselaere 2014) as well as studies on the spread of particular decorative practices (Livingston Smith 2007) or shaping techniques (e.g.
-Sterner and David 2003; Mayor 2011a).
-Three main features of the wider region's ceramic record are important to this study: 1) the relative lack of clear temporal change, 2) high contemporary variability, and 3) pottery as exchange items.
-
 ### Lack of temporal markers:
-
-Susan K.
-McIntosh's analyses of the ceramics of Jenné-jeno and surrounding sites resulted in a first detailed pottery chronology on the Middle Niger.
-In two studies (McIntosh and McIntosh 1980; McIntosh 1995), McIntosh was able to show how some pottery shapes, fabrics and decorations showed clear temporal trends.
-There was, however, also quite strong stability, particularly in many of the decorative techniques (McIntosh and McIntosh 1980, pp.
-233--244).
-Subsequent studies on other sites have also shown change in ceramic production, and particularly in its decorative aspects, to have been very gradual, and stability to have been the norm (e.g.
-Schmidt et al.
-2005).
-This contrasts with the archaeological record in many other regions of the world, where the outward appearance of pottery has been far more changeable over time and can therefore be used as chronological markers.
-This has not been possible for most periods and areas of the Middle Niger region, where decorative aspects tend only to vary in their proportion through time.
 
@@
 
-The more striking differences between archaeological ceramics in the area are not diachronic, but contemporary.
-Pottery from different areas varies in its forming techniques, its shapes (though the most pottery is globular or ovoid in its basic shape), and in decorative practices.
-Occasionally, as at Tongo Maaré Diabal (Gestrich and MacDonald 2018) or Kumbi Saleh (van Doosselaere 2014), different productions can be clearly separated.
-The mechanisms underlying these differences have been the subject of extensive ethnoarchaeological and archaeological research by the University of Geneva.
-Led from the outset by the goal of relating ceramic production to ethnic groups (Gallay et al.
-1998, pp.
-15--22), this extensive research documented how certain parts of the ceramic production -- notably base forming techniques and tools employed in the potting process -- could be employed to create typologies that were patterned by the ethnic identity, economic milieu, and ecological zone of the producers (Gallay et al.
-1998, pp.
-106--107; Mayor 2011b, p.
-14).
-But certain decorative techniques were also found to have a certain correlation to the ethnicity-based typology created by the Geneva team (Mayor 2011b, pp.
-40--78).
-The main limiting factor here, as Mayor stresses, is that "stylistic borrowings are relatively easy and frequent" (Mayor 2011b, p.
-78), but given the fact that some of the decors do correlate with ethnic groups --- or at least with the types assigned to ethnic groups --- Mayor hints at the suggestion that decorative techniques can historically be assigned to one or another group from which they originated (ibid.).
-
-Based on their extensive fieldwork, Mayor's hypothesis is that pottery making is learnt within stable and relatively impermeable family relationships.
-This is in keeping with some of Gosselain's results from Niger, where he has suggested that this initial approach to the shaping of a pot was the most influenced by teaching and learning (Gosselain 2000, p.
-193).
-Since this usually happens from an early age within the family, it makes sense that kinship relationships, and thus often group and linguistic belonging, would influence the propagation of such techniques.
-These processes of knowledge exchange, limited to endogamous groups and ethnic milieus, are what explains the commonalities that underlie the typologies created by the Geneva group.
-Exchange outside of these limits is seen as the exception, rather than the norm.
-Nevertheless, Mayor gives the following scenarios for the diffusion of pottery traits (Mayor 2011b, p.
-83): Firstly, when a technique is simple and expedient, thus easy and profitable to learn.
-Secondly, when an ethnic group is in the minority in a certain area or when communities find themselves at the geographical margins of their ethnic group.
-Thirdly, rare cases in which the social order is transgressed by marriage outside of the ethnic group.
-Mayor thus expects that shaping and decorative techniques would have had a single and ethnically distinct origin and that it is "several centuries of borrowing" (ibid.) that have led to the less demarcated situation that the Geneva team observed in the 1990s and 2000s.
-
 ### Ceramics as trade goods:
 
-Even though most ceramic production appears to have been made for local markets, there is also evidence of pottery being used in trade over longer distances.
-Historically, the practice of pottery being moved on the Niger by boat and sold widely throughout the region has been documented in an account from the 18^th^ century (Park 1799, p.
-219).
-From the archaeological evidence, it seems as though this was not confined to later periods, but at least occasionally practiced also in the early second millennium and before.
-Finds of imported ceramics have, for instance, been made at Essouk (Nixon 2017, p.
-143,146-147), and at Kumbi Saleh (van Doosselaere 2014, pp.
-228--233).
-We must therefore always consider the possibility that archaeological assemblages do not only reflect local production, and add the indirect contact from long-distance trade to the possible mechanisms of change in ceramic practices.
-
-If the goal of the work by Gallay and Mayor and their teams was to identify the borders between groups in the past, this study asks a different question: how was pottery knowledge transmitted?
-What were the factors limiting or encouraging the flow of information?
-We thus begin from a null hypothesis of unlimited sharing, rather than of delimited groups, to see how the results might complement those of the Geneva research teams.
-Using pottery décor as a relatively easily borrowed attribute, we aim to generate hypotheses on how pottery making knowledge spread along the ancient Middle Niger.
-Further, we wish to demonstrate the use of network analyses for approaching West African archaeological ceramics and for making such connections visible and explorable.
-This technique presents a change of perspective on the ceramics of this region, allowing questions of connection to take precedence over those of boundaries.
-
 ## Ceramic decors as networks
-
-In recent years, the use of network analysis in archaeology has been growing significantly (see e.g.
-Knappett 2021).
-A large number of these studies focus on ceramics, but geographically most important are North America (Hart 2012; Mills et al.
-2013; Crabtree 2015; e.g.
-Barker and Young 2017; Vavrasek 2020) and the Mediterranean (Torrence and Swadling 2008; Brughmans and Poblome 2016; Iacono 2016; de Groot 2019; Sternini 2019).
-The archaeological records of these regions are amongst the best and longest studied in the world, with fine-grained pottery typochronologies and extensive excavations.
-To our knowledge, this is the first application of a network approach to archaeological ceramic material from sub-Saharan Africa, where the conditions are somewhat different (McIntosh 1995, pp.
-130--131).
-That is not to say that archaeological pottery on the Middle Niger is poorly studied.
-Most researchers in the wider area have detailed publications on their ceramic assemblages, efforts have been made at systematizing nomenclature (Haour et al.
-2010) and at synthetic overviews (Mayor 2011b).
-A large body of ethnographic and ethnoarchaeological literature provides the theoretical basis that informs the network model.
-
-Ethnographic studies of pottery manufacturing in Africa suggest that the way that pottery is made depends largely on systems of learning and teaching the craft (Gosselain 1998), on local market consumer choices (Dietler and Herbich 1994), or on marriage patterns among manufacturing groups (ibid.).
-In short, the processes underlying the propagation of technical practices through time and across geographical space are dependent on communication and, more specifically, knowledge exchange.
-Such exchange occurs between humans and we can therefore understand similarities in the *production* as evidence that knowledge was shared.
-However, different elements of the pottery making process are thought to be reflective of different kinds of social interaction.
-Gosselain (2000) concludes that, on the whole, forming techniques are more stable, reflective of durable cultural ties, sometimes even ethnic groups.
-On the other end of the spectrum, decorations are thought to be influenced by far more ephemeral contact, in which seeing someone else's pot might be enough to change a potter's production, as could indirect contact via the wishes of customers.
-
-Decoration on pottery vessels is a topic that has been central to archaeological debates for a very long time.
-To construct an overly simplified history of ceramic decors in African archaeology, we might distinguish an early period analogous to other parts of the world, in which decors (as a main constituent of typologies) served to distinguish human groups.
-This position became more elaborate and changed rationale over time, arguably culminating in Hodder's "Symbols in Action" (1982).
-David et al.'s "Why pots are decorated" (1988) did much to remove confidence in this idea and to situate decorative practices as part of chains of production and expressions of often unconscious propagation of social values.
-For them, "designs on pottery, far from being 'mere decoration,' art for art's sake, or messages consciously emblemic of ethnicity, are low-technology channels through which society implants its values in the individual --- every day at mealtimes" (David et al.
-1988, p.
-379).
-Over the intervening years, the comparatively strong ethnoarchaeological research in Africa has developed an increasingly complex picture of ceramic décor.
-Gosselain's (2011) "pourquoi le decorer", part review, part programmatic article, gives a good overview : Decorations on pottery occur within cultural constraints and logics, are expressions of individuality, are market-oriented, have practical considerations, are made for visual effect, have profound meanings while being thought of as trivial, are subject to fashion and taste, stand for group membership, mean different things to different people.
-In short, pottery decoration is not anchored in any stable way to any cultural phenomenon.
-More importantly for this study, however, Gosselain points out that the propagation, the process of spreading, of decorative practices is a more fruitful topic for study (ibid.: 12).
-Various decorative practices presuppose various kinds of contact.
-Some of these, as in the case of a particularly intricate rouletted tool, require the sharing of cultural knowledge and probably reflect shared practices in other cultural domains as well.
-Others are transmitted more easily, by sight alone.
-Whatever these mechanisms, there are constraints on transmission and innovation.
-As Herbich (1987, p.
-202) notes, potters innovate and express themselves with reference to other potters they interact with.
-Therefore, the innovation and spread of decorative practices does not occur freely, but is helped by geographical proximity, social contact, and cultural factors.
-It is likewise hindered by geographic distance and cultural differences.
-
-This, then, is the basis on which our social network model is conceived.
-It supposes that co-occurrence of a given practice of decorating pottery is emblematic of a form of sustained social contacts, within the limitations mentioned above.
-These contacts are not as close as we would expect from a teaching and learning environment, which would be better reflected in forming techniques, but are sustained and meaningful enough for there to have been an exchange of cultural aesthetic values and an exchange of knowledge of how to decorate pottery.
-These exchanges are therefore what the ties in our network represent.
-
-In choosing sites as the nodes of the network, the data come mainly from contexts in which pottery was consumed, rather than produced.
-The network agents in the model are therefore communities of consumers.
-Their ties, based on shared decors on their pottery, are indirect: they must be understood to have been connected via producers (individuals and communities) and possibly also merchant middlemen which remain largely unseen in the archaeological record.
-Since we have established at least occasional long-distance trade in pottery in our research area (see above), this is clearly a factor limiting the model's explanatory power.
-Nevertheless, we can use the work carried out by the Geneva MAESAO project to suggest that the majority of pottery found at a site will generally be locally produced (Mayor 2011b, p.
-84), which might reasonably allow some conflation between pottery producers and consumers in the network nodes.
-A low level of imported ceramics can be adjusted for in the network model, but should not be excluded, as these "weak ties" have been shown to be important for introducing innovation in social networks (Granovetter 1973), including technical innovation in ceramic production (Roux et al.
-2018).
-
-The analysis of the network will attempt to generate hypotheses on how such knowledge exchanges were structured, the level of interaction between inferred producers of ceramics, and the overall level and structure of cultural contact in the region.
-This differs somewhat from the rationales for pottery décor networks in studies from other parts of the world.
-Hart and Engelbrecht (2012, pp.
-324--330) justify their building of networks based on pottery décor similarity by arguing for décors as signals in which both group membership and individual skill are expressed.
-Iacono (2016), on the other hand, takes the co-occurrence of pottery decoration as analogous to other forms of social encounters, and thus as representative of them, conveying the agency of individuals "simply through sensory recognition" (Iacono 2016, p.
-126).
-We would like to stress here that these differences are in no small part due to the history of research and the nature of the archaeological record in different parts of the world, and that the building of network models on archaeological data always requires a well-theorised foundation appropriate to the historical context of the studied material.
 
@@
 ## Data
-
-In this paper, we use pottery decor data from contexts occupied in the 12th century AD to attempt a reconstruction of networks through which pottery decorative practices were shared.
-The network is constructed from standard archaeological frequency data, in which the proportions of decorative practices are represented across the selected site phases (the raw data are available as Annex 1).
-The decorative practices encode both the tool used and the action performed, for instance a twisted cord (tool) roulette (action).
-This data is partly taken from existing literature, partly relies on our own fieldwork (TMD and TOG2) or raw data raised by other researchers (TBZ), and was partly documented in collections (Gao-Saney, Tellem).
-This time slice, which represents a cumulative view of a great many individual interactions, will then be compared to earlier contexts in the same wider area, in order to demonstrate the diachronic possibilities of the approach.
-The sites we have been able to include in the network model are shown in \@ref(fig:sitemap).
-We can see here that sites are geographically unequally distributed.
-This is due to research history as well as past cultural dynamics.
-Certainly, the IND and the areas bordering it (notably the Lakes region) were strongly populated at the time, maybe more so than other areas.
-The cluster of sites around the Bandiagara escarpment is due to the extraordinarily detailed and long-term research of the University of Geneva's research teams.
-Blank spaces do not imply a lack of people, in this case people making and using pottery.
-Global politics and research priorities over the last 50 years have meant that these are areas we do not currently have excavated and dated ceramics for, though we know that sites exist, often in abundance.
-This means that in addition to the inherent difficulty of archaeological network models -the actual network is not known, only its effects, from which the network can be partially reconstructed (Sindbaek (2013)-, the effects of missing areas or sites on the network need to be calculated (Peeples et al.
-2016).Accordingly, this study should be understood as an exercise in the reconstruction of a possible network that led to the data preserved in the archaeological record: a tiny, unrepresentative sample of a tiny unrepresentative sample, influenced by an unknown amount of mostly unknowable pre- and post-depositional biases.
-Nevertheless, we will attempt to analyse the reconstructed network to formulate some hypotheses about the nature of innovation and spread in ceramic decors.
-Future generations of researchers are encouraged to improve upon this initial attempt.
 
@@
 
-Our theoretical approach to what the network represents informs the way in which the network model is constructed.
-So far, the most frequent form of making ties in archaeological ceramic data has been by calculating the statistical similarity between assemblages.
-(Hart and Engelbrecht 2012) use the Brainerd-Robinson similarity coefficient, as do, for instance, (Mills et al.
-2015), while (de Groot 2019) uses the Jaccard and Kulczyinski-2 similarity measures.
-The similarity value is then used as the basis and weight for forming a tie between the two assemblage nodes.
-Since we did not only want to base our network on overall similarity but wanted to be able to closer analyze the networks of individual decors, we have constructed a multiplex network.
-This is a network in which edges can have different types, and thus the network can be envisaged as having layers.
-A multiplex network is distinct from a multilayer network, in which connections can also exist between layers, and which has recently seen its first application to archaeological ceramics by (Upton 2019).
-In our multiplex network, each decorative practice is represented as a distinct layer.
-The layers share the site assemblages as nodes, though not every node is necessarily represented on every layer.
-The ties are weighted by the product of the percentages of a given décor in two assemblages.
-This network construction is implemented in R (R Core Team 2022) using the tidygraph package (Pedersen 2022).
+## Data Quality
 
-## quality of data
 ```{r iNext overview, echo = FALSE}
@@
 
-As is shown in \@ref(fig:total-network), the initial structure of the data leads to a very tightly connected network, known as a "hairball".
-This is normal for archaeological networks based on co-occurrence or similarity, since even rare occurrences lead to a tie.
-However, the multiplex nature of the network allows us to make an important initial observations on its structure: the networks in which individual decor types are propagated are only partly overlapping.
-There are no two decor networks that cover one another entirely.
-This pattern emerges more clearly in \@ref(fig:total-network-sample), when the number of decorative practices represented on the graph is reduced.
-This structure suggests that the networks allowing decorative practices to propagate were at least partially different in every case.
-Remembering the constraints on spread identified by Gosselain (2011, see section 2), we might propose that external factors, such as geography and cultural differences, provided "frontiers" to the spread, albeit different ones in each case.
-Some decors are very widespread, and thus had fewer inhibitions to their spread, while others occur less frequently.
- We can now proceed to further analyse the network in order to identify where such frontiers and catalysts of spread might have been and how they were structured.
-
 ```{r make decor network, echo = FALSE}
@@
 ggraph(decor_graph, layout = "stress") +
-  coloured_edges +
+  geom_edge_fan(colour = "grey") +
+  geom_node_point()+
+  geom_node_text(aes(label = label))+
   theme_graph()
@@
 ```
-
-In the first place, we will determine whether there are decors that share similar knowledge exchange networks and others that are dissimilar.
-Using the multinet package (Magnani et al.
-2021), the statistical distance (Jaccard similarity) between the different decors' networks were calculated based on the nodes they contained.
-
-\@ref(fig:layer-comparison) shows that some networks are more similar to one another than others.
-For instance, the practice of rouletting with a braided cord  (McIntosh and Gueye 2010) appears to have a significant shared network with that of rouletting with a folded strip (Haour and Keita 2010).
-This means that knowledge of the two tools necessary for these roulettes (for details, see Haour et al.
-2010) spread in a similar geographic and cultural sphere, without being impeded.
-On the other hand, there appears to have been quite strong boundaries in place between the networks sharing knowledge on rouletting with a braided strip and dragging a comb along the surface of the pot or using painted decoration.
-There thus appear to have been groups of decors that were shared within similar networks, and we can thus ask whether there might have been communities in the network predominantly sharing specific decors.
 
@@
 
-While the hyperconnected hairball network has shown us key elements of the structure of pottery décor sharing, it is of less use for more traditional network analyses.
-These include finding key nodes in the network and key routes through it.
-In our case, such network statistics might tell us which of the sites, on the basis of their assemblages, were influential in the propagation of pottery decorations, and through which channels the knowledge of different décors might have preferentially passed.
-In order to extract such measures, the number of connections needs to be reduced.
-Network sparsification or 'backbone extraction', as this process is also called, is a common challenge in network analysis across its different applications (Serrano et al.
-2009; Saha et al.
-2013).
-In archaeological networks, this has frequently been achieved through thresholding, i.e.
-retaining only the strongest connections and discarding the others (as implemented for example in Hart & Engelbrecht 2012).
-There are, however, a number of other recent approaches to network sparsification.
-For our purposes, we have found the best results to come from a process called the "disparity filter" (Serrano et al.
-2009).
-This evaluates the weight of each edge against a null model in which the overall edge weights of a node are evenly distributed among its edges.
-If an edge is found to diverge from the null model at a set probability, it is retained.
-The backbone package (Domagalski et al.
-2021) implements this in R.
-In order to use it, we must first 'flatten' the multiplex graph, by adding the weights of the edges between each pair of nodes on all layers.
-We then extract the backbone of the graph using the disparity filter at different probability thresholds (\@ref(fig:backbone-extraction)).
-
 ```{r flatten}
@@
 ```
+
 ```{r weighted flat decor}
@@
 
-The three communities from the previous section are also visible in the extracted backbone, albeit with some major differences in configuration.
-The spread-out cluster of long-distance contact now also includes the sites at Akumbu and Kawinza which were placed in the IND cluster.
-The latter is now reduced to the sites of the Eastern IND, while the Dia sites take a central position within the network, and the "Dogon" cluster remains largely intact.
-Dammassogou and Kokolo NE 2 are isolated from the network, and remain so until the significance is lowered to 0.2.
-We will use the backbone at alpha = 0.18 to proceed with our analysis.
-If this selection of significance values seems rather arbitrary, we point out that analogous situations exist in other attempts at archaeological network simplifications.
-Hart & Engelbrecht (2012, p.
-339), for instance, threshold their similarity network to the point where all nodes form a network but for a single outlier.
-Higher significance values also do not necessarily mean we get a more reliable network model, especially in a situation where sampling and recovery bias is as extreme as here.
-
 ### Influential nodes
-
-On this backbone, whose ties are no longer weighted, we can perform network calculations.
-We will begin by asking which nodes are the most influential in the network, i.e. possess the highest potential for spreading information.
-This is calculated through eigenvalue centrality, which is based on the number of nodes a node is linked to and how well those nodes are linked themselves.
-Eigenvalue centrality gives us a first idea of key sites in the spread of pottery decor knowledge.
 
@@
 
-The result (\@ref(fig:backbone-eigenvalue)) shows that the highest influence on the network is exerted by the sites in the Eastern IND and Lakes cluster.
-This reinforces the pattern already seen in clustering and shows the importance of the urban sites of the IND to Middle Niger culture in this period.
-We should not forget that the network is partial, i.e. many connections to many sites are missing.
-While the IND and the Bandiagara plateau and the Seno plain are both well researched and the networks therefore comparable, sites like Gao-Saney might have been influential in an undocumented local network.
- Nevertheless, it suggests this region specifically as a center through which information and innovation in pottery decoration were disseminated in this period.
-
 ### Brokers
-
-Betweenness centrality shows nodes in brokerage positions within the network.
-These are interesting to us in finding out where the different network cliques connect, and which sites are at crossroads positions in the pottery knowledge exchange.
-These are likely to be places with diverse markets, where producers belonging to different networks meet, and where information thus "jumps" between network cliques.
 
@@
 
-Particularly interesting among the results (\@ref(fig:backbone-betweenness) is the position of the two sites at Dia, Shoma and Mara (Bedaux et al.
-2005), as providing a connection between the wider Niger cluster and the Eastern IND.
-It is notable that Shoma clusters more with the former, whereas Mara is more closely linked to the latter, via the site of Toguéré Galia (Bedaux et al.
-1978).
-The Eastern IND sites that had such high eigenvector centrality have particularly low betweenness centrality.
-Also of note is the brokerage position of the Sanga cave sites (Tellem) in relation to the Bandiagara and Seno cluster.
-As the assemblage here notionally stems from a funerary context, their high betweenness centrality might indicate that ceramics using unusual or imported decorations were preferentially deposited here, while the ceramics from contemporary habitation sites tended to have a more restricted repertoire.
-
 ### Evaluating measurements
-
-In order to know whether these centrality measures are valid, rather than random effects of the backbone sparsification, we implement a variation on Peeples' (2017) suggested method for estimating the robustness of centrality measures in archaeological networks.
-In order to do so, the edges of the original graph are sampled at 10 % intervals, from 10 to 90 %.
-The binary backbone of each of these samples is then extracted and the centrality measures calculated.
-A Spearman's rank correlation test is run between the samples to see how the deletion of edges influences the centrality within the backbone.
-This is repeated 100 times.
-\@ref(fig:validity-test) shows that the correlation between the eigenvector and betweenness ranks of nodes in the original graph and the mean of the sampled graphs decreases quite rapidly.
-This shows us that our network is rather fragile, and the inferences we have made on the basis of the extracted backbone should be treated with the appropriate caution.
 
@@
 ## Community detection
-
-We approach this question by clustering, or community analysis, which a number of recent studies have successfully used on archaeological netoworks.
-Radivojevic and Grujic (2018) have clustered sites on the Balkan based on the source of their copper, Mazzucato (2019) has identified communities of building sites in Çatalhöyük to analyse the process of urbanisation, while Vargas et al.
-(2020) used community detection to find distinctive clusters of rock art sites in Patagonia.
-These studies differ significantly as regards the object of research and their questions, but are similar in data structure and methodological approach.
-All three used the Louvain clustering algorithm (Blondel et al.
-2008) to detect groups of sites that share a lot of edges with each other and have fewer edges with nodes outside of their community.
-The Louvain method is especially useful for archaeological datasets that typically contain few but heavily connected nodes (Blondel et al.
-2008; Yang et al.
-2016).
 
@@
 ```
-
-Here, we employ the same clustering algorithm, which results in three communities with a modularity of `r modularity`.
-[Modularity shows how clearly the communities can be defined]{.underline} on a scale [from -1 to 1]{.underline} (1 [being a network with completely separated sub-networks]{.underline})[.]{.underline} As is shown in \@ref(fig:plot-clusters), community 1 contains seven sites in an area centered on the Inland Niger Delta.
-Community 2 also contains seven sites, but located in the so-called "Dogon Country", on the eastern edge of the IND, on the Bandiagara plateau and in the Seno plain.
-Community 3 has the fewest nodes, but geographically the largest extent, containing the southernmost site of the study near Ségou, Togu Missiri, as well as the northernmost at Essouk in the Sahara.
-The nodes of community 3 show frequent connections to sites of communities 1 and 2, while those communities are not as strongly connected between each other.
 
@@
 
-The clustering of the network appears to be influenced quite strongly by geographic features.
-In the case of community 1, this is the IND and its adjacent areas, which were heavily populated at this time (McIntosh 1995, pp.
-373--377).
-The population density is likely to have been a significant factor in the spread of pottery decors, as Gosselain has suggested.
-\@ref(fig:cluster1) shows that there is a very well connected group of geographically close sites in the eastern IND (Jenné-jeno (J-J), Hambarketolo (Hamb), and Kaniana are part of the same wider site cluster).
-This community mainly shared braided cord roulettes, dragged comb incisions.Interestingly, this also includes the site of Kawinza in the Malian Lakes Region (Raimbault and Sanogo 1991).
-Sites on the western margin of the IND, Dia and Akumbu, are less well connected in this period, conforming to the observations of McIntosh (1995, pp.
-369--370), but these remain connected via folded strip roulettes and painted decors.
-
 ```{r cluster1, fig.cap="Internal connections within community 1"}
@@
 
-The main geographic factor affecting community 2, and isolating it from community 1, is the mountainous region of the Bandiagara plateau.
-The cluster has strong links, mostly because of the common use of braided strip roulettes and impressions of a straight-woven cord mat (\@ref(fig:cluster2).
-The networks exchanging these decors, which are in part linked to base forming techniques, appear to have had a spatial component, with mat impressions linking sites in the Seno plain while braided strip roulettes are more strongly shared on the Bandiagara plateau.
-These spheres, however, had considerable overlap, most notably at the Sangha cave sites (Tellem) and at Sadia.
-
 ```{r cluster2, fig.cap="Internal connections within community 2"}
@@
 
-Community 3 appears to be linked through the wider Niger river system and the long-distance exchange routes that linked into it.
-This cluster is the most diverse one regarding the types of decoration techniques shared (\@ref(fig:cluster3)), yet the weak nature of the resulting ties shows that this might be due to more exchange of artifacts and less in common shared knowledge across artisans.
-This is particularly visible at Essouk, which has the weakest connections that, due to its position and cultural history (Nixon 2017), are probably due to imports more than a shared cultural sphere.
-All of the sites in the community have been linked to long-distance trade (Insoll 1996; Nixon 2017; Gestrich and MacDonald 2018; Gestrich et al.
-2021), though this applies just as well to some of the sites in the other communities.
-The fact that these cluster together is likely due to the fact that their closer local communities are not part of the network model.
-This, however, allows us to see such ephemeral, long-distance connections based on weaker ties that might be obscured by more intense local exchange.
-
 ```{r cluster3, fig.cap="Internal connections within community 3"}
@@
 ```
+
 ```{r cluster4, fig.cap="Internal connections within community 3"}
@@
 ```
-
-The measures of network centralization and the mean weight of the edges relative to their number show us that communities 1 and 2 are more equally connected with higher weights, whereas community 3's network is more centralized, thus more uneven, but also much less intense (@ref(fig:cluster-comparison)).
-This suggests to us a fundamental difference in the nature of the networks.
-Communities 1 and 2 may have been founded on close cultural ties, frequent exchanges, and maybe even kinship bonds within specialized potter communities.
-Cultural frontiers, accentuated by geographical boundaries, hindered the spread of decorative techniques between them.
-Community 3 seems to have been a network based on far less intense contacts, such as sporadic movement of individuals in the context of trade or migration or indirect, down-the-line, exchange of knowledge.
-\@ref(fig:cluster_centralities) shows the distribution of the weighted degree and eigenvector centralities for each community within the overall network.
-Clearly, community 1 consists of nodes that both have more and stronger connections and consequently a higher potential for influence on the network.
-This reflects a number of points in the archaeological record.
-Firstly, the ceramics in the IND are highly decorated with diverse decors during this period.
-The clustered urban sites in this area are likely to have been culturally diverse, both drawing in populations from surrounding areas and exerting cultural influence across the wider region.
-McIntosh sees the Eastern IND as becoming increasingly culturally distinct from the Western and Northern IND during this period (McIntosh 1995, p.
-372), and the emergence of new material cultural practices may well be reflected in these measures.
 
@@
 
-# Graph of centralization of the different communities
+# Graph of edge diversity of the different communities
 g2 <- ggplot(mean_edge_weight) +
@@
 
-These three communities, however, were by no means separate from one another.
-If we consider only those edges that "cross" between communities, we can identify the decors that passed between communities, but also the sites that were important crossing points (\@ref(fig:crossings).
-These nodes are indicated by their weighted degree centrality, that is, the sum of the weights of all edges pertaining to the node.
-The sites of cluster 1 are generally better connected, both within the cluster and to sites outside their cluster.
-This might be a measure of the cultural influence and diversity of the IND's urbanized society in this period.
-The strongest node in this measure represents the assemblage from TMD (Tongo Maaré Diabal, Horizon 5 (Gestrich and MacDonald 2018)), which provides a crossing between all three communities.
-These results complement Gestrich's (2013, MacDonald & Gestrich 2018) result from correspondence analysis from which he claims that Tongo Maaré Diabal's ceramic assemblage reflects varying geographic influences.
-Even without the edges that tie the individual communities together, the network of crossing edges remains densely connected, demonstrating just how well connected the populations of the Middle Niger region must have been in this period, and that the limits of the communities we have detected in the model are far from being clear-cut cultural boundaries.
-
 ```{r crossings, fig.cap="Network of edges which cross between clusters"}
@@
 ## Diachronic analysis
-
-The network of interactions that we have been working with up to this point is built on data taken from archaeological contexts whose date range includes the 12^th^ century AD.
-This means that, in some cases, the period under consideration is actually much longer.
-Change over time within this period is to be expected, but is hard to distinguish given the nature of the data.
-Within wider brackets, the issue of change can be approached from a network perspective.
-While a more systematic diachronic view will be the subject of future publications, we will here consider an early 9^th^ century AD sample in order to show how the temporal development of the network can be approached.
 
@@
 
+# put cumulative edge weights back into the backbone
+flat_edges <- early_flat |> 
+  get.data.frame("edges")
+
+early_flat_bb <- early_flat_bb |> 
+  activate(edges) |> 
+  select(-weight) |> 
+  left_join(flat_edges)
+
 #plot
 ggraph(early_flat_bb, layout = "stress") +
-  geom_node_point() +
-  geom_edge_fan(color = "grey", alpha = 0.8) +
+  geom_node_point(aes(size = centrality_betweenness())) +
+  geom_edge_fan(aes(width = weight), color = "grey", alpha = 0.8) +
   scale_edge_width(range = c(0.2, 3)) +
@@
 
-The analytical steps run on the 12^th^ century network were carried out identically on the earlier assemblages.
-The main differences are best visualized on the extracted backbone (\@ref(fig:early-backbone).
-Here, only two network cliques remain.
-The first takes sites within the IND and on its margins, including TMD and Akumbu.
-The other group consists of sites on the Bandiagara plateau and in the Seno plain.
-The latter group is more strongly internally connected than it is in the 12^th^ century network, while the IND group is less closely connected.
-Essouk is isolated in the backbone until alpha is increased to 0.37, indicating that connections to Essouk are not very reliable.
-
-A more in-depth treatment of the temporal development of the pottery networks in this area will be the subject of future publications.
-For now, we might draw the following preliminary conclusions: The comparison of the two networks appears to show several processes.
-One, observable on the Bandiagara-Seno cluster, is a network that is becoming less tightly connected, in other words it opens up to more outside influences between the two periods.
-This, in a sense, is reflective of Gallay et al.'s vision of increasingly mixed traditions, discussed above.
-The second process is observable in the IND cluster, which is less strongly connected in the 9^th^ century network and becomes more coherent and also internally diversified by the 12^th^ century.
-This is more reflective of a dynamic of an area in the process of cultural genesis, rather than fragmentation.
-Thirdly, the long-distance connections emerging from the 12^th^ century networks are very weak, so as not to be statistically detectable, in the 9^th^ century.
-While there are issues of unequal samples between the two periods, we can take this to suggest that the impact of long-distance exchange along the Middle Niger on pottery decoration was not yet very strong in the earlier period.
-
 # Discussion
 
-Potter communities and the propagation of knowledge within them, have been the subject of several very detailed ethnoarchaeological and archaeological studies in this region and within Africa more widely.
-Their boundaries might be seen in other elements of the pottery making sequence, notably in the forming technique, but decorative practices are recognized not to be straightforwardly related to social identities (Gosselain 2011, p.
-11).
-Our network model is intended to represent communication of knowledge about pottery that transcends smaller communities of practice.
-As such, it permits us to explore what mechanisms of exchange may have been at the basis of the patterns in archaeological ceramic decors on sites along the Middle Niger.
-
-The results of our analyses suggest that the propagation of pottery decor motifs occurred within networks of differing geographic extent.
-Some of these networks were local, geographically, and possibly also culturally, bounded.
-This can, for instance, be seen in the group of sites on the Bandiagara massif and in the Seno plain, or in the eastern part of the IND.
-Cross-cutting such local exchange, however, were long-distance contacts along the Niger River, which led to similarities between very distant sites.
-This underlines a key aspect of past cultural dynamics in the wider region: there was, and continues to be, both a high level of cultural diversity, of human groups and languages, and overarching regional connectivity.
-These two processes, the local and the global, interacted to shape the Middle Niger cultural landscape.
-Every site assemblage can therefore be seen to be reflective of how strongly innovation and tradition influenced local potting communities.
-
-Network analysis allows us to explore the roles of nodes within the network, and thus to show note where this interaction between network groups takes place.
-While the archaeological record is necessarily only a sample and many nodes are missing, we can nevertheless gain a preliminary insight into the channels along which information on pottery decoration was shared and identify those sites within the record which occupy key positions in this exchange.
-In the case of the 12^th^ century Middle Niger, our network model suggests that the clustered urban sites of the eastern IND exerted a large influence on the sharing of pottery decoration techniques, even though other sites, such as Dia, were more important to their spread.
-
-The diachronic comparison of two networks from different periods begins to show us processes of innovation, and how these change networks.
-This must be seen as a work in progress, and a fuller picture will be the subject of future publications.
-Nevertheless, we can detect how the sites within the IND became a major centre for diversification and diffusion of decorative techniques, while the close-knit and formerly more isolated "Bandiagara cluster" opened up to outside influences.
-Overall, the later period is marked by more ties, and therefore an intensifying network.
-This is likely due to an increased frequency of contact between heterogenous potter communities, perhaps as a result of a combination of increases in population size, the establishment of new markets, and an increased flow of goods, some of which will have been contained in pottery.
-Mayor (2011b, p.
-208) has also suggested political factors behind the spread of some decors.
-The difference between the networks is less likely to stem from an overall higher [long-range]{.underline} mobility of individual potters, since the overlapping structure of the individual technique networks suggests gradual change.
-
-Chronological and spatial approaches to pottery decor in the study region have not often yielded satisfying results, with changes only in percentages of decors which rarely provided the basis for clear phase changes or cultural boundaries.
-Our approach here emphasises that this picture of gradual differences points to a strongly connected society, in which pottery producers were routinely exposed to their neighbours' productions, and in which decorative techniques were mostly able to propagate without impediment.
-Nevertheless, within this picture, local traditions can be discerned as cliques within the networks, whose degree of isolation varies, but which are never without outside contact.
-
-We should note here that the network has artificial boundaries which are likely to impact on how well the model reflects actual social contacts in the period in question.
-Our sample is geographically centered on the IND, where we have one of the network cliques and the most abundant data.
-Especially towards the geographical edges of the sample, the model is probably much less reliable.
-Thus, we know that the Bandiagara -- Seno cluster is only a small part of what would likely be a much larger and internally differentiated cluster below the Niger Bend.
-If this were fully included in the model, it is certain that it would provide new and different linkages to the northeastern sites of our model, Gao and Essouk.
-The latter two sites also have extensive cultural links further north and east, which cannot be captured in this model.
-The same is true of sites in the south and west (Togu Missiri, Akumbu, Kawinza), whose main local networks are probably simply not reflected in this sample.
-All of this has led to network clusters of different kinds, as discussed above: the geographically large Niger river cluster is very likely reflective of very different forms of contact than the smaller, more tightly knit IND cluster is.
-We hypothesise that such smaller clusters could probably be found anywhere in the area if there were enough suitable data, and that the historical spread of pottery decors would have been influenced by both closer local and looser regional networks.
-In Gosselain's (2011) sense, the flow of information in the larger regional network would most likely have been impeded by geographic distance, but also by the force of the closer local networks.
-
 # Conclusion
-
-Modelling pottery decorative techniques along the Middle Niger as a knowledge exchange network presents a change of perspective on the subject.
-Previous work, including the author's own, has focused on distribution areas and relative prevalence at different sites.
-We have taken the same information to indicate the degree to which pottery producers supplying the various sites were part of contact networks in which knowledge of decors could spread.
-While this approach is certainly as reductive as any other, we draw the novel conclusion that the relatively static record of pottery decor in this area is due to a diverse and very well connected population of pottery producers.
-They exchanged their techniques most frequently on a local level, where geographically bounded closer networks shared intensively.
-Such exchange might have taken many forms, such as marriage-related mobility, kinship-based learning networks, much more ephemeral exchanges in marketplaces, or even indirect contact by seeing another potter's product.
-The latter form is the most likely basis for the way in which these local networks corresponded to a larger one: all along the Middle Niger, from our southernmost sample in the Ségou area to Gao and the Saharan entrepot of Essouk, pottery and the knowledge of certain decorative techniques appears to have travelled.
-
-The study of archaeological material as networks shifts the focus "from spatial relations to social relations"(Knappett 2011).
-Herein lies the main difference of the approach taken in this paper to previous synthetic studies of archaeological ceramics on the Middle Niger.
-Mayor's (2011b) study, in particular, has admirably shown how the ceramics of the wider region can reflect a number of cultural groups and historical influences.
-These preliminary results of our network study, on the other hand, begin to show how the area's pottery reflects webs of exchange, both at smaller local levels and at a larger regional level.
-It makes visible and understandable the way in which the Middle Niger was both an integrated cultural sphere, albeit one consisting of significant differences.
-These two approaches do not invalidate each other but should be seen as complementary.
-The results feed into long-standing discussions on the Middle Niger, and West Africa in general, as a highly connected and interdependent cultural space in the first and early second millennium AD (for recent discussions on this topic, see Gestrich et al.
-2021; Dueppen and Gallagher 2023).
 
@@
 The work presented here was undertaken as part of the project "Borrowed words and shared objects" within the German Research Foundation's (DFG) Priority Programme "Entangled Africa".
-The authors wish to thank the Institut des Sciences Humaines du Mali, the Musée National du Mali, and the Rijksmuseum voor Volkenkunde in Leiden for access to their pottery collections, and Sam Nixon and Douglas Post Park for providing primary data.
+Nikolas Gestrich was responsible for the research design and implementation.
+He is the main author of the analysis code and the manuscript.
+Juan-Marco Puerta Schardt aided in collating the data and co-authored the analysis code, particularly the section on community detection.
+He also contributed to the draft versions of the manuscript.
+The authors wish to thank the Institut des Sciences Humaines du Mali, the Musée National du Mali, and the Rijksmuseum voor Volkenkunde in Leiden for access to their pottery collections, and Susan McIntosh and Douglas Post Park for providing primary data.
 The R compendium for this article was created using the rrtools package.
@@
 <!-- The following line inserts a page break  -->
-
-\newpage
-
-# References
-
-<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
-
-Barker CS, Young LC (2017) Networks of Ceramic Exchange: Comparing Homol’ovi Pueblo III Pithouse and Pueblo Communities. KIVA
-
-Bedaux R (1980) The geographic distribution of footed bowls in the upper and middle Niger region.
-In: Swartz, B., Dumett, R.
-(eds), West African Culture Dynamics: Archaeological and Historical Perspectives.
-pp 247–258
-
-Bedaux R, Lange A (1983) Tellem, reconnaissance archéologique d’une culture de l’Ouest africain au Moyen-Age: la poterie. Journal des Africanistes 53:5–59
-
-Bedaux R, Polet J, Sanogo K (2005) Recherches archéologiques à Dia dans le delta intérieur du Niger (Mali): Bilan des saisons de fouilles 1998-2003. CNWS Publications, Leiden
-
-Bedaux RMA, Constandse-Westermann TS, Haquebord L, et al (1978) Recherches archeologiques dans le Delta Interieur du Niger (Mali). Palaeohistoria Bussum 20:91–220
-
-Blondel VD, Guillaume J-L, Lambiotte R, Lefebvre E (2008) Fast unfolding of communities in large networks. J Stat Mech 2008:P10008. https://doi.org/10.1088/1742-5468/2008/10/P10008
-
-Brughmans T, Poblome J (2016) Pots in space: understanding Roman pottery distribution from confronting exploratory and geographical network analyses.
-In: Barker E, Bouzarovski S, Pelling C, Isaksen L (eds) New worlds from old texts: revisiting ancient space and place.
-Oxford University Press, Oxford, pp 255–280
-
-Crabtree SA (2015) Inferring Ancestral Pueblo Social Networks from Simulation in the Central Mesa Verde. J Archaeol Method Theory 22:144–181. https://doi.org/10.1007/s10816-014-9233-8
-
-Curdy P (1982) Tiébala (Mali): un complexe céramique du 6e siècle après J.-C. Archives suisses d’anthropologie générale 46:183–198
-
-David N, Sterner J, Gavua K (1988) Why pots are decorated. Current Anthropology 29:365–389
-
-de Groot BG (2019) A Diachronic Study of Networks of Ceramic Assemblage Similarity in Neolithic Western Anatolia, the Aegean and the Balkans (c.6600–5500 bc).
-Archaeometry 61:600–613.
-https://doi.org/10.1111/arcm.12450
-
-Dietler M, Herbich I (1994) Ceramics and thnic identity : ethnoarchaeological observations on the distibution of pottery styles and the relationship between the social contexts of production and consumption.
-In: Terre cuite et société : céramique, document technique, économique, culturel.
-Rencontres internationales d’archéologie et d’histoire (14, 21- 23 oct.
-1993, Antibes).
-Eds APDCA (Assoc.
-pour la promotion et la diffusion des connaissances archaéologiques).
-Juan-les-Pins, pp 459–472
-
-Domagalski R, Neal ZP, Sagan B (2021) Backbone: An R package for extracting the backbone of bipartite projections. PLoS ONE 16:e0244363. https://doi.org/10.1371/journal.pone.0244363
-
-Dueppen SA, Gallagher D (2023) Networked Farmers, Ancestral Rituals, Regional Marketplaces, and Salt: New Insights into the Complexity of First Millennium BC/AD Farming Societies in West Africa.
-Afr Archaeol Rev.
-https://doi.org/10.1007/s10437-022-09509-2
-
-Gallay A, Huysecom E, Mayor A (1998) Peuples et céramiques du Delta intérieur du Niger (Mali): un bilan de cinq années de missions (1988-1993). Mainz
-
-Gallin A (2011) Les styles céramiques de Kobadi. Analyse comparative et implications chronoculturelles au Néolithique du Sahel malien. Frankfurt
-
-Gestrich N, Champion L, Keita D, Coulibaly N (2021) Evidence of an 11th century CE cola nitida trade into the Middle Niger region. African Archaeological Review 38:403–418
-
-Gestrich N, MacDonald KC (2018) On the Margins of Ghana and Kawkaw: Four Seasons of Excavation at Tongo Maaré Diabal (AD 500-1150), Mali. Journal of African Archaeology 16:1–30
-
-Gosselain O (2000) Materializing identities : an African perspective. Journal of Archaeological Method and Theory 7:187–217
-
-Gosselain OP (1998) Social and Technical Identity in a Clay Crystal Ball. In: Stark MT (ed) The Archaeology of Social Boundaries
-
-Gosselain OP (2011) Pourquoi le décorer? Quelques observations sur le décor céramique  en Afrique. Azania: Archaeological Research in Africa 46:3–19
-
-Granovetter MS (1973) The Strength of Weak Ties. American Journal of Sociology 78:1360–80
-
-Haour A, Keita D (2010) Folded Strip Roulette/ Roulette de fibre plate pliée.
-In: Haour A, Manning K, Arazi N, et al.
-(eds) African Pottery Roulettes Past and Present.
-Techniques, Identification  and Distribution.
-Oxbow, Oxford, pp 169–176
-
-Haour A, Manning K, Arazi N, et al (eds) (2010) African pottery roulettes past and present: techniques, identification and distribution. Oxbow Books, Oxford
-
-Hart JP (2012) The effects of geographical distances on pottery assemblage similarities: a case study from Northern Iroquoia.
-Journal of Archaeological Science 39:128–134.
-https://doi.org/10.1016/j.jas.2011.09.010
-
-Hart JP, Engelbrecht W (2012) Northern Iroquoian Ethnic Evolution: A Social Network Analysis. J Archaeol Method Theory 19:322–349. https://doi.org/10.1007/s10816-011-9116-1
-
-Herbich I (1987) Learning patterns, potter interaction and ceramic style among the  Luo of Kenya. African Archaeological Review 5:193
-
-Hodder I (1982) Symbols in Action: Ethnoarchaeological Studies of Material Culture. Cambridge University Press, Cambridge
-
-Iacono F (2016) From Networks to Society: Pottery Style and Hegemony in Bronze Age Southern Italy. Cambridge Archaeological Journal 26:121–140. https://doi.org/10.1017/S0959774315000190
-
-Insoll T (1996) Islam, Archaeology and History. Gao Region (Mali) ca. AD 900-1250. Tempus Reparatum, Cambridge Monographs in African Archaeology 39. BAR S647. Oxford
-
-Knappett C (2011) An archaeology of interaction: network perspectives on material culture and society. Oxford University Press, Oxford
-
-Knappett C (2021) Networks in Archaeology. The Oxford Handbook of Social Networks 443–466. https://doi.org/10.1093/oxfordhb/9780190251765.013.26
-
-Livingston Smith A (2007) Histoire du décor à la roulette en Afrique subsaharienne. Journal of African Archaeology 5:189–216
-
-Magnani M, Rossi L, Vega D (2021) Analysis of Multiplex Social Networks with R. Journal of Statistical Software 98:1–30
-
-Mayor A (2011a) Impressions de vanneries et technique du martelage sur forme concave:  Anthropologie et histoire d’une technique dans la boucle du Niger. Azania 46:1:88–109
-
-Mayor A (2011b) Traditions céramiques dans la boucle du Niger. Africa Magna, Cologne
-
-Mazzucato C (2019) Socio-Material Archaeological Networks at Çatalhöyük a Community Detection Approach. Frontiers in Digital Humanities 6:. https://doi.org/10.3389/fdigh.2019.00008
-
-McIntosh RJ (1998) The peoples of the Middle Niger. The island of gold. Blackwell Publishers, Malden, Mass.; London
-
-McIntosh SK (1995) Excavations at Jenné-Jeno, Hambarketolo, and Kaniana, Inland Niger delta, Mali, the 1981 season. University of California press, Berkeley (Calif.) Los Angeles (Calif.) London
-
-McIntosh SK (ed) (1999) Beyond Chiefdoms. Pathways to Complexity in Africa. Cambridge
-
-McIntosh SK, Gueye NS (2010) Braided cord roulette (simple and composite).
-In: Haour A, Manning K, Arazi N, et al.
-(eds) African Pottery Roulettes Past and Present.
-Techniques, Identification  and Distribution.
-Oxbow, Oxford
-
-McIntosh SK, McIntosh RJ (1980) Prehistoric Investigations in the Region of Jenne, Mali. BAR, Oxford
-
-Mills BJ, Peeples MA, Haas, Jr. WR, et al (2015) Multiscalar Perspectives on Social Networks in the Late Prehispanic Southwest. American Antiquity 80:3–24. https://doi.org/10.7183/0002-7316.79.4.3
-
-Mills BJ, Roberts JM, Clark JJ, et al (2013) The dynamics of social networks in the Late Prehispanic U.S.
-Southwest.
-In: Knappett C (ed) Network analysis in archaeology.
-New approaches to regional interaction.
-Oxford University Press, Oxford, pp 181–202
-
-Nixon S (2017) Essouk-Tadmekka: An Early Islamic Trans-Saharan Market Town. Brill
-
-Park M (1799) Travels in the Interior of Africa. W. Bulmer & Co., London
-
-Pedersen T (2022) tidygraph: A Tidy API for Graph Manipulation. https://tidygraph.data-imaginist.com, https://github.com/thomasp85/tidygraph
-
-Peeples MA (2017) Network Science and Statistical Techniques for Dealing with Uncertainties in Archaeological Datasets. http://www.mattpeeples.net/netstats.html. Accessed 18 Oct 2022
-
-Peeples MA, Mills BJ, Haas WRJr, et al (2016) Analytical Challenges for the Application of Social Network Analysis in Archaeology.
-In: Brughmans T, Collar A, Coward F (eds) The Connected Past: challenges to network studies in archaeology and history.
-Oxford University Press, Oxford, pp 59–84
-
-R Core Team (2022) R: A Language and Environment for Statistical Computing. R Foundation for Statistical Computing, Vienna, Austria
-
-Radivojević M, Grujić J (2018) Community structure of copper supply networks in the prehistoric Balkans: An independent evaluation of the archaeological record from the 7th to the 4th millennium BC.
-Journal of Complex Networks 6:106–124.
-https://doi.org/10.1093/comnet/cnx013
-
-Raimbault M, Sanogo K (eds) (1991) Recherches Archéologiques au Mali: Prospections et inventaire,  fouilles et études analytiques en Zone lacustre. Karthala, Paris
-
-Roux V, Bril B, Karasik A (2018) Weak Ties and Expertise: Crossing Technological Boundaries. J Archaeol Method Theory 25:1024–1050. https://doi.org/10.1007/s10816-018-9397-8
-
-Saha T, Rangwala H, Domeniconi C (2013) Sparsification and Sampling of Networks for Collective Classification.
-In: Greenberg AM, Kennedy WG, Bos ND (eds) Social Computing, Behavioral-Cultural Modeling and Prediction.
-Springer, Berlin, Heidelberg, pp 293–302
-
-Schmidt A, Arazi N, MacDonald KC, et al (2005) La poterie.
-In: Bedaux RMA, Polet J, Sanogo K, Schmidt A (eds) Recherches archéologiques a Dia dans le Delta intérieur du Niger  (Mali): bilan des saisons de fouilles 1998-2003.
-CNWS, Leiden, pp 191–256
-
-Serrano MÁ, Boguñá M, Vespignani A (2009) Extracting the multiscale backbone of complex weighted networks.
-Proceedings of the National Academy of Sciences 106:6483–6488.
-https://doi.org/10.1073/pnas.0808904106
-
-Sindbæk SM (2013) Broken links and black boxes: material affiliations and contextual network synthesis in the Viking world.
-Network Analysis in Archaeology: New Approaches to Regional Interaction 71–94
-
-Sterner J, David N (2003) Action on matter: the unique African tamper and concave anvil pot-forming  technique. Journal of African Archaeology 1:3–38
-
-Sternini M (2019) The production centres and river network of Italian terra sigillata between the Arno and Tiber valleys: a geographical point of view.
-Journal of Roman Archaeology 32:485–494.
-https://doi.org/10.1017/S1047759419000254
-
-Tamari T (1991) The development of case systems in West Africa. Journal of African History 32:221–250
-
-Tamari T (1995) Linguistic Evidence for the History of West African "Castes. In: Conrad DC, Frank BE (eds) Status and Identity in West Africa. Indiana University Press, Bloomington, pp 61–85
-
-Upton AJ (2019) Multilayer Network Relationships and Culture Contact in Mississippian West-Central Illinois, A.D. 1200 - 1450. Ph.D., Michigan State University
-
-van Doosselaere B (2014) Le Roi et le Potier: Étude technologique de l’assemblage céramique de Koumbi Saleh, Mauritanie (5e/6e - 17e siècles AD). Africa Magna, Frankfurt am Main
-
-Vargas FE, Lanata JL, Abramson G, et al (2020) Digging the topology of rock art in northwestern Patagonia. Journal of Complex Networks 8:cnz033. https://doi.org/10.1093/comnet/cnz033
-
-Vavrasek JL (2020) Interactions Between Jefferson County Iroquoian Women and Men as Explored Through Social Network Analysis of Pottery and Pipe Decorative Motif Traits.
-Ph.D., State University of New York at Albany
-
-Yang Z, Algesheimer R, Tessone CJ (2016) A Comparative Analysis of Community Detection Algorithms on Artificial Networks. Sci Rep 6:30750. https://doi.org/10.1038/srep30750
 

### 2024-06-03 analysis/paper/paper.Rmd
last update inext

--- 2024-05-15 analysis/paper/paper.Rmd
@@
 ## quality of data
+
+The pottery samples vary strongly in size.
+This might effect the number of decorations present and might lead to (especially rare) decoration techniques beeing absent in some sites due to the small amount of sherds excavated.
+We use the iNext-package (Hsieh et al.
+2016) to estimate the effect the sample size has on the diversity of the sample.
+This way we can decide how the different abundance of sherds might impact our network.
 ```{r iNext overview, echo = FALSE}
@@
 
-i
+d <- i$iNextEst$size_based |> 
+slice(c(seq(1, n(), by = 10),seq(20, n(), by = 40))) |> 
+  arrange(Assemblage, m)
+
+d
 
@@
 # getting a table of the results of the analysis
-iNEXT(decors_abs, q=0, datatype="abundance")
+e <- iNEXT(decors_abs, q=0, datatype="abundance")
+
+e <- e$iNextEst$size_based |> 
+slice(c(seq(1, n(), by = 10),seq(20, n(), by = 40))) |> 
+  arrange(Assemblage, m)
+
+e
+#creating datatable of both periods
+#rbind(d,e) |> 
+ # mutate(across(where(is.numeric), ~ round(.x, 4))) |> 
+  #write.csv2(here::here("analysis/data/derived_data/inext.csv"))
+
 ```
@@
                      "paint",
-                     "straight_mat_impression")) |>
+                     "straight_cord_mat")) |>
 
@@
 decor_05 <- extract_backbone(decor_flat, 0.05)
-decor_15 <- extract_backbone(decor_flat, 0.14)
+decor_15 <- extract_backbone(decor_flat, 0.15)
 decor_20 <- extract_backbone(decor_flat, 0.20) #20 ideale anzahl?

### 2024-06-03 analysis/supplementary_materials/data preparation.qmd
changes accroding to paper revisions
last update inext

@@
 
+There have been several excavations at different sites at Gao; however, many of the older ones did not document ceramics in any detail (Cissé et al.
+2013).
+Of the more recent excavations by Insoll (in 1993 & 1996) and the Malian Direction Nationale de Patrimoine Culturel (2001, 2002 & 2006), only the latter have documented ceramics in a way that can be used in our analysis.
+While we tried to re-record Insoll’s pottery which is stored at the Institut de Sciences Humaines in Bamako in 2020, selection errors meant that we did not have a suitably datable sample for this analysis.
+Our use of the ceramics reported by Cisse are complicated by the fact that there is an inconsistency between dates produced from different laboratories at the site, with those of the University of Waikato being consistently around 100 years older than the two related dates analysed by Beta Analytic (Cisse 2011:139).
+A further complication is added by Cisse's decision to present the ceramic data in three aggregated assemblages (Final, Upper, Lower) for each unit, making chronological control of the samples even more difficult.
+Furthermore, only pottery from ACGS can be used for our model as the roulette decorations have been sufficiently documented in this unit.
+
+In Unit ACGS, the final levels (1-10) were made up of extensive trash deposits.
+Their upper component is dated to 995-1217 cal.
+CE (Beta 261360, 950 ± 60 BP).
+We include the "Final" assemblage in our twelfth century network as reflective of occupation at Gao, but there are points of caution: No architectural remains are known from this period at Gao Saney, the single date has a wide calibrated range and a date in neighbouring GS1, 40 cm lower, has a University of Waikato date with a calibrated range at the end of the first millennium CE.
+As only the top and the bottom of the ACGS sequence were dated, we cannot estimate how old the lower layers included in the sample are.
+
+All of the units excavated by Cisse were probably inhabited during the ninth century, but to isolate the contexts is rather difficult.
+A simple sequence model in Oxcal 4.4 refines the date from GS3 Level 5 to 803-979 cal CE, and that from Level 12 to 772 - 939 cal CE, suggesting that the bulk of the deposits dates to the ninth and the first half of the 10th centuries.
+In the presentation of the ceramic data, however, the "Upper" section comprises of Levels 2-12 and its date range probably extends into the 11th century.
+In GS1, the "Upper" assemblage is narrower, uniting Levels 5-9.
+These are dated by a single date in Level 7b, to a modelled age of 696-884 CE (648-876 cal CE), suggesting an admixture of 8th and possibly even 7th-century materials.
+In order to limit the chronological range of the sample, only the Upper assemblage from GS1 has been used.
+
+Although GS1 itself cannot be utilized, its proximity to ACGS enables a comparative analysis of the stratigraphy between the two units.
+Consequently, despite the absence of direct dating for the "Upper" Horizon of ACGS, we consider it the most suitable horizon for our ninth century network.
+
+Both datasets contain cord roulette decorations that were not specified; 15.5% of the decorations documented for lower layers and 17.5% of the decorations documented for the final layers.
+
 ```{r}
@@
 
-The site was excavated as part of the \"Human Population and Paleoenvironment in Africa\" by Huysecom et al..
+The site was excavated as part of the "Human Population and Paleoenvironment in Africa" by Huysecom et al..
 The dates of the different stratigraphic ensembles of the site are the results of a bayesian model based on 27 radiocarbon dates.

### 2024-06-03 analysis/paper/paper.Rmd
fixes to release 0.1.0

--- 2024-05-29 analysis/paper/paper.Rmd
@@
 The full text of the article is available in the published version in "Azania: archaeological research in Africa".
-This document contains the code for all analyses and figures in the text,as well as some supplementary analzses and figures.
+This document contains the code for all analyses and figures in the text,as well as some supplementary analyses and figures.
 It is structured by section to ease overview.
@@
   geom_node_text(aes(label = label), col = "black", repel = TRUE) +
+  guides(color = "legend", edge_width = "none", size = "none")+
   theme_graph()
@@
   ggraph(layout = "graphopt") +
+  scale_edge_width(range = c(0.2, 2)) +
   geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
   scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
-  scale_edge_width(range = c(0.2, 2)) +
   geom_node_point(aes(size = centrality_degree(weights = weight)), colour = "yellow") +
   geom_node_text(aes(label = label), col = "black", repel = TRUE) +
+  guides(color = "legend", edge_width = "none", size = "none")+
   theme_graph()

### 2024-06-07 analysis/paper/paper.Rmd
figure changes

@@
  iNEXT(q=c(0), datatype="abundance")
- ggiNEXT(i, type=1, facet.var="Assemblage")+
+ p1 <- ggiNEXT(i, type=1, facet.var="Assemblage")+
   labs(title = "Doupwil")
@@
  iNEXT(q=c(0), datatype="abundance")
-  ggiNEXT(i, type=1, facet.var="Assemblage")+
+ p2 <-  ggiNEXT(i, type=1, facet.var="Assemblage")+
   labs(title = "Jenne-jeno")
@@
 ggraph(decor_20, layout = "nicely") +
-  geom_node_point(aes(size = centrality_eigen())) +
   geom_edge_fan(color = "grey", alpha = 0.8, aes(edge_width = weight)) +
+   geom_node_point(aes(size = centrality_eigen(weights = weight))) +
   scale_edge_width(range = c(0.2, 3)) +
@@
 ggraph(decor_20, layout = "nicely") +
-  geom_node_point(aes(size = centrality_betweenness())) +
    geom_edge_fan(color = "grey", alpha = 0.8, aes(edge_width = weight)) +
+    geom_node_point(aes(size = centrality_betweenness(weights = weight))) +
   scale_edge_width(range = c(0.2, 3)) +
@@
 ggraph(early_flat_bb, layout = "stress") +
+  geom_edge_fan(aes(width = weight), color = "grey", alpha = 0.8) +
     geom_node_point(aes(size = centrality_betweenness())) +
-  geom_edge_fan(aes(width = weight), color = "grey", alpha = 0.8) +
   scale_edge_width(range = c(0.2, 3)) +
   geom_node_text(aes(label = label), repel = TRUE) +
-  theme_graph()
+  theme_graph()+
+  labs(size = "betweenness centrality")
 ```

### 2024-06-10 analysis/paper/paper.Rmd
300 dpi picks

--- 2024-06-03 analysis/paper/paper.Rmd
@@
       #- --lua-filter=../templates/pagebreak.lua
+      dpi: 300
 abstract: |
@@
   Mali; network analysis; pottery; archaeological ceramics
+
 ---
@@
 
+## quality of data
 ```{r iNext overview, echo = FALSE}
@@
   geom_node_text(aes(label = label), col = "black", repel = TRUE) +
-  guides(color = "legend", edge_width = "none", size = "none")+
   theme_graph()
@@
   ggraph(layout = "graphopt") +
-  scale_edge_width(range = c(0.2, 2)) +
   geom_edge_fan(aes(color = name, edge_width = weight, edge_alpha = weight)) +
   scale_edge_color_manual(values = decorcolor[E(decor_graph)$name]) +
+  scale_edge_width(range = c(0.2, 2)) +
   geom_node_point(aes(size = centrality_degree(weights = weight)), colour = "yellow") +
   geom_node_text(aes(label = label), col = "black", repel = TRUE) +
-  guides(color = "legend", edge_width = "none", size = "none")+
   theme_graph()

### 2024-06-10 analysis/paper/paper.Rmd

--- 2024-06-07 analysis/paper/paper.Rmd
@@
  iNEXT(q=c(0), datatype="abundance")
+
  p1 <- ggiNEXT(i, type=1, facet.var="Assemblage")+
@@
  iNEXT(q=c(0), datatype="abundance")
+
  p2 <-  ggiNEXT(i, type=1, facet.var="Assemblage")+
